\chapter*{Acknowledgements} 
\thispagestyle{plain}
This project has been under development since December 2006.  I should like to
express my warm thanks to my academic advisors, Professor Lars Hellan and
Professor Mary Esther Kropp Dakubu,  for their guidance and advice throughout
the years. I cannot thank them enough for their support and  linguistic insight.
Since I do not speak Chakali, I am especially indebted to  Daniel Kanganu
Karija,  Afia Kala Tangu, Fuseini Mba Zien,  Baaba Tangu, Awie Bakuri Ahmed, 
Seidu Kassim Tangu, Henry Seidu Daannaa, Godfrey Bayon Tangu,  Kpersi Naa, 
Kotia Nwabipe, `Spenta' and Sakara Baduong for their instruction and
enlightenment. To my
friends and colleagues, Benjamin Waldron, Eva Léonie Mutuzo, Peter Liem,
Franklin Arhinful, Cynthia Henewaa Kumah,  Mari Uuskula, Signe Rix Berthelin,
Kaja Borthen,
Jacques Koreman, Assibi Apatewon Amidu and Cesare Poppi, who kindly suggested
corrections, commented on the text, or helped me to solve problems I
encountered,  I owe a special thanks. Thanks too to  Sandra Foldvik for help
with editing the final draft. I am very grateful to the GILLBT office in Tamale
for the facilities I enjoyed there and for  allowing me to access their
archives, in particular to Hannes Hertzel for helping me with technical details,
and Anthony Naden for insightful recommendations. Mary Esther Kropp Dakubu,
James Saanchi, Godfrey Bayon Tangu and many other Ghanaians are responsible for
the smooth logistics  which made the research possible. Finally, with immense
gratitude, I want to thank everyone in the villages where I stayed for their
warm welcome and cooperation. 
\thispagestyle{plain}

%Lucy Skinner