\chapter*{Acknowledgement}
\label{sec:acknow}

%jungle
%sumo naa
%teacha wusa
%nana
%mansa bele

The completion of this project was dependent upon a multitude of factors, the 
most important being  the  knowledge and generosity of local residents whose 
mother tongue is Chakali. I am  indebted to  Daniel Kanganu Karija,   Fuseini 
Mba Zien, Afia Kala Tangu,  Awie Bakuri Ahmed,   Baaba Tangu, Seidu Kassim 
Tangu, Henry Seidu Daannaa, Godfrey Bayon Tangu,  Kpersi-Naa, Kotia Nwabipe, 
Amoa Bari-Naa,  and Sakara Baduong, among many others,  for their instructions 
and enlightenments. I owe a special thanks to Daniel Kanganu Karija and  Fuseini 
Mba Zien for their loyal contribution since the very beginning of my learning 
journey. A special thank you goes  to Jonas Kpierekoh, a specialist in 
agroforestery and principal programme officer at the  Environmental Protection 
Agency, who spent time with me and local experts in Ducie working on the 
scientific name of many tree species.  With immense gratitude, I  thank 
everyone in the villages where I stayed -- especially Ducie and Gurumbele --  
for their warm welcome, hospitality,  and cooperation. 

I would like to express my deep thanks to  Mary Esther Kropp Dakubu,  for her 
guidance, support,  advice,  and  linguistic insight throughout the years.   I 
wish to thank Eva Léonie Mutuzo, Benjamin Waldron,  Felix K. Ameka,  Kaja 
Borthen, Albert 
Awedoba,  Tyson Farrell, Lars Hellan, and Assibi Apatewon Amidu who kindly 
suggested 
corrections, and commented on  earlier versions.  Thanks as well to many people 
on the mailing list of the FieldWorks Language Explorer software, and  
Sebastian Nordhoff at  Language Science Press,  for helping me with 
technical details. [add reviewers and series editors]



% 
% \footnote{https://groups.google.com/forum/\#!forum/flex-list}

