\chapter{Grammar Outline}
\label{sec:gramsketch}


\section{Introduction}
\label{sec:Introduction}

This chapter  provides a broad outline of the grammar and introduces those 
aspects needed to understand the formations of words and sentences found in the 
dictionary. Further, it acts as a preliminary grammar of the language, which is 
and will always be essential for future description and analysis since it sets 
forth claims to be confirmed, rejected, challenged,  or improved.   First, the 
common clause structure, the main elements of syntax and clause coordination and 
subordination are presented. Then, the nominal syntax and morphology are 
introduced,  followed by the verbal syntax and morphology. Finally, elements of 
grammatical pragmatics and language usage phenomena are examined.  The work is 
descriptive and employs theory grounded in traditional grammar but  influenced 
by  recent work in linguistic typology. When necessary, the relevant theoretical 
assumptions are introduced and the relevant literature provided. Recall that the 
full list of glossing tags is available on page \pageref{sec-ABB} and the 
glossing convention is discussed in Section \ref{sec:intro-outline}.

\section{Clause}
\label{sec:GRM-nom}
I assume that a clause is a  grammatical unit that can express a proposition. A  clause which 
can stand as a complete utterance is an independent clause. When 
a grammatically correct clause cannot stand on its own, it is dependent on  a 
main clause.  Three sorts of speech act are covered: a statement, a question, or 
a command. The former is encoded in a declarative clause (Section 
\ref{sec:GRM-decl-clause}), and the latter two are encoded in interrogative 
clauses (Section \ref{sec:GRM-interr-clause}) and imperative clauses  (Section 
\ref{sec:GRM-imper-clause}) respectively.  Constructions are treated as 
clause-types; constructions are persistent formal and semantic frames which are 
conventionalized and display both compositional and non-compositional 
characteristics. In this section  the components of the common and frequent 
independent  clauses and the major constructions encountered are presented.  In 
Section \ref{GRM-clause-coord-subord},  clause coordination and subordination 
are introduced. Section \ref{sec:GRM-adjuncts} covers the adjunct constituents
responsible for modifying a main predicate and the function of the postposition.



\subsection{Declarative clause}
\label{sec:GRM-decl-clause}

Statements are expressed by a series of declarative clause types. The structure
of most common clauses consists of  a simple predicate, one or two arguments and
an optional adjunct. This structure is represented in
(\ref{ex:GRM-clause-frame})


\ea\label{ex:GRM-clause-frame}
 {\sc s|a}  $+$ {\sc p} $\pm$ {\sc o} $\pm$ {\sc adj} 
\z

where the symbol $+$  requires for the presence of the element preceding
and following it,  whereas  $\pm$ means that the term following it may be
optional.
The
symbol {\sc s} stands for the subject of an intransitive clause,  {\sc a} 
for the subject of a transitive clause, {\sc p}  for the predicate,  {\sc
o}  for the object of a transitive clause, and finally {\sc adj}
stands for an adjunct to a clause. The main instantiations  of this common
structure are shown in  (\ref{ex:GRM-cl-fr-inst}).\footnote{The way I represent
the components of a clause is inspired by \citet[31]{Bonv88}.}


\ea\label{ex:GRM-cl-fr-inst}

\ea\label{ex:GRM-cl-fr-inst-s-p}
 {\sc s}  $+$ {\sc p} 
\ex\label{ex:GRM-cl-fr-inst-s-p-o}
 {\sc a}  $+$ {\sc p} $+$ {\sc o}
\ex\label{ex:GRM-cl-fr-inst-s-p-adj}
 {\sc s}  $+$ {\sc p}  $+$ {\sc adj} 
 \ex\label{ex:GRM-cl-fr-inst-s-p-o-adj}
 {\sc a}  $+$ {\sc p} $+$ {\sc o} $+$ {\sc adj} 

\z
\z


The predicate ({\sc p})  is represented by a verbal syntactic constituent ({\it
v}) whereas  the arguments ({\sc s, a, o}) are represented by nominal syntactic
constituents   ({\it n}).  The adjunct constituent  ({\sc adj}) may consist of 
words or phrases referring to time, location, manner of action, etc.  (see
Section \ref{sec:GRM-adjuncts} on adjunct types).  An
argument may be seen as core or peripheral.  The core
argument of an intransitive clause is realized in the subject position ({\sc
s}), which
precedes the predicate. 



 \ea
\begin{multicols}{2}

\ea\label{ex:GRM-core-S-A-O}{
\glll àfɪ́á díjōō.\\
 {\sc s}  {\sc p}\\
{\it n} {\it v}.{\sc foc}\\
\glt `Afia ate.'
}

\ex\label{ex:GRM-core-S-O}{
\glll àfɪ́á díjōō kɪ̀ŋkáŋ̀.\\
 {\sc s}  {\sc p} {\sc adj}\\
{\it n} {\it v}.{\sc foc}  {\it qual}\\
\glt `Afia ate a lot.'
}


\ex\label{ex:GRM-core-A-O}{
\glll àfɪ́á dí sɪ̀ɪ̀máá rā.\\
 {\sc a}  {\sc p}  {\sc o} {}\\
{\it n} {\it v} {\it n} {\sc foc}\\
\glt `Afia ate food.'
}

\ex\label{ex:GRM-core-A-O}{
\glll àfɪ́á dí sɪ̀ɪ̀máá  kɪ̀ŋkáŋ nà.\\
 {\sc a}  {\sc p}  {\sc o}  {\sc adj}  {}\\
{\it n} {\it v} {\it n}  {\it qual} {\sc foc}\\
\glt `Afia ate food a lot.'
}

\z 
\end{multicols}
 \z


The core arguments of a transitive clause are realized
in the subject ({\sc a}) and object ({\sc o}), the former preceding and the
latter following the predicate in their canonical positions. These
characteristics are illustrated in 
(\ref{ex:GRM-core-S-A-O}).\footnote{Focus ({\sc foc}) may be 
integrated into the verb or coded in a focus particle.  Section 
\ref{sec:focus-forms} presents  the
various forms the focus particle can take.}



%  \end{minipage}
% \vspace*{15pt}



% Nominal
% and verbal syntactic constituents are discussed in
% Section \ref{sec:GRM-nom} and \ref{sec:GRM-verbals} respectively, whereas
% adjuncts are presented in section
% \ref{sec:GRM-adverbs}.

The grammatical relations are primarily determined by
constituent order. Thus, the subject and object functions are not
morphologically
marked,  except that the subject pronouns in {\sc s} and {\sc a} positions  can 
have  strong or  weak forms (see Section \ref{sec:GRM-personal-pronouns}). This
is extraneous to the marking of grammatical functions but pertinent to the
emphasis put on  an  event's participant. A peripheral argument  consists of a
constituent foreign to the core predication, that is, an argument which is not
part of the core participant(s) typically attributed to a predicate.  


As peripheral argument,  an  adjunct  ({\sc adj}) may be realized by a single 
word or a phrase. Reference to space, manner, and time are the typical  
denotations of peripheral arguments.  Adjuncts will be discussed in brief here, 
but details are offered in Section \ref{sec:GRM-adjuncts}.  Adjuncts are 
optional to the main predication and can be added to both intransitive and 
transitive clauses, as shown in  (\ref{ex:vp26.12}), and (\ref{ex:GRM-core-S-O}) 
and (\ref{ex:GRM-core-A-O}) above (see Sections \ref{sec:SPA-blc}, 
\ref{sec:SPA-postp},  and \ref{sec:GRM-obl-phrase} for  discussions on the 
postposition).

\ea

\ea\label{ex:vp26.12}{\rm Manner expression  in intransitive clause}\\
\gll ʊ̀ ɲʊ̃́ã́ làɣá nɪ̀.\\
   {\psg} drink {\ideo} {\postp}\\
\glt  `He drank quickly.' 
% 
\ex\label{ex:vp26.13.}{\rm  Manner expression in transitive clause}\\
\gll ʊ̀ ɲʊ̃́ã́ à nɪ́ɪ́  làɣálàɣá nɪ̀.\\
   {\psg} drink {\art} water  {\ideo} {\postp}\\
\glt  `He drank the water quickly.' 

\z 
 \z



A variation of the prototype  clause in (\ref{ex:GRM-clause-frame}) is a
clause containing an additional core argument.  \citet[116]{Dixo10b} calls  a
clause which contains an
additional core argument, that is,  an extended argument (i.e. {\sc e}), an
{\it extended} (intransitive or transitive) clause. The
difference between an adjunct and an additional core argument is not a clear-cut
one;   still,   the locative phrase in (\ref{ex:GRM-add-arg-e}) is treated as
  an additional core argument of the predicate {\sls bile} `put'. In section
\ref{sec:GRM-obl-phrase}, I call a clause constituent whose semantics is
characterized by an  affected or effected object, although realized in a
postpositional phrase, an oblique object phrase. Thus, the extended argument
in (\ref{ex:GRM-add-arg-e}) should be treated as an oblique object. 


\ea\label{ex:GRM-add-arg-e}{{\sc a} $+$ {\sc p}  $+$  {\sc o} $+$   {\sc e}}\\
\glll ŋmɛ́ŋtɛ́l {sìì à bìlè}  {ʊ̀  kùó}  {tìwìzéŋ nʊ̀ã̀  
nɪ̄}.\\
{\sc a} {\sc p}   {\sc o}   {\sc e}\\
spider {raise.up   {\conn} put} {{3.\sg.\poss}   farm } 
{road.large  {\reln}  {\postp}}\\

\glt  `Spider went establish his farm by a main road.' (LB 003)

\z

For the remaining, a ditransitive clause consists of a transitive clause with an
additional core argument.  In Chakali, the verb {\sls tɪɛ} `give', a predicate 
that conceptually implies both a Recipient (R)  and a Theme (T) and is typically
associated with ditransitive clauses, restricts its (right-) adjacent argument
in object position as  beneficiary of the situation. The thing transfered
(i.e. Theme) can never follow the verb if the beneficiary of the transfer
(Recipient) is realized. This is shown in (\ref{ex:GRM-arg-e-ditrans}).

\ea\label{ex:GRM-arg-e-ditrans}

 \ea\label{ex:GRM-arg-e-ditrans-ben-the-1}
\glll kàlá tɪ́ɛ́ àfɪ́á {à lɔ́ɔ́lɪ̀}.\\
{\sc a} {\sc p} {\sc o}$_{R}$ {\sc e}$_{T}$\\
K. give A.  {{\art} car}\\

\glt  `Kala gave Afia the car.' 

 \ex\label{ex:GRM-arg-e-ditrans-ben-the-2}
\glll  kàlá tɪ́ɛ́ ʊ̄  {à lɔ́ɔ́lɪ̀}.\\
{\sc a} {\sc p} {\sc o}$_{R}$ {\sc e}$_{T}$\\
K. give {\sc 3.sg}  {{\art} car}\\

  `Kala gave her  the car.' 

 \ex\label{ex:GRM-arg-e-ditrans-the-ben-1}
 *Kala tɪɛ a lɔɔlɪ Afia 
 \ex\label{ex:GRM-arg-e-ditrans-the-ben-2}
*Kala tɪɛ ʊ Afia 

\z 
 \z

The assumption is that the verb {\sls tɪɛ} `give'  is transitive and its
extended argument is always the tranfered entity (i.e.
Theme) in a ditransitive clause. This is supported by the extensive use of the 
{\it manipulative serial verb construction} (see Section
\ref{sec:GRM-multi-verb-clause}), used as an alternative strategy,  in order to
express transfer of
possession  and information.



\ea\label{ex:GRM-m-svc-give}
\glll  kàlá kpá  {à lɔ́ɔ́rɪ̀ / ʊ̄} tɪ̀ɛ̀ áfɪ́á.\\
{\sc a} {\sc p} {\sc o}$_{T}$  {\sc p}  {\sc o}$_{R}$\\
K. take  {{\art} car / 3.\sg} give A.\\

\glt  `Kala gave  the car/it to Afia.' ({\it lit.} Kala take the car/it give
Afia.)
\z

The extended argument in sentence (\ref{ex:GRM-arg-e-ditrans-ben-the-1})  and
(\ref{ex:GRM-arg-e-ditrans-ben-the-2})  above  is the \is{Theme}Theme argument 
of the verb
{\sls kpa} `take'   in a \is{serial verb construction}serial verb construction  
 
in
(\ref{ex:GRM-m-svc-give}).   Ditransitive clauses are
very rare in the text corpus despite their grammaticality.  Multi-verb
clauses, which are discussed in Section \ref{sec:GRM-multi-verb-clause},
may offer  better strategies to arrange arguments and predicates than
ditransitive clauses. The following subsections present various clause types and
constructions which are based on the declarative clause structure introduced
above.  





\subsubsection{Identificational clause}
\label{sec:GRM-ident-cl}


An identificational clause can express generic and ordinary categorizations, or
assert the identity  of two expressions. Generic categorization involves the 
classification of a subset to a set (e.g. Farmers are hard-working),
whereas an ordinary categorization holds between a specific entity and a generic
set  (e.g.  Wusa is a farmer). The clause can assert the identity of the
referents of two specific entities, a clause also known as equative or identity
(e.g. Wusa is the farmer). The examples in
(\ref{ex:GRM-ident-cl}) illustrate the
distinctions. 

\ea\label{ex:GRM-ident-cl}

\ea\label{ex:GRM-ident-gen-cat}{\rm Generic categorization}\\
\gll
 bɔ̀là jáá kɔ̀sásēl lē\\
 elephant {\ident}  bush.animal {\foc}\\
\glt `The/An elephant is a bush animal.'

\ex\label{ex:GRM-ident-ord-cat}{\rm Ordinary categorization}\\
\gll
wʊ̀sá jáá pápátá rá\\
W. {\ident} farmer {\foc}\\
\glt `Wusa is a farmer.'

\ex\label{ex:GRM-ident-tk-id}{\rm Identity}\\

\ea
\gll
wʊ̀sá jáá à tɔ́ɔ̀tɪ̀ɪ̀ná\\
W. {\ident} {\art} landlord\\
\glt `Wusa is the landlord.'
\ex
\gll
wʊ̀sá jáá  à báàl tɪ̀ŋ ká sáŋɛ̃̄ɛ̃̄ kéŋ̀\\
W. {\ident} {\art} man {\art} {\egr} sit.{\pfv} {\advm}\\
\glt `Wusa is the man sitting like this.'
\ex
\gll
à báàl tɪ̀ŋ kà sáŋɛ̃̄ɛ̃̄ kéŋ̀  jáá wʊ̀sá\\
 {\art} man {\art} {\egr} sit.{\pfv} {\advm} {\ident}   W.\\
\glt `The man sitting like this is Wusa.'


%\ex\label{ex:GRM-ident-}{\it }

\z 
\z
 \z

The verb {\sls jaa}  ({\it gl.} {\ident}) always  occurs between two nominal  
expressions,  and, as shown in the last two examples in 
(\ref{ex:GRM-ident-tk-id}),  their order  does not matter, except for the 
generic categorization where the order is always [{\it hyponym} {\sls jaa} {\it 
hypernym}].  So,  the sentences {\sls papata ra jaa  wʊsa} `farmer {\sc foc} 
is Wusa' and 
{\it 
a 
tɔɔtɪɪna  jaa  wʊsa} `landlord {\sc foc} is Wusa'   are  as acceptable as in 
the 
order given in (\ref{ex:GRM-ident-ord-cat}) and the first example in 
(\ref{ex:GRM-ident-tk-id}). 


%nin na 
%re keng

\subsubsection{Existential clause} 
\label{sec:GRM-loc-cl}

One type of existential clause is the basic  locative construction, which is
described in Section \ref{sec:SPA-blc}. Its
two main characteristics are the obligatory presence of the postposition {\sls 
nɪ},  which signals that the phrase contains the conceptual ground, and the
presence of a locative predicate or the general existential predicate {\sls 
dʊa}. An example is provided in (\ref{ex:GRM-loc-cl}).

\ea\label{ex:GRM-loc-cl}{\rm Basic Locative construction}\\
\gll à báál dʊ́ɔ́ à dɪ̀à nɪ̄.\\
{\art}  man be.at {\art} house {\postp}\\
 \glt  `The man is at/in the house.'
\z

The existential predicate {\sls dʊa} is glossed `be at', but it is not the case
that it is only used in spatial description. For instance,  adhering to a
religion may be expressed using the existential predicate {\sls dʊa} and the
postposition {\sls nɪ}, e.g.  {\sls ʊ̀ dʊ́á jàrɪ́ɪ́ nɪ̄} `he/she is a 
Muslim', 
even
though no space reference is involved in such an utterance. 

An existential clause is also used in order to express that something is at
hand, accessible or obtainable. The clause in (\ref{ex:GRM-avail-cl}) is called
here 
the availability construction. It slightly differs from the
locative
construction in (\ref{ex:GRM-no-avail-cl}) because of  the absence of the
postposition
{\sls nɪ}.

\ea\label{ex:GRM-avail-vs-loc}

\ea\label{ex:GRM-avail-cl}{\rm Availability construction}\\

\gll à mòlèbíí dʊ́á dé.\\
{\art}  money be.at {\dem}\\
\glt  `There is money (available).'

\ex\label{ex:GRM-no-avail-cl}
\gll à mòlèbíí dʊ̄ā dé nɪ̀\\
{\art}  money be.at {\dem}  {\postp}\\
\glt `The money is there.'

\z 
 \z


Another use is the attribution of a property ascribed to a participant. The
example in (\ref{ex:GRM-loc-propascr}) reads literally `a sickness is at Wojo', 
i.e. a person named Wojo is sick.  In addition to the clause presented in
(\ref{ex:GRM-loc-propascr}), ascribed property may also be conveyed in a
possessive clause (see Section \ref{sec:GRM-poss-cl}). 


\ea\label{ex:GRM-loc-propascr}
\gll gàràgá dʊ́á wòjò nɪ̄.\\
sickness be.at W. {\postp}\\
\glt  `Wojo is sick.'
\z

 The
verb {\sls dʊa} is the only verb with an allolexe (i.e. a combinatorial 
variant) used only in the 
negative.
Consider (\ref{ex:GRM-allolexe}).

\ea\label{ex:GRM-allolexe}

\ea\label{ex:GRM-allolexe-pos}
\gll ʊ̀  dʊ́á dɪ̀à nɪ̄.\\
{\sc 3.sg} be.at house {\postp}\\
\glt  `She is in the house.'

\ex\label{ex:GRM-allolexe-neg}
\gll ʊ̀  wáá tùwò dɪ̀à nɪ̀.\\
{\sc 3.sg} {\neg} {\neg}.be.at house {\postp}\\
\glt  `She is not in the house.'


\ex\label{ex:GRM-allolexe-pos-out}
 \textasteriskcentered ʊ  tuwo dɪa nɪ
\ex\label{ex:GRM-allolexe-neg-out}
 \textasteriskcentered ʊ  wa dʊa dɪa nɪ

\z 
 \z



\subsubsection{Possessive clause}
\label{sec:GRM-poss-cl}

A possessive clause expresses a relation between  a
possessor and a possessed.  Generally,  the  {\it
have-}construction  is used to convey a possessive relation. It consists of
the verb {\sls kpaga} `have',  and two nominal expressions acting as subject and
object; the former being the possessor (\psor) of the relation, while  the
latter being  the possessed
(\psed).

\ea\label{ex:GRM-poss-have}
\glll kàlá kpágá nã̀ɔ̃̀ rā.\\
K. have cow {\foc}\\
  {\psor} {}   {\psed} {}\\
\glt  `Kala has a cow'
\z

Example (\ref{ex:GRM-poss-have}) says that an animate alienable possession
relates  Kala (possessor) and a cow (possessed).  Since the  {\it
have-}construction does not encode animacy or alienability features,   staple
food can `have' lumps, i.e. {\sls kapala kpaga bie}, and someone can `have' a
senior brother, i.e. {\sls ʊ kpaga bɪɛrɪ}.  Abstract possession may also be
conveyed using the {\it have-}construction. In (\ref{ex:GRM-poss-have-abst}),
  shame, hunger,  thirst, and sickness are conceived as the possessors, the
possessed being the person experiencing these feelings. 



\ea\label{ex:GRM-poss-have-abst}

 \ea\label{ex:GRM-poss-have-abst-1}
\gll hɪ̃̀ɪ̃̀sá kpàgà   à   hã́ã̀ŋ kɪ̀ŋkáŋ̀.\\
shame    have {\art}   woman much\\
\glt `The woman was ashamed ...' (CB 034)
\ex\label{ex:GRM-poss-have-abst-2}
\gll lʊ̀sá kpágáń̩ nà.\\
hunger have.{1.\sg} {\foc}\\
\glt `I am hungry.'
\ex\label{ex:GRM-poss-have-abst-3}
\gll  nɪ́ɪ́ɲɔ̀ksá kpágán̩ nà\\
thirst have.{1.\sg} {\foc}\\
\glt  `I am thirsty.'
\ex\label{ex:GRM-poss-have-abst-4}
\gll  gàràgá kpágán̩ nà\\
sickness have.{1.\sg} {\foc}\\
\glt  `I am sick.'

\z 
 \z

Some characteristics ascribed to animate entitites are expressed by  the
word {\sls tɪɪna} `owner' following the possessed.  However it
is an existential clause (\ref{ex:GRM-poss-owner-exist}), rather than the  {\it
have-}construction, which carries the
possessive phrase {\sc psed}{\sls -tɪɪna}, as (\ref{ex:GRM-poss-owner})
illustrates.



\ea\label{ex:GRM-poss-owner}

 \ea\label{ex:GRM-poss-owner-exist}
\glll ʊ̀ jáá sísɪ́ámà-tɪ́ɪ́ná.\\
{3.\sg} {\ident} seriousness-owner\\
  {\psor} {}   {\psed}\\
\glt `He is serious'

 \ex\label{ex:GRM-poss-owner-have}
\gll ʊ̀ kpágá sísɪ́ámà rá.\\
{3.\sg} have {seriousness} {\foc}\\
\glt `He is serious'

\z 
 \z

Another way to express possession is by using a non-verb clause which
  exclusively identifies   the possessor. For instance, a speaker may utter 
{\sls 
mɪ́n nà} `it is
mine' in order to say that a
certain thing belongs to him or her. This utterance consists solely of the third
singular strong pronoun followed by
the focus particle (see Section \ref{sec:GRM-pronouns} on pronouns).


\subsubsection{Non-verb clause}
\label{sec:GRM-noverb}

As its name suggests, a non-verb clause is a clause without verbal elements. 
Its
main function is to identify or assert the (non-) existence of 
something.  The examples in (\ref{ex:GRM-noverb}) assert the (non-) existence
of a
referent with a single nominal expression, followed by the focus particle in
the affirmative and the negative particle in the negative (see Section
\ref{sec:GRM-foc-neg} on focus and negation). 


\begin{multicols}{2}
\ea\label{ex:GRM-noverb}
 \ea\label{ex:GRM-noverb-aff-1}
\gll fʊ́n ná.\\
knife {\foc}\\
\glt `It is a shaving knife.'
 \ex\label{ex:GRM-noverb-aff-poss}
\gll ǹ̩ fʊ́n ná.\\
{\sc 1.sg.poss} knife {\foc}\\
 \glt `It is my shaving knife.'
 \ex\label{ex:GRM-noverb-neg-1}
\gll fʊ́n lɛ̀ɪ́.\\
knife {\neg}\\
 \glt `It is not a shaving knife.'
 \ex\label{ex:GRM-noverb-neg-poss}
\gll ǹ̩  fʊ́n lɛ̀ɪ́.\\
{\sc 1.sg.poss} knife {\neg}\\
 \glt `It is not my shaving knife.'

\z 
 \z
\end{multicols}

Correspondingly the manner deictics {\sls keŋ} and {\sls  nɪŋ} are also found 
in 
non-verb
clause. For instance, {\sls kéŋ né} means `That is it!', but the same string 
is
more often heard as {\sls kéŋ nȅȅ} `Is that so/it?',  i.e.
constructed  as a polar question (see Section \ref{sec:GRM-interr-polar} on 
polar questions, and Section \ref{sec:GRM-adv-pro}  on  {\sls keŋ} 
and
{\sls  nɪŋ}).


%see dakubu p22 The sentence is a two-argument topic-comment proposition with no
%verb.

\subsubsection{Multi-verb clause}
\label{sec:GRM-multi-verb-clause}



A multi-verb clause is a clause containing more than one verb. The main type of
multi-verb clause is the serial verb construction (SVC), the definition of which
is still subject to contention. Let us start by stating that the SVC in Chakali
has the following properties: (i) a SVC is a sequence of verbs which act
together as a single predicate, (ii) each verb in the series could make up a
predicate on its own, (iii)  no connectives  surface (coordination or
subordination), (iv)  tense, aspect, mood, and/or polarity are marked only once,
(v)  a verb involved in a SVC may be formally shortened,  (vi)  transitivity is
common to the series, so arguments are shared (one argument obligatorily), (vii)
the verbs in the series are not necessarily contiguous, and  (viii) the grammar
does not limit the number of verbs. These characteristics are not uncommon for 
SVCs in West-Africa \citep{Amek05a}. In this section, the SVC in Chakali is
identified using representative examples. 


Even though the construction has more than one
verb, it describes a single event and does not contain  markers of
subordination or coordination. The first sequence of verbs in
(\ref{ex:GRM-mvc-svc}) illustrates the phenomenon.



\ea\label{ex:GRM-mvc-svc}
\glll à  kɪ̀rɪ̀nsá   m̩̀   màsɪ̀   kpʊ́ àká dʊ̀gʊ̀nɪ̀ tá.\\
{\art}  tsetse.fly.{\pl}   {1.\sg}    beat   kill   {\conn} 
chase   let.free\\
 {} {} {}  {} [{\it v} {\it v}]  {} [{\it v} {\it v}]\\
\glt `I beat and killed the tsetse flies, and drove them away.' (CB 023)
\z

Together,  the verbs {\sls masɪ} `beat' and  {\sls kpʊ} `kill'  in 
(\ref{ex:GRM-mvc-svc})  constitute a single event.  The same can be said about 
the verbs {\sls dʊgʊnɪ} `chase' and {\sls ta} `let free' in the second clause
following the connective.   If the clause following the connective   {\sls aka}
lacks a subject,  the subject of the preceding clause shares its reference in
the two clauses   (see Section \ref{GRM-clause-coord-ka-aka} on the connective 
{\sls aka}). What we have in (\ref{ex:GRM-mvc-svc}) is one SVC separated from
another multi-verb clause by the connective {\sls aka},  and the three verbs 
{\sls 
masɪ},  {\sls kpʊ} and {\sls dʊgʊnɪ}  share the reference of the  nominal {\sls 
a
kɪrɪnsa} `the tsetse flies' as their Theme argument and {\sls m̩̀} as their 
Agent
argument, i.e. {\sc o} and {\sc s} respectively. The role of  the verb {\sls ta}
in the sentence depicted in  (\ref{ex:GRM-mvc-svc}) is discussed at the end of
this section.

Tense/aspect (\ref{ex:GRM-svc-tense}), mood (\ref{ex:GRM-svc-aspect}), and/or
polarity value (\ref{ex:GRM-svc-negation}) are marked only once, usually with
preverb particles. This means that they are not repeated for each verb of which
a predicate is composed. The preverb particles are discussed in Section
\ref{sec:GRM-precerv}.


\ea\label{ex:GRM-svc-preverb}
 \ea\label{ex:GRM-svc-tense}{
\gll  ǹ̩ tʃɪ́ kàá màsɪ̀   kpʊ́   à  kɪ̀rɪ̀nsá rá.\\
{1.\sg} {\cras} {\fut.\prog}   beat   kill {\art} tsetse.fly.{\pl} {\sc foc} 
\\ 
\glt `I will be beating and killing the tsetse flies tomorrow.'
}
 \ex\label{ex:GRM-svc-aspect}{
\gll  ǹ̩  há màsɪ̀   kpʊ́   à  kɪ̀rɪ̀nsá rá.\\
  {1.\sg}  {\mod}   beat   kill {\art} tsetse.fly.{\pl} {\sc foc}\\ 
 \glt `I am still beating  and killing the tsetse flies.'
 }
 \ex\label{ex:GRM-svc-negation}{
\gll  ǹ̩   wà másɪ́   kpʊ́   à  kɪ̀rɪ̀nsá.\\
  {1.\sg}  {\neg}   beat   kill {\art} tsetse.fly.{\pl}\\
\glt `I did not beat and kill the tsetse flies.'
 }

\z 
 \z
%wàá will not

SVCs must share at least one core  argument. The example 
(\ref{ex:GRM-arg-sh-objsubj}) is an instance of argument sharing: the two verbs
in the construction share the (referent of the) noun {\sls foto} `picture' and
are not contiguous. The
transitive verb {\sls tawa} `pierce' takes  {\sls foto} as its object, whereas 
{\sls laga} takes  {\sls foto} as its subject. A representation of 
object-subject
sharing (or switch sharing) appears under the free translation in
(\ref{ex:GRM-arg-sh-objsubj}).
%\footnote{The label {\it object-subject sharing}
%is borrowed from \citet[20]{Osam03}.} 

\ea\label{ex:GRM-arg-sh-objsubj}{\rm Object-subject sharing}\\
\glll  hɛ̀mbɪ́ɪ́ táwá fótò làgà dáá nɪ́.\\
nail pierce picture hang wood  {\postp}\\
{} {\it v} {}  {\it v}\\
 \glt `A picture hangs from a nail on a wooden pole.'

{\it foto} $<x>$\\
{\it tawa} $<${\sc a}$ =  y$ ,  {\sc o}$=x$  $> $\\
{\it laga} $<${\sc a}$ = x$ , {\sc o} $= z $  $ >$\\
\z

%consistent with identifying function



Subject-subject and object-object sharing are more common than object-subject
sharing. In example
(\ref{ex:GRM-mvc-svc}), which is repeated below, the nominal expression {\sls a 
 kɪrɪnsa} is the shared object of three verbs, i.e. {\sls masɪ}, {\sls kpʊ} and 
{\sls dʊgʊnɪ}, whereas the pronoun {\sls m̩} is the shared subject for the same 
three
verbs. However, only {\sls masɪ} and {\sls kpʊ}  make up the SVC. 

\begin{exe}


\ex\label{ex:GRM-mvc-svc}{\rm Subject-subject and Object-object sharing}\\
\gll à  kɪ̀rɪ̀nsá   m̩̀   màsɪ̀   kpʊ́ àká dʊ̀gʊ̀nɪ̀ tá.\\
{\art}  tsetse.fly.{\pl}   {1.\sg}    beat   kill   {\conn} 
chase   let.free\\
\glt `I beat and killed the tsetse flies, and drove them away.'

{\it m̩} $<x>$\\
{\it kɪrɪnsa} $<y>$\\
{\it masɪ} $<${\sc a}$ =  x$ ,  {\sc o}$=y$  $> $\\
%{\it  kpʊ} $<${\sc subj}$ = x$ , {\sc obj} $= y $  $ >$\\
{\it dʊgʊnɪ} $<${\sc a}$ = x$ , {\sc o} $= y $  $ >$\\
 
\end{exe}

SVCs often involve two verbs, but there can be three or more verbs involved. 
Examples of three-verb and four-verb sequences are given in
(\ref{ex:GRM-mvc-3-4}). Each of the verbs involved can otherwise act alone as
main
predicate. Notice that the free translations provided do not accommodate well
the idea that
the two examples in (\ref{ex:GRM-mvc-3-4}) are conceived as single event.
In Section \ref{GRM-clause-coord-subord},  it will be shown that connectives
are usually present  when one wishes to distinguish events.


\ea\label{ex:GRM-mvc-3-4}

\ea{
\glll ʊ̀ síí kààlɪ̄ nà.\\
{3.\sg} rise go see\\
{}   {\it v}$_{1}$  {\it v}$_{2}$  {\it v}$_{3}$\\
\glt {\it lit.}  `She stood, went, and saw'
}

\ex{
\glll ʊ̀ brá tùù tʃɔ́ kààlɪ̀.\\
{3.\sg} turn go.down run go\\
{} {\it v}$_{1}$  {\it v}$_{2}$  {\it v}$_{3}$  {\it v}$_{4}$\\
\glt `She return down and ran away' (from a tree top or hill)
}

\z 
 \z

A  manipulative serial verb construction \citep[378]{Amek06} is a SVC
which  expresses a transfer of possession (e.g. give, bring, put)  or  
information (e.g. tell). It consists of the verb {\sls kpa} `take' and another
verb following it. The example in (\ref{ex:GRM-m-svc-give}), repeated in 
(\ref{ex:GRM-m-svc-give-1}), 
illustrates a transfer of possession. 

\begin{exe}

\ex\label{ex:GRM-m-svc-give-1}{\rm Manipulative serial verb construction}\\
%{\it Manipulative serial verb construction}\\
\glll  kàlá kpá  {à lɔ́ɔ́lɪ̀ / ʊ̄} tɪ̀ɛ̀ áfɪ́á.\\
K. take  {{\art} car / 3.\sg} give A.\\
{} {\it v} {}  {\it v} {}\\
\glt  `Kala gave the car/it to Afia' ({\it lit.} Kala take the car give Afia.)
 
\end{exe}


Frequent co-locations of the type presented in (\ref{ex:GRM-m-svc-give}) are 
{\sls kpa wa}, {\it lit.}  take come,  `bring',  {\sls kpa kaalɪ}, {\it lit.} 
take 
go,
`send', {\sls kpa pɛ}, {\it lit.} take add,  `add', {\sls kpa ta}, {\it lit.} 
take
let free, `remove', {\sls kpa bile}, {\it lit.} take put,  `put (on)'  and 
{\sls 
kpa
dʊ}, {\it lit.} take put,  `put (in)'. The two verbs may or may not be
contiguous;  usually the Theme argument of the  verb {\sls kpa} `take'  is found
between the two verbs.






Finally, some multi-verb clauses are not  SVCs.  There are a few verbs
which
bear a
relation to the main predication and  contribute  aspects of the phase of
execution or scope of an event.\footnote{These verbs are similar 
to what \citet[108]{Bonv88}
calls {\it auxiliant}.} For instance, a {\it
terminative}  construction describes an event coming to an end or reaching a
termination, and  a {\it relinquishment} construction describes an event whose
result is the release or abandonment of someone or something.  The verbs {\sls 
peti}
`finish' and {\sls ta} `abandon' in (\ref{ex:GRM-mvc-pha-3.1}) and
(\ref{ex:GRM-mvc-pha-help}), together with a non-stative predication, determine
each construction. 



\ea\label{ex:GRM-mvc-phase}

\ea\label{ex:GRM-mvc-pha-3.1} {\rm Terminative construction}\\
\glll làɣálàɣá hán nɪ̀ ǹ̩ kʊ̀tɪ̀ à ʔã́ã́ pétí.\\
{\advt} {\dem} {\postp} {1.\sg} {skin} {\art} bushbuck  finish\\
{} {} {} {}  {\it v} {} {} {\it v}\\
\glt `I  just finished skinning the bushbuck.'

\ex\label{ex:GRM-mvc-pha-40.3}
\gll  m̩̀ pétì à tʊ́má rá.\\
{1.\sg} finish {\art} work {\foc}\\
\glt `I have finished the work.'


\ex\label{ex:GRM-mvc-pha-help}{\rm Relinquishment construction}\\
\glll  kpá ǹ̩ néŋ tà.\\
take {1.\sg} hand let.free\\
 {\it v} {}  {} {\it v}\\
\glt `Leave my hand' (Let me go!)

\ex\label{ex:GRM-mvc-pha-relish} 
\gll  à bʊ̃́ʊ̃́ŋ tá ʊ̀ʊ̀ bìè rē.\\
{\art} goat abandon {3.\sg.\poss} child {\foc}\\
\glt `The goat abandoned its kids.'

\z 
 \z

The examples  in (\ref{ex:GRM-mvc-pha-3.1}) and (\ref{ex:GRM-mvc-pha-help}),
which may be called  {\it phasal  constructions},\footnote{The analysis of the
progressive and prospective in Ewe and Dangme in \citet{Amek08} influences the
way I approach and name the phenomenon.}  are treated as multi-verb clauses
since the predication is expressed with more than one verb. Yet, they are not
SVCs because the second verb in each example only specifies aspects of the
process
of the event  and does not contribute to the main predication as verb sequences
in SVCs do. Nonetheless, these verbs can function otherwise as main predicate,
as shown in (\ref{ex:GRM-mvc-pha-40.3}) and (\ref{ex:GRM-mvc-pha-relish}).
Similarly, the verb {\sls baga} `attempt to no avail'  conveys
nonachievement, e.g. {\sls ʊ buure kisie baɣa}, {\it lit.} he look.for knife 
fail,
`he did not find the knife',  and the verb {\sls na} `see' conveys confirmation 
or
verification, e.g. {\sls sʊɔrɛ dɪsa na}, {\it lit.} smell soup see, `smell the
soup'. Going back to example (\ref{ex:GRM-mvc-svc}) above, the verb {\sls ta}
contributes to a {\it relinquishment} multi-verb construction, similar to
(\ref{ex:GRM-mvc-pha-help}) above, and not to a SVC. 
 
% 

\subsubsection{Basic locative construction}
\label{sec:SPA-blc}


The \is{basic locative construction} basic locative construction  of a language 
is  the prototypical  and predominant construction used to locate a figure with 
respect to a ground \citep[15]{Levi06}. In Chakali it resembles the 
construction given in (\ref{ex:PSPV4}), although some sentences produced in 
elicitation contexts appear with the focus particle following the 
postposition {\sls nɪ}.  The \is{focus particle}focus particle  is a pragmatic 
marker which identifies for the addressee the topical subject (i.e. may be 
distinct from the grammatical subject) and does not convey locative meaning 
(Section \ref{sec:GRM-focus}). The focus particle will be ignored in the 
discussion. The third  line  in (\ref{ex:PSPV4}) associates parts of the 
sentence with a `conceptual level'. On that line one can find notions such as 
{\it figure} and {\it ground},  and \textsc{trm}, which stands for 
\is{topological relation marker} topological relation marker 
\citep[see][]{Brin12}. These are the linguistic expressions which convey  the 
spatial relationships in Chakali.  The figure {\sls a gar} `the cloth'  
functions as subject and the phrase {\sls a teebul ɲuu nɪ} `on the table' 
functions as  oblique object  of the main predicate. The last line is a free 
translation which captures  the general meaning of the situation. It is 
accompanied by a reference to the illustration which the first line 
describes.\footnote{See footnote \ref{ft:spatial-corpus} for information on the 
relevant corpus.}


\begin{exe}
\ex\label{ex:PSPV4}
\glll  [à gár] [ságá] [à {téébùl ɲúù} nɪ̀].\\
{\art} cloth be {\art}  {table {\reln}}  {\postp}\\
 \textit{figure} {} \textsc{trm}  {} {\textit{ground}+\textsc{trm}}  {} \\
% \textsc{s}   \textsc{pred} \textsc{obl}\\
\glt `The cloth is on the table.' (PSPV 4)
\end{exe}


In (\ref{ex:PSPV4}), the spatial relation is expressed via three topological
relation markers, that is,  the main predicate {\sls saga} `be on' or `sit', the
postposition  {\sls nɪ} and the relational nominal predicate {\sls ɲuu} `top 
of'.
The main predicate  {\sls saga}  denotes a stative event which  localizes the
figure with respect to the ground.  The postposition  {\sls nɪ} has no other
function than to signal that the oblique object is a locative phrase. The
relational nominal predicate {\sls ɲuu} designates the search domain and depends
on the reference entity of the ground (i.e. {\sls teebul}). The latter 
two topological
relation markers are discussed in more detail in Section 
\ref{sec:SPA-postp} and  \ref{sec:SPA-relnoun}.



\subsubsection{Comparative construction}
\label{sec:GRM-compar-ct}

A comparative construction has the semantic function of assigning a graded
position on a predicative scale to two (possibly complex) objects.
The comparative construction of inaquality can be expressed with the
transitive predicate {\sls kaalɪ} `exceed, surpass', whose  two arguments are
the objects compared.\footnote{\citet{Brin05} presents a lexical-functional
grammar (LFG)  account of the comparative construction in Gã, a language also
exhibiting an  `exceed'-  or `surpass'-comparative.}  One of the arguments 
represents the
standard
against which the other is
measured and found to be unequal.  The nominal expression in subject position is
the {\it comparee}, i.e. the objective of comparison, whereas the
one in object position is the {\it standard}, i.e. the object that
serves as yardstick for comparison \citep{Stas08}. The gradable
predicative scale is verbal and is normally adjacent to  the comparee, but may
be repeated adjacent to the standard. Given that both the scale and the
transitive predicate {\sls kaalɪ} are verbs, a comparative construction is  a
 type of multi-verb clause.  If the predicative scale is absent, as in
(\ref{ex:GRM-comp-tr-sca-abs}),  one
may still interpret the construction as a comparative one, in which case both
the
context
and the meaning of  the nominals involved would provide the property on which
the
comparison  is made. These characteristics are illustrated in
(\ref{ex:GRM-comp-tr}).


\ea\label{ex:GRM-comp-tr}{\rm Comparative transitive construction}\\
\ea\label{ex:GRM-comp-tr-sca-pres}{
\glll wʊ̀sáá zɪ́ŋá kààlɪ̀ áfɪ́á.\\
  W. grow surpass A.\\
[{\it n}]$_{comparee}$  [{\it v}]$_{scale}$ {\it v} [{\it n}]$_{standard}$\\
\glt `Wusa is taller than Afia.'
}

\ex\label{ex:GRM-comp-tr-sca-abs}{
\glll wʊ̀sá bàtʃɔ́lɪ́ káálɪ́ kàlá bàtʃɔ́lɪ́.\\
W.  running surpass K. running\\
[{\it n}  {\it n}]  {\it v} [{\it n} {\it n}]\\
\glt `Wusa's running is better/faster than Kala's running.'
}

\z 
 \z

Another way to compose a comparative construction of inequality is with the
identificational clause bounded with a postpositional phrase.  It is referred
to as a
comparative intransitive construction since the standard is not encoded in the
grammatical object of a transitive verb. Instead, the predicative scale is
embedded in a nominalized property following the identificational verb {\sls 
jaa}
(see Section \ref{sec:classifier} on classifiers).


\ea\label{ex:GRM}{\rm Comparative intransitive construction}\\

\glll wʊ̀sá jáá nɪ́hɪ̃̀ɛ̃̂ àfɪ̀á nɪ́.\\
W.  {\ident} old A. {\postp}\\
[{\it n}]$_{comparee}$   {\it v} [{\it v}]$_{scale}$  [{\it n}]$_{standard}$  
{}\\
\glt `Wusa is older than Afia.'
\z

The same  two strategies are used to
express a superlative degree: surpassing or being superior to all others is
explicitly expressed by a phrase containing the pronoun {\sls ba} `they, them'.
This
is shown in (\ref{ex:GRM-super}).


\ea\label{ex:GRM-super}{\rm Superlative construction}\\

\ea
\glll wʊ̀sá zɪ́ŋá kāālɪ́ bá.\\
W. grow surpass {\sc 3.pl}\\
{} {\it v} {\it v} {}\\
\glt  `Wusa is the tallest.'
\ex 
\gll wʊ̀sá jáá nɪ́hɪ̃̀ɛ̃̂ bà nɪ́.\\
W. {\ident} old {\sc 3.pl} {\postp}\\
\glt  `Wusa is the oldest.'

\z 
 \z

A comparison of equality (i.e. X is same as Y) consists of a subject
phrase containing both objects to be  compared joined by the  connective {\sls 
(a)nɪ} followed by the scale, the verb {\sls maase} `equal, enough, ever' and 
the
reciprocal word {\sls dɔŋa} `each other'  (see Section 
\ref{sec:GRM-recipro-reflex} on reciprocity
 and reflexivity). This is shown in (\ref{ex:GRM-comp-equal}).

\ea\label{ex:GRM-comp-equal}{\rm  Comparison of equality construction}\\

\gll wʊ̀sá nɪ́ àfɪ̄ā bɪ̀nsá máásé dɔ́ŋá rā.\\
W. {\conn} A.  year equal {\recp} {\foc}\\

\glt `Wusa is as old as Afia.'
\z


Finally,  the verb {\sls bɔ} in (\ref{ex:GRM-comp-verb}) is a comparative
transitive verb which can be translated with the English comparative adjective
and prepositon `better than'.


\ea\label{ex:GRM-comp-verb}
\glll zàáŋ tʊ́má bɔ́ dɪ̀àrè tɪ̀ŋ tʊ̄mā.\\
today work better.than yesterday {\art} work\\
{}  {}  {\it v} {} {}   {}\\
\glt `Today's work is better than yesterday's work'
\z


\subsubsection{Modal clause}
\label{sec:GRM-compar-ct}

A modal clause is a clause expressing  ability, possibility,   obligation, desire, 
etc. 


\paragraph{Ability-possibility}
\label{sec:GRM-ability-possibility}

An
ability-possibility construction is a clause containing the word {\sls kɪŋ} 
immediately preceding the main verb(s).  The construction conveys either
the
physical or mental
ability of something or someone, or probability or possibility under some
circumstances. The construction is more frequent in the negative, but affirming
an ability or possibility is also possible in the positive using the
construction. The word {\sls kɪŋ} is glossed  {\abl} to refer to
`ability'.\footnote{The word {\sls kɪŋ} has a  nominal homophone meaning 
`thing' (and a classifier derived from the noun, see Section
\ref{sec:classifier}).  Its distribution would suggest that it is a kind of 
preverb (Section \ref{sec:GRM-precerv}), although it seems premature to
categorize it.}

\ea
\label{ex:}
{\upshape Ability-Possibility construction}\\

\ea
\label{ex:GRM-modal-12.2}
\gll ʊ̀ wà kɪ́ŋ wàà.\\
{3.\sg} {\neg} {\abl} come\\
\glt  `He is not able to come.'

\ex  
\gll ɪ̀ kàá kɪ̀ŋ kààlʊ̄ʊ́.\\
 {2.\sg} {\fut} {\abl} go.{\foc}\\
\glt  `You may go.'

\ex\label{ex:GRM-modal-13.1}
\gll ǹ̩ kàá kɪ̀ŋ wàʊ̀ tʃȉȁ.\\
 {1.\sg}  {\fut} {\abl} come.{\foc} tomorrow\\
\glt `May I come tomorrow?'

\z
\z



The phrase {\sls a bɔnɪɛ̃ nɪ} `perhaps' is an adjunct phrase used when the
occurence of a situation  or an achievement  is in doubt. The word {\sls bɔnɪɛ̃}
is not
used in any other context in the text corpus or lexicon. Thus, the dubitative
modality construction is a construction marked by the presence of the phrase
{\sls a bɔnɪɛ̃ nɪ} clause-initially.



\ea
\label{ex:GRM-modal}{\rm Dubitative construction}\\

\ea
\label{ex:GRM-modal-45.5}
\gll {à bɔ́nɪ̃́ɛ̃́ nɪ́}  dʊ́ɔ́ŋ kàá wàʊ̄.\\
{\dub} rain {\fut} come.{\foc}\\
\glt  `Perhaps it is going to rain.'


\ex
\label{ex:GRM-modal-45.3}
\gll  {à bɔ́nɪ̃́ɛ̃́ nɪ́}   ʊ̀ dɪ̀ wááwáʊ́.\\
{\dub}  {3.\sg} {\hest} come.{\pfv.\foc}\\
\glt `Perhaps he came yesterday.'

\z
\z



In some contexts, a speaker may prefer  to use a cognitive verb in a phrase
like {\sls n̩ lisie} `I think (...)'  or
the phrase {\sls a kʊ̃ʊ̃ n̩ na}, {\it lit.} it tires me {\sc foc},  `I wonder
(...)' as alternative to the dubitative
construction. 



\paragraph{Desiderative}
\label{sec:GRM-desiderative mood}

As an independent verb {\sls ŋmá} means `say'. The same verb can also 
function in a construction [NP {\sls ŋma} [NP VP]]  conveying a desiderative 
mood,  corresponding to the English modal expression `want to'.



\ea\label{ex:dsdrtv-2}
\gll ŋ̀ ŋmá [ŋ́ káálɪ̀ dùsèè tʃɪ̄ā].\\
  {\sc 1.sg} say  {\sc 1.sg} go D. tomorrow\\
\glt  `I want to go to Ducie tomorrow.'
\z 

 
 Notice the high tone on the  {\sc 1.sg} pronoun subject  in (\ref{ex:dsdrtv-2}),  which suggests 
that the embedded clause is in the subjunctive mood (Section \ref{sec:GRM-subjunctive}).



\subsection{Interrogative clause}
\label{sec:GRM-interr-clause}

An interrogative clause consists either of a clause (i) with an initial
interrogative word/phrase (Section \ref{sec:GRM-interg-pro} on pro-form 
interrogatives), or (ii) with the absence of an initial interrogative
word but the presence of an extra-low tone at the end of the clause. The former
is called a `content' question and the latter a `polar' question. 

\subsubsection{Content question}
\label{sec:GRM-interr-content}

A content question contains an interrogative word/phrase whose typical position
is clause-initial. In (\ref{ex:GRM-inter-content}), {\sls bàáŋ} `what' 
replaces
the complement of the verb {\sls jaa}, whereas {\sls (a)àŋ́} `who'  replaces
the subject constituent of the clause. The inventory of interrogative
words/phrases can
be found in Section \ref{sec:GRM-interg-pro}.

\ea\label{ex:GRM-inter-content}

\ea\label{ex:GRM-inter-content-what}
\gll bááŋ ʊ̀ kàà jáà.\\
{\q} {3.\sg} {\ipfv} do\\
 \glt  `What is he doing?' 
\ex\label{ex:GRM-inter-content-who}
\gll àáŋ káá wáá báŋ̄.\\
{\q}  {\ipfv} come here\\
\glt  `Who is coming here?'

\z 
 \z

When an interrogative word/phrase cannot be located clause-initially,  it is
found at the canonical position of the constituent replaced. There are
rare cases, but in (\ref{ex:GRM-inter-content-who-rev-bear-in}), which is
equivalent to  (\ref{ex:GRM-inter-content-who-rev-bear-ex}),  the question
word {\sls aŋ} `who' appears in the object position following the transitive 
verb
{\sls maŋa} `beat' and is slightly lengthened. 


\ea\label{ex:GRM-inter-content-who-rev-bear}

\begin{multicols}{2}
 \ea\label{ex:GRM-inter-content-who-rev-bear-in}
\gll  zɪ̀ɛ́n máŋá àŋ́ŋ (?).\\
 Z. beat {\q}\\
 \glt  `Zien beat who?'

 \ex\label{ex:GRM-inter-content-who-rev-bear-ex}
\gll  àŋ́ zɪ̀ɛ̀n mȁŋȁ.\\
{\q} Z. beat\\
 \glt  `Who did Zien beat?'

\z 
\end{multicols}
 \z





\subsubsection{Polar question}
\label{sec:GRM-interr-polar}

A polar question is characterized by an interrogative intonation, consisting of
an extra-low tone at the end of the utterance. Additionally, lengthening  on the
penultimate vocalic segment takes place. The properties differentiating an
assertive clause from a polar question are illustrated in
(\ref{ex:GRM-inter-polar}). The
extra-low tone is represented with a double grave accent (i.e.  ̏). 

\ea\label{ex:GRM-inter-polar}{\rm Assertion vs. question}\\

\begin{multicols}{2}
\ea
\gll ʊ̀ wááʊ̀.\\
{3.\sg} come.{\ipfv .\foc}\\
\glt  `He is coming.'
\ex 
\gll ʊ̀ wááʊ̏ʊ̏.\\
{3.\sg} come.{\ipfv .\foc}\\
\glt `Is he coming?'%22.1.1

\z 
\end{multicols}
 \z

Possibly common to all Ghanaian languages, the  agreeing response
to a  negative polar interrogative
  takes into account the logical negation, as 
(\ref{ex:GRM-inter-polar-neg-rep}) illustrates. 



\ea\label{ex:GRM-inter-polar-neg-rep}

\begin{multicols}{2}
\ea\label{ex:GRM-inter-polar-neg-rep-S}{\rm Speaker}\\
\gll  ɪ̀ wàà kāālɪ̏ɪ̏.\\
{2\sg} {\neg} go.{\q}\\
\glt `Aren't you going?'

\ex\label{ex:GRM-inter-polar-neg-rep-A}{\rm Addressee}\\
\gll ɛ̃̀ɛ̃́ɛ̃̀.\\
yes\\
\glt `No' ({\it lit.} Yes, I am not going)


\z 
\end{multicols}
 \z

A negative polar interrogative in English usually asks about the
positive proposition, i.e. with `Aren't you going?' the speaker presupposes 
  that the addressee is going,   while in Chakali it questions the
negative proposition, i.e. with {\sls  ɪ̀ wàá káálɪ̏ɪ̏} the speaker's belief
is that the addressee is not going. That is probably why we get `yes' in Chakali
and `no' in English for a corresponding negative polar interrogative.



\subsection{Imperative clause}
\label{sec:GRM-imper-clause}

An imperative clause consists of a clause  expressing direct commands, 
requests, and prohibitions. It can be an exclusively addressee-oriented clause 
or  can include the speaker as well. This distinction, i.e. 
exclusive-inclusive, is rendered in  (\ref{ex:GRM-imperative-exc-inc}). In 
(\ref{ex:GRM-imperative-exc}) the  speaker excludes herself  from the 
performers of the action, i.e., only the adressee(s) is urge to perform the 
action,  while in  (\ref{ex:GRM-imperative-inc}) the speaker includes 
herself among the performers.

\ea\label{ex:GRM-imperative-exc-inc}

\ea\label{ex:GRM-imperative-exc}{\rm Exclusive}\\
\gll fùùrì à díŋ dʊ̀sɪ̀.\\
blow {\art} fire quench\\
\glt `Blow on this flame (to extinguish it).'

\ex\label{ex:GRM-imperative-inc}{\rm Inclusive}\\
\gll tɪ̀ɛ̀ jà mùŋ làɣàmɛ̀ kààlɪ̀ tɔ́ʊ́tɪ́ɪ́ná  pé.\\
give {1\pl} all gather go landlord end\\
\glt `Let's all go to the landlord  together.'

\z
\z

When an order is given directly to the addressee, as in 
(\ref{ex:GRM-imper-exc-var}), the clause may be introduced with the particle 
{\sls dɪ}. Some consultants believe that omitting the particle may be perceived 
as rude.  The particle {\sls dɪ} can also code the information that performing 
the action is requested by someone else than the speaker.\footnote{It is not 
known whether these `{\sls dɪ}-strategies' give rise to multiple 
interpretations}.  In addressing a command to a group the second person plural 
subject pronoun  usually appears in its canonical subject position, but it may 
be absent if the speaker believes that the context allows a single 
interpretation.\footnote{If A asks `What does he want?', B may reply {\sls dɪ́  
má dɪ́ wāā} `That you ({\it pl.}) should be coming'. In this case the first 
{\sls dɪ} heads a  clause which introduces indirect speech and the second is an 
imperfective particle,  the latter being covered in Section 
\ref{sec:GRM-ipfv-part}. }


\begin{multicols}{2}
 

\ea\label{ex:GRM-imper-exc-var}
\ea\label{ex:GRM-imper-exc-var-sg}
\gll (dɪ̀)  wàà.\\
 {\comp} come\\
\glt `Come!'

\ex\label{ex:GRM-imper}  
\gll dɪ̀ wáá.\\
 {\comp} come\\
 \glt `Come!' (keep coming! or follow me!)
 
\ex\label{ex:GRM-imper} 
\gll máá wáà.\\
 {\sc 2.pl} come\\
\glt `Come!'

\ex\label{ex:GRM-imper-exc-var-out}  
\gll dɪ́ máá wāā.\\
 {\comp}  {\sc 2.pl}  come\\
\glt `Come!' (requested by someone else than speaker)
\z
\z

\end{multicols}

Example (\ref{ex:GRM-hortative-vp11.3}) expresses a wish of the speaker and no
addressees are called for. Such a meaning is sometimes associated with 
\is{optative} optative
mood. Similarly but not  identically,  an utterance like the one in
(\ref{ex:GRM-hortativ-vp11.4})  assumes one or more addressees, yet the desired
state of affairs is not in the control of anyone in particular but of everyone. 
 As in (\ref{ex:GRM-imperative-inc}), the  strategy in both cases is
to use the verb {\sls tɪɛ} `give'.  


\ea\label{ex:GRM-hortative}
 
\ea\label{ex:GRM-hortative-vp11.3}{\rm Optative}\\
\gll tɪ̀ɛ̀ m̩̀ mɪ̀bʊ̀à bírgì.\\
    give {\sc 1.sg.poss} life delay\\
\glt  `Let me live long!' 

\ex\label{ex:GRM-hortativ-vp11.4}{\rm Hortative}\\
\gll tɪ̀ɛ̀ à gʊ̀à píílé.\\
  give {\art} dance start\\
\glt  `Let the dance begin!'

\z 
 \z

A prohibitive clause consists of  a negated proposition conveying an imperative 
(or hortative) mood. It is marked by the negative particle {\sls tɪ}/{\sls te} 
`not'   ({\it gl.} {\sc neg.imp}) occurring clause-initially.


\ea\label{GRM-neg-imp-vp15.10.}
\gll té káálíí, dʊ́ɔ́ŋ kàà wáʊ̀.\\
  {\neg.\imp} go rain ?{\ipfv}? come.{\foc}\\
\glt  `Don't go, it's going to rain.' 
\z
 
The prohibitive also involves a high front vowel  suffixed to its verb. The 
quality of the vowel, i.e. {\sls -ɪ}/{\sls -i}, is determined by the quality of 
the verbal stem.

\begin{multicols}{2}
\ea\label{ex:GRM-neg-imperative}
\ea 
\gll gó.\\
circle\\
\glt  `Move in a circle around.'
\ex 
\gll  té   góìí.\\
{\neg.\imp} circle\\
\glt  `Don't move in a circle around.'
\columnbreak
\ex  
\gll kpʊ́.\\
kill\\
\glt `Kill.'
\ex
\gll tɪ́ kpʊ́ɪ̀ɪ́.\\
{\neg.\imp} kill\\
\glt `Don't kill.'
\z 

 \z
 \end{multicols}

In addition, a distinction within the prohibitive can be made
between a prohibition (or advice) for a future situation  
(\ref{ex:GRM-neg-fut}), 
and  for an on-going situation (\ref{ex:GRM-neg-pres}). 


\begin{multicols}{2}
 
 

\ea\label{ex:GRM-neg-fut-pres}

 \ea\label{ex:GRM-neg-fut}
\gll  kʊ̀ɔ̀rɪ̀ sɪ̀ɪ̀mã́ã̀.\\
  make food\\
 \glt `Make food.'
 
 
 \ex\label{ex:GRM-neg-pres}
\gll  té kʊ́ɔ́rɪ́ sɪ̀ɪ̀mã́ã̀\\ 
{\sc neg}  make food\\
 \glt `Do not make food.' (addressee not in the process)
 \vfill
 \columnbreak
 
  \ex\label{ex:GRM-neg-pres}
  \gll  tɪ́ɪ́ kʊ̄ɔ̄rɪ̄ɪ̄.\\
 {\sc neg.imp}  make\\
 \glt `Do not make (food).' (addressee in the process of making)
   \ex\label{ex:GRM-neg-pres}
  \gll  tíí kʊ̄ɔ̄rɪ̀  sɪ̀ɪ̀mã́ã̀.\\
 {\sc neg.imp}  make  food\\
 \glt `Do not make food.' (addressee in the process of making)
\z
\z

\end{multicols}



\subsection{Exclamative clause}
\label{sec:GRM-excla-clause}

It is generally known that the difference between a declarative and an exclamative  clause is 
that the former is meant to be informative and the latter expressive. One criterion for determining 
the class of exclamative clause is the use of  exclamatory codas  \citep[242]{alla14}, i.e.  
exclamatory words or particles such as  {\sls woo} in (\ref{ex:GRM-excla-clause}), which modify the 
illocution of the clause and are usually found clause-finally.

 \ea\label{ex:GRM-excla-clause}
\ea\label{ex:GRM-excla-clause-1}
\gll  bɛ̀lɛ̀ɛ́ tɪ́ŋ mùŋ nè kéŋ wòòò \\
G. {\sc art} meaning {\sc foc} {\sc dxm}  {\sc interj}\\
\glt `‎‎That is the meaning of Gurumbele!' (Bele-history 1.16)

\ex\label{ex:GRM-excla-clause-2}
\gll dʊ̃́ʊ̃́ ɲú kpárá rá wōōō\\
python head double {\sc foc} {\sc interj}\\
\glt Python is double-headed wooo (so it can see more 
than us)!  (Python 1.74)

\z
\z


After offering a chronicle on the history of his village and the reason why it has its name, the 
speaker 
uttered (\ref{ex:GRM-excla-clause-1}) to intensify his stance in the presence of other community 
members. In  (\ref{ex:GRM-excla-clause-2}), the narrator of the folktale wants to mark the 
surprising fact that the python is equipped with extraordinary visual power.



\ea\label{ex:GRM-excla-clause-3-4}
\ea\label{ex:GRM-excla-clause-3}
\gll  ʊ̀ kà báŋ wà zú dìà, ʊ̀ bàŋ ŋmá dɪ́  ɛ̃̀hɛ̃́ɛ̃̄ɛ̃̀\\
{\sc 3.sg} {\sc ipfv} just come enter room  {\sc 3.sg} just say {\sc comp}  {\sc interj}  \\
\glt `When he entered  the room,  she said: ``yes!''' (Python 1.8)\\

\ex\label{ex:GRM-excla-clause-4}
\gll dɪ́ ʔábbā!, dɪ́  ʊ̀ʊ̀ bàmbíí nár wááwáʊ́\\
{\sc comp}  {\sc interj} {\sc comp}  {\sc 3.sg.poss} heart person come.{\sc pfv.foc}\\
\glt (Mother said) `Indeed,   her love has finally come.'  (Python 1.9)\\

\z
\z

In  (\ref{ex:GRM-excla-clause-3}), the  speaker is a mature girl who waited a long time and met 
many aspirants to finally encounter the right man to marry. In this context, the exclamative word  
{\sls ɛ̃hɛ̃}, which generally code a positive reaction (Section \ref{sec:GRM-greet}),  can be 
translated into `yes, exactly, finally' and paraphrased as  `this is the person I like'.   The 
speaker confirms that the man is the right one, with a strong emotional reaction,  allowing the 
addressee -- in this case her parents -- to know about her stance and feeling. The sequence in  
(\ref{ex:GRM-excla-clause-4})  is the reaction of her mother who confirms the daughter's reaction. 
Notice however  that the interjections {\sls ɛ̃hɛ̃}, {\sls ʔabba}, and {\sls woo} are not specific 
to Chakali: they are found in most, if not all, languages of the area.



\subsection{Clause coordination and subordination}
\label{GRM-clause-coord-subord}
%embedded clause, subordinate clause) cannot stand alone as a sentence

A relation between two clauses is signaled with or without a morpheme,  and 
various  structures and morphemes  are used to relate clauses.  Two
relations are discussed below: coordination and subordination. 

\subsubsection{Coordination}
\label{GRM-clause-coord}


%conjunct ka, aka, a
The distribution of four clausal connectives which are used in coordinating
clauses is presented: these are {\sls a}, {\sls ka}, {\sls aka} and {\sls 
dɪ}.\footnote{See \citet[143--149]{Mcgi99} for an account of similar clausal
connectives in Pasaale.}  


\paragraph{Connective {\sls a}}
\label{GRM-clause-coord-a}


The connective {\sls a} `and'  introduces a clause without overt subject.  When 
it is used between two clauses, the subject of the first clause must 
cross-refer to the covert subject of the second clause  (and subsequent 
clauses). It links a sequence of closely related events carried out by the same 
agent, and the events are encoded in  verb phrases denoting temporally distinct 
events. The example in (\ref{ex:GRM-coor-vp8.1}) is  an illustration of four 
consecutive clauses introduced by the connective  {\sls a}.   This phenomenon 
is often referred to as `clause chaining'. 

\ea\label{ex:GRM-coor-vp8.1}
\gll dɪ̀àrɪ̀ tɪ̀ŋ ǹ̩ dɪ̀ káálɪ́ bɛ̀lɛ̀ɛ̀ rá, à [jàwà nàmɪ̃̀ɛ̃́], à 
[kpá wàà dɪ̀á], à  [wà tɪ̀ɛ̀ ǹ̩ hã́ã̀ŋ], à [ŋmá tɪ̀ɛ̀ ǹ̩ hã́ã̀ŋ] 
dɪ́ ʊ́ʊ́ tɔ́ŋà. ʊ̀ tɔ̀ŋà jà dí\\ 
{\advt} {\art} {\sc 1.sg} {\hest} {go} {G.}  {\foc} {\conn}   buy meat
{\conn}  take come home {\conn}  come give my wife  {\conn}  say give my wife
{\comp}  {\sc 3.sg}   {cook} {\sc 3.sg}   {cook}  {\sc 1.pl} {eat}\\
\glt  `Yesterday I went to Gurumbele,  bought some meat, brought it
home to my wife, told her to cook it. She cooked and we ate.'
\z


\paragraph{Connectives {\sls ka} and {\sls aka}}
\label{GRM-clause-coord-ka-aka}

Generalizing from the examples available, for both the connectives {\sls ka} 
and 
{\sls aka}, either (i) the subject of the clause preceding the connective is 
inferred in the second clause, i.e. as for  the connective  {\sls a} above, or 
(ii) a different subject surfaces in the second clause. Each case is shown in 
(\ref{GRM-clause-conn-ka-1-subj}) and (\ref{GRM-clause-conn-(a)ka-2-subj}) 
respectively.   


\ea\label{GRM-clause-conn-ka-1-subj} 
\gll [ŋmɛ́ŋtɛ́l   láá nʊ̃̀ã̀  nɪ́] ká  [ŋmá dɪ́ ʊ́ʊ́  
wá  
ɲʊ̃̀ã̀ nɪ́ɪ́]\\
spider collect mouth {\postp}  {\conn} say {\comp}  
{\sc 3.sg}  come   drink water\\
\glt  `(Monkey went to spider's farm to greet him.) Spider accepted
(the
greetings) and (Spider) asked him (Monkey) to come and drink water.'  (LB 011)
 \z



\ea\label{GRM-clause-conn-(a)ka-2-subj} 

\ea\label{GRM-clause-conn-ka-2-subj} 
\gll  [dɪ̀  ɪ̀    wáà    párà]  ká [kìrìmá  wà 
dʊ́mɪ́ɪ́]\\
{\conn} {\sc 2.sg} {\ingr}  farm {\conn} 
tstse.fly.{\pl}  {\ingr}  bite.{\sc 2.sg}\\
\glt  `When  you are doing the weeding and  tsetse flies bite
you (...)' (CB 003) 

\ex\label{GRM-clause-conn-aka-2-subj} 
\gll  [dɪ́   námùŋ tɪ́  bɪ́ wàà   jɪ́rà kɪ̀ŋkùrùgíé
ŋmɛ́ŋtɛ́l sɔ́ŋ] àká [ɪ̀ jɪ̀rà   kéŋ̀]\\
{\comp}   anyone     {\neg}   {\itr} {\ingr}  
   call   enumeration    
eight name   {\conn}   {\sc 2.sg}   call  {\advm}\\
\glt  `(The monkey said:  ``They said) that anyone should not say the
number eight and you have called the number eight''.' (LB
017) 

\z 
 \z


Secondly, the connectives  {\sls ka} and {\sls aka} `and'  may encode a 
`logical' or `natural' sequence of events.   For instance, in 
(\ref{GRM-clause-conn-ka-1-subj}), someone traveling (or coming from the road) 
expects to be offered water to drink after the greetings are exchanged. 
Thirdly, the  connectives  {\sls ka} and {\sls aka} `and' 
 also suggest a causal relation between interdependent clauses. In 
(\ref{GRM-clause-conn-a}), it is the counting of the mounds which caused Spider
to be confused, which can be seen as an unexpected outcome.  

% % % but the left-hand conjunct in the {\sls  (a)ka}-construction in 
% % % (\ref{GRM-clause-conn-a})  behaves somehow  like a subordinate clause. Once 
% % % again, the fact that the subordinate causal  clause is a VP adjunct, thus a 
% % % syntactic constituent of the matrix sentence, explains its behaviour 
% % concerning % focus marking structures. The impossibility of ocusing the 
% second % segment of a % Justification construction through negation  reinforces 
% its  % syntactic % peripheral status.



\ea\label{GRM-clause-conn-a} 
\gll   ʊ́ʊ́wà  ŋmɛ́ŋtɛ́l   já  kùrò àkà bùtì\\
{\sc 3.sg.emph}   spider do count  {\conn}  confuse\\
\glt  `(Because) he himself (Spider) did count and he became confused'
(LB  007) 
 \z


Nevertheless the connectives  {\sls ka} and {\sls aka}  can 
introduce a clause denoting an event
which is not necessarily related to the event of the previous clause. It looks
as if  the connectives in (\ref{GRM-clause-conn-ka-then}) is used to
integrate an unrelated event to  the overall situation.   

\ea\label{GRM-clause-conn-ka-then} 
\gll [nànsá sú bárá múŋ̀.] ká [dʊ̃́ʊ̃́ tɪ̀ŋ ŋmá dɪ́ kɪ̀ndɪ́gɪ́ɪ́ 
dʊ́ɔ́ à dɪ̀ā nɪ́]\\
 meat fill place all  {\conn} python {\art} say {\comp} something is  {\art}
house {\postp}\\
\glt `Meat was all over the place. Then,  Python said: ``there is something in
the room''.' (PYTH??)
 \z

\ea\label{GRM-clause-conn-ka-transition} 
\gll  [à  bìpɔ̀lɪ́ɪ́  sìì     tʃɪ́ŋá]  àká   [ŋmá,  
ámɪɛ̃̀ɛ̃̀   ɪ̀      ɲɪ́ná] {...}\\
{\art} young.man   raise   stand {\conn} said,  {\adv}   {\sc 2.sg.poss} 
father  {...}\\
\glt `The young man stood up and said:  ``So, when your father (...)''.' (CB
010)
 \z

Notice that the `standing' and `saying' events in
(\ref{GRM-clause-conn-ka-transition}) are strictly transitional, but this is not
the case in (\ref{GRM-clause-conn-ka-then}). The connective {\sls ka} in
(\ref{GRM-clause-conn-ka-then}) opens a sentence which marks a shift from a
scene description (i.e.  `there was meat all over the place') to a character's
intervention (i.e. `Python speaking').  

Perceived event integration  seems to be what predicts the choice between 
{\sls ka} and {\sls aka}, but no firm conclusions can be drawn. 

\ea\label{GRM-ev-int-1} 

\ea\label{GRM-ev-int-1-ka} 
\gll Kàlá káálɪ́ jàwá ká jàwà múrò rō.\\
K. go market {\sc conn} bought rice {\sc foc}\\
\glt  `Kala went to the market and bougth rice.'

\ex\label{GRM-ev-int-1-aka} 
\gll  Kàlá káálɪ́ jàwá àká pɪ̀ɛ̀sɪ̀ bùlèŋà tíísà.\\
K. go market {\sc conn} ask B. station\\
\glt  `Kala went to the market and asked for the Bulenga station.' 

\z 
 \z


The cause-consequence relation in (\ref{GRM-ev-int-1-ka}) may be seen as 
`tighter' than the relation betwen the clauses in (\ref{GRM-ev-int-1-aka}). 
Buying items is a stronger effect of going to the market than 
looking for a location; market is where buying items happens. The examples in 
(\ref{GRM-ev-int-1}) suggest that {\sls aka} connects `less-integrated' 
clauses. 
 
\ea\label{GRM-ev-int-2} 
 
\ea\label{GRM-ev-int-2-ka-1} 
\gll ʊ̀ zʊ́ʊ́ dɪ̀á ká dí sɪ̀ɪ̀máá rā.\\
 {\sc 3.sg} enter house {\sc conn} eat food  {\sc foc}\\
\glt `She entered the house and ate the food.' (expected)

\ex\label{GRM-ev-int-2-aka} 
\gll  ʊ̀ zʊ́ʊ́ dɪ̀á àká vrà sɪ̀ɪ̀máá rā.\\
{\sc 3.sg} enter house {\sc conn} knock food  {\sc foc}\\
\glt `She entered the house and knock the food.' (unexpected)

\ex\label{GRM-ev-int-2-ka} 
\gll ʊ̀ zʊ́ʊ́ dɪ̀á ká vrà sɪ̀ɪ̀máá rā.\\
{\sc 3.sg} enter house {\sc conn} knock food  {\sc foc}\\
\glt `She entered the house and knock the food.' (knowledge of intention)

\z 
 \z

Commenting on each hypothetical situation in which (\ref{GRM-ev-int-2}) may be 
uttered, one consultant agreed 
that in (\ref{GRM-ev-int-2-ka}) the intention of the subject's referent are 
known and confirmed in the second clause, which is not the case in   
(\ref{GRM-ev-int-2-aka}). The  events expressed in the second cluase in  
(\ref{GRM-ev-int-2-ka-1}) and  (\ref{GRM-ev-int-2-ka}) are perceived as more 
predictable given the first clause (and world knowledge) than the event 
expressed in the second clause  in (\ref{GRM-ev-int-2-aka}).\footnote{The 
connectives {\sls aŋ}  and {\sls ka} in Pasaale \citep{mcgi99} offers a good 
baseline for comparison.}  




\paragraph{Connective dɪ}
\label{GRM-clause-coord-di}
The clausal connective {\sls dɪ} `and' or `while'  is homophonous with a
complementizer particle (Section \ref{GRM-clause-comp-di}), a connective used in
conditional constructions (Section \ref{GRM-clause-subord}),   and a preverb
particle signaling the imperfective aspect (Section \ref{sec:GRM-ipfv-part}). It
connects two clauses which encode different events, yet these events must be
interpreted as occurring simultaneously.  A clause introduced by the connective
{\sls dɪ} has no overt subject, instead the subject is inferred, as it has the
same referent as the subject of the preceding clause. Two examples are
provided in (\ref{GRM-clause-conn-di}). 

% Some examples of the clausal connective
% {\sls dɪ} in the corpus may be argued to convey intention or purpose, e.g.
% (\ref{GRM-clause-conn-di-3}). 

\ea\label{GRM-clause-conn-di}

\ea\label{GRM-clause-conn-di-vp22.4.9.}
\gll líé ʊ̀ kààlɪ̀ dɪ̀ wá.\\
  {\q} {\sc 3.sg} go {\conn} come\\
\glt  ` Where is he coming from?' ({\it lit.} where he left and  come)

\ex\label{GRM-clause-conn-di-vp47.2.9.}
\gll kpá sɪ̀ɪ̀má háŋ̀ dɪ̀ káálɪ̀.\\
 take   food {\dem}  {\conn} go\\
\glt  ` Take this food away! ({\it lit.} take this food and go)

%  \ex\label{GRM-clause-conn-di-3}
% \gll ʊ dʊa ja gantal nɪ dɪ wa.\\
%  {\sc 3.sg} be.at {\sc 1.pl.poss} back {\postp} {\sc conn} come\\
% `She is following us' ({\it lit.}  she is at our back and come)

\z 
 \z



\subsubsection{Subordination}
\label{GRM-clause-subord}

The morpheme {\sls tɪŋ} is mainly used as a  determiner in noun phrases  (see
Section \ref{sec:GRM-np-def}).  However, there are instances where the discourse
following {\sls tɪŋ} must be treated as subordinated and related to the noun
phrase of which  {\sls tɪŋ} is part. One may argue that the morpheme {\sls tɪŋ} 
can
function  as a relativizer. Consider (\ref{GRM-clause-subord-rel}).


\ea\label{GRM-clause-subord-rel} 
\gll  kúrò [píé tɪ́ŋ]$_{NP}$  ʊ̀ kà tɔ́ à kùò nɪ́ kéŋ̀ tɪ̀ɛ̀ʊ́\\
 count yam.mound.{\sc pl}  {\art} {\sc 3.sg} {\egr} cover {\art} farm {\postp} 
{\advm} give.{\sc 3.sg}\\
\glt  `(Spider$_{x}$ asked Buffalo to) count  for him$_{x}$ the yam mounds which
he$_{x}$ covered at the farm.' (LB 006)
 \z


In (\ref{GRM-clause-subord-rel}), the phrase {\sls ʊ̀ kà tɔ́ à kùò nɪ́
kéŋ̀} is (i) in apposition to the noun phrase {\sls píé tɪ́ŋ}, and (ii)
in a subordination relation with the noun phrase.\footnote{Examples 
LB 004, LB 012, CB 019 and  CB 026 in the appendix display the same sort of
subordination.}


In a conditional construction like the one in
(\ref{GRM-clause-subord-if-vp46.11}), the subordinate
clause is headed by the particle {\sls dɪ},  whereas the main clause follows
the subordinate clause. Proverbs are typically  conditional
constructions.  An example is given in (\ref{GRM-clause-subord-proverb}).


\ea\label{GRM-clause-subord-di}

\ea\label{GRM-clause-subord-if-vp46.11}

\gll dɪ̀ ǹ̩ fɪ̀ tú kààlɪ̀ dē, bà kàá  tùgúǹ nō.\\
   {\sc conn}  {\sc 1.sg} {\mod} {go.down} go {\adv} {\sc 3.pl.h+} {\fut}
beat.{\sc 1.sg} {\foc}\\
\glt  `If I was to go down there, they will beat me.' 

\ex\label{GRM-clause-subord-proverb}
\gll dɪ̀ ɪ̀ zíŋ wā zɪ̀ŋà,  ɪ̀ wàá kɪ̀ŋ gáálɪ́ díŋ nɪ̄.\\
  {\sc conn} {\sc 2.sg}  tail {\ingr} long  {\sc 2.sg} {\neg} {\abl} be.over
fire
{\postp}\\
`If you have a long tail, you cannot cross fire.'


 \z 
 \z
 

Adverbial expressions are used as connectives in similar clause-relating
functions. The subordinate clause of a concessive construction is introduced by
the expression  {\sls anɪ amuŋ}, {\it lit.} and-all, `despite',  `in spite
of', `although' or `even though'. A subordinate clause  which conveys a
consequence or a justification of the proposition in the main clause  is
introduced by the expressions {\sls aɲuunɪ} or {\sls awɪɛ}, {\it lit.} 
the-head-on
 and  the-matter,  respectively,  `therefore' or
`because'. Examples are shown in (\ref{GRM-clause-conces-consec}).

\ea\label{GRM-clause-conces-consec}

\ea\label{GRM-clause-conces}

\gll ʊ̀ wááwáʊ́ {ànɪ́ ámùŋ} dɪ́ ʊ̀ wɪ́ɪ́ʊ̀.\\
    {\sc 3.sg} come.{\pfv .\foc} {\conn} {\comp}  {\sc 3.sg} sick.{\foc}\\
\glt  `He came in spite of his illness.' 


\ex\label{GRM-clause-consec-1}

\gll ǹ̩ wà kpágá sákɪ̀r, {àɲúúnɪ̀} ǹ̩ dɪ̀ válà nã̀ã̀sá.\\
{\sc 1.sg} {\neg} have bicycle {\conn} {\sc 1.sg}   {\ipfv} walk
leg.{\pl}\\
\glt `I don't have a bicycle, therefore I am
walking.'


 \z 
 \z
 

\paragraph{Complementizer dɪ}
\label{GRM-clause-comp-di}


Example (\ref{GRM-clause-comp-inds}) shows that the complementizer {\sls dɪ}
introduces indirect speech. 
\begin{exe}
 \ex\label{GRM-clause-comp-inds}
 \gll kùórù   bìnɪ̀hã́ã̀ŋ    ŋmá   dɪ́  ɛ̃̀ɛ̃́ɛ̃́ɛ̃̀.\\
 chief   young.girl   say   {\comp} yes\\
 \glt  `The chief's daughter answered yes.'  (CB 011) 
 \z


Direct speech is usually introduced by a speech
verb only, e.g. {\sls ŋma (tɪɛ)} `say (give)',   {\sls tʃagalɪ} `teach, show,
indicate', {\sls hẽsi} `announce', etc.  This is shown in 
(\ref{GRM-clause-comp-ds}) with {\sls hẽsi}.

\begin{exe}
 \ex\label{GRM-clause-comp-ds}
 \gll tɔ́ʊ́tɪ́ɪ́ná ŋmá dɪ́ bá hẽ́sí má ká pàrà kùó.\\
 landlord say  {\comp} {\sc 3.pl.g}b  announce {\sc 2.pl} {\egr} farm farm\\
 \glt  `The landowner says that they announced:  ``You go and work at the
farm''.' 
 \z

In (\ref{GRM-clause-comp-2-int}),  the complementizer {\sls dɪ} introduces a
clause which conveys the intention of the event in the main clause. In a literal
sense, the husband {\sls lala} `open'  the wife {\it in order to} have her 
{\sls 
sii} `raise up'. In (\ref{GRM-clause-comp-2-pur}) it is shown that a purpose (or
an intention) can be encoded when {\sls dɪ} introduces the goal. In the latter 
case however consultants say that the complementizer {\sls dɪ} is optional.



\ea\label{GRM-clause-comp-2}

 \ea\label{GRM-clause-comp-2-int}
\gll  tʃʊ̀ɔ̀sá   pɪ́sɪ̀, ʊ̀   báàl tɪ̀ŋ té lálá à hã́ã̀ŋ   
 dɪ̀  ʊ́   síí  dùò nɪ̀.\\
  morning scatter    {\sc 3.sg.poss}   husband  {\art} 
early  wake.up 
{\art}  wife   {\comp}   {\sc 3.sg}   raise.up asleep  
{\postp}\\
 \glt  `Early in the morning her husband woke up the wife from sleep.' ({\it
lit.} that she should/must stand up)  (CB 030)

 \ex\label{GRM-clause-comp-2-pur}
 \gll ʊ̀ káálɪ́ (dɪ́) ʊ́ʊ́ ká ɲʊ̃̀ã̀ nɪ̄ɪ̄.\\
{\sc 3.sg} go   {\comp}  {\sc 3.sg}   {\egr} drink  water\\
 \glt  `He went to have a drink of water.' 


  \z 
 \z






 \paragraph{Clause apposition}
 \label{GRM-dep-comp-clause}
% 
In (\ref{GRM-clause-appo}) it is shown that a desire can be
encoded by two clauses in apposition. 
% 
 \begin{exe}
 \ex\label{GRM-clause-appo}
 \gll jà búúrè nɪ̄ɪ̄ rā já ɲʊ̃́ã̀.\\
{\sc 2.pl} want water {\foc} {\sc 2.pl} drink\\
 \glt  `We want some water to drink.' 
 \z


 
 
\subsection{Adjunct adverbials and postposition}
\label{sec:GRM-adjuncts}


The notion `adverbial' is used in the sense of  `modifying main predicate', 
that 
is,  a manner, a place, or a time in which a state of affair is carried out. An 
adjunct adverbial is thus an expression, clause or non-clause, which is not an 
argument of  the main predicate,  modifies the main predicate, and is 
positioned 
at the periphery 
in 
an adjunct constituent  ({\sc adj}). The clause frame in 
(\ref{ex:GRM-clause-frame}) is repeated in (\ref{ex:GRM-clause-frame-1}). 

\begin{exe}
\ex\label{ex:GRM-clause-frame-1}
 {\sc adj}  $\pm$ {\sc s|a}  $+$ {\sc p} $\pm$ {\sc o} $\pm$ {\sc adj} 
\end{exe}



Adjuncts are usually found following the core constituent(s), but may also be
found at the beginning of a clause. As
shown in (\ref{ex:GRM-pre-adj}), 
reference to time may be found at the beginning of a clause before 
the 
subject.

%to time (\advt), location (\advl) or manner (\advm).

\ea\label{ex:GRM-pre-adj}
{{\sc adj} $+$ {\sc s}  $+$ {\sc p}  $+$  {\sc o}}\\
\glll  {[tʃʊ̀ɔ̀sá  pɪ̀sɪ̀]}   {à bìpɔ̀lɪ́ɪ̀}  kpá {ʊ̀ páŕ}\\ 
 {\sc adj}  {\sc s}  {\sc p} {\sc o}\\
{morning   scatter}   {{\art} young.man} take {{3.\sg.\poss} hoe}\\
  \glt  `The following day the young man took his hoe along...' (CB 005)
\z

 In Section \ref{sec:GRM-compar-ct}, the dubitative construction was 
identified with the phrase  {\sls à bɔ́nɪ̃́ɛ̃́ nɪ́}  `perhaps'  opening the 
clause. There are other constructions in which temporal, locative, manner, or 
tense-aspect-mood meaning is signaled by the presence of an adjunct adverbial  
initially that introduces new information.  

\ea\label{ex:GRM-phra-adv}

\ea\label{ex:GRM-phra-adv-time}{\rm Temporal}\\
\gll [tàmá fìníì] ʊ̀ fɪ̀ sʊ́wá.\\
few little {\sc 3.sg} {\mod} die\\
\glt `A little longer and she would  have died.'


\ex\label{ex:GRM-phra-adv-evi}{\rm Evidential}\\
\gll [wɪ́dɪ́ɪ́ŋ ná] dɪ́ ʊ̀ náʊ́ rā.\\
truth {\foc} {\comp} {\sc 3.sg} see.{\sc 3.sg} {\foc}\\
\glt  `It is certain that he saw him.

\z 
 \z



In (\ref{ex:GRM-phra-adv-time}), the phrase {\sls tama finii} is not inherently 
temporal, but must be interpreted as such in the given context. In 
(\ref{ex:GRM-phra-adv-evi}) the verbless clause {\sls wɪdɪɪŋ na}  can be seen as 
adding an illocutionary force; it additionally signifes that the speaker has 
evidence and/or wish to convince the addressee about the proposition. In the  
next sections,  temporal and manner adverbials, then the postposition {\sls nɪ} 
and the oblique object phrase are discussed.


% 
% % They are glossed {\sc dem} (i.e. adverb locative), {\sc dxm}  (i.e.
% % adverb manner) and {\sc advt}  (i.e. adverb time) respectively. Examples are
% % provided below.

\subsubsection{Temporal  adjunct}
\label{sec:GRM-manner-adv}

A temporal nominal adjunct  ({\it gl.} {\sc advt}) is an expression which 
typically indicates when  an event occurs. In Section 
\ref{sec:GRM-preverb-three-int-tense}, the three-interval tense system is 
introduced.  It was shown that the temporal nominal {\sls  dɪare} `yesterday' 
and 
{\sls tʃɪa} `tomorrow'  have preverbs counterpart.  The  temporal nominal  
{\sls 
zaaŋ} (or {\sls zalaŋ}) expresses `today',  and   {\sls tɔmʊsʊ} can express 
either 
 `the day before yesterday' or  `the day after tomorrow',   yet neither {\sls 
zaaŋ} and {\sls tɔmʊsʊ}   have a corresponding preverb. Thus {\sls  dɪare} 
`yesterday',  {\sls tʃɪa} `tomorrow',   and  {\sls zaaŋ} `today', which  
typically 
function as adjunct  and can be disjunctively connected by the nominal 
connective {\sls anɪ}  are  treated as nominals.
%connected with what connectives?

\ea\label{ex:GRM-adj-temp-adv}

\ea\label{ex:GRM-adj-temp-adv-thatday}

\gll àwʊ̀zʊ́ʊ́rɪ̀ ǹ̩ wà tùwò nɪ̄.\\
{\advt} {\sc 1.sg} {\neg} {be.at} {\postp}\\
\glt `That day I wasn't there.'


\ex\label{ex:GRM-adj-temp-adv-LB5}

\gll àwʊ̀zʊ́ʊ́rɪ̀  dɪ́gɪ́ɪ́     kɔ̀sánã́ɔ̃́   válá (...).\\
{\advt}  one     buffalo walked {}\\
\glt `One day a buffalo walked (by and greeted the spider) (...)'. (LB 005)

\ex\label{ex:GRM-adj-temp-adv-CB17}
\gll [dénɪ̀],   [sáŋà   dɪ́gɪ́ɪ́]   à   
hã́ã̀ŋ já   pàà à  báál   zōmō  (...)\\
{\advt} time   one     {\art}  wife  {\hab}   take.{\pl} 
{\art} husband insult.{\pl} {} \\
\glt `[During their life, it happened] on one occasion that the woman
did insult  the man (...)' .  (CB 017)

\ex\label{ex:GRM-adj-temp-adv-everyday}
\gll  ǹ̩ já kààlɪ̀ ʊ̀ pé rè [{tʃɔ̀pɪ̀sɪ̀} bɪ́ɪ́-mùŋ].\\
 {\sc 1.sg} {\hab} go {\sc 3.sg} end {\foc} day.break {\itr}-all\\
\glt `I visit him every day.'

\ex\label{ex:GRM-adj-temp-adv-nownow}
\gll [làɣálàɣá háǹ nɪ̄] ǹ̩ kʊ̀tɪ̀ à ʔã́ã́ pétí.\\
{\advt} {\dem} {\postp} {1.\sg} {skin} {\art} bushbuck  finish\\
\glt `I  just finished skinning the bushbuck.'


\z 
 \z


Some expressions tagged as temporal nominal are treated as complex, though
opaque, expressions. For instance,  {\sls awʊzʊʊrɪ} is translated into  `that
day' in (\ref{ex:GRM-adj-temp-adv}), but the forms {\sls wʊ̀sá} `sun' and
{\sls zʊʊ} `enter'  are perceptible. The phrase {\sls laɣalaɣa han nɪ} in
(\ref{ex:GRM-adj-temp-adv-nownow}) literally
means `now.now this on' ({\advt} {\dem} {\postp}), but `only a moment
ago'  is a better translation.  Similarly, {\sls denɪ} is analyzed as a
temporal nominal, but usually functions as a connective. It is made from  the
spatial demonstrative {\sls de} and the potsposition {\sls nɪ}, and is 
translated 
to
English as `thereupon', `after that', `at that point', or simply `then'. It is
mainly used at the beginning of a sentence to signal a transition  between the
preceding  and the following situations;
(\ref{ex:GRM-adj-temp-thereupon}) suggests a transition of the resultative type.
The appendix contains other examples of {\sls denɪ}, e.g.  CB (008, 017, 019)
and 
LB (006, 016).


\ea\label{ex:GRM-adj-temp-thereupon}
\gll dénɪ̄   rè,     ʊ̀ʊ̀      hã́ã́ŋ   tɪ̀ŋ 
ŋmá   dɪ́  ààí, (...)\\
 {\advt}   {\foc}  {\sc 3.sg.poss}  wife  {\art}  say   {\comp} 
no {}\\
\glt [The man said: `Don't cry, if you tell your father that I drove the tsetse
flies away,  weeded the farm and took you as a wife, I will also tell your
father you are freeing yourself in bed.']  `Then, the wife said: ``No, (I won't
say
anything to my father)'''  (CB 036)
\z




\subsubsection{Manner adjunct}
\label{sec:GRM-manner-adv}

A manner  expression describes the way the event denoted by
the verb(s) is carried out. The examples in (\ref{ex:GRM-adj-mann}) illustrate
the meaning and distribution of  manner expressions.


\ea\label{ex:GRM-adj-mann}


\ea\label{ex:GRM-adj-mann-carefully}
\gll dɪ̀ sã́ã́ bʊ̃̀ɛ̃̀ɪ̃̀bʊ̃̀ɛ̃̀ɪ̃̀.\\
{\comp} drive {carefully}\\
\glt `Drive carefully.'

\ex\label{ex:GRM-adj-mann-slowly}
\gll dɪ̀ ŋmà bʊ̃̀ɛ̃̀ɪ̃̀bʊ̃̀ɛ̃̀ɪ̃̀.\\
{\comp} talk {slowly}\\
\glt `Talk slowly.'

\ex\label{ex:GRM-adj-mann-lighly}
\gll ʊ̀ tʃɔ́jɛ̄ kààlɪ̀ félfél.\\
 {\sc 3.sg} run.{\pfv} go {lightly}\\
\glt `She ran away lightly (manner of movement, as a light weight
entity).'

\ex\label{ex:GRM-adj-mann-silently}
\gll  ǹ̩ kàà wáá dɪ̀ à   hã́ã́ŋ  sáŋà   tʃérím.\\
{\sc 1.sg} {\sc ipfv} come {\sc comp} {\sc art} woman sit quietly\\
\glt `When I was coming, the woman sat quietly' 

\z 
 \z

It is common for an \is{ideophone}ideophone to function as a manner expression 
 (Section \ref{sec:GRM-onoma}). One could argue that  all the manner 
expressions in
(\ref{ex:GRM-adj-mann}) are ideophones, i.e. they display reduplicated forms
and {\sls tʃerim} is one of a few words which ends with a bilabial nasal. The
examples in (\ref{ex:GRM-adj-mann-ideo-adv}) show the repetition of two
expressions; one is an \is{ideophone}ideophone, i.e. {\sls kaŋkalaŋ} `crawl of 
a 
snake', and the
other  a  reduplicated  manner expression,  i.e. {\sls 
laɣalaɣa} 
`quickly' from {\sls laɣa} 
`now'. The
repetition of {\sls kaŋkalaŋ} and {\sls  laɣalaɣa} conveys that the motion was
({\sls kpa} `taken'  and) occurring with great speed.

%inceptive meaning of kpa


\ea\label{ex:GRM-adj-mann-ideo-adv}


\ea\label{ex:GRM-adj-mann-ideo}
\gll à bààŋ kpá {kàŋkàlàŋ kàŋkàlàŋ kàŋkàlàŋ}.\\
{\conn} just take crawl.rapidly\\
\glt `(She was after the python) but (he) started to crawl away like a shot.' 
(REF?)

\ex\label{ex:GRM-adj-mann-adv}
\gll  kà bààŋ kpá làɣàlàɣà làɣàlàɣà.\\
{\conn} just take {\advm} {\advm}\\
\glt `(She) started to (walk) quickly.'


\z 
 \z


The manner adverb {\sls kɪŋkaŋ} `abundantly',  which is composed of the 
classifier
{\sls kɪn} and the verb {\sls kana} `abundant',  typically quantifies or 
intensifies
the
event and always comes after the word encoding the event.  Notice in
(\ref{ex:GRM-adj-mann-alot-v})  and (\ref{ex:GRM-adj-mann-alot-n})   that {\sls 
kɪŋkaŋ} follows a verb and a nominalized verb respectively. However, in
(\ref{ex:GRM-adj-mann-alot-quant}), {\sls kɪŋkaŋ} does not function as a
manner adverb but as a quantifier.



\ea\label{ex:GRM-adj-mann-alot}

\ea\label{ex:GRM-adj-mann-alot-v}
\gll gbɪ̃̀ã́     ɪ̀    jáárɪ́jɛ́  kɪ́ŋkāŋ    nà
(...)\\ 
monkey   you  unable.{\pfv}  {\advm} {\foc} {}\\
\glt `Monkey, you are so incompetent, (...).' (LB 016)

\ex\label{ex:GRM-adj-mann-alot-n}

\gll dúó tʃʊ̄ɔ̄ɪ̀ kɪ́ŋkāŋ wà wíré.\\
asleep lie.{\nmlz} {\advm} {\neg} good\\
\glt `Sleeping too much is not good.'

\ex\label{ex:GRM-adj-mann-alot-quant}
\gll kùórù   kùò tɪ́ŋ   kà   kpágá kìrìnsá  kɪ́ŋkāŋ, dé rē jà 
kààlɪ̀\\
 chief farm {\art} {\rel} have tsetse.fly.{\sc pl}  {\quant}.many {\sc dem} {\sc 
foc} {\sc 1.pl} go\\
\glt   `The chief's farm that has  many tsetse flies, there we went' 


\z 
 \z



\subsubsection{Postposition {\sls nɪ} and (non-) locative adjunct}
\label{sec:SPA-postp}

First, the postposition {\sls nɪ} signals  that the constituent in which it  
appears  is locative.  In fact, in the majority of the  utterances in the 
 corpus,\footnote{\label{ft:spatial-corpus}The {\it Chakali Location and 
Position Corpus} 
\citep{JAB-space-C-10} contains the audio material of the elicitation 
procedure described in \citet{Bowe93, Amek99, Meir01a, Meir01b}.}  the  
postposition {\sls nɪ} is present irrespective of the
locative verb involved or whether or not a relational noun occurs. Only a  few
exceptions can be found,   and they are
systematically accounted for by two factors: (i) some verbs  do not co-occur 
with {\sls nɪ}, e.g. {\sls tɔ} `cover', {\sls kpaga} `have' and {\sls su} 
`fill',  
and (ii) some situations are described using an intransitive clause, e.g.   
{\sls 
à bónsó tʃíégìàō} `the cup is broken'  (TRPS 26).  
\citet[370]{Amek06} present the Ewe verb {\sls le}, glossed  `be at',  which is 
used in the majority of the sentences denoting the situations of the TRPS.   
The 
translation of  Ewe {\sls le} to Chakali would then be equivalent to {\sls dʊa 
{\rm NP} nɪ}.\footnote{The Ewe verb {\sls le} may also function as predicator 
of 
qualities \citep[373]{Amek06}. In Chakali,  it was shown  in Sections 
\ref{sec:GRM-ident-cl} and \ref{sec:classifier} that   {\sls jaa} predicates 
over 
qualities,  not  {\sls dʊa}.}



Second, the postposition {\sls nɪ} identifies an oblique object phrase, and
 conveys that the oblique object phrase contains the ground object with which a 
figure related. The complement precedes the  postposition. The examples  
displayed in ({\ref{ex:postp-corres}) show that the complement of  {\sls nɪ} is 
a 
noun phrase (see {\it RelP} in Section \ref{sec:SPA-relnoun}). Since there are 
no prepositions in the language, the abbreviation PP in 
({\ref{ex:postp-corres}}) unambiguously stands for Postposition Phrase.

\ea\label{ex:postp-corres}
 \ea {[}{[}{[}a dɪa{]}$_{NP}$ ɲuu{]}$_{RelP}$ nɪ{]}$_{PP}$  {\rm `on the roof 
of the 
house'}\\
 \ex {[}{[}a dɪa{]}$_{NP}$ nɪ{]}$_{PP}$ {\rm  `in/at the house'}\\
 \ex  {[}{[}baŋ{]}$_{NP}$ nɪ{]}$_{PP}$ {\rm  `here'}\\
 \ex {[}{[}de{]}$_{NP}$ nɪ{]}$_{PP}$  {\rm `there'}\\
 \ex {[}{[}ʊ{]}$_{NP}$ nɪ{]}$_{PP}$  {\rm `at/on/in him/her/it'}\\
\z
\z

Nevertheless, the postposition does not inform the addressee on any of the 
elementary topological spatial notions; none of the concepts of proximity, 
contiguity, or containment is encoded in  the postposition {\sls nɪ}. It never 
selects particular figure-ground configurations but must be present for all of 
them. 

Besides the description of static configurations, the postposition {\sls nɪ}  
is 
used frequently in adverbial/connective expressions: {\sls à bɔ́nɪ̃́ɛ̃́ nɪ́} 
`maybe, perhaps', {\sls à ɲúú nī} `therefore', {\sls búŋbúŋ ní} `at 
first', etc. These expressions do not have a purely locative function, but are 
rather used as clausal adjuncts,  or to introduce logical conclusion (see 
Sections   \ref{GRM-clause-subord} and \ref{sec:GRM-adjuncts}). 



Interestingly, if the postposition  occurs between the focus particle 
(Sections \ref{sec:GRM-foc-neg} and \ref{sec:GRM-focus}) 
and
the preceding nominal, one would expect  the focus particle to surface in its
default form, i.e. {\sls ra},  since the required adjacency is no longer 
satisfied
(Section  \ref{sec:focus-forms}). 


\ea\label{ex:GRM-focus-form}

\ea\label{ex:GRM-foc-form-X}
\TExt{\Txt{$\alpha$}\COn{rr}\ \  & \Txt{\it nɪ} & \ \  \TXT{\foc}}
 
\ex\label{ex:GRM-foc-form-1}
\gll  à máŋkɪ́sɪ̀ ɲúú nɪ̄ rò/rè.\\
    {\art} {match} {\reln} {\postp} {\foc}\\
\glt `on the top of the matchbox'

\ex\label{ex:GRM-foc-form-2}
\gll  à  pùl nɪ́ rō/rē.\\
    {\sc art} {river} {\sc postp}  {\sc foc}\\
\glt `on/at the river'

\z 
 \z

However, on several occasions, the postposition becomes `transparent' and 
vowel-harmony can still operate (i.e. though not the place 
assimilation of consonant). The
phenomenon is shown in (\ref{ex:GRM-focus-form}).\footnote{A more extreme case
is found in example (\ref{ex:GRM-obl-obj-no-spa-foc}).}



\subsubsection{Oblique object phrase}
\label{sec:GRM-obl-phrase}

The oblique object phrase ({\sc obl}) is an element of a clause whose  
semantics is
characterized by an  affected or effected object, although realized by a
postpositional phrase.  In Section \ref{sec:SPA-postp},  it is claimed that the
postposition {\sls nɪ} (i) identifies an oblique object phrase, (ii) conveys 
that
the oblique object phrase contains the ground object in localization, and
(iii) follows its complement. 

While localization is
the main function of  {\sls nɪ}, the postposition can also be found when there
is no  reference to space. For instance, in Section \ref{sec:GRM-manner-adv}, I
discuss the connective {\sls denɪ} (i.e. {\dem}+{\postp}), whose role in
discourse is to signal a temporal transition, not a spatial one.  The
examples in (\ref{ex:GRM-obl-obj-no-spa}) illustrate some of the non-spatial
uses
of the oblique object phrase headed by {\sls nɪ}.


\ea\label{ex:GRM-obl-obj-no-spa}

\ea
\gll ʊ̀ ɲʊ̃́ã́  [làɣálàɣá nɪ̄].\\
   {\psg} drink {\advm} {\postp}\\
\glt  `He drinks quickly.' 

\ex
\gll bàáŋ ɪ̀ fɪ́ kàà sʊ́ɔ́gɪ̀ [tʃʊ̀ɔ̀sá tɪ́n nɪ̄].\\
 {\q} {\sc 2.pl} {\pst} {\egr} crush  morning {\art} {\postp}\\
\glt  `What were you crushing this morning?' 

\ex\label{ex:GRM-obl-obj-no-spa-foc}
\gll à kùórù ŋmá dɪ́ ʊ̀ bááŋ káá sīī [ǹ̩ nɪ́] rē.\\
{\art} chief say {\comp} {\sc 3sg.poss} temper {\egr} raise {\sc 1.sg} {\postp}
{\foc}\\
\glt  `The chief told me that he was very angry with me.' 


\z 
 \z

 


%\begin{exe}
% \ex\label{GRM-clause-appo}
% \gll\\
%\\
% \glt  `' 
%  \z

% 
% \ex\label{ex:vp19.2.}
% \gll anɪ a muŋ dɪ ʊ dia boloo , n̩ ja ka lɪ ʊ pe re tʃʊɔpɪsɪ bɪ muŋ.\\
%\\
% \glt  ` Although his room is far away, I visit him every day.  ' 




\section{Nominals}
\label{sec:GRM-nom}


The term `nominal'  identifies  a formal and functional  syntactic level and
lexemic level. At the syntactic level, a noun phrase is a nominal  which can
either function as core or peripheral argument.  Its composition may
vary from a single pronoun to a noun with modifier or series of
modifiers. At the lexeme level, two categories of lexemes are assumed:
nominal and verbal. These two types correspond roughly to the semantic division
{\it entity} and {\it event}, but do not correspond to the syntactic categories
{\it noun} and {\it verb}. That is because lexemes are assumed to not be
specified for syntactic category. The diversity  of forms and functions of
nominals is presented below. 


\subsection{Noun phrases}
\label{sec:GRM-verb-phrases}

A noun phrase (NP)  consist of a nominal head, and optionally, its dependent(s).
In this section,  the internal components of noun phrases and the roles these
components have within the noun phrase are described. First,   indefinite and
definite noun phrases are considered. Then, the elements which can be found in
the noun phrase are introduced. 

\subsubsection{Indefinite noun phrase}
\label{sec:GRM-np-indef}

Indefinite noun phrases are used when ``the speaker invites the addressee to 
construe a referent [which conforms with] the properties specified in the term'' 
\citep[184]{Dik97}.  In Chakali, a noun standing alone can  constitute a noun 
phrase (N = NP). Such a noun phrase can be interpreted as indefinite, i.e. the 
noun phrase is a non-referring expression,  or   generic, i.e. the noun phrase 
denotes  a kind or class of entity  as opposed to an individual.  In rare cases, 
a definite noun phrase can be interpreted from a single noun  (i.e. lacking  an 
article). Each interpretation is obviously dependent on the context of the 
utterance in which the noun occurs.

\ea\label{GRM-np-type-indef}{\rm  N = NP}\\

 
\ea\label{GRM-np-indef-1}
\gll  kàlá jáwá   [pɪ́ɛ́ŋ]$_{NP}$ ná.\\
  Kala buy mat {\sc foc}\\
\glt  `Kala bought a MAT' 


\ex\label{GRM-np-indef-2}
\gll  [dʒɛ̀tɪ̀]$_{NP}$ kɪ̀m-bɔ́n  ná.\\
  {lion.{\sc sg}} {\sc clf}-dangerous.{\sc sg} {\sc foc}\\
\glt  `A lion is DANGEROUS'



\z 
 \z

In (\ref{GRM-np-type-indef}),  the noun phrase {\sls  pɪɛŋ}  describes any mat
and
is interpreted as a novelty in the addressee's knowledge of Kala, while
{\sls dʒɛtɪ} describes the entire class of lions. 

Noun phrases  containing the
numeral {\sls dɪ́gɪ́ɪ́} `one'  may be translated as
English `a certain'. The
expression {\sls badɪgɪɪ} can be translated as `one of them', `someone' or
`anyone'  (e.g. {\sls ʊ wa ja badɪgɪɪ}, {\it lit.} he-not-be-one.of.them, `he is
an
illegitimate child'). 

%Other strategies to introduce a specific object or set of
%objects that is believed to be new to the addressee is the 


\subsubsection{Definite noun phrase}
\label{sec:GRM-np-def}

Definite noun phrases are employed when ``the speaker invites the addressee to
identify a referent which he (the speaker) presumes is available to the
addressee'' \citep[184]{Dik97}. A definite noun phrase may consist of  a single
pronoun (pro = NP), as
shown in (\ref{GRM-np-type-pro}).

\begin{exe}
 \ex\label{GRM-np-type-pro}{\rm pro = NP}\\
\gll  [ʊ̀]$_{NP}$  sʊ́wáʊ́.\\
   {\sc 3.sg} die\\
\glt  `She died'
\z


A possessive noun phrase is always definite. A possessive pronoun followed by a
noun is analyzed as a succession of a noun phrase and a noun. Thus,  the noun
phrase in (\ref{GRM-np-type-pro-n})  is analyzed as a sequence of the noun
phrase {\sls ʊ} and the  noun  {\sls mãã} (pro + N = NP). 


\begin{exe}

 \ex\label{GRM-np-type-pro-n}{\rm pro + N = NP}\\
\gll   [ʊ̀ mã̀ã̀]$_{NP}$ ŋmá dɪ́ ʔői̋.\\
   {\sc 3.sg.poss} mother say  {\sc comp} {\sc interj}\\
\glt  `Her mother said, ``Oi!''.'
\z


The treatment  NP+N for possessive noun phrase is   motivated  by the 
possibility of  recursion of  an attributive possession relation. The expression 
{\sls lɔɔlɪgbɛrbɪɪ}$_{N}$  `car key' is the head in the three possessive noun 
phrases   {\sls karedʒa lɔɔlɪgbɛrbɪɪ} `Kareja's car key', {\sls karedʒa ɲɪna 
lɔɔlɪgbɛrbɪɪ} `Kareja's father's car key' and {\sls karedʒa ɲɪna bɪɛrɪ 
lɔɔlɪgbɛrbɪɪ} `Kareja's father's senior brother's car  key'.  Notice that in 
these examples the nominal head consists of the right-most element in the noun 
phrase. The compound noun  {\sls  lɔɔlɪgbɛrbɪɪ}  `car-key', and correspondingly, 
the head of the left daughter of the  possessive noun phrase is the right-most 
element, here underlined,  e.g.   [[{\sls karedʒa ɲɪna 
\underline{bɪɛrɪ}}]$_{NP}$ [{\sls lɔɔlɪgbɛrbɪɪ}]$_{N}$]$_{NP}$, {\it lit.} `the 
key of the brother of the father of Kareja'. 




% % The syntactic tree in (\ref{ex:GRM-poss-def-tree}) illustrates the structure of 
% % this definite noun phrase. %add lɔɔlɪgbɛrbɪɪ to dict 
% % \ea\label{ex:GRM-poss-def-tree} \Tree[-1]{  &&&&&\it q{NP} \Bq{dlll}\Bq{drrr} 
% % &&&&\\ &&\it q{NP} \Bq{dll}\Bq{drr} &&&&&& \it q{N}\Bq{dl}\Bq{dr} &\\ \it 
% % q{karedʒa} &&&& \it q{NP} \Bq{dl}\Bq{dr} &&& \it q{lɔɔlɪ}  &&\it q{gbɛrbɪɪ}\\ 
% % &&& \it q{ɲɪna} && \it q{bɪɛrɪ} &&&&} \z





\paragraph{Articles a and tɪŋ}
\label{sec:GRM-np-def-articles}
%This treatment predicts that the head of a possessive noun phrase 

There are two articles in Chakali; one which encodes specificity and the other
definiteness. The first one is the 
article {\sls a} ({\it gl.} {\sc art1}) and the other is  {\sls tɪŋ}  ({\it
gl.} {\sc art2}).  

The  article {\sls a} is translated with the English article
`the'.\footnote{A pre-nominal article is not found in Tampulma or Pasaale. The
fact that
Waali and Dagaare use an identical  article suggests that the definite article
{\sls a}
is a contact-induced innovation.} It must precede the head noun and cannot
co-occur with the possessive pronoun.  In the context of
(\ref{GRM-np-type-det1}), the speaker assumes that the addressee is informed 
about
Kala's interest in buying a mat. 


\begin{exe}
 \ex\label{GRM-np-type-det1}{\rm  a + N = NP}\\
\gll  kàlá jáwá  [à pɪ́ɛ́ŋ]$_{NP}$ ná.\\
  Kala buy {\sc art1}  mat {\sc foc}\\
\glt  `Kala bought the MAT' 
\z


The type of mat,  its color or the location where Kala bought the mat and so on
are not necessarily shared pieces of information between the speaker and 
addressee
in (\ref{GRM-np-type-det1}).  The only information the speaker believes they
have in common is Kala's interest in purchasing a mat. The article {\sls a} is 
treated as a functional word which makes the noun phrase specific but not
necessarily
definite.  When a noun phrase is  specific, the speaker should have a particular
referent in mind whereas the addressee may or may not share this knowledge.


The article {\sls tɪŋ}  ({\it gl.} {\sc art2}) can also be seen to correspond to
English `the',  but a preferable paraphrase would be `as referred previously' or
 `this (one)'.  The article {\sls  tɪŋ} appears when the speaker knows that the
addressee will be able to identify the referent of the noun phrase. In that 
sense,
the referent is familiar.\footnote{In the giveness hierarchy of
\citet[278]{Gund93}, the status {\it familiar} is reached when ``the addressee
is able to uniquely identify the intended referent because he already has a
representation of it in memory.''}   When {\sls tɪŋ} follows a noun, the 
referent
must either have been mentioned previously or the speaker and addressee have an
identifiable referent in mind. Thus, compared to the examples
(\ref{GRM-np-type-indef}) and (\ref{GRM-np-type-det1}) above, a proper
interpretation of example (\ref{GRM-np-type-det2}) requires that both the
speaker and addressee have a particular mat in mind. In terms of word order, the
article  {\sls a}  initiates the noun phrase and  the article {\sls tɪŋ}  
belongs
near the end of the noun phrase. 
 

\begin{exe}
 \ex\label{GRM-np-type-det2}{\rm   (a +) N + tɪŋ = NP}\\
\gll  kàlá jáwá  [à pɪ́ɛ́ŋ  tɪ́ŋ]$_{NP}$ nā.\\
  Kala buy {\sc art1}  mat {\sc art2} {\sc foc}\\
\glt  `Kala bought the MAT'
\z

Consider the slight meaning difference between
(\ref{GRM-np-type-det2-a}) and (\ref{GRM-np-type-det2-b}).


\ea\label{GRM-np-type-det2-ab}
% \vspace{-12pt}
 
  \ea\label{GRM-np-type-det2-a}
\gll ɲɪ̀nɪ̃̀ɛ̃́ ɪ̀ ɲɪ́ná kà dʊ́.\\
 {\sc q}.how {\sc 2.sg.poss} father {\sc  egr} be\\
\glt  `How is your father?'

  \ex\label{GRM-np-type-det2-b}
\gll ɲɪ̀nɪ̃̀ɛ̃́ ɪ̀ ɲɪ́ná tɪ́ŋ kà dʊ́.\\
   {\sc q}.how {\sc 2.sg.poss} father {\sc art2} {\sc  egr} be\\
\glt  `How is your father?'
  
 
\z 
 \z


Both sentences may be translated with `How is your father?'. However, whereas 
the sentence (\ref{GRM-np-type-det2-a}) can request  a general description
of the father (i.e. skin color, size, general health, etc.), the sentence
in (\ref{GRM-np-type-det2-b}) asks for a particular aspect of the
father's condition which both the speaker and the addressee are aware of, for
instance the father's sickness. As sketched above, the article {\sls tɪŋ}  in
(\ref{GRM-np-type-det2-b}) establishes that a particular disposition of the
father is known  by both the speaker and the addressee,  and the speaker
asks, with the question word {\sls ɲɪnɪ̃ɛ̃} `how',   for details. 

The two  articles {\sls a} and {\sls tɪŋ}  are not in complementary 
distribution.
The article {\sls tɪŋ} may occur following the head of a possessive noun phrase,
although it is not attested  following a weak pronoun. When the articles {\sls 
a} and {\sls tɪŋ} co-occur,  language consultants could omit
the preposed {\sls a}  without affecting the interpretation of the
 proposition. 

Notice that any of the terms {\it article}, {\it determiner} and
{\it demonstrative} could have been used to identify {\sls a} and {\sls tɪŋ}.
Similar forms/functions are labelled differently in the
literature; for instance \citet[47]{Bodo97} calls the prenominal {\sls a} in
Dagaare an article which he translates into Eng. `the'  and  {\sls nyɛ},  a 
demonstrative, into `this'.  


Yet, an analysis of the paradigms in (\ref{GRM-np-det-dem}),  in which Central 
Dagaare is included for illustration,\footnote{Central Dagaare as it is spoken 
in Nadowli, Sombo, Dafɪama, etc. Thanks to John Gaanaa for providing the 
examples. } has never been offered. It is shown that a sequence of two 
demonstratives (in \citeauthor{Bodo97}'s term)  can occur in postnominal 
postition.


% \begin{minipage}[h]{12cm}
\ea\label{GRM-np-det-dem}{\rm Corresponding  noun phrases in Chakali and
Central  Dagaare} 

\begin{multicols}{2}

 \ea\label{GRM-np-det-dem-dag}{\rm Central Dagaare}\\
à bíé   `the child'\\
à bíé ŋà   `the child this'\\
à bíé  ŋánɛ́ɛ́ ŋà  `the child this this'\\
 \textasteriskcentered  a bie  ŋa ŋanɛɛ  

 \ex\label{GRM-np-det-dem-cli}{\rm Chakali}  (no frame, see p133 for frame)\\
à bìé   `the child'\\
à bìè tɪ́ŋ̀   `the child this'\\
à bìè háŋ̀ tɪ̀ŋ  `the child this this'\\
 \textasteriskcentered a bie  tɪŋ haŋ  

 
\z
\end{multicols}
 \z




The question raised by paradigm  (\ref{GRM-np-det-dem}) is  whether one
should call {\sls tɪŋ} a demonstrative (and not an article) based  on the 
English
glosses  supplied by  \citet[47]{Bodo97}, or show that Chakali and Central
Dagaare stack determiner-like function words in an unusual way and that this is
a problem for the general description of noun phrases.\footnote{The
demonstrative {\sls haŋ} `this' is presented in Section \ref {sec:GRM-demons}.}
The latter position is
chosen and a description of  {\sls a} and {\sls tɪŋ} is provided above. The
noun phrase {\sls a bie haŋ tɪŋ} means `this child of which we (both speaker and
addressee)   have a familiar/common representation'.  The discourse implications
 of  {\sls a} and  {\sls tɪŋ}  need 
further study. 

Now that the indefinite and definite noun phrases have been presented, the
subsequent sections introduce the elements which can compose  either  indefinite
or  definite noun  phrases.


% The position taken here is
% that only article and demonstrative exist in the language; as opposed to the
% latter,  the former lacks deictic power and cannot function as noun
% phrase on its own.  


\subsection{Nouns}
\label{sec:GRM-noun}

In this section the elements admitted in the
schematic representation (\ref{sec:GRM-noun-strcut}) are discussed.

\ea\label{sec:GRM-noun-strcut}
[[ {\sc lexeme}]$_{stem}$ - [{\sc noun class}]]$_{n}$
\z

A stem may have 
nominal or verbal lexeme status. The latter has either a state (i.e. stative) or
a process (i.e. active) meaning.  A stem can be either atomic or complex and a
noun class suffix may be overt or covert.  In a
 process which turns a lexeme into a noun-word,  the noun class provides the
syntactic category {\it noun}. 




\subsubsection{Noun classes}
\label{sec:GRM-noun-classes}

The accepted view is that ``the Gurunsi languages, and indeed all Gur languages,
had historically a system of nominal classification which was reflected in
agreement. The third person pronominal forms and other parts of speech were at a
certain time a reflection of the nominal classification''  \citep{Nade89}.
 Similar affirmations are present in \citet{Mane69b, Waa71, Nade82, Nade98,
Tcha07}.  In this section and in Section
\ref{sec:GRM-gender}, it is suggested that
an eroded form of this `reflection' is still observable in Chakali.
\citet{brin08c} claims that in Chakali inflectional class
(i.e. noun class) and agreement class (i.e. gender) should be distinguished and
analyzed as separate phenomena at a synchronic level.

 The identification of noun classes is based on non-syntagmatic evidence; noun
class is a type of inflectional  affix, independent of agreement
phenomena, where the values of number
and class are exposed. In Chakali, as in all  other Southwestern Grusi  
languages,\footnote{\citet[136]{Nade98} state that ``[i]n
Vagla most traces
of this [noun-class system where paired singular/plural noun affixes correlate
with concording pronouns and other items] system have been lost. The
morphological declensions of nominal pluralization have not yielded to a clear
analysis''.  Even though the authors do not attempt to allot nouns into classes,
Marjorie Crouch's field notes (1963, Ghana Institute for Linguistics, Literacy
and Bible Translation (GILLBT)) present seven classes. Nominal classifications
are proposed for other SWG languages (number of classes for each language in
parenthesis): Sisaala of Funsi in \citet{Rowl66} (2), Sisaala-Pasaale in
\citet{Mcgi99} (5) and Isaalo in \citet{Mora06} (4).  The number of classes is 
of
course determined by the linguist's analysis.\label{foot:noun-class}}  the
values are exposed by
suffixes: number refers to either singular or plural, and class can be regarded
as phonological and/or semantic features encoded in the lexemes for the
selection
of the proper pair of singular and plural suffixes. This will be considered in
Section \ref{sec:GRM-sem-ass-crit}. 



 \begin{table}[!h]
 \caption{The five most frequent noun classes \label{tab:GRM-synop-nc}}
   \centering
   \begin{Itabular}{lccccc}

 \lsptoprule
    &  {\sc cl.1} & {\sc cl.2}  & {\sc cl.3} & {\sc cl.4} & {\sc cl.5} 
\\  [1ex] \midrule
{\sc sing} & -V&  \O&  \O& -V  & \O\\
{\sc plur} & -sV& -sV & -V & -V  & -nV\\ 
 \lspbottomrule
   \end{Itabular}
 \end{table}



One method used to identify the noun classes of a language appears in
\citet[23]{Rowl66}. The author writes that ``[t]he nouns in Sissala may be
assigned to groups on the basis of the suffixes for singular and plural''. 
 According to this definition, there are nine noun 
classes, of which four are rare.   A synopsis is displayed in Table 
\ref{tab:GRM-synop-nc}, and each
of them is discussed below. 



\paragraph{Class 1}
\label{sec:class1}

Class 1 allows a variety of stems:  CV, CVC, CVVCV,  and CVCV are possible.
It gathers the nouns whose singular is formed by a single vowel
suffix {\it -V} and plural by a
light syllable {\sls -sV}.


\begin{table}[h]

\caption{Class 1 \label{tab:freq-noun-class-1}}
\centering
%\subfloat[][{\sc class 1}]{
 \begin{Itabular}{lllll}
  \lsptoprule
{\sc class} & Stem & {\sc sg.} &   {\sc pl.} & Gloss\\ [1ex] 
\midrule
{\sc cl.1}  &   va   &  váà   &  vá{\T ꜜ}sá  & dog\\ 

%{\sc cl.1}  &  hɛn   &  hɛ̀ná   &  hɛ̀nsá  & bowl\\
%{\sc cl.1}  &  da   &  dáá   &  dààsá & tree\\

{\sc cl.1}  &  pɛn   &  pɛ̀ná   &  pɛ̀nsá  & moon\\
{\sc cl.1}  &  gun   &  gùnó   &  gùnsó  & cotton\\
{\sc cl.1}  &  tʃuom   & tʃùòmó  & tʃùònsó   & togo hare\\
{\sc cl.1}  &  bi   &  bìé   &  bìsé  & child\\
{\sc cl.1}  &  gbiegi   &gbìègíè   &  gbìègísé  & type of hawk\\

  \lspbottomrule
 \end{Itabular} 
 %}

\end{table} 

 The quality of the vowels of the singular and plural is
determined by
the quality of the stem vowel and the harmony rules in operation. The rules were
stated in Section \ref{sec:vowel-harmony} and correspond to the noun class
realization rules given in 
(\ref{ex:GRM-Hrules}).


\ea\label{ex:GRM-Hrules}

\ea\label{ex:mod-front-suffix}
-(C)V$_{nc}$  $>$ [ $\beta${\sc ro},  {\sc +atr}, {\sc --hi}]  / [ $\beta${\sc 
ro}, 
{\sc +atr}] C* \_\\

A noun class suffix vowel becomes {\sc +atr} if preceded by a {\sc +atr}
stem vowel, and shares the same value for the
feature {\sc ro}  as the one specified on the preceding (stem) vowel. A noun
class suffix is always {\sc --hi}.

 \ex\label{ex:low-suffix}
-(C)V$_{nc}$  $>$ {\sc +lo}  / {\sc --atr} C* \_\\

A noun class suffix vowel becomes {\sc +lo} if the preceding stem vowel is 
either
{\sls ɪ}, {\sls ɛ}, {\sls ɔ}, {\sls ʊ} or {\sls a}.\\

\z 
 \z


 



 
 \paragraph{Class 2}
\label{sec:class2}
 
Table \ref{tab:freq-noun-class-2} displays  nouns assigned to
class 2. Typically, this class consists of nouns whose stems are CVV or CVCV.
While the singular form  displays no overt
suffix,  {\sls -sV} is suffixed onto the stem to form the plural.  

\begin{table}[h]


\caption{Class 2 \label{tab:freq-noun-class-2}}
\centering
%\subfloat[][{\sc class 2}]{
 \begin{Itabular}{lllll}
  \lsptoprule
{\sc class} & Stem & {\sc sg.} &   {\sc pl.} & Gloss\\ [1ex] 
\midrule

%{\sc cl.2}  &nãã &  nã̀ã̀ & nã̀ã̀sá & leg\\
{\sc cl.2}  &  daa   &  dáá   &  dààsá & tree\\
%{\sc cl.2} &  tii  &  tìì   &  tìsè  & akee tree\\
{\sc cl.2}  &  bɔla & bɔ̀là   &  bɔ̀làsá  &  elephant\\
%{\sc cl.2}  &  bʊɔ &  bʊ̀ɔ́   &  bʊ̀ɔ́sá  &  hole\\
%{\sc cl.2}  &  joŋ   &  jòŋ́   &  jósó  & slave\\
%{\sc cl.2} &  ziŋ  &  zíN   &  zísé  & tail\\
{\sc cl.2} &  kuoru &  kùórù   &  kùórùsó  & chief\\
{\sc cl.2} &tomo &tòmó & tòmòsó& type of tree\\

%{\sc cl.2} &  ŋmɛŋ   &  ŋmɛ̀ŋ   &  ŋmɛ̀sà  & rope\\
%{\sc cl.2}  &   tuto &tútò   &  tùdùsó  & mortar\\
{\sc cl.2} &  bele  & bèlè  &  bèlèsé &type of wild dog\\
{\sc cl.2} & tii   &  tíì &  tíìsè & type of tree\\
%put as type of tree
  \lspbottomrule
 \end{Itabular} 
% }

\end{table}



 The rules in  (\ref{ex:GRM-Hrules}) capture the majority of the
singular/plural pairs of class 1 and 2. However, it is insufficient in some
cases, that is, there are cases which raise uncertainty in the allotment of
the pairs into one class or the other. Consider the examples in
Table \ref{tab:uncer-noun-class}.


\begin{table}[h]
\caption{Uncertain class 1 or 2 \label{tab:uncer-noun-class}}
\centering
%\subfloat[][{\sc class 3}]{
 \begin{Itabular}{lllll}
  \lsptoprule
 {\sc sg.} &   {\sc pl.} & Gloss\\ [1ex] 
\midrule
dʊ̃́ʊ̃̀  & dʊ̃́s{\T ꜜ}á  &   python\\
kìrìmá & kɪ̀rɪ̀nsá & tsetse fly\\
kɔ̀wɪ̀ɛ́   & kɔ̀wɪ̀sá   & soap\\
lɛ́hɛ́ɛ́  & lɛ̀hɛ̀sá&  cheek\\
%hõõ  & hõsa  & grasshopper\\
  \lspbottomrule
 \end{Itabular} 


\end{table}


 Two questions are raised by looking at the data in table
\ref{tab:uncer-noun-class}: (i) What is the stem of these nouns?  (ii) Is
there a good reason to favor final vowel deletion instead of insertion, e.g.
 /kɪrɪma/ vs. /kɪrɪm/?
Addressing  the first question, consider the first pair of words of table
\ref{tab:uncer-noun-class}, i.e. {\sls dʊ̃ʊ̃}  and {\sls dʊ̃sa}. On the one 
hand, 
if
 {\sls dʊ̃} is treated as   the stem and  the word for `python' is assigned to
class
1,   the refutation of the rule in   (\ref{ex:GRM-Hrules}) must be explained,
i.e.
vowel suffixes are always {\sc -hi}.  On the other hand, if  the stem
is  {\sls dʊ̃ʊ̃},  a deletion rule which reduces the length of the 
vowel, i.e. {\sls /dʊ̃ʊ̃-sa/}  $\rightarrow${\sls [dʊ̃́s{\T ꜜ}á]},  must be 
stated.
Such a decision  would
assign
the word for `python' to class 2.  The decision taken here is to respect the
rule in
(\ref{ex:GRM-Hrules}), which is empirically supported, and assume an {\it ad
hoc} deletion rule. The deletion rule may be explained by general prosody,
something which is not considered here. The word pairs in table
\ref{tab:uncer-noun-class} are assigned the following classes: `python' is in
class 2 and the last stem vowel is deleted in the plural, `tsetse fly' is in
class 1 and its stem is /kirim/, `soap' is in  class 1 and its stem is /kɔwɪ/,
and finally  `cheek' is in class 2 and the last stem vowel is
deleted in the plural.



 \paragraph{Class 3}
\label{sec:class3}

Nouns in class 3 generally have a sonorant coda consonant, i.e. {\sls l}, {\sls 
n}, 
{\sls r}, etc. Class 3 contains nouns whose singular forms have no overt
suffix and plural forms  which have a single vowel as suffix. As for class 1 and
2, the
plural vowel suffix of class 3 is determined by the harmony rule given in
(\ref{ex:GRM-Hrules}).



\begin{table}[h]
\caption{Class 3 \label{tab:freq-noun-class-3}}
\centering
%\subfloat[][{\sc class 3}]{
 \begin{Itabular}{lllll}
  \lsptoprule
{\sc class} & Stem & {\sc sg.} &   {\sc pl.} & Gloss\\ [1ex] 
\midrule

{\sc cl.3}  &  nɔn &  nɔ́ŋ   &  nɔ́nã́  &  fruit\\
%{\sc cl.3}  &  ɲiŋ &  ɲíŋ   &  ɲíŋá  &  tooth\\
%{\sc cl.3}  &  par &  pár   &  párá  &  hoe\\
%{\sc cl.3  &  kUr &  k\'Ur &  k\'Ur\'U  & bench\\
{\sc cl.3}  &  hããn & hã́ã̀ŋ   &  hã́ã̀nà  & woman\\
{\sc cl.3}  &  gɔŋ &  gɔ́ŋ  &  gɔ́ŋá  & river\\
%{\sc cl.3}  &  hamoŋ &  hàmõ̀ŋ  &  hàmõ̀nà  & child\\
{\sc cl.3}  &  nar &  nár   &  nárá  &  person\\
%{\sc cl.3  &  bUɔ ŋ   &  bʊʊ̀ŋ &  b\'Uná  & goat\\
%{\sc cl.3}  &  tɔn &  tɔ́ŋ &  tɔ́ná  & skin/book\\
%{\sc cl.3}  &  sɔŋ &  sɔ́ŋ &  sɔna  &  name\\
{\sc cl.3}  &  ʔol  &  ʔól &  ʔóló  & type of mouse\\
%{\sc cl.3}  &  ʔul  &  ʔúl &  ʔúló  &  navel\\
{\sc cl.3}  & butet &   bùtérː  &  bùtété & turtle\\
{\sc cl.3}  &   sel  &   sélː  & sélé  & animal\\
  \lspbottomrule
 \end{Itabular} 
 %}
 

\end{table}
 
 
 
 \paragraph{Class 4}
\label{sec:class4}

The major characteristic of class 4 is that all the stems have a final
syllable consisting of  {\sc [+hi, -ro]} vowel(s) (see table
\ref{tab:freq-noun-class-4}) .  Class 4 is analyzed in the following way: in
both  the singular and the plural, a  vowel is added to the stem, i.e. V]\# $>$
V]-V\#. The suffix vowel of the singular is always an exact copy of the stem
vowel.  If the stem vowel is {\sc [+atr, +hi]} the plural suffix vowel is {\sls 
-e},
 and if the stem vowel is  {\sc  [-atr, +hi]}, the  plural suffix vowel  {\sls 
-a}.
This low vowel is then raised due to the height of the stem vowel. In normal
speech, one can perceive either  {\sls -a} or {\sls -ɛ} in that position. Given 
the
rules in (\ref{ex:GRM-Hrules}),  class 4 is certainly the most problematic in
terms of uniformity. However,  class 4 is productive. A similar noun class was
found in both Tampulma, Vagla,  and  Dɛg. 


 
 \begin{table}[h]
\caption{Class 4 \label{tab:freq-noun-class-4}}
\centering

 \begin{Itabular}{lllll}
  \lsptoprule

{\sc class} & Stem & {\sc sg.} &   {\sc pl.} & Gloss\\ [1ex] 
\midrule

{\sc cl.4}  &  begi   &  bégíí &  bégíé  & heart\\
{\sc cl.4}  &  si   &  síí &  síé  & eye\\
{\sc cl.4}  &fili &fílíí&fílíé&bearing tray\\
{\sc cl.4}  &  bɪ   &  bɪ́ɪ́ &  bɪ́á  & stone\\
{\sc cl.4}  &  wɪ   &  wɪ́ɪ́ &  wɪ́ɛ́  & matter, thing\\
{\sc cl.4}  &  wɪlɪ   & wɪ́lɪ́ɪ́   &  wɪ́lɪ́ɛ́  & star\\
  \lspbottomrule
 \end{Itabular}
\end{table} 



Class 4 also includes nominalized verbal lexemes.  In Section
\ref{sec:GRM-verb-act-stem},  it is shown that one way to make  a noun from a
verbal lexeme is
to suffix a  high-front vowel to the verbal stem. For instance,  the lexeme  
{\sls zɪn} may be translated into English `drive', `ride' or `climb'. In the 
word  
{\sls kɪ́nzɪ̀nɪ́ɪ́} `horse', {\it lit.} thing-riding, the suffix  -[{\sc +hi,
-ro}]  is added to the verbal lexeme {\sls zɪn} making it nominal.
Consequently,
the plural of {\sls kɪ́nzɪ̀nɪ́ɪ́} `horse'  is {\sls kɪ́nzɪ̀nɪ́ɛ́}. The 
sequences 
{\sls -ie} and {\sls -ɪɛ} of class 4  often coalesce and may be  perceived as 
{\sls -ee}
and {\sls -ɛɛ}
respectively. 
 
 
 \paragraph{Class 5}
\label{sec:class5}


 The monosyllabic stems of class 5  nouns can either be CVV or CVC. Class 5
consists of nouns which  form their singular with no overt suffix and form their
plural with the suffix {\sls -nV}. The quality of the consonant is determined by
the stem and the place assimilation rules introduced in Section
\ref{sec:focus-forms}, some of which are repeated in  (\ref{GRM-cl-5}). The
vowel of the plural suffix is determined by the stem vowel and the rules in 
(\ref{ex:GRM-Hrules}). 
 


\ea\label{GRM-cl-5}
{\it Class 5 suffix -/nV/ surfaces -[lV] if the  coda consonant of the stem is
[l].}\\
{\sc -/[nasal]V/}$_{nc}$  $>$  {\sc -/[lateral]V/}$_{nc}$  /  {\sc [lateral]} 
\_\\

\z


 
 \begin{table}[h]
 \caption{Class 5 \label{tab:freq-noun-class-5}}
\centering
 \begin{Itabular}{lllll}
  \lsptoprule
{\sc class} & Stem & {\sc sg.} &   {\sc pl.} & Gloss\\ [1ex] 
\midrule

{\sc cl.5}  &  zɪn &  zɪ̀ŋ́ &  zɪ́nná  &  type of bat\\
{\sc cl.5}  &hʊ̃n&hʊ̃̀ŋ́&hʊ̃́nná& farmer or hunter gear\\
{\sc cl.5}  &  kuo &  kùó   &  kùónò  & farm\\
{\sc cl.5}  &  ɲuu &  ɲúù   &  ɲúúnò  & head\\
%{\sc cl.5}  &  sũũ &  sũ̀ũ̀   &  sũ̀ũ̀nó  &  guinea fowl\\
{\sc cl.5}  &  vii & víí   &  vííné &   type of cooking pot\\
{\sc cl.5}  &din&díŋ & dínné & fire\\
{\sc cl.5}  &pel &pél & péllé & burial specialist\\

  \lspbottomrule
 \end{Itabular}
\end{table} 
 

 \paragraph{Nasals in noun classes}
\label{sec:gene-sum}


 
Apart from the singular of class 4,  much of the same vocalic morpho-phonology
is found in all classes. This was reduced to the two rules in
(\ref{ex:GRM-Hrules}). Furthermore, in all the noun classes, the nasal
consonants surface differently depending on the phonological context. The rules
in  (\ref{ex:GRM-nrules}) predict the observed outputs and are derived from the
nasal assimilation rules in Section \ref{sec:internal-sandhi-nasal-place}.
 
\ea\label{ex:GRM-nrules}\textit{Possible outputs of  nasals}\\

\ea\label{ex:GRM-Nrules}
 C[{\sc +nasal}]\   $>$ ŋ / \_ \#\\
 /hããn/  $>$  [hã́ã̀ŋ]  `female' {\sc cl.3.sg}   


\ex\label{ex:GRM-Msrules}
 /m/ $>$ n / \_  C [{\sc -labial, -velar}]\\
 /tʃuom/   $>$ [tʃùònsó]   `togo hares'  {\sc cl.1.pl}  

\ex\label{ex:GRM-NGsrules}
 /ŋ/ $>$ n / \_  C [{\sc -labial, -velar}]\\
/kɔlʊ̃ŋ/ $>$  [kɔ̀lʊ̀nsá]  `wells'   {\sc cl.2.pl} 


\z 
 \z

The rule in  (\ref{ex:GRM-Nrules})  says that  any nasal consonant occurring
word finally becomes [ŋ]. The rule in (\ref{ex:GRM-Msrules}) changes a bilabial
nasal into an alveolar when it precedes a non-labial and non-velar consonantal
segment. The rule in (\ref{ex:GRM-NGsrules}) changes a velar nasal into an
alveolar in the same environment.

 
 \paragraph{Generalization and summary}
\label{sec:gene-sum}

While the method proposed suggests that one should look for pairs of forms, the
present classification treats phonologically empty suffixes as `exponents'. What
counts as a noun class is the paradigm determined by the  inflectional
pattern of the lexeme. The five  most frequent pairs were presented in tables
\ref{tab:freq-noun-class-1} to \ref{tab:freq-noun-class-5} and the exponents are
gathered in  Table \ref{tab:GRM-nc-exponent}.\footnote{The percentage is based
on a list
of 978 singular/plural pairs  (lexicon 02/10/10 version). The five classes in
Table \ref{tab:GRM-nc-exponent} make up 88\% of the nouns which are assigned a
class in the lexicon.} 
%see all numbering/percentage as it was changed

 \begin{table}[!h]
 \caption{The five most frequent noun classes   \label{tab:GRM-nc-exponent}}
   \centering
   \begin{Itabular}{lccccc}
 \lsptoprule
    &  {\sc cl.1} & {\sc cl.2}  & {\sc cl.3} & {\sc cl.4} & {\sc cl.5} 
\\  [1ex] \midrule
{\sc sing} & -V&  \O&  \O& -V  & \O\\
{\sc plur} & -sV& -sV & -V & -V  & -nV\\ \midrule
    &   8\%&  32\%  &  23\% &   17\%   & 8\%\\
 \lspbottomrule
   \end{Itabular}
 \end{table}



In practice the most productive and regular patterns are those recognized as
noun classes. However, some words do not fit perfectly into the patterns
described
above but are not totally alien to genetically related languages and the
reconstructions of Proto-Grusi in \citet{Mane69a, Mane69b} and
Proto-Grusi-Kirma-Tyurama  in \citet{Mane82}.   In fact, there are more
possibilities
and surfaces forms when the  classes  ({\it sg./pl.})  {\sls \O/\O},  {\sls 
\O/ta},
{\sls \O/ma} and {\sls ŋ/sV} are included in the classification. Examples are 
given  in table
\ref{tab:GRM-less-pro-nc}.  
 
\begin{table}[h]
\caption{Noun classes 6, 7, 8, and 9 \label{tab:GRM-less-pro-nc}}
\centering
  \begin{Itabular}{lllll}
  \lsptoprule
{\sc class} & Stem & {\sc sg.} &   {\sc pl.} & Gloss\\ [1ex] 
\midrule

{\sc cl.6}  & dʒɪɛnsa & dʒɪ́ɛ̀nsá & dʒɪ́ɛ̀nsá & twin\\
{\sc cl.6}  &kapʊsɪɛ &  kápʊ̀sɪ́ɛ́ & kápʊ̀sɪ́ɛ́ & kola nut\\
{\sc cl.6}  & kpibii & kpìbíí & kpìbíí & louse\\[0.2ex] \midrule

{\sc cl.7}  & kuo & kúó &kùòtó  &roan antelope\\
{\sc cl.7}  &kie  &kìé & kìété & half of a bird\\
{\sc cl.7}  &fɔʊ̃ &fɔ́ʊ̃̀& fɔ̀tá & baboon\\[0.2ex] \midrule
%{\sc cl.7}  &  taa &  tàá &  tàátá  & language\\
{\sc cl.8}  & naal &náàl&nááləmà&grand-father\\
{\sc cl.8}  &ɲɪna &ɲɪ́nà&ɲɪ́námà &father\\
{\sc cl.8}  &  hɪɛŋ & hɪ́ɛ́ŋ &hɪ́ɛ́mbá &relative\\[0.2ex] \midrule

 {\sc cl.9}  &  jo   &jóŋ̀ & jósò  & slave\\
{\sc cl.9}  & zi &zíŋ̀ &zísè &tail\\
{\sc cl.9}  & ŋmɛ&ŋmɛ́ŋ̀&ŋmɛ́sà&rope\\

  \lspbottomrule
 \end{Itabular} 

\end{table} 


 The nouns in class 6 do not formally differentiate singular and plural.   Those
in class 7 mark their plural with the suffix {\sls -tV} and  class 8 with the 
suffix {\sls -mV}.  The singular exponent of class 7 and 8 is covert. Finally,
the nouns of class 9 have a suffix {\sls -ŋ} in the
singular and {\sls -sV} in the plural. In Table \ref{tab:l-leasttive-class},  
the
percentage of occurence of the less productive noun classes 6, 7, 8
and 9 is given.
 
  
 \begin{table}[h]
   \caption{Less productive  noun classes 
\label{tab:l-leasttive-class}}
   \centering
   \begin{Itabular}{lcccc}

  
 \lsptoprule
&  {\sc cl.6}   & {\sc cl.7}  &  {\sc cl.8} & {\sc cl.9}\\
[1ex] \midrule
 {\sc sing}   & \O  & \O   & \O & -N\\
{\sc plur}   & \O & -tV  & -mV & -sV\\ \midrule
    &   7\%&   1.8\% &   0.9\%   & 0.8\%\\
 \lspbottomrule
  
   \end{Itabular}
 \end{table}

In addition, there are pairs which can only imperfectly be reduced  to the nine
classes presented until now. However, the problem lies in the stem and not in
the inflectional patterns. For example the color terms ({\it sg./pl.}) {\sls 
pʊ̀mmá}/{\sls pʊ̀lʊ̀nsá} `white' and {\sls búmmó}/{\sls bùlùnsó} `black' 
 do
not have comparable pairs and do not fit the noun classes described above. One
would expect *{\sls pʊmmasa} to be the plural form for  `white' (also *tɪɪnama 
for
{\sls tɪ̀ɪ̀ná}/{\sls tʊ́mà} `owner'). Other examples are the  pairs {\sls 
tɪ́ɛ̀}/{\sls tɛ́sà} `foetus' and {\sls túò}/{\sls tósó} `bow'. Also here, 
one 
expects  the
last vowel to delete in each of the plural forms instead of the penultimate one.
Moreover,  inconsistent class assignment across speakers, across villages, and
surprisingly different forms (predominantly in the plural) from the same speaker
on different elicitation sessions do arise, but the latter case rarely occurs. 
 

%also tʊ̀ɔ́nɪ́ã̀ tʊ̀ɔ́nsà  type of genet
%also kùòlíè  kùòlúsò  type of tree 
%apocope?
%  hɔ̃́ʊ̃̀
%  type of grasshopper
% \cl 2
%  hɔ̃́sà


 
\paragraph{Semantic assigment criteria}
\label{sec:GRM-sem-ass-crit}

Several authors have presented their views on  the semantic classification of
nominals.   The general idea is that there must be an underlying system which
can explain, first, why some words display identical number morphology, and
second, how these words are related in `meaning'. \citet[23]{Tcha07} shows that
Tem organizes its nominals on the basis of semantic values such as humanness,
size, and countability. \citet[41]{Awed07} argues that nominal groupings  in
Kasem should take into consideration phonological and semantic characteristics,
in addition to other more cultural factors.  Similarly, \citet{Assi07}
argues at length on the shortcoming of traditional semantic rules and argues for
abandoning them. 

The semantic value of the noun class suffixes has proven difficult to
establish. It is possible that there are analogies in class assignment based on
semantic criteria but it is more likely that synchronically (i) the phonological
shape of the stem triggers the suffix type, and that (ii) some classes can be
identified as residues of former semantic assignment. Let me comment on each of 
these
points. 

First, most class 3 nouns have  a sonorant consonant in the coda position,
the stems of  class 4 nouns must have their last vowel specified for  [{\sc -hi,
-ro}] and a typical class 2 noun is either   CVV or CVCV.  These are some of
the characteristics  described for the noun classes. It seems that the
phonological
shape of the stem plays a role in class assignment and that there is no
productive class
where most of its  members are assigned to a particular semantic domain.  Using
four features of the animacy hierarchy
of  \citet{Comr89}, i.e.  human $[${\sc
hum}$]$, animal (exclude human) or other-animate
and insects $[${\sc anim}$]$, concrete inanimate $[${\sc conc}$]$ and abstract
(inanimate) $[${\sc abst}$]$,  \citet{brin08} shows that the noun
classes  do not encode any of these distinctions. Such
distinctions may have
been  expected given the nominal classification of other Gur languages. For
instance in Dagaare, an Oti-Volta language and linguistic `neighbor' of
Chakali, \citet[124]{Bodo94} presents the Class 2 (V/ba) as ``unique in that it
is the only class that has exclusively [+human] nouns in it''. From a
diachronic point of view, this suggests that Chakali has dropped all animacy
distinctions in the noun class system while preserving one distinction in
agreement (see Section \ref{sec:GRM-gender}).


Secondly, languages related to Chakali, i.e. geographically and genetically,  
had/have noun class systems whose classifications were based, at least 
partially, on semantic criteria. To my knowledge, the most conservative system 
today  within Grusi is Tem (see {\it identification sémantique} in 
\citet{Tcha07}). When and how the speakers of Chakali  classified nouns based on 
semantic criteria is impossible to know,  but traces can be detected in  the   
{\it less productive noun classes}, that is class 6, 7, 8, and 9. Some members 
of class 6 consist of nouns with mass or abstract  denotations, i.e. rice,  
louse, struggle, profit, etc.  Class 7 also contains mass and abstract nouns, 
i.e. oil,  honey, water, and taboo, but also bush animals such as bushbuck, 
waterbuck, baboon, roan antelope and hartebeest. Class 7 represents 
approximately 2\% of the noun sample (see Table \ref{tab:l-leasttive-class}) and 
 mass/abstract nouns and bush animals represent each  30\% of class 7 
membership. Class 8 is likely to be the class where kinship and human 
classification terms were assigned, as mother, father, and `owner of' are among 
remnant members of that class.  Finally, a  common trait of class 9 may be 
`elongated things', since words referring to  rope, arm, tail, and ladder are 
members. Yet, only eight nouns are assigned to class 9. Despite the arbitrary 
nature of the semantic assignment of class 9,  \citet[94]{Mane75} maintains that 
there are Oti-Volta languages which show relics of  the Proto Oti-Volta class 
{\sls *ŋu- *u}, which is  itself a remnant of Proto-Gur class 3   according to 
\citet[11]{Mieh07}, and that this class contains ``les noms du bâton, du pilon, 
du balai, de la corde, de la peau et du chemin''.  Although these nouns seem to 
denote `elongated  things',   Manessy claims that they cannot contribute to an 
hypothesis. Generally, howewer, the fact that members of classes 6, 7, 8, and 9 
are similarly clustered in other languages suggests that these classes are 
remnants of a more productive semantic assignment system. Beside semantic 
domains, the simple empirical fact that homonyms are found with different 
suffixes excludes a purely phonologically-based class assignment. There is no 
way a speaker can correctly pluralize the stems {\sls tíì} `type of tree' and 
{\it tíí} `type of ants' based entirely on their (segmental) phonological 
shape.\footnote{I put segmental in parenthesis since  homonyms {\it with the 
same tonal melody} belonging to two different classes have not yet been  found. 
The pair {\it pól} ({\sc cl.5}) `river' and {\it pól} ({\sc cl.3}) `vein' may 
be treated as one example, but their meanings point to a common etymology. 
Nevertheless, \citet{Bonv88}, \citet{Awed07} and \citet{Tcha07} provide data to 
support a similar claim.} 

% % A combination of both (i) and (ii) seems consistent with the data observed. Chakali speakers 
seem to 
% % acquire the noun classes as French or Dutch speakers acquire the grammatical gender of 
inanimate 
% % % entities which lack natural gender. %mostly semantically arbitrary %a healthy mix of lexical 
storage 
% % % and %rule-based semantic and phonologica prediction
% 
% Finally, class assignment in complex stem nouns indicates  that the 
% denotation of a word plays no role in determining its noun
% class (see Section \ref{sec:GRM-com-stem-noun}). 
% The class of a complex stem noun is always determined by the rightmost stem.
% Given that compounding is highly productive,  this purely  formal
% process suggests that semantic criteria in
% noun class assignment are inoperative.

\paragraph{Tone patterns of noun classes}
\label{sec:GRM-tone-p}

Some tonal melodies are identified. One of them is the  general tendency for
the singular and  plural words in a pair to display the same tonal melody. For
instance, a HL melody may be associated with both the singular and the plural,
e.g.  {\sls zíŋ̀}/{\sls zísè} `tail' ({\sc cl.9}) 
and   {\sls lʊ́l̀}/{\sls lʊ́là} `biological relation'  ({\sc cl.3}). These 
cases 
are tonally regular. 
Another common pattern is when a singular noun displays a H melody, but the
plural a LH melody, e.g.  {\sls dáá}/{\sls dààsá} `tree' ({\sc cl.2}). 
While 
it
 seems that  the
plural suffix -{\sls sV}  depresses a preceding H,  it does not do so in class 9
nouns.
The majority of class 4 nouns in the data available are high tone irrespective
of the number of moras and they are all tonally regular. Some cases involving
singular CVC words with moraic
coda exhibit the deletion of a low tone;  {\sls zɪ̀ŋ́}/{\sls zɪ́nná} `bat' 
({\sc 
cl.5}),   {\sls gèŕ}/{\sls gété} `lizard' ({\sc cl.3}) and
{\sls sàĺ}/{\sls sállá} `flat roof' ({\sc cl.3})  have a LH tonal melody in 
the
singular but  H in the plural. The downstep rule (Section
\ref{sec:tone-intonation})  predicts that a high tone preceded by a low tone is
perceived as lower than a preceding high
tone, e.g. {\sls váà}  {HL},  {\it vá{\T ꜜ}sá}  {HLH}  `dog' 
({\sc cl.1}).  
In
spite of some variations,  it seems that there are recurrent
melodies. Representative examples are presented in
Table \ref{tab:GRM-tm-nc-1-5}.
 

 \begin{table}[htb!]
   \caption{Tonal melodies in noun classes 1--5
\label{tab:GRM-tm-nc-1-5}}
   \centering
   \begin{Itabular}{lp{1cm}lp{1cm}ll}

 \lsptoprule
{\sc class} &  Tone  melody {\it sg.}  &   Singular   &  Tone  melody  {\it
pl.} &   Plural & Gloss
\\ [1ex]

\midrule

{\sc cl.1}   &   HL   & váà & HLH &   vá{\T ꜜ}sá  & dog\\
  &  LH &  gùnó&  LH& gùnsó  & cotton\\
& HL & tʃíníè & HL &  tʃínísè  & type of climber\\
  & L &  dɪ̀gɪ̀nà  & LH & dɪ̀gɪ̀nsá&  ear\\[0.2ex] \midrule


{\sc cl.2}  &  H  &   síé &  LH   &
sìèsé&  face\\
&  L   &  bɔ̀là   & LH&   bɔ̀làsá  &
elephant\\
&  LH &  tòmó  &  LH &  tòmòsó  &  type of
tree\\
&   LH   & sòntògó &  LH &   sòntògòsó & base\\
&  HL &  júò  &  HLH   &  júòsó  &
quarrel\\[0.2ex] \midrule


{\sc cl.3} &   H &   hóg  &   H   &   hógó  &   bone\\
&   H   &  sɔ́nná  &  H  &  sɔ́nnəsá
& lover\\
 &  LH &  gèŕ  &  HH  &  gété  
& lizard\\
 &  LH &  pààtʃák  &  LH   &  pààtʃàgá 
& leaf\\[0.2ex] \midrule


{\sc cl.4} &  H &  síí  &  H& síé  &  type of dance\\
 &  H &  bégíí &  H & bégíé
& heart\\
& H & tʃɪ̃́ɪ̃́  & H & tʃɪ̃́ã́ & dawadawa seed\\
& LH& sòkìé & LH & sòkìété & type of tree\\[0.2ex] \midrule




{\sc cl.5}&H &  víí  &  H &  vííné
& cooking 
pot\\
  &  L &  bɔ̀g  &  L &  bɔ̀ɣənà   &
type of tree\\
  &  HL &  bámpɛ̀g  &  HL&   bámpɛ̀gənà  &
half of nut\\
  &  LH &  kùó  &  LHL &  kùónò  &  farm\\
  &  LH &   zɪ̀ŋ́  &  H&   zɪ́nná  &  bat\\
  &  L &  tʃàl   &  LH&   tʃàllá  &  blood\\
 &  LH &  sàĺ  &  H &  sállá &  flat
roof\\

  
 \lspbottomrule
   \end{Itabular}
 \end{table}





\paragraph{Noun class reconstruction}
\label{sec:GRM-noun-class-recons}
The numerical labeling of the noun classes in Table \ref{tab:GRM-nc-exponent}
and \ref{tab:l-leasttive-class} is arbitrary. Given  the state of the
documentation on nominal classifications in other SWG languages, and the fact
that  almost all singular suffixes
have disappeared in today's SWG languages, a reconstruction  is practically
impossible. Nonetheless,  some 
observations  on similarities between the noun class
system in Chakali and other SWG noun class systems can be put forward. The
information sources are my field notes on neighboring languages, the
reconstruction of the
noun class suffixes of Grusi in \citet{Mane69a, Mane69b},  and the 
reconstruction
of noun classes in Gur in \citet{Mieh07}; the latter being for the most part an
up-date and synthesis of Manessy's work \citep{Mane69a, Mane69b, Mane75, Mane79,
Mane82, Mane99}. Needless to say, the following statements are first
impressions.

%need name of pasaale girl

Field notes on neighboring languages, supported with unpublished
material produced
by GILLBT's staff,\footnote{\label{ft:GRM-naden-donate}In 2008, Tony Naden gave
me  a copy of his ongoing
Vagla and Dɛg lexicons. I am also 
indebted to: Kofi Mensa (New
Longoro) for Dɛg, Modesta
Kanjiti  (Bole) for Vagla and Dɛg, Pastor Alex Kippo (Tuossa) for Vagla and 
Yusseh Jamani (Bowina) for Tampulma.}  provided relevant information on
the
(dis-)similarities of Chakali with other SWG languages. As in all SWG languages,
a typical Vagla noun class is characterized by  suffixation.
The most frequent plural markers in Vagla are {\sls -zi}, {\sls -nɪ} and
{\sls -ri}. The pattern found in Chakali  class 4 is similar to the one found in
Vagla, i.e.
({\it sg.}/{\it pl.}) {\sls bàmpírí}/{\sls bàmpíré} `chest',  {\sls 
hūbí}/{\sls hūbé}  `bee' and   {\sls gíngímí}/{\sls gíngímé} `hill'.  
In 
Dɛg,   the most
frequent plural markers are mid vowel suffixes, often rounded,  and
the {\sls -rV}, {\sls -nV} and  {\sls -lV} suffixes, with which the vowel 
harmonizes
in
roundness and {\sc atr} with the stem vowel. Both Vagla and Deg display 
miscellaneous classes which are characterized by  a simple difference
in vowel quality between the last vowel of the singular and the plural, e.g. Dɛg
{\sls dala}/{\sls dale} `cooking place'. Attested
alternations  ({\it sg.}/{\it pl.}) in Vagla are {\sls -i}/{\sls -e},  {\sls 
-i}/{\sls -a},  {\sls -a}/{\sls -i}, 
{\sls -u}/{\sls -a},  {\sls -o}/{\sls -i} and  {\sls -e}/{\sls -i},   and in 
Dɛg {\sls 
-a}/{\sls -e}, {\sls -e}/{\sls -a}, {\sls -i}/{\sls -e}, {\sls -o}/{\sls -i} 
and  {\sls 
-i}/{\sls -a}.\footnote{These singular/plural pairings are extracted from the 
Vagla and
Deg lexicons (fn. \ref{ft:GRM-naden-donate}) and are not exhaustive.}  The noun
classes of Tampulma and Pasaale
correspond more to those of Chakali. Tampulma has at least
the
following class suffix pairs ({\it sg.}/{\it pl.}): {\sls \O}/{\sls -V}, {\sls 
-i}/{\sls -e}, {\sls \O}/{\sls -nV},  {\sls  \O}/{\sls -sV}, {\sls  -V}/{\sls 
-sV},  
{\sls -hV}/{\sls -sV} and  
{\sls \O}/{\sls -tV}.
Tampulma displays similar harmony rules to those found in Chakali. Apart from
the singular suffix {\sls -hV}, all the noun class suffixes in Tampulma are
manifested in Chakali.  Correspondingly, Pasaale reveals  pairs and
harmony rules similar to those of Chakali and Tampulma.\footnote{As mentioned in
footnote
\ref{foot:noun-class}, the number of noun classes is determined by the
linguist's analysis.  \citet[5--12]{Mcgi99} is a good example of 
the consequence of analyzing noun classes differently. For instance,  
 \citet[7]{Mcgi99} postulate a subclass  ({\it sg.}/{\it pl.})  {\sls -l/-lA} 
for 
word pairs like {\sls baal/baala} `man', {\sls gul/gulo} `group', 
{\sls miibol/miibolo} `nostril' and  {\sls mɔl/mɔlɔ} `stalk'. If these words 
were 
part
of the
Chakali data, they would have been alloted to class 2 ({\sls -\O/-V}), that is,
I would have treated the /l/ as a coda consonant of the stem instead of a noun
class
suffix consonant. In addition, whereas I derive the quality of the vowel
entirely from harmony rules,  \citeauthor{Mcgi99} assume archiphonemes, like A
and E, which surface depending on harmony rules.}

It is important to keep in mind that the analysis in \citet{Mane69a, Mane69b} is
based on a very
limited set of SWG data,  most of the data being extracted from
\citet{Bend65}. He
stresses often the tentative nature of his claims and  sets forth more than one
 hypothesis on several occasions. The Chakali plural suffix of class 8 {\sls 
-mV}
may be treated as a descendant of the Proto-Grusi Class {\sls *B$_{1}$A}
\citep[32]{Mane69b}, class 9 {\sls -ŋ} as a descendant of the Proto-Grusi Class
{\sls *NE}   \citep[37, 41]{Mane69b}, class 1 {\sls -V} as a descendant of the
Proto-Grusi
Class {\sls *K$_{1}$A}  \citep[39]{Mane69b}, classes 1, 2, and 9 {\sls -sV} as 
descendants
of the Proto-Grusi Class  {\sls *SE}  \citep[39]{Mane69b} and class 7 {\sls 
-tV} 
as
a descendant of the Proto-Grusi Class {\sls *TE/O}  \citep[43]{Mane69b}. The 
vowel
suffixes of class 1 and 4 may also descend from the Proto-Grusi Class {\sls *YA}
\citep[34]{Mane69b}. 

In consulting \citet[7--22]{Mieh07}, Chakali's  most frequent plural suffix  
{\sls -sV}, found in class 1, 2, and 9, would seem to correspond to Proto-Gur 
Class 13
*{\sls -sɪ}, the plural suffix of class 5 {\sls -nV} to  Proto-Gur Class 2a 
*{\sls 
-n.ba}
or Proto-Gur Class 10 *{\sls -ni}, class 7 {\sls -tV} to Proto-Gur Class 21 
*{\sls 
-tʊ} and class 8  {\sls -mV}  to Proto-Gur Class 2 *{\sls -ba}. The singular 
suffix
{\sls -ŋ} would correspond to Proto-Gur Class 3 *{\sls -ŋʊ}.


Needless to say, these observations  deserve further investigation. Even though
there is  literature to support the reconstruction of the Gur classes, little
can be done in the SWG
area unless descriptions of  nominal classifications in the languages  Winyé,
Vagla, Tampulma, Phuie,  Deg, Siti,  and the dialects of Sisaala  are made 
available. A synthesis of these
descriptions could be
compared to  `better-documented' nominal classfications of Grusi languages 
such as Kasem (Northern Grusi, \citet{Awed79, Bonv88, Awed03}),  Lyélé (Northern
Grusi, \citet{Delp79}),  Lama  (Eastern Grusi, \citet{Arit87, Ours89}), Kabiyé
(Eastern Grusi, \citet{Tcha07}),  Chala   (Eastern Grusi, \citet{Klei00}) and 
Tem
(Eastern Grusi, \citet{Tcha72, Tcha07}), to evaluate the Proto-Grusi noun class
suffixes of \citet{Mane69b} and Proto-Gur of \citet{Mieh07}, and to reconstruct
the nominal classifications of SWG  languages.


\subsubsection{Atomic stem nouns}
\label{sec:GRM-sim-bas-noun}

The notion of stem in the present context refers to the host of a noun class
suffix or the  host of a nominalizer, i.e. the element which conveys the lexical
meaning and  to which affixes attach. A stem can be either irreducible or
reducible morphologically: they are referred to as atomic  and complex stem
respectively.  Complex stems are presented in  Section
\ref{sec:GRM-com-stem-noun}.   An atomic stem is always a  nominal or a verbal
lexeme.  A verbal lexeme may either be of the type `state' or `process'. Three
types of nominalization formation (i.e. nominalizers) are attested: suffixation,
prefixation, and reduplication.  

%A few prefixes are attested. Some are treated as classifiers (see Section
%\ref{sec:}). Other formal means for nominalization are suffixation and
%reduplication. 

\paragraph{Nominal  stem}
\label{sec:GRM-nom-stem}
A nominal stem is a predicate denoting a class of entities.   Nouns composed by
the combination of  a nominal stem and a noun class affix are the most common. A
nominal stem can be juxtaposed with various noun class affixes, yielding forms
with
different meanings. For instance, the lexeme {\sls baal} is associated with the
general meaning `man'. In a context where the lexeme is used in the singular,
{\sls baal} can mean either `a man' or `a husband'. Given the same context but
used in the plural, the lexeme {\sls baal} is disambiguated by the 
plural suffix it takes;  {\sls baala} `men'  ({\sc cl.3}) and  {\sls baalsa}
`husbands'  ({\sc cl.2}). Another
example is the lexeme {\sls natɔʊ} `shoe'. The word {\sls natɔwa} refers to `a 
pair
of shoes' whereas {\sls natɔʊsa} refers to `pairs of shoes' or `unsorted shoes'.
Evidence from other Grusi languages suggests that the situation where lexemes
are found in different noun classes was certainly a   more common
phenomenon than it is today \citep[126--128]{Bonv88}. This may coincide
with semantically richer
noun class suffixes. In addition, for many noun classes the singular forms are
not overtly marked and the plural forms are by and large less frequent. This
situation makes it difficult to provide the necessary evidence which would
demonstrate that nominal stems are found together with different noun classes.  

Nominal stems exist in opposition to the verbal ones. To classify a stem in such
a dichotomy, the linguistic test carried out consists of placing the stem in
several core predicative positions, i.e. positions where an
argument must appear. If the stem is perceived as grammatical in the given
context by 
language consultants, it cannot be nominal. For instance, in French the word
{\sls bille}  `marble' cannot take a nominal argument in a non-genitive
predication, e.g. *Marie billait/a billé  `Mary marbled/has
marbled'.\footnote{\label{ft:GRM-fre-eng-deri}French and English are not 
appropriate languages to use
for the point I want to make  because of their preference  for the categorial
derivation n$>$v. I believe that
Chakali cannot as easily derive a verb from a noun.}

%whether this is grammar or social convention is not answered


The examples in
(\ref{GRM-nom-or-verb}) illustrate a simple classification procedure. It uses a
frame where the predicate is in the 
perfective aspect and  the same 
predicate, as opposed to the argument,  is in focus.

\ea\label{GRM-nom-or-verb}
 
\ea /di/   {\rm`eat'}  $\rightarrow$ {\it ʊ̀  díjōó}  {\rm |{\sc 3.sg} 
eat.{\sc pfv.foc}| 
`he
ate'}

  \ex /kpeg/  {\rm `hard'}  $\rightarrow$ {\it ʊ̀ kpégéó} {\rm  |{\sc 3.sg} 
hard.{\sc
pfv.foc}| `he is strong'}

   \ex /sɪama/  {\rm`red'}  $\rightarrow$ *{\it ʊ sɪamao}, {\rm but}  {\it ʊ̀ 
sɪ́árēó} 
 {\rm |{\sc
3.sg} red.{\sc
pfv.foc}|  `it is
red'} 
   \ex /bi/  {\rm `child'} $\rightarrow$  *{\it ʊ bio} 
 % \ex //
 

\z 
 \z


The test displayed in (\ref{GRM-nom-or-verb}) shows that  {\sls di} and  {\sls 
kpeg} are verbal,  whereas {\sls sɪama} and {\sls bi} are not. In
Section \ref{}, it will be shown that color terms change forms
depending on whether
they occur  in a nominal or verbal context (TO DO).  


\paragraph{Verbal process stem}
\label{sec:GRM-verb-act-stem}

%Somewhere should be included a section on derivative affixes (and/or other
%types of nominalizer) 

Verbal process stems denote non-stative events. Table \ref{tab:GRM-nom-process} 
displays  two types of nominalization formation -- suffixation and 
reduplication -- involving verbal process stems,  `agent of X' and `action of 
X', where X replaces the meaning of the verbal process stem. 

\begin{table}[htb!]

\centering
\caption{Examples of nominalization of verbal process stem
\label{tab:GRM-nom-process}}
 \begin{Itabular}{llll}
 \lsptoprule
Sem. value & Verb. process stem & Nmlz & Form\\
 \midrule

Agent of X &  gʊ̀ɔ̀ `dance' &  -/r/ & gʊ́ɔ́r `dancer'\\
Agent of X &  kpʊ́  `kill' &   -/r/  & kpʊ́ʊ́râ  `killer'\\
Agent of X &   búól   `sing' &  reduplication &   bùòlbúóló  `singer'\\
Agent of X &   sùmmè `help' &  reduplication &   súsúmmá 
`helper'\\[1ex]\midrule

Action of X  & gʊ̀ɔ̀ `dance' &  -/[{\sc +hi, -bk}]/ & gʊ́ɔ́ɪ́ɪ́ `dancing'\\
Action of X &  kpʊ́  `kill' &  -/[{\sc +hi, -bk}]/  & kpʊ́ɪ̀ɪ́  `killing'\\
Action of X  &  búól   `sing'  &-/[{\sc +hi, -bk}]/  & búólíí 
`singing'\\
Action of X  &   sùmmè `help'   &-/[{\sc +hi, -bk}]/  &  súmmíí  `helping'
\\
\lspbottomrule
 
 \end{Itabular} 

\end{table} 


% The first column describes in prose the meaning of each nominalization,
%the second column provides the stem, the third column provides  the
%nominalization formations and the fourth provides the  translation.


 In Table \ref{tab:GRM-nom-process}, the column entitled semantic value (i.e.
Sem. value)  identifies the meaning of the verbal nominalization. In such a
context, `agent of X'  refers to the instigator or doer of the state of affairs
denoted by  the predicate X and
the nominalization is generally accomplished by the suffix -/{\sls r(a)}/.  
However,  there are some expressions with the equivalent
agentive denotation which
do
not suffix -/{\sls r}/  to the  predicate, e.g.  {\sls ʔɔra} `to sew' vs.
{\sls ʔɔta} `sewer' and {\sls maŋa} `to beat' vs.  {\sls kɪŋmaŋana} `drummer'.  
The
singular forms  are given in the fourth column: 
the plural of agent nominals  of
this type, i.e. nominalized by the suffix -/{\sls r}/, is made by  a single 
vowel
suffix  ({\sc cl.3}) whose surface
form  depends on harmony rules.\footnote{One language consultant
had a problem retrieving the plural of some agent nouns. He often repeated the
singular entry
for the plural. I interpret this as  either a situation where agent nouns do
not show differences in the singular and plural ({\sc cl. 6}), or different {\it
sg.}/{\it pl.}
forms exist but he could not retrieve them. The pair {\sls kpʊra}/{\sls kpʊrəsa}
`killer(s)'   is unusual.  The word {\sls  sãsaar} means
`woodcarver' and not `car driver'  even though {\sls sãã} can mean both 
`carve'
and `drive vehicle'.  People usually use  {\sls lɔ́ɔ́lɪ̀sã́ã́r}, or the 
English
word {\sls dərávɛ̀}, 
 which is common all over Ghana, to refer to a `car driver'. } Another verbal
nominalization process conveying `agent of X' is reduplication. The evidence
suggests that  only the first syllable  is reduplicated.

The second nominalization process is  interpreted as `action of X' or
`process of X' and
consists of the suffixation of a  high front vowel to the verbal
 stem.\footnote{The nominalization `the process X' is often not distinguishable 
from
`the result of a process X'.  Does `dancing'  refer to `the process of dance', 
`the result of the process of dance' or both?} The surface form of the vowel
depends
on the quality of the stem vowel and {\sc atr}-harmony  (see
rule \ref{RULE-atr} in Section \ref{sec:vowel-harmony}). 
Consider example (\ref{ex:vp36.1.}).


\ea\label{ex:vp36.1.}
\glll ʊ̀ píílì wáɪ́ɪ́ rá.\\
ʊ pile wa-ɪ-ɪ ra\\
  {\sc 3.sg} start come-{\sc nmlz}-{\sc cl.4} {\sc foc}\\
\glt  `He begins coming' 
\z

 The final vowels in the
words referring to `the process of X' are analyzed as a sequence of two
vowels: first a nominalizer suffix (i.e. {\sc nmlz})  on the verbal stem,  and
second,  a noun class suffix.  Such nominalized verbal stems are alloted to
noun class 4;  their singular suffix is a copy of the {\sc nmlz} vowel,
and their  plural suffix is the low vowel {\sls a}, raised to a mid height, e.g.
{\sls pɛrɪɪ}/{\sls pɛrɪɛ} `weaving(s)'  ($<$ {\sls pɛra} `weave', see class 4
in Section \ref{sec:class4}).

%The infinite forms are always prefixed by the low vowels. The transcription
%shows harmony in some instances but it is a topic to further investigate. 


\paragraph{Verbal state stem}
\label{sec:GRM-verb-state-stem}

Verbal state stems  denote static events. They generally
function  as verbs, but they can take the role of attributive modifiers in
noun phrases, referred to as  `qualifiers' in Section \ref{sec:GRM-qualifier}.
In that role, their semantic value is similar to the value of adjectives in
English: they denote a property  assigned to a referent.  To function as  a
qualifier, some verbal state predicates must be nominalized. As with verbal 
process stems,  verbal state stems are found in nouns which
have
been nominalized by suffixation of a  high front vowel, i.e. `the state of X'.
For instance, the verbal state predicate {\sls kpeg} has a general meaning 
which 
can be
translated into English as `hard' and `strong'. The expression {\sls kpégíí} 
in
{\sls a teebul kpegii dʊa de} `The hard table is there' functions as qualifier in
the noun phrase {\sls a teebul kpegii} {\it lit.} `the table hard'. 


\ea\label{exːGRM-v-sta-p-hard}{\rm Verbal state stem {\sls kpeg}  `hard' in
complex stem nouns}\\
 
 \ea\label{exːGRM-v-sta-p-hard-head}
{ɲúú{\T ꜜ}kpég} $<$ {\sc head-hard} {\rm `stubborness'} 
%pl. ɲuukpegse

 \ex\label{exːGRM-v-sta-p-hard-arm}
{nékpég} $<$ {\sc arm-hard} {\rm `stingy'} 

 \ex\label{exːGRM-v-sta-p-hard-tree}
{dààkpég} $<$ {\sc wood-hard} {\rm `strong wood'}
 
\z 
 \z
 

 Examples  are provided in
(\ref{exːGRM-v-sta-p-hard}) using  {\sls kpeg} again for
the sake of illustration.  Notice that only (\ref{exːGRM-v-sta-p-hard-tree}) has
a transparent and compositional meaning. Verbal state stems are 
usually found in complex stem nouns (Section \ref{sec:GRM-com-stem-noun}). 





%In a nominal context, the form in occurs in a  modifier position, i.e. ,  and
%the form in is found in compound formation, i.e. .



% Nevertheless there are  verbal state predicates which cannot be
% nominalized. For instance, the verbs {\sls dʊa} and  {\sls jaa} cannot
%list verb state predicate
%locative; dua
%identificational; ja


 


\subsubsection{Complex stem nouns}
\label{sec:GRM-com-stem-noun}


A  complex stem noun, as opposed to an atomic one,  is formed by the
combination of two stems (XY). Either X or Y in a  XY-complex stem noun may be 
atomic or complex.  Nominal stems ({\sc ns}), verbal state stems ({\sc ss}) and
verbal process stems ({\sc ps}), together with a single noun class  suffix 
(and/or other
types of nominalizer) are
the linguistic elements which take part in the
formation of complex stem nouns. 


\ea\label{exːGRM-cplx-stm}
 
 %\ex\label{exːGRM-cplx-stm-SS-AS} {\sls deŋlii}  `straight'  (deŋ|lii
%= `single'|`thither' 
 % $>$  {\sc ss} + {\sc ps} + {\sc cl.4}.sg)

  \ea\label{exːGRM-cplx-stm-NS-NS-1}%
{\sls nébíí} {\rm  `finger'}\\  %
ne-bi-i  $>$  {\sc arm-seed}\\%
 {\sc ns} + {\sc ns} + {\sc cl.3.sg}
 

  \ex\label{exːGRM-cplx-stm-NS-NS-2}
 {\sls pàtʃɪ́gɪ́búmmò} {\rm  `secretive'}\\ %
patʃɪgɪ-bummo  $>$  {\sc stomach-black}\\  %
 {\sc ns} + {\sc ns} (+ {\sc cl.1.sg})

 \ex\label{exːGRM-cplx-stm-NS-SS}
 {\sls ŋmɛ́ŋhʊ̀lɪ́ɪ̀} {\rm   `dried okro'}\\ %
ŋmɛŋ-hʊl-ɪ-ɪ $>$   {\sc okro-dry}\\  %
 {\sc ns} + {\sc ss} + {\sc nmlz}+ {\sc cl.4.sg}
 
 \ex\label{exːGRM-cplx-stm-PS-PS}
 {\sls jàwàdír̄} {\rm  `business person'}\\ %
jawa-di-r  $>$ {\sc buy-eat-agent}\\  %
 {\sc ps} + {\sc ps} + {\sc nmlz} (+ {\sc cl.3.sg})

 
\z 
 \z



In (\ref{exːGRM-cplx-stm-NS-NS-1}) and (\ref{exːGRM-cplx-stm-NS-NS-2}),  all 
stems are nominal. In (\ref{exːGRM-cplx-stm-NS-SS}),  the verbal state stem 
{\sls 
hʊl} `dry'  follows a nominal stem,  and  in  (\ref{exːGRM-cplx-stm-PS-PS}) 
both 
stems are of the type verbal process.  In these stem appositions, it is the 
noun 
class suffix of the rightmost stem which appears. Further, stems are lexemes, 
as 
opposed to nouns or verbs.  This is readily apparent in  
(\ref{exːGRM-cplx-stm-NS-NS-1}) and (\ref{exːGRM-cplx-stm-NS-NS-2}), in which 
the leftmost stems {\sls ne} and {\sls patʃɪgɪ} would appear as {\sls néŋ̀} 
and 
{\sls pàtʃɪ́gɪ́ɪ́} if they were full-fledged nouns. Thus, although complex 
stem 
nouns contain more than one stem, there is only  one noun class associated with 
the noun and it is always the noun class associated with the rightmost stem.  
This was mentioned in Section \ref{sec:GRM-sem-ass-crit} to support the claim  
that semantic criteria in noun class assignment may be  nonexistant. 


If  stems are treated as lexemes, there is still a problem in accounting for the
 `reduced' form of  some lexemes when they occur in stem appositions. That
is, the first stem  of a complex stem noun is often reduced to a single
syllable in the case of a polysyllabic lexeme, or a monosyllabic lexeme of the
type CVV is reduced to CV. For
example,  {\sls lúhò}  and  {\sls lúhòsó} are respectively the singular and
plural forms for 
`funeral' ({\sc cl.2}).  The expectation is that
when the lexeme takes part in position X of a XY complex stem noun,
it should exhibit its lexemic form, i.e.   {\sls lúhò}. Yet, the word for 
`last
funeral'  is {\sls lúsɪ́nnà}, {\it lit.} funeral-drink,  and not {\sls 
*luhosɪnna}.  Not all  lexemes get reduced in that particular 
environment, nevertheless, it is  more common (and visible)  for polysyllabic 
lexemes or
monosyllabic ones built on a heavy syllable. Moreover, some lexemes are more
frequent in that environment than others.


The relation between the stems in a complex stem noun is asymmetric.  The 
relation is defined in terms of what the referents of the stems and the complex 
noun as a whole have to do with each other.  As in a syntactic relation between 
a head and a modifier, one of the stems modifies while the other stem is 
modified. The semantic relations between the stems  are of two types: 
`completive' modification and  `qualitative' modification. These distinctions 
are discussed in  Sections \ref{sec:GRM-comp-completive} and 
\ref{sec:GRM-comp-quality} below.


\paragraph{Completive modification}
\label{sec:GRM-comp-completive}

A completive modification in a complex stem noun XY can translate as `Y of X' 
of 
which Y is the head. For instance {\sls sììpʊ́ŋ}   `eyelash', {\it lit.} 
eye-hair, is a kind of hair and not a kind of eye. And {\sls ʔɪ̀lnʊ̃̀ã̀} 
`nipple', {\it lit.} breast-mouth, is most likely seen as a kind of orifice 
than 
as  a kind of breast.  In both cases, the noun class is suffixed to the 
rightmost stem, incidentally to the head of the morphological construction, 
i.e. 
{\sls sììpʊ́ŋ}/{\sls sììpʊ́ná} {\sc (cl.3)} and {\sls ʔɪ̀lnʊ̃̀ã̀}/{\sls 
ʔɪ̀lnʊ̃̀ã̀sá} {\sc (cl.2)}. As mentioned earlier,  either X or Y  in a 
complex 
noun XY can be complex. The word {\sls nèpɪ́ɛ̀lpàtʃɪ́gɪ́ɪ́} `palm of the 
hand' 
is an example of two completive modifications. It consists of a complex stem 
{\sls nepɪɛl} `hand', which is composed of  {\sls ne} `arm' and {\sls pɪɛl} 
`flat', 
and the atomic stem {\sls patʃɪgɪ} `stomach', yielding in turn  `flat of hand' 
and then `inside of flat of hand'. 


\paragraph{Qualitative modification}
\label{sec:GRM-comp-quality}

A qualitative modification in a complex stem noun is the same as the  syntactic
modification  noun-modifier. The difference lies in the formal status of the
elements: when the
relation is held at a syntactic level the elements are words, whereas at the
morphological level they are lexemes. As mentioned earlier,  either X or Y  in a
complex noun XY can be complex. For instance, the word {\sls nebiwie} consists 
of
the combination of {\sls ne}  `arm' ({\sc cl.9}) and {\sls bi} `seed' ({\sc
cl.4}), then the combination of {\sls nebi} `finger' and {\sls wi} `small'. The 
noun class  of {\sls wi} `small'  is {\sc cl.1}, so the singular and plural
forms for the word `little finger' are {\sls nèbìwìé} and  {\sls 
nèbìwìsé}
respectively. The first relation involved is a completive modification, i.e.
`seed  of arm', while the second is a qualitative one, i.e. `small seed  of
arm' or `small finger'.  A qualitative modification in a complex noun XY can
translate as `X has
the property Y'  of which X is the head. Therefore, unlike many languages,  it
is not necessarily the head of the morphological construction which determines
the type of inflection.


\begin{table}[htb!]

\centering
\caption[Distinction between completive and qualitative
modification]{Distinction between completive and qualitative
modification using /daa/ `tree' or `wood'.  Abbreviations: {\sc h}= head,  {\sc
m}=
modifier, {\sc ns}= nominal stem, {\sc ss}= verbal state stems,  {\sc ps}=
verbal process stem, \label{tab:GRM-complet-and-qualit}}
\begin{Itabular}{llllll}
\lsptoprule
&  \multicolumn{3}{c}{Structure} & Stems &  Word\\  \cline{2-4}
 & Lex. type &  Function & Semantic& &\\[1ex]\midrule
%\multirow{4}{5mm}{\begin{sideways}\parbox{15mm}{Completive}\end{sideways}}
\multirow{4}{5mm}{\begin{sideways}\parbox{20mm}{Completive}\end{sideways}}

& {\sc ns-ns} & {\sc m-h}&{\sc whole-part}&/daa/-/luto/&   dààlútó\\
& &&&`tree'-`root'&`root of tree'\\[1ex]

&{\sc ns-ss}&{\sc m-h}&{\sc whole-part}&/daa/-/pɛtɪ/ &dààpɛ́tɪ́\\
&&&&`tree'-`end'& `bark'\\[1ex]

&{\sc ns-ns}&{\sc m-h}&{\sc whole-part}&/kpõŋkpõŋ/-/daa/&kpõ̀ŋkpõ̀ŋdāā\\
&&&&`cassava'-`wood' &`cassava plant'\\[1ex] \midrule


 \multirow{4}{5mm}{\begin{sideways}\parbox{20mm}{Qualitative}\end{sideways}}

&{\sc ns-ns}&{\sc h-m}&{\sc thing-charac}&/daa/-/sɔta/& dààsɔ̀tá\\
&&&&`tree'-`thorn'& `type of tree'\\[1ex]

&{\sc ns-ns}&{\sc h-m}&{\sc thing-charac}&/ɲin/-/daa/& ɲíndáá\\
&&&&`tooth'-`wood'&  `horn'\\[1ex]

&{\sc ps-ns}&{\sc h-m}&{\sc purpose-thing}&/tʃaasa/-/daa/&tʃáásádàà\\
&&&&`comb'-`wood'&`wooden comb'\\[1ex]

\lspbottomrule
\end{Itabular} 
\end{table} 


The examples in Table \ref{tab:GRM-complet-and-qualit} illustrate the
distinction between the completive and qualitative modification. The form {\sls 
daa}  conveys either the meaning `tree' or `wood'. Both meanings may function as
head or as modifier.  If the head stem follows its modifier, it is a completive
modification, and vice-versa for the qualitative modification. A semantic
relation between the stems may  be a whole-part relation, a characteristic added
to define an entity or a purpose  associated with an entity. 

So far,   XY-complex stem nouns were assumed to be  endocentric compounds whose
head is X in qualitative modification and the head is Y in completive
modification.  However, a word such as {\sls patʃɪgɪbummo} `secretive', {\it 
lit.}
stomach-black, suggests that some XY-complex stem nouns may  either lack a head
or have more than one head. These possibilities are not ignored, but in this
particular case the complex stem noun may be seen as involving  the abstract
senses of {\sls patʃɪgɪɪ} and {\sls bummo}, that is  `essence' and 
`subtle/restrained' respectively, making {\sls patʃɪgɪbummo} a qualitative
modification  which can be formulated literally as `subtle/restrained essence',
i.e.   a property applicable to humans. Thus, the stem {\sls patʃɪgɪɪ} is 
treated
as the head, and {\sls bummo} as the stem functioning as the qualitative 
modifier.
Another example is {\sls dààdùgó}. This word consists of the stems {\sls 
daa}
`tree' and  {\sls dugo} `infest'   and refers to a type   of insect. Unlike the
analyzed expressions displayed in  Table \ref{tab:GRM-complet-and-qualit}  none
of the stems can be treated as the head of the expression and the meaning of the
whole noun cannot be transparently  predicted from its constituent parts. This
leads me to provisionally consider the expression {\sls dààdùgó}  as an
exocentric compound, i.e. a complex stem noun without a head.


 
\paragraph{Compound or circumlocution}
\label{sec:GRM-comp-vs-circum}

For a few expressions,  it is hard to tell whether they are compounds, i.e. the
results of  morphological operations, or circumlocutions, i.e.  the results of
syntactic operations \citep[165]{Alla01}. Clear cases of circumlocution
nevertheless exist. For instance,  the word {\sls kpatakpalɪ} `type of hyena'  
is
treated by one language consultant as {\sls kpa ta kpa lɪɪ} {\it lit.}
 `take let.free take leave'.\footnote{Yet {\sls kpatakpari} is the word for
`hunting trap' in Gonja \citep{Rytz66}.}  Another example is {\sls 
sʊ́wàkándíkùró} `parasitic plant'. This expression refers to a type of
parasitic plant lacking a  root which grows upon and survives from the
nutrients provided by its  hosts. The word-level expression originates from the
sentence  {\sls sʊwa ka n̩ di kuoro}, {\it lit.}  die-and-I-eat-chief, `Die so
that I can become the chief'. It is common to find names of individuals being
constructed in this way: the oldest woman in Ducie is known as {\sls n̩wabɪpɛ}, 
{\it lit.}  {\sls n̩ wa bɪ pɛ}  `I-not-again-add'. Since two successive 
husbands 
died early,  she used to say that she will never marry again. For that reason
people call her {\sls ǹ̩wábɪ̀pɛ̄}.  



\subsubsection{Derivational morphology}
\label{sec:GRM-der-morph}

A derivational morpheme is an affix which combines with a stem to form a word.
The meaning it carries combines with the meaning of the stem.  By definition,
a derivational morpheme is a bound affix, and  thus 
cannot exist on its own as a word. This property keeps apart complex
stem nouns and derived nouns. Yet, the distinction between a bound affix and
a lexeme is not obvious, mainly because some bound affixes were probably lexemes
at a previous stage, or still are today. 
% \paragraph{}
% \label{sec:GRM-der-}


\paragraph{Maturity and sex of animate entities}
\label{sec:GRM-der-matur}


The specification of the maturity
and sex of an animate entity is accomplished in the following
way: male, female, young, and adult are organized in morphemes encoding
one or two distinctions. These morphemes are suffixed to the rightmost stem.
To distinguish between male and female, the morphemes ({\it sg.}/{\it pl.}) 
{\sls wal/wala} `male'  and {\sls nɪɪ/nɪɪta} `female'  are used as
(\ref{exːGRM-sex-ent}) illustrates.


\ea\label{exːGRM-sex-ent}
 
 \ea\label{exːGRM-sex-en} {\sls bɔ̀là-wál-\O} /  {\sls bɔ̀là-wál-á}\\
{\rm elephant-male-sg / elephant-male-pl   ({\sc cl.3})}
 \ex\label{exːGRM-sex-en}  {\sls bɔ̀là-nɪ́ɪ́-\O}  / {\sls bɔ̀là-nɪ̀ɪ̀-tá}\\
{\rm elephant-female-sg / elephant-female-pl ({\sc cl.7})}
 
 
\z 
 \z


The language employs two strategies to express the distinction between  the
adult animal and its young, which is  called here 
`maturity'.  The first
is to simply add the morpheme {\sls -bi} `child'  to the head,
e.g. {\sls bɔlabie/bɔlabise} `young elephant(s)'. In the second strategy 
both the sex and maturity distinctions are conveyed by the morpheme.  This is
shown in Table \ref{GRM-maturity-sex}. 

%componential-

\begin{table}[htb!]

\caption{Morphemes encoding maturity and sex of animate entities}
\centering
 \begin{Itabular}{lcc}
\lsptoprule
&\textsc{male}&\textsc{female}\\
\midrule
\textsc{young}& -w(a|e)lee & -lor\\
\textsc{adult} & -wal & -nɪɪ\\
\lspbottomrule

  
 \end{Itabular} 

\label{GRM-maturity-sex}
\end{table} 

Some examples are more opaque than others. For instance, the onset consonant of
the morpheme {\sls wal/wala} `male' may surface as a bilabial plosive,  e.g. 
{\sls bʊ̃́ʊ̃̀mbāl} `male goat'.  One can also observe a difference in form 
between
the word {\sls pìèsíí} `sheep',   {\sls pèmbál}  `male sheep' and  {\sls 
pènɪ̀ɪ́} `female sheep'. The words displayed in the first three rows of table
\ref{tab:GRM-matur-sex-ex} show the least transparent derivations.  The
annotation of tone is a first impression and the symbol - indicates that the
consultants could not associate a word for the intended meaning. 


%\ex\label{exːGRM-sex-en}  {\sls bɔlɛbie}  /  {\sls bɔlɛbise}
 %\ex\label{exːGRM-sex-en}  {\sls bɔlɛbiwal}  /  {\sls bɔlɛbiwala}
%. a case of qualitative
%modification


\begin{table}[htb!]

\caption{Maturity and sex/gender of animals}
\centering
 \begin{Itabular}{l>{\slshape}l>{\slshape}l>{\slshape}l>{\slshape}l>{\slshape}l}
\lsptoprule
Animal & {\rm Generic} & \multicolumn{2}{c}{{\rm Adult}} 
& \multicolumn{2}{c}{{\rm Young}}\\ \cline{3-4} \cline{5-6}
 && \multicolumn{1}{c}{{\rm Male}} &  \multicolumn{1}{c}{{\rm Female}} & 
\multicolumn{1}{c}{{\rm Male}} &   \multicolumn{1}{c}{{\rm Female}}\\
\midrule


fowl  &zál̀& zím{\T ꜜ}bál & zápúò&
zímbéléè  & zápúwìé\\

sheep   &píésíí &pèmbál & pènɪ̀ɪ́&
pémbéléè&pélòŕ\\

goat  &bʊ̃́ʊ̃̀ŋ & bʊ̃́ʊ̃́mbál   & bʊ̃̀nɪ̀ɪ́ &  bʊ̃̀mbéléè & 
bʊ̃̀ʊ̃̀lòŕ\\

pouched rat &sàpùhĩ́ẽ̀ & sàpúwál & sàpúnɪ́ɪ̀&
 sàpúwáléè& sàpúlòr\\

antilope   &ʔã́ã́ &ʔã̀ã̀wál  
&ʔã́ã́nɪ́ɪ́&ʔã̀ã̀wéléè&ʔã̀ã̀lòŕ\\

dog   &váà  &váwàl &vánɪ̀ɪ̀&
váwáléè&válòŕ\\

cat  &dìébìé &díèbəwál   
&díèbənɪ̀ɪ̀&díèbəwáléè&díèbəlòr\\

cow    &nã̀ɔ̃́ &nɔ̃̀wál  
&nɔ̃̀nɪ̀ɪ́&nɔ̃̀wáléè&nɔ̃̀lòŕ\\

elephant &bɔ̀là &bɔ̀lwàl & bɔ̀lənɪ̀ɪ́&  -  &bɔ̀llòŕ\\

guinea fowl &sũ̀ṹ&sũ̀wál &sũ̀nɪ̀ɪ́& - & -\\
bush mouse &ʔól&ʔólwál &  ʔólnɪ̀ɪ́ & - & -\\
%house mouse&dàgbòŋó&dagboŋowal  &dagboŋonɪɪ& - & -\\
lizard  &gèŕ &géwál &génɪ̀ɪ́&  -& -\\

\lspbottomrule
 \end{Itabular}
\label{tab:GRM-matur-sex-ex}
\end{table} 



\paragraph{Inhabitant of ...}
\label{sec:GRM-inhabitant-of}

To express `I am from X',  where `be from X' refers  to the place where
someone is born and/or the place where someone lives, the verb {\sls lɪ̀ɪ̀} is
used, e.g. {\sls sɔɣla n̩ lɪɪ} `I am from Sogola'.  

Expressions with the meaning `inhabitant of X'  can be  noun words referring to
this same idea, that is  `being from X'. The examples in table
\ref{tab:inhabitant-of} show that the meaning is captured in suffixes {\sls 
-(l)ɪɪ/(l)ɛɛ/la} which display vowel qualities in the singular and plural
similar to those found in noun class 4. 


\begin{table}[htb!]

\caption{Inhabitant of ... \label{tab:inhabitant-of}}
\centering
 \begin{Itabular}{l>{\slshape}l>{\slshape}ll>{\slshape}l>{\slshape}l}
\lsptoprule
Location &  {\rm sg.}  &  {\rm pl.}   & Location & {\rm sg.} &  {\rm pl.} \\[1ex] \midrule
Chakali   & tʃàkálɪ́ɪ́   & tʃàkálɛ́ɛ́ &
Katua  &kàtʊ́ɔ́lɪ́ɪ́   &kàtʊ́ɔ́lɛ́ɛ́\\
Motigu   &mòtígíí  &mòtígíé &%
Tiisa &  tíísàlí  &tíísàlá\\
Ducie   &dùsélíí  &dùséléé&%
Chasia  &tʃàsɪ́lɪ́ɪ́ & tʃàsɪ́lɛ́ɛ́\\
  Bulinga  & búléŋíí & búléŋéé&%
Wa &  wáálɪ́ɪ́  &wáálà\\
Gurumbele&grʊ̀mbɛ̀lɪ́lɪ́ɪ́& grʊ̀mbɛ̀lɪ́lɛ́ɛ́&%
Tuosa  &  tʊ̀ɔ̀sálɪ́ɪ́  & tʊ̀ɔ̀sálá\\
\lspbottomrule

 \end{Itabular}
\end{table} 
 

%({\sls mũ̀sóró} `type of food ingredient') , 
%({\sls dárà} `type of game')



\paragraph{Category switch}
\label{sec:GRM-der-cat-switch}

The
phenomenon called `category switch' refers to a derivational process whereby two
words with  related meanings and composed of the same segments change category
based entirely on their tonal melody. Examples are provided in 
(\ref{exːGRM-der-cat-switch}).

 \ea\label{exːGRM-der-cat-switch}
 
 \begin{tabular}{lllllll}

tʊ̀mà & ({\it v})   & {\rm `work'}  &  $\leftrightarrow$ &  tʊ́má  & ({\it n})  &  {\rm `work'}\\
gʊ̀à  & ({\it v})  &   {\rm `dance'}  & $\leftrightarrow$ &  gʊ̀á  & ({\it n})  &  {\rm `dance'}\\
jɔ̀wà &  ({\it v})  &{\rm `buy'}  & $\leftrightarrow$& jɔ̀wá&  ({\it n})   &{\rm `market'}\\
mʊ̀mà & ({\it v}) & {\rm `laugh'}&  $\leftrightarrow$ &mʊ̀má &({\it n}) &{\rm  `laughter'} \\
gòrò &({\it v}) & {\rm `circle'} & $\leftrightarrow$& góró &({\it n}) & {\rm `bent'}  
 \end{tabular}

\z


\paragraph{Agent- and event-denoting nominalizations}
\label{sec:GRM-der-agent}

Agent- and event-denoting nominalizations were discussed in
Section \ref{sec:GRM-verb-act-stem} in connection with the licensing of verbal
stems in atomic noun formation. Apart from their roles in complex stem nouns, it
was shown that both verbal state and verbal process stems undergo these two
nominalizations processes in order to function as atomic nouns.
The two
processes are summarized in (\ref{exːGRM-der-agent}) and 
(\ref{exːGRM-der-action}). 

\ea\label{exːGRM-der-agent}{\rm Agent nominalization}\\

\ea\label{exːGRM-der-agent-suffix}
{\rm A verb stem takes the suffix -[r]  to express agent-denoting
nominalization.}\\
{\sls  sʊ̃̀ã̀sʊ́ɔ́r} / {\sls sʊ̃̀ã̀sʊ́ɔ́rá} {\rm ({\sc cl.3}) `weaver(s)'}\\
  $\leftarrow$  {\sls sʊ̃̀ã̀} ({\it v}) {\rm  `weave'}\\
{\sls  lúlíbùmmùjár} / {\sls lúlíbùmmùjárá}  {\rm ({\sc cl.3})  `healer(s)' }
\\
 $\leftarrow$ {\sls lulibummo} ({\it n}) {\rm `medicine'} + {\sls ja} ({\it v})  {\rm `do'}


 \ex\label{exːGRM-der-agent-redup}
{\rm A verb stem gets partially reduplicated   to express
agent-denoting nominalization.}\\
{\sls  súsúmmá} / {\sls súsúmmə́sá} ({\sc cl.3})   {\rm`helper(s)'}\\
 $\leftarrow${\sls sùmmè} ({\it v})  {\rm `help'}\\
{\sls  sã́sáár} / {\sls sã́sáárá} ({\sc cl.3})  {\rm  `carver(s)'}\\
 $\leftarrow$ {\sls sã̀ã̀} ({\it v})  {\rm `carve' }

\z 
 \z


\begin{exe}
 \ex\label{exːGRM-der-action}{\rm Event nominalization}\\
{\rm A verb stem takes the suffix -/[{\sc +hi, -bk}]/   to express
event-denoting
nominalization.}\\
{\sls lʊ́lɪ́ɪ́ } / {\sls  lʊ́lɪ́ɛ́} ({\sc cl.4})  {\rm `giving birth'} \\
 $\leftarrow$ {\sls  lʊla} ({\it v})  {\rm `give birth' } \\
{\sls kpégíí} / {\sls   kpégíé}  ({\sc cl.4})  {\rm  `hard'  or
`strong'} \\
  $\leftarrow$ {\sls  kpeg} ({\it v})  {\rm `hard' or `strong'} \\

\z


Notice in (\ref{exːGRM-der-agent}) that the two
agent-denoting nominalizations can occur on a single stem. Also, the
noun class does not seem to be  determined by the suffix
-[r].  While one consultant prefers that all agent nouns end as ({\it
sg.}/{\it pl.}) {\sls -r/-rV}, another consultant varies between {\sls -r/-rV} 
({\sc cl.3}) and  {\sls -ra/-rəsV} ({\sc cl.1}).  In addition, there is another
agent-denoting word formation which simply adds the word {\sls kʊɔrɪ} `make' to
the noun denoting the product, e.g. {\sls nã̀ã̀tɔ̀ʊ̀kʊ́ɔ́rá} / {\sls 
nã̀ã̀tɔ̀ʊ̀kʊ́ɔ́rəsá} ({\sc cl.1}) `shoemaker(s)' $<$ {\sls nããtɔʊ} ({\it 
n})
`shoe' + {\sls kʊɔrɪ} ({\it v}) `make'.


\subsubsection{Proper nouns}
\label{sec:GRM-prop-noun}

% 
% Are these the names these people were given by their families? they look like
% nick-named if they are not actually titles.  The whole paragraph sounds
%strange. %  Either get more ethnographic information on naming or cut some of
%this out.

%Proper nouns  are  usually characterized by their  inability  to inflect for
%number and to co-occur with articles and modifiers. 

As a rule,   proper nouns
have  unique referents:  they  name people, places, spirits, and so on.  So in
the
area where Chakali is spoken, there is only one river named {\sls gòlógòló}, 
only
one hill named {\sls dɔ̀lbɪ́ɪ́}, one
village
named {\sls mòtìgú},  only one shrine named {\sls dàbàŋtʊ́lʊ́gʊ́}, etc.  
Nevertheless more than one person can have the same
name, and the same applies to a lesser extent to villages. For instance,
{\sls sɔ̀ɣlá} may refer to the Chakali village situated between Tuosa and 
Motigu, 
or to
a Vagla village situated at the junction of the Bole-Wa and Damongo-Wa road. To
identify the former, one must say {\sls tʃàkàlsɔ̀ɣlá}. 


A  Chakali person may bear two or three names: his/her father's name, the name
of his/her grandfather or great-grandfather, and his own (common) name. In the
case of the (great-) grandfather's name, it is a feature of the newborn or an
external sign which suggests the child's name.  The common name may be changed
in
the course of one 's life. Today, regardless of whether a  person is Muslim or
not,
common names are mainly from Arabic, Hausa, and Gonja origins, probably due to
the
Islamization of Chakali (see Section \ref{sec:SOC-why-cli-still}).


Common names among the  elders (over 50 years old) consist of the name of a
non-Chakali village,  together with {\sls nàà} `chief'. In Tuosa, Ducie, and
Gurumbele, one finds one or more Kpersi Naa, Mangwe Naa, Jayiri Naa, Wa Naa, 
Sing Naa,  Busa Naa, etc. The next generation (approximately below 50 YO) tend to
have either `Muslim' names or `English-title' names. Common Muslim names are
Idrissu, Fuseini, Mohamedu, Ahmed, Mohadini, etc.  Typical 
`English-title' names are {\sls Spɛ́ntà} `inspector',  {\sls Dɔ́ktà} 
`doctor', 
{\sls Títʃà} `teacher', etc. Apart from `teacher',  which can identify actual
teachers in communities in which schools are present, none of the individuals
are actual teachers, doctors or inspectors. The same can be said about the older
generation, none of them are/were chief of those places. The villages are not
Chakali villages and these individuals have no connections with the villages
used in their names. It seems that these common names are trendy  nicknames
peers  assign to each other. One consultant claims that the elders can be
ranked in terms of power and influence according to their nicknames. 

In Chakali society, one may have two additional names, a drumming name and a
Sigu name. A drumming name is used in drummed messages sent to other villages
about weddings or deaths,  while a {\sls  sígù} name is a name one receives
when initiated to the shrine {\sls  dààbàŋtólúgú} at {\sls dáámúŋ}. 

Because of their pragmatic function,  proper nouns  are rarely observed in a
plural form, but some contexts may allow this. In
(\ref{ex:GRM-propn-noun-plur}), the proper name {\sls Gbolo} takes the plural
marker {\sls -sV}.\footnote{The context of (\ref{ex:GRM-propn-noun-plur}) makes
sense when one understands that the name `Gbolo' has a particular meaning.  It
is understood that when a couple has a  fertility problem,  it is common to
travel to the community of Mankuma and to consult their shrine. If the woman
gets pregnant after the visit, they must return to Mankuma to appease the
shrine. Subsequently, the child must be named `Gbolo' and automatically acquires
 the Red Patas monkey as  totem.}


  \begin{exe}
   \ex\label{ex:GRM-propn-noun-plur}
\gll  gbòlò-só bá-ŋmɛ̀nàá ká dʊ̀à dùsèè ní.\\
gbolo.({\sc g.}b)-{\sc pl}  {\sc g.}b-{\sc q}.many {\sc  egr} exist Ducie {\sc
postp}\\
\glt  `How many Gbolos are there in Ducie?' 
 \z

Finally, circumlocution is a common process found in names of people and dogs 
 (e.g. the example of {\sls n̩wabɪpɛ}, 
{\it lit.}  {\sls n̩ wa bɪ pɛ}  `I-not-again-add', was given in Section
\ref{sec:GRM-comp-vs-circum}).   A few examples of dog names are given in
(\ref{ex:GRM-propn-dog-name}).


  \ea\label{ex:GRM-propn-dog-name}{\rm Examples of dog names}\\
 
% \ex\label{ex:GRM-propn-dog-name-} 

 \ea\label{ex:GRM-propn-dog-name-1} 
{\sls jàsáŋábʊ́ɛ̃̀ɪ̀} {\rm  `Let's keep peace'}\\ 
ja-saŋa-bʊɛ̃ɪ   $>$  {\rm we-sit-slowly } 
 \ex\label{ex:GRM-propn-dog-name-2} 
{\sls ǹ̩nʊ̀ã́wàjàhóò}  {\rm  `I will not open my mouth again'}\\ 
 n̩-nʊã-wa-ja-hoo  $>$ {\rm my-mouth-not-do-hoo}
 \ex\label{ex:GRM-propn-dog-name-3} 
{\sls kùósòzɪ́má}   {\rm `God knows'}\\  
kuoso-zɪma $>$ {\rm  god-know}

\z 
 \z



% In many folktales, animals are the characters and their acts evoke
% human beings' behaviors. There is a tendency in folktales to switch between
% using the animal name as common noun and using it as proper noun. (example)


%1. born in village but moved from there
%2. not necesarily chief but from royal family
%3. gan naa= `more than a chief' grand father of John
%4. pure nick name

\subsubsection{Loan nouns}
\label{sec:GRM-borr-noun}


 A loan noun,  or more generally a loanword, can be defined as  ``a word that at
some point came into a language by transfer from another language''
\citep[58]{Hasp08}.  When a word is found in both Chakali and in another
language, many loan scenarios are conceivable. However,  for some semantic
domains such as   bicycle or car parts, school material and so on, the past and
present sociolinguistic situations  suggest that Chakali is
the recipient language and Waali, English, Hausa, Akan, and Dagaare are the
donor
languages.  Loan scenarios differ and are harder to establish when other SWG
languages are involved. It is often unfeasible to demonstrate whether the same
form/meaning in two languages was inherited from a common ancestor, or  
borrowed by one and subsequently passed on to other SGW languages. Moreover, it
may be unwise to assume that in all cases Chakali is  the recipient language,
especially for loan words in domains which  were in the past fundamental in
Chakali lifestyle,  but to a lesser degree for neighboring ethnic groups. 
Thus, Chakali as a donor language can be evaluated in a wider Grusi-Oti Volta
genesis, or  at a micro-level where the influence of Chakali on Bulengi is
established.

It is unlikely that Chakali borrowed from English through contact. And Ghanaian
English, in Wa town and Chakali communities,  is not an effective mode of
communication, at least in social spheres where Chakali men and women 
interact (see discussion in Section \ref{sec:vital}).  Nonetheless, the
situation is different for school children  who are
exposed to Ghanaian English on a regular basis. I believe that Ghanaian English
spoken by native speakers of Waali, Dagaare, or Chakali  is the only potential
variety of English which can function as a donor language. Examples of words 
ultimately  from English origin are: {\sls ʔàbəlù} `blue', {\sls 
ʔásɪ̀bɪ́tɪ̀} 
`hospital', {\sls dɔ́ktà} `doctor', {\sls bàlúù} `balloon', {\sls 
bátərbɪ́ɪ́} 
`battery (-stone)', {\sls bɛ́lɛ́ntɪ̀} `belt',  {\sls təráádʒà} `trouser',  
{\sls 
détì} `date', {\sls mɪ́ntɪ̀} `minute',   {\sls dʒánsè} `type of dance',  
{\sls 
kàpɛ́ntà} `carpenter', {\sls kɔ́lpɔ̀tɛ̀} `coal pot', {\sls kɔ́tà} `quarter', 
 
{\sls 
lɔ́ɔ́lɪ̀} `lorry (any four-wheel vehicle)',   {\sls sákə̀r} `bicycle',  {\sls 
pɛ̀n} 
`pen', {\sls sùkúù} `school',   {\sls tʃítʃà/títʃà} `teacher' and many 
more.  
There 
is a recurrent falling tonal melody (i.e. HL) among the loan nouns of  
ultimately English origins. Many of them,  if not all, can be found in other 
languages of the area \citep{sisa75, daku07}. 

People are aware of the linguistic fragility of Chakali; some of the
language
consultants confirm that many people do not bother making  the effort to use
Chakali and that many
prefer to use Waali expressions.  The knowledge and interest of our
language consultants in their language made  it possible  to reduce the number
of
Waali
expressions in the dictionary in Part \ref{sec:cli-eng-dic}.  Despite 
that,
when a word is found both in
Waali and Chakali, it is not automatically classified as borrowed from Waali,
yet it is only suspected to be non-Chakali.  Examples such
{\sls dʒɪ̀ɛ̀rá} `sieve', {\sls dʒùmbùrò} `type of medicine', {\sls gbàgbá}
`duck', {\sls kákádùrò} (Akan)  `ginger', {\sls kàpálà} `fufu', {\sls 
káṹ}  `mixture
of sodium carbonate' , {\sls nāsárá}  (Hausa) `white man', and  {\sls 
sɛ̀nsɛ́nná}
`prostitute' are some of the Waali/Chakali nouns  found in transcribed texts, or
by chance. 

The weekdays are from Arabic (probably via Hausa or Oti-Volta languages).
Vagla and
Tumulung Sisaala,  but  not Dɛg, use similar expressions \citep[60]{Nade96}: 
{\sls atanɛa} `Monday', {\sls atalata} `Tuesday', {\sls alarba} `Wednesday',   
{\sls alamussa} `Thursday',  {\sls ardʒɛmaa} `Friday' , {\sls asɛbɛtɛ} 
`Saturday'   and
{\sls allahadi} `Sunday'.  The expressions for the lunar months seem to be
borrowed from Waali, but Dagbani and Mamprusi have similar expressions. In
these
Oti-Volta languages, some of the names  correspond to important festivals, i.e.
1, 3, 7, 9, 10, and 12 below. In Chakali, only {\sls dʒɪ́mbɛ̄ntʊ̀} is celebrated
and  is considered the first month.\footnote{Dagbani {\sls buɣum} and Waali 
{\sls  dʒɪmbɛntɪ} are both treated as first month by the speakers of these
languages.} The lunar months are: {\sls dʒɪ́mbɛ̄ntʊ̀} `first month
(1)', {\it sífə̀rà}  `second month (2)', {\it dùmbá} `third month (3)', 
{\it dùmbáfúlánààn} `fourth month (4)',  {\it dùmbákókórìkó} `fifth 
month
(5)', {\it kpínítʃùmààŋkùná} `sixth month (6)', {\it kpínítʃù} 
`seventh
month (7)', {\it ʔàndʒèlìndʒé} `eighth month (8)', {\it sʊ́ŋkàrɛ̀} `ninth
month (9)', {\it tʃíŋsùŋù} `tenth month (10)', {\it dùŋúmààŋkùnà}
`eleventh month (11)' and {\it dùŋú} `twelth month (12)'.  It was understood
that
these terms and concepts are not known by the majority.\footnote{Thanks to Tony
Naden, who helped me clarify some issues relating to this topic.} 




\subsubsection{Relational Nouns}
\label{sec:SPA-relnoun}

%              Some relational nouns are de-
% rived from body parts, e.g., ..., but not all, e.g.,
% -baamban??

Many  languages present formal identity between body parts terms and expressions
used to designate elements of space. The widely accepted view is that
diachronically  spatial relational nouns  -- sometimes called spatial nominals, 
\citep[895]{Hell07},  or adpositions,  \citep[137]{Hein97} --  are ``the result of
functional split'' and that ``they are derived from nouns denoting body parts or
locative concepts through syntactic reanalysis'' \citep[256]{Hein84}. In Ewe
(Kwa) for instance, it is claimed that  the prepositions have evolved from verbs
and postpositions from nouns \citep[367--369]{Amek06}. 

% 
% In the theory of grammaticalization's terms
% \citep[137]{Hein97} 
% In  \citet{Amek06}
% it is claimed that i
% 
% What we call a relational noun has received several other labels in the
% literature; spatial nominal \citep[895]{Hell07}, postposition
% \citep[256]{Hein84}. 


Chakali relational nouns are formally identical to body part nouns although 
not all body part nouns have a relational noun counterpart. For instance,
whereas {\it 
ɲuu} can have both  a spatial meaning, i.e. `on top of X', and  a body part one,
i.e. `head',  the body part terms {\it bembii} `heart', {\it hog} `bone'  or 
{\it 
fʊ̃ʊ̃} `lower back', among others,  cannot convey  spatial meanings. Table
\ref{tab:relvsbody} displays the body parts found in the data which
convey spatial meaning.\footnote{The body part term {\it gàntàĺ}
`back' is from
the Ducie lect and corresponds to {\it hàbʊ̀á} in the Motigu, Gurumbele,
Katua, Tiisa, and Tuosa lects.}



\begin{table}[h!]
\caption[Spatial nominal relations and body part nouns]{Spatial nominal
relations and body part nouns: similar forms and different, but related,
meanings\label{tab:relvsbody}}
\centering
\begin{small}
 \begin{Qtabular}{lllll}
\lsptoprule
Projection& Spatial relation & PoS: {\it reln}  & Body parts &
 PoS: {\it n}\\   \midrule
Intrinsic & &&&\\

& \textsc{top}  & {  ɲuu(x,y)} & head & {  ɲuu(x)}\\
&\textsc{containment} &  {patʃɪgɪɪ(x,y)}  & stomach & {
patʃɪgɪɪ(x)}\\

& \textsc{side} &  { logun(x,y)}& rib & { logun(x)}\\
& \textsc{mouth} &  { nʊã(x,y)}& mouth & { nʊã(x)}\\
& \textsc{base/under} &  { muŋ(x,y)}& arse & { muŋ(x)}\\
& \textsc{middle} &  { bambaaŋ(x,y)}& chest box & { bambaaŋ(x)}\\
\cline{1-3}

Relative  & &&&\\
& \textsc{left} &  {neŋgal(x,y)}& left hand &  {neŋgal(x)}\\
& \textsc{right}  & {nendul(x,y)}  & right hand &{nendul(x)}\\
& \textsc{back} &  {gantal(x,y)}  & dorsum & {gantal(x)}\\
& \textsc{front} & {sʊʊ(x,y)}  & front  & {sʊʊ(x)}\\

 \lspbottomrule
 \end{Qtabular}
\end{small}

\end{table} 







The contrast between an intrinsic and a relative frame of reference was brought
up in Section \ref{sec:SPA-exper1-sum} where I confirmed that both  an 
intrinsic and a relative exocentric and egocentric  frames of reference were
being used. An intrinsic frame of reference is  invariant no matter how the
spatial relation is viewed by the speaker or the addressee, whereas a
relative
frame of reference depends on
how a spatial relation is  viewed. 


% Sure that is right. Also suppose that you put the cutlass you used to cut the
% head on the head. Now you can say, 'a karintie saga a nyuu nyu ni'. In your
% given example, the kpulikpuli is on the head. Let's change it and say the
% kpulikpuli is under the head. Then we would have, 'a kpulikpuli dua a nyu mun
% ni'. I'm not sure how these will allow, could you give examples? gantal gantal
% 'behind the back'
% nua nua  'at the entrance of the
% mouth' pachigi pachigi 'inside the stomach' (Kasim)

How can we distinguish a relational
noun from a noun?  Above all,  the differentiation between relational
nouns and body part nouns cannot rely solely on surface syntax criteria,
precisely
because the configuration of a possessive noun phrase and a
relational noun phrase are identical. This is shown in
 (\ref{ex:SPA-reln-vs-bpn}). 

 
\ea\label{ex:SPA-reln-vs-bpn}
\ea\label{ex:SPA-bpn}{\rm Possessive attributive phrase}\\
 {[\textsc{n}$_{1}$-\textsc{n}$_{2}$]}$_{NP}$ where \textsc{n}$_{2}$=body 
part,   e.g. {\sls báál  ɲúù} {\rm `a man's head'}

\ex\label{ex:SPA-reln}{\rm Spatial nominal  phrase}\\
 {[{\sc n}$_{1}$-{\sc n}$_{2}$]}$_{NP}$ where {\sc n}$_{2}$=spatial relation,   
e.g. {\sls téébùl  ɲúù}  {\rm  `top of the table'}
\z
\z

Even though the two corresponding nominal structures may cause ambiguities,
the
interpretation is generally disclosed by the meaning of the nominal preceding
the \textsc{n}$_{2}$ in  (\ref{ex:SPA-reln-vs-bpn}). The term  {\sls  ɲuu}, 
for
instance, can only mean `top of' in a 
phrase in which it follows another nominal and refers to a projected
location of \textsc{n}$_{1}$'s referent. In (\ref{ex:SPA-bpn}), even
though {\sls ɲuu}
 immediately follows a nominal,  it would not normally refer to the projected
location `on the top' but only to the man's head. Nevertheless, despite any
attempts to identify  structural characteristics which may contribute to the
disambiguation of a phrase involving a body part term,  
ambiguities may still arise.


%a kpulikpuli dua a statue nyuu nyuu ni






Another aspect of body part terms is their different function in  morphological
and syntactic structure. While a relational noun is a syntactic word,  body
part terms may also function as morphemes in compound nouns to express a
specific
part-whole relationship or a conventionalized metaphor \citep[141]{Hein97}. 
Whereas the distinction may be formally and semantically hard to distinguish,
the number of body
part terms which can be the stem in a compound noun is larger than those
functioning as relational nouns. Some examples are shown in Table
\ref{tab:SPA-bpt-compound}.

%fix the etymology

\begin{table}[h]
\caption{Body part terms in compound nouns\label{tab:SPA-bpt-compound}}
\centering
%\begin{small}
 \begin{Qtabular}{l>{\slshape}llp{1.6in}}
\lsptoprule
Body part term & {\rm Compound noun}  & Morph. gloss & Gloss\\ \midrule


eye & tɔ́ʊ́-{\T ꜜ}síí  & village-eye & village's center\\
 & kpã̀ã̀n-síí & yam-eye & yam stem\\
& nɪ̀ɪ̀-síí & water-eye & deepest area of a  river\\
 &  nã̀ã̀-síí & leg-eye & ankle bump\\

 mouth &   gɔ̀ŋ-nʊ̃̀ã́ & river-mouth & river bank\\
		&   ʔɪ̀l-nʊ̃̀ã́   & breast-mouth& nipple\\
    &   dɪ́à-nʊ̃́ã̀  & house-mouth& door\\

leg &   gɔ́n-{\T ꜜ}nã́ã́   & river-leg & split of a river\\ 

&   dáá-{\T ꜜ}nã́ã́  & tree-leg & branch\\




 head &   kùósò-ɲúù  & god-head & sky\\
    &   tìì-ɲúù  &land-head  ({\it etym})& west\\
%stomach&   \sls tɔ̀ʊ́pàtʃɪ́gɪ́ɪ́  & village-stomach & inside the village\\

 arse &  tìì-múŋ &land-arse  ({\it etym}) & east\\

neck &  vìì-báɣə̆́ná & pot-neck & neck of a container\\

testicle &   mááfà-lúró  &  gun-testicle &  gun powder container\\

penis &  mááfà-péŋ  & gun-penis &gun trigger\\

ear &   mááfà-dɪ́gɪ́ná   & gun-ear & flintlock frizzen\\

arm &  fàlá-néŋ̀ & calabash-arm & calabash stem\\
 
navel & fà-ʔúl & calabash-navel & calabash node\\

nose & píí-mɪ́ɪ́sà & yam mound-nose &  part of a yam mound\\
liver & tɔ́ʊ́-pʊ́ɔ̀l & village-liver &  important community member\\
\lspbottomrule
 \end{Qtabular}
%\end{small}

\end{table} 

Ignoring for the moment the structure in which they are involved, there seem to
be two types
of spatial interpretation accessible with body part terms. And there also seems
to be a gray zone between the two.\footnote{This gray zone may receive a
diachronic interpretation.  In
\citet[1072]{Amek07c},  the postpositions in Sɛkpɛlé are seen as evolving
``from body part and environment terms''  and have a similar, but not
identical, 
function as those of Chakali relational nouns. For instance, Sɛkpɛlé's  
postpositions ``cannot be modified'' nor can they vary ``with respect to number
 marking''.} The first interpretation is the literal
attribution of human
characteristics (i.e. anthropomorphic) in  reference to parts of object. In a
such a
case, a body part term refers to a part of an object in analogy to an animate
entity. For instance, a trigger of a gun (i.e. the lever that activates the
firing
mechanism) is  attributed to the penis to characterize its physical appearance.
The
second interpretation does not designate a fixed part of an object
but a location projected from a part of an object.  In such a case it designates
a spatial environment in contact with or detached from an object
\citep[44]{Hein97}. To make the
distinction clear,  in the sentence `a label is glued on the neck of the bottle'
the body part term `neck' designates a breakable part of the bottle, 
whereas
in the sentence `John is standing at the back of the car' the body part term
`back' does not designate any part of the car but a relative spatial
location, the area behind the car. 


 In \citet[44]{Hein97}, the variety of denotations of body part terms is
accounted for in a diachronic perspective. The  claim is sketched in
(\ref{ex:Heine-stage}).


\ea\label{ex:Heine-stage}{\rm From body part to spatial concept: A four-stage
scenario \citep[44]{Hein97}}\\
Stage 1: a region of the human body\\
Stage 2: a region of an (inanimate) object\\
Stage 3: a region in contact with the object\\ 
Stage 4:  a region detached from the object\\
\z


However, synchronically each of these stages is observable: A  Chakali 
relational noun is more easily interpreted as  a region in contact with or
detached from an object, a body part term in a compound
noun designates
a part of an object, and a body part term used as
a full-fledged noun is associated with a part of the human body. 
Nevertheless, the examples provided in Table \ref{tab:SPA-bpt-compound} show
that the distinction is not a clear-cut one: Does the expression {\sls teebul 
ɲuu}
designate a spatial environment  in contact with or detached from a table or  a
part of table? Both interpretations seem acceptable.

% % For instance, {\sls dáámúŋ}, literally `tree under', can  mean either a
% % resting area or a location for the initiation rite (neither being 
%%obligatorily
% % under a tree), and the sentence {\sls ʊ dʊa daa muŋ nɪ} can only be 
% interpreted
% %as
% % `it is under the tree', but never as `it is at the resting area' nor as  `it
% %is
% % at the location for the initiation rite'.


Relational nouns
are rarely found in their plural form: but on grammatical grounds, nothing
prevents them from being expressed in the plural. 
To describe a situation where  for every bench there is a calabash sitting on
it, 
the sentence in (\ref{ex:reln-plu-head}) is appropirate.


\ea\label{ex:reln-plu-head}
\gll à fàlàsá ságá à kóró ɲúúnó nɪ̄.\\
{\art} calabash.{\pl}  sit {\art} bench.{\pl}  {\reln .\pl} {\postp}\\
\glt `The calabashes sit on top of the benches.' 
\z

One may argue that the `top of a bench' is a spatial environment  in contact
with the bench, even a physical part of the bench, so pluralization may simply
suggest that the   `top of a bench' is a word refering to an entity,  and not a
locative phrase. Two pieces of evidence go against this view: first,  notice
that {\sls koro} `bench'  in {\sls koro ɲuuno} is plural.  Recall Section
\ref{sec:GRM-com-stem-noun} in which a noun class ({\it sg./pl.} marking) was
argued to  appear only at the end of a word. If  `top of a bench' was a word and
not a phrase, we would expect its plural form to be  *{\sls korɲuuno}. 
Secondly, 
deciding whether or not the `top of' is indeed in contact with or detached from
the bench is not conclusive. To describe a situation where several balls are
under several tables, one may use the sentence in (\ref{ex:reln-plu-stomach}), 
in which case it cannot be argued that  under of the table is a physical part of
the table.\footnote{One may argue that it is indeed a part of the table,
identical to the interior space of a container.}

\ea\label{ex:reln-plu-stomach}
\gll à bɔ́lsā dʊ́á à téébùlsō pátʃɪ̀gɪ̄ɛ̄ nɪ̄.\\
{\art} ball.{\pl}  be.at {\art}   table.{\pl}  {\reln .\pl} {\postp}\\
\glt `The balls are under the tables.'
\z



% kopu suguli a teebul nyu- logun ni
% a haan dua u gantal ni 'the woman is at his back'
% The above sentences are right. But in the case of "a haana dua ba gantala ni"
%we  only pularise the subject (haana) but keep the location (gantal) singular
% because 'ba' has already indicated that they are many. So we have 'a haana dua
% ba gantal ni' NOT 'a haana dua ba gantala ni'. However, a balsa dua teebulSO
% pachigiE ni is right. Why because, we are referring to objects and not humans
%in  which case we cannot use 'ba' indicate it's a single object or many. For
% instance, we can have " a fota saga dasa nyuuni ni" (the baboons are on the
% trees) 'a zinni laga da nyu ni' (the bats are hanging on a tree).

Another aspect of relational nouns and oblique phrases in general is that they are structurally 
very rigid, that is, they are not easily extracted or preposed. The sentences in  
(\ref{ex:reln-extract-in-1}) and  (\ref{ex:reln-extract-in-2})  are nevertheless acceptable. 


\ea\label{ex:reln-extract}

\ea\label{ex:reln-extract-in-1}
\gll  à téébùl  ɲúú nī, à fàlá sàgà.\\
 {\art } table {\reln} {\postp}  {\art } calabash sit\\
\glt `On top of the table, the calabash sits.'

 \ex\label{ex:reln-extract-in-2}
\gll  téébùl lō, à fàlá ságá ʊ̀   ɲúú nī.\\
 table {\foc} {\art} calabash sit {3.\sg.\poss} {\reln. top} {\postp}\\
\glt `Table, the calabash sits on top of it.' ({\it lit.} `sits on its head')

 \ex\label{ex:reln-extract-out-3}
* teebul lo, a fala saga  ɲuu nɪ

 \ex\label{ex:reln-extract-out-1}
 * ʊ   ɲuu nɪ, a fala saga teebul
 
 \ex\label{ex:reln-extract-out-2}
 *ɲuu nɪ, a fala saga teebul

\z
\z

The sentence in  (\ref{ex:reln-extract-in-2}) is acceptable but odd. It shows
that the nominal complement of the relational noun
{\sls ɲuu} can be uttered at the beginning of the sentence while the possessive
pronoun {\sls ʊ}  is located  in the complement slot of   the relational noun, 
functioning as anaphora. The sentence is ungrammatical if the
pronoun is absent {\it in situ} (\ref{ex:reln-extract-out-3}),  or if the
oblique object phrase is preposed but the nominal {\sls teebul} stranded, 
whether
an anaphora referring to {\sls teebul} is present  
(\ref{ex:reln-extract-out-1}) 
or absent
(\ref{ex:reln-extract-out-2}). 



We now have  evidence for treating the relational nouns as
members of a closed class of lexical items whose function is to localize the
figure to a
search domain.  It is not only that body part terms acquire spatial meaning
following a noun referring to inanimate entities, but that, in diachrony, 
only a limited set of body part
terms has acquired that spatial meaning,  and, in synchrony, 
they form a subtype of nominal identified as relational noun. They are 
nouns since they can pluralize, but they acquire the status of functional words
since they constitute a formal class with limited membership where each of
the members expresses spatial meaning and requires a nominal complement.

\ea\label{ex:postp-struct}
 [[[a dɪa]$_{NP}$ ɲuu]$_{RelP}$ nɪ]$_{PP}$   `on the roof of the house' 
\z

In the structure  of the BLC oblique  phrase  in (\ref{ex:postp-struct}), the 
relational noun {\sls ɲuu} is within the complement phrase of the postposition 
{\sls 
nɪ}.  A relational noun phrase (RelP) consists of a head  and noun phrase 
complement.  We are now in a  better position to state that the complement 
phrase of the postposition is a (nominal) phrase which corresponds to the 
conceptual ground. 


To summarize, on a diachronical basis, it is believed that the function of 
relational nouns  as locative adpositions originates from their purely `entity'
meaning
through grammaticalization \citep[44, 83]{Hein84}. The form of Chakali body part
terms supports the claim.   On a synchronical basis, only  {\sls patʃɪgɪɪ}
`stomach',  {\sls logun } `rib',  {\sls gantal} `dorso', {\sls muŋ}   `arse', 
{\sls 
nʊã} `mouth',  {\sls sʊʊ} `front', {\sls bambaaŋ} `chest box'  and {\sls ɲuu} 
`head'
are relational nouns. Relational nouns are  nouns which lack the 
referential power of the default interpretation of  body part term  (i.e.
interpreted in  isolation), and which take a
complement which must obligatorily be filled by an entity capable of projecting
a spatial environment.

%\citep[137]{Hein97}
%inherent vs. relative spatial terms
%check with Linguistic Semantic book, examples...


\subsection{Pronouns and pro-forms}
\label{sec:GRM-pronouns}

 A pronoun is a type of pro-form.  The difference between pronouns and 
pro-forms depends on whether they can be 
anaphors of nominal arguments. In 
this section, the personal, impersonal,
demonstrative, and possessive pronouns are introduced, followed by the
expressions used to convey reciprocity and reflexivity.   In
Section \ref{sec:GRM-adv-pro},  the adverbial pro-forms are  introduced.


\subsubsection{Personal pronouns}
\label{sec:GRM-personal-pronouns}
Table \ref{tab:GRM-pers-pro} gives an overview of the personal pronoun forms.

%check tones all
\begin{table}[h]
 \caption{Weak, Strong, and Emphatic forms of personal 
pronouns\label{tab:GRM-pers-pro}}
  \centering
  \begin{Itabular}{l>{\slshape}l>{\slshape}l>{\slshape}l}
\lsptoprule 
Pronoun & {\rm Weak ({\sc wk})}   &  {\rm Strong ({\sc st})}  &  {\rm  Emphatic 
({\sc emph})} \\
Gram. function  & {\sc s|a} {\rm and} {\sc o}  &  {\sc s|a} & {\sc s|a}\\[1ex]
\midrule
{\sc 1.sg.} &  n̩ &   mɪ́ŋ & ń̩wà\\
{\sc 2.sg.}  &   ɪ& hɪ́ŋ & ɪ́ɪ́wà\\
{\sc  3.sg.}  &  ʊ&  wáá & ʊ́ʊ́wà\\
{\sc 1.pl.}  &   ja&  jáwáá & jáwà\\
{\sc 2.pl.} & ma &   máwáá & máwà\\
{\sc  3.pl.g}a &  a  &   áwáá & áwà\\
{\sc 3.pl.g}b  &   ba&   báwáá & báwà\\
 
\lspbottomrule
  \end{Itabular}
\end{table}


  
  The weak form first person singular pronoun is a syllabic nasal which assimilates its 
place feature from the following phonological material (see Section 
\ref{sec:ext-nasal-place}).  All weak forms may be lengthened in the 
imperfective (\ref{sec:GRM-trans-intran}). The  personal pronouns  do not encode a gender 
distinction in the singular but  an animacy  distinction is made between non-human and  human in 
the 
plural. They are glossed {\sc 3.pl.g}a and {\sc  3.pl.g}b  respectively (see Section 
\ref{sec:GRM-gender} for the role of gender in agreement).  The weak forms can surface either with 
a 
low or high tone; when an action  has not yet occurred or a wish is expressed the pronoun is 
perceived with a high tone (see Section \ref{sec:GRM-subjunctive}). Otherwise the weak forms 
normally have low tones.  The strong and emphatic forms are attested with the melodies with which 
they are associated in Table \ref{tab:GRM-pers-pro}. 



  \ea\label{ex:GRM-pro-WSE}  \begin{multicols}{2}
   \ea\label{ex:GRM-pro-W}
\gll ʊ̀ dí kʊ̄ʊ̄ rā.\\
  {\sc 3.sg} eat t.z. {\sc foc}\\
\glt  `She ate t.z.' 
   
   \ex\label{ex:GRM-pro-S}
   \gll wáá dí kʊ̄ʊ̄ (*ra).\\
  {\sc 3.sg.st} eat t.z.  {\sc foc}\\
\glt  `SHE ate t.z.' 
   
 \ex\label{ex:GRM-pro-E}  
   \gll ʊ́ʊ́wà dí kʊ̄ʊ̄ rā.\\
  {\sc 3.sg.emph} eat t.z.  {\sc foc}\\
\glt  `It is her who ate t.z.' 

 \ex\label{ex:GRM-pro-W-cleft}  
 * ʊ m̩ maŋʊʊ ra
  
  \ex\label{ex:GRM-pro-S-cleft}  
   \gll  wáá  m̩̀ màŋà (*ra).\\
  {\sc 3.sg.st} {\sc 1.sg}  beat  {\sc foc}\\
 \glt   `Him, I beat.'
  \ex\label{ex:GRM-pro-E-cleft}  
   \gll  ʊ́ʊ́wà  m̩̀ máŋʊ́ʊ́ rā.\\
  {\sc 3.sg.emph}  {\sc 1.sg}  beat.{\sc 3.sg}   {\sc foc}\\
 \glt   `It is him who I beat.'

  \z  
   \end{multicols}
  \z
  
The sentences in (\ref{ex:GRM-pro-W}-\ref{ex:GRM-pro-E}) show that while a weak or 
an emphatic pronoun can co-occur with a  \is{focus particle}focus particle,  a strong pronoun 
cannot. In addition, (\ref{ex:GRM-pro-W-cleft}-\ref{ex:GRM-pro-E-cleft}) confirm that both 
emphatic and strong pronouns may be fronted but weak pronouns cannot.  While both emphatic and 
strong pronouns typically appear  at the beginning of a sentence, their difference is that  the   
\is{emphatic pronoun}emphatic pronoun is always coreferential with a weak pronoun in the clause,  
as 
the pair (\ref{ex:GRM-pro-S-cleft}-\ref{ex:GRM-pro-E-cleft}) demonstrates.   The  distinction 
between weak 
({\sc wk}) and strong 
({\sc st})  is relevant when pronouns function as subject only.  Their proper use is 
conditioned by the emphasis  placed on the participant(s) of the event or the 
event itself, and by the polarity of the clause in which they appear.\footnote{The purpose of such 
distinctions derives mainly from the articulation of information. \citet{Purv07} offers an analysis 
for Dagbani whereby personal pronoun forms vary depending on their position in relation to their 
lexical host.}  In this way, strong pronouns cannot co-occur in a sentence in which another
constituent is in focus, that is a nominal phrase  flanked by the focus
marker  or   a
verb ending with the assertive suffix vowel   {\sc -[+ro,  +hi]} (compare 
examples 
(\ref{ex:GRM-out-STR-FOC-buy}) and (\ref{ex:GRM-out-STR-FOC-finish}) with 
(\ref{ex:GRM-out-STR-FOC-buy-g}) and  (\ref{ex:GRM-out-STR-FOC-finish-g})). 
In
addition,  in
sentences 
where a negative operator occurs, strong pronouns are disallowed, as  
(\ref{ex:GRM-out-STR-NEG-buy}) and   (\ref{ex:GRM-out-STR-NEG-finish}) show.

\ea\label{ex:GRM-weak-strong-arg}\begin{multicols}{2}
   \ea\label{ex:GRM-out-STR-FOC-buy-g}
\gll   mɪ́ŋ   jáwàà   kɪ̀nzɪ́nɪ́ɪ̀.\\
  {\sc 1.sg.st}  buy  horse\\
\glt  `I bought a horse.' 

   \ex\label{ex:GRM-out-STR-FOC-buy}
 *mɪŋ   jawa   kɪnzɪnɪɪ ra
   
   
%    \ex\label{ex:}
% \gll ǹ jáwá kɪ̀nzɪ́nɪ́ɪ́ rà.\\
%   {\sc 1.sg.wk}  buy  horse  {\sc foc}\\
% \glt  `I bought a HORSE.'
%    
   \ex\label{ex:}
\gll ǹ̩ wà jáwá kɪ̀nzɪ́nɪ́ɪ̀.\\
 {\sc 1.sg.wk} {\sc neg} buy  horse\\
\glt  `I did not buy a horse.' 
  

   \ex\label{ex:GRM-out-STR-NEG-buy}
  *mɪŋ wa  jawa   kɪnzɪnɪɪ 
  
     \ex\label{ex:GRM-out-STR-FOC-finish-g}
\gll ǹ̩   pétījó.\\
 {\sc 1.sg.wk}  {terminate.{\sc pfv.foc}}\\
\glt  `I finished.' 


   \ex\label{ex:GRM-out-STR-FOC-finish}
  *mɪŋ   petijo

   \ex\label{ex:}
\gll mɪ́ŋ   pétījé.\\
  {\sc  1.sg.st} {terminate.{\sc pfv}}\\
\glt   `I finished.'  

   \ex\label{ex:GRM-out-STR-NEG-finish}
  *mɪŋ   wa petije
 
   
  \z 
     \end{multicols}
 \z
%clitizised pronoun in objetc position here. Bring a CVVV example
  
  
 \subsubsection{Impersonal pronouns}
 \label{sec:GRM-impers-pro}

An impersonal pronoun does  not refer to a particular person or thing. The form
{\sls a} is treated as an impersonal pronoun  in some
particular context.

      
\begin{exe} 
\ex\label{ex:imps-pro-sing}
\gll à mááséjó kéŋ̀.\\
  {\sc 3.sg.imps}  enough.{\sc pfv.foc} {\sc dxm}\\
\glt `That's enough' or `That's it' or `Stop'
\z

Example (\ref{ex:imps-pro-sing}) is treated as a type of impersonal construction.  It is 
characterized by its  subject position being  occupied by the pronoun {\sls a}, which may be seen 
as 
referring to the situation,  but not to any participant: this particular example is appropriate in 
contexts involving pouring  liquids or giving food on a plate, or when people are quarrelling. In 
these hypothetical contexts, using the personal pronoun {\sls ʊ} instead of  the  impersonal 
pronoun 
 {\sls a} would be unacceptable.  

The language does not have a passive construction as one finds in English, for example. 
Nonetheless, 
an argument  can be demoted  by placing it in object position, here as \is{{\sc o}-clitic}{\sc 
o}-clitic  (see Section \ref{sec:GRM-morph-opro}). This is shown in (\ref{ex:GRM-vp23.3.}).


\ea\label{ex:GRM-vp23.3.}
\gll ká à nàmɪ̃̀ã́?  bà tíéú rò.\\
   {\sc q}.where {\sc art} meat {\sc 3.pl.g}b  eat.{\sc pfv}.{\sc 3.sg.o} 
{\sc
foc}\\
\glt  `Where is the meat? It has been eaten.'
\z


This type of impersonal
construction illustrated in  (\ref{ex:GRM-vp23.3.})  is characterized by the
personal pronoun {\sls ba} ({\sc 3.pl.g}b)  in
subject position. In this context,  the subject is not a known agent and the 
pronoun {\sls ba} does not refer to anyone/anything in 
particular. Therefore,  the pair {\sls a}/{\sls ba} is treated  as the singular 
and
plural impersonal pronouns, only when they occur in impersonal constructions, 
as shown above.






\subsubsection{Demonstrative pronouns}
\label{sec:GRM-demons-pro}


In the examples (\ref{ex:GRM-demons-pro-reply-1}) to
(\ref{ex:GRM-demons-pro-quest}),  the demonstrative pronouns  
function as noun phrases. All the examples below were accompanied with
 pointing gestures when uttered.



\ea\label{ex:GRM-demons-pro-reply-1}{\rm Replies to the question: Which cloth
has she chosen?}
 
 
  \ea\label{ex:GRM-demons-pro-reply-1sg} 
 \gll háǹ nā.\\
   {\sc dem.sg} {\sc foc}\\
 \glt `It is this one' 
   
   \ex\label{ex:GRM-demons-pro-reply-1pl}
    \gll hámà rā.\\
   {\sc dem.pl} {\sc foc}\\
 \glt `It is these ones' 
 
\z 
 \z



\ea\label{ex:GRM-demons-pro-quest}{\rm The speaker asks the addressee whether 
he had moved a certain object.} 

 \gll   ɪ̀ jáá háǹ nȁ.\\
  {\sc 2.sg} do {\sc dem.sg}  {\sc foc}\\
 \glt `You did THIS?' 

\z


\ea\label{ex:GRM-demons-pro-quest}{\rm Explanation on how the fingers cooperate
when they scoop t.z. from a bowl.}

 \gll hámàā ká zɪ̀ pɛ́jɛ̀ɛ̀ à zɪ́ já wà tììsè  háŋ̀.\\
 {\sc dem.pl} {\sc egr} then add.{\sc pfv} {\sc conn}  then do come support {\sc
dem.sg}\\
 
\glt `These (two fingers) are then added,
and then they come to support  this one.' 

\z



The expressions {\sls háŋ̀} ({\it sg.}) and {\sls hámà}
({\it pl.}) are employed for spatial deixis, specifically as proximal
demonstratives, corresponding to English `this' and `these' respectively. The
language does not offer another set for distal demonstratives.



\subsubsection{Interrogative words}
\label{sec:GRM-interg-pro}


Interrogative constructions are of two types:  yes/no interrogatives and  pro-form interrogatives 
(see Section \ref{sec:GRM-interr-clause}). The former type, as the dichotomy suggests, requires  
a `yes' or a `no' answer.  A pro-form interrogative  uses  an interrogative word which identifies 
the sort of information requested. In Chakali,  some interrogative words may be treated as 
pronouns, while others may be treated as the combination of a noun and a pronoun.  Table 
\ref{tab:GRM-interg-pro} gives a list of interrogative words, together with an approximate English 
translation,  the sort of information requested by each  and a link to an illustrative example of 
pro-form interrogatives.  Some examples are listed in (\ref{ex:GRM-interg-pro}) where the question 
words are marked as  {\sc q}  together with a compatible gloss.


\begin{table}[htb!]
 \caption{Interrogative pronouns \label{tab:GRM-interg-pro}}
  \centering
  \begin{Itabular}{>{\slshape}llll}
\lsptoprule 
{\rm Pronoun} & Gloss  & Meaning requested & Example\\[1ex] \midrule
bàáŋ́ & what &  non-animate entity, event & \ref{ex:vp1.11.a}\\
 àŋ́ & who & animate entity & \ref{ex:vp2.5}\\
 lìé & where & location & \ref{ex:vp9.25}\\
ɲɪ̀nɪ̃̀ɛ̃́ & why/how & condition, reason& \ref{ex:vp22.4.1}\\
(ba/a)wèŋ́  & which &  entity, event & \ref{ex:vp22.4.4}\\
 (ba/a)ŋmɛ̀nà & (how) much/many & entity, event & \ref{ex:vp22.4.10}\\
 sáŋ(a)-wèŋ́ & when & time & \ref{ex:vp22.4.15}\\
\lspbottomrule
  \end{Itabular}
\end{table}


  
  \ea\label{ex:GRM-interg-pro}
  
\ea\label{ex:vp1.11.a}
\gll bàáŋ́ ɪ̀ kàà jáà.\\
{\sc q}.what  {\sc 2.sg} {\sc egr} do\\
\glt `What are you doing?' 


\ex\label{ex:vp2.5}
\gll  àŋ́ ɪ̀ kà ná à tɔ́ʊ́ nɪ̄.\\
  {\sc q}.who {\sc 2.sg}  {\sc egr}  see {\sc art} village {\sc postp}\\
\glt  `Whom did you see in the village?' 

%check comp here
\ex\label{ex:vp9.25}
\gll lìé nī dɪ̀ tʃʊ̀ɔ̀lɪ́ɪ́ kà dʊ̀ɔ̀.\\
  {\sc q}.where {\sc postp} {\sc comp} sleeping.room   {\sc egr} exist\\
\glt  `Where is the room for sleeping?' 



\ex\label{ex:vp22.4.1}
\gll ɲɪ̀nɪ̃̀ɛ̃́ ɪ̀ já kà jááʊ́.\\
   {\sc q}.how  {\sc 2.sg} {\sc hab}   {\sc egr} do.{\sc 3.sg.o}\\
\glt  `How do you do it?' 



\ex\label{ex:vp22.4.4}
\gll áwèŋ́ ɪ̀ kà kpàɣà.\\
   {\sc q}.which   {\sc 2.sg}  {\sc  egr} catch\\
\glt  `Which one did you catch?' 


\ex\label{ex:vp22.4.10}
\gll àŋmɛ̀ná ɪ̀ kà kpàgàsɪ̀.\\
   {\sc q}.many {\sc 2.sg}  {\sc  egr}  catch.{\sc pv}\\
\glt  `How many of them did you catch? (non-human reference)' 

\ex\label{ex:vp22.4.15}
\gll {sáŋáwèŋ́} ɪ̀ kàà wáá.\\
    {\sc q}.when {\sc 2.sg} {\sc  egr} come\\
\glt  `When are you coming?' 
  
   
  \z 
 \z


When the question word {\sls lie} `where' is  followed by the locative 
postposition {\sls nɪ},  a request for a particular location is interpreted. 
This 
question word can also be followed by the noun  {\sls pe} `end' in which case 
it 
should be interpreted as `where-towards' or `where-by', e.g. {\sls líé pé ɪ̀ kà 
válà?} `Where did you go by?'.  Another form used to request information on a 
location is {\sls ká(á)}. This form is neither specific to Chakali nor to 
location {\it per se}:  Waali uses it for the same 
purpose and the form is even used to request other types  of information. For 
instance, {\sls káá tʊ́má?} means `how is work?' in the two languages. It 
might be that Chakali borrowed the form from Waali.  It was employed 
consistently in an experiment I carried out, which is discussed  in Section 
\ref{sec:SPA-exper1}. Example  (\ref{ex:GRM-vp23.3.}),  repeated in  
(\ref{ex:GRM-vp23.3.-1}), illustrates the use of {\sls ka(a)} as interrogative 
word.



\begin{exe}
\ex\label{ex:GRM-vp23.3.-1}
\gll ká à nàmɪ̃̀ã́?  bà tíéú rò.\\
   {\sc q}.where {\sc art} meat {\sc 3.pl.b} chew.{\sc pfv}.{\sc 3.sg.o} {\sc
foc}\\
\glt  `Where is the meat? It has been eaten.'

\end{exe}





When they stand alone as interrogative words, the
expressions {\sls weŋ} and {\sls ŋmɛna}, roughly
corresponding to English `which' and `how much/many', must be prefixed by either
{\sls a-} or {\sls ba-} reflecting a distinction between non-human and human 
entities respectively (see Section \ref{sec:GRM-gender}). The expression {\sls 
saŋa-weŋ} in (\ref{ex:vp22.4.15}) is
literally translated as `time which'.  The question word {\sls baaŋ} can be 
used together with {\sls wɪɪ} to correspond to English `why', i.e.
{\sls bááŋ wɪ́ɪ́ ká wàà ɪ̀ dɪ̀ wíì?}  `Why are you crying?'.  The expression {\sls 
baaŋ
wɪɪ} is equivalent to English `what matter'. 





%\begin{exe}
%\ex\label{ex:vp23.1.}
%\glll  àná ká tūgùù?.\\
 %aŋ ka tuga-u\\
 %{\sc qw}   {\sc  egr} beat-{\sc 3.sg.o}\\
%\glt  `By whom is he being beaten.'
%\z

%The question words may be followed by the focus particle. This is shown in
%example (\ref{ex:vp23.1.}) with the question word {\sls aŋ}  `who'.




\subsubsection{Possessive pronouns}
\label{secːGRM-poss-pro}

The possessive pronouns are displayed in Table \ref{tab:posspro}. 

\begin{table}[h!]
  \caption{Possessive pronouns \label{tab:posspro}}
  \centering
  \begin{Itabular}{l>{\slshape}l}
\lsptoprule 
Pronoun    &  {\rm Form}\\
Gram. function & {\rm   Possessive} \\[1ex] \midrule
{\sc 1.sg.poss}  & n̩(ː)\\
{\sc 2.sg.poss}   &   ɪ(ː)\\
{\sc  3.sg.poss}   &  ʊ(ː)\\
{\sc 1.pl.poss}   &   ja\\
{\sc 2.pl.poss}    & ma\\
{\sc  3.pl.a.poss} &  a(ː)\\
{\sc 3.pl.b.poss}   &  ba\\
 
\lspbottomrule
  \end{Itabular}
\end{table}

A possessive pronoun with a form C or V  tend to be lengthened,  although their 
length have no meaning. These pro-forms can function as possessor  ({\sc 
psor}), but never as
 possessed  ({\sc psed}),  in  an attributive possessive relation. This
is shown in (\ref{ex:vp7.15}). 

\ea\label{ex:vp7.15}
\gll à kùórù ŋmá dɪ́ ʊ̀ʊ̀ hã́ã́ŋ tʃɔ́jɛ̄ʊ́.\\
  {\sc art} chief say {\sc comp} {\sc psor}.{3.sg.poss}  {\sc psed}.wife ran.{\sc pfv.foc}\\
\glt  `The chief said that his wife ran away.' 
\z

The  weak personal pronouns and the possessive pronouns have the same forms, the
differences between the two being their respective syntactic positions and their
argument structures. The weak pronoun normally precedes a verb while the
possessive pronoun normally precedes a noun, and the weak pronoun is an
argument of a verbal predicate while the possessive pronoun can only be the
possessor in a possessive attributive construction. 
%tone differences


\ea\label{ex:vp7.15}
\gll                
{mɪ́n{\T ꜜ} ná}\\
{\sc 1.sg.st.} {\sc foc}\\

\glt  `It is MINE.' 
\z


Phrasal possessives, as in English `it is mine, yours, etc.', are expressed with the
strong personal pronoun  in a verbless identificational
construction. This is shown in (\ref{ex:vp7.15}).


\subsubsection{Reciprocity and reflexivity}
\label{sec:GRM-recipro-reflex}


Reflexive and reciprocal pronouns do not exist in Chakali.  Instead,
reciprocity and reflexivity  are
encoded in  the nominals {\sls dɔŋa}   and {\sls tɪntɪn}, which are glossed in 
the
texts as {\sc recp}  and {\sc refl} respectively.   Reciprocity is illustrated 
in
(\ref{ex:GRM-recipro}) and reflexivity in (\ref{ex:GRM-reflex}). 

%more like emphasis than reflexivity

% They are categorized nominals since {\sls % dɔŋa}  is believed to 

  \ea\label{ex:GRM-recipro}
   
   
\ea\label{ex:vp24.1.}
\gll à nɪ̀báálá bálɪ̀ɛ̀ kpʊ́ dɔ́ŋá wā.\\
  {\sc art} men two kill {\sc recp}  {\sc foc}\\
\glt  `The two men killed EACH OTHER.' 

\ex\label{ex:vp24.2.}
\gll jà kàá kpʊ́ dɔ́ŋá wá.\\
   {\sc 1.pl} {\sc fut} kill  {\sc recp}   {\sc foc}\\
\glt  `We will kill  EACH OTHER.' 

\ex\label{ex:vp24.3.}
\gll à hàmṍwísè káá júó dɔ́ŋá rā.\\
  {\sc art} children {\sc  egr} fight {\sc recp}  {\sc foc}\\
\glt  `The children are fighting against one another.' 
 
   
  \z 
 \z



 \ea\label{ex:GRM-reflex}
  
   
\ea\label{ex:vp25.1.}
\gll  à báál kpʊ̄ ʊ̀ tɪ̀ntɪ̀ŋ.\\
   {\sc art} man kill  {\sc 3.sg.poss} {\sc refl.sg}\\
\glt  `The man killed himself.' 

\ex\label{ex:vp25.2.}
\gll jà kàá kpʊ̄ jà tɪ̀ntɪ̀nsá wá.\\
  {\sc 1.pl}  {\sc fut} kill {\sc 1.pl.poss}  {\sc refl.pl} {\sc foc}\\
\glt  `We shall kill OURSELVES  .'

\ex\label{ex:vp25.4.}
\gll à bìé kpá kísìé dʊ̄ ʊ̀ʊ̀ tɪ̀ntɪ̀ŋ dáŋɪ́ɪ́.\\
    {\sc art} child take knife put  {\sc 3.sg.poss}   {\sc
refl.sg} wound.{\sc nmlz}\\
\glt  `The child wounded himself with his knife.' 
  
   
  \z 
 \z




\subsection{Qualifiers}
\label{sec:GRM-qualifier}


Since qualifiers display singular/plural pairs (as  do nouns) and verbs do not 
inflect for number (but see pluractional verb in  Section 
\ref{sec:GRM-PluralVerb}), qualifiers are treated  as part of the nominal domain. 
Examples are 
presented in (\ref{ex:GRM-qual}).


\ea\label{ex:GRM-qual}
 
  \ea\label{ex:GRM-qual-red}
sɪ̀àmá {\it sg.}, sɪ̀ànsá {\it pl.}   {\rm ({\sc cl.}1) `red'}
  \ex\label{ex:GRM-qual-bad}
 bɔ́ŋ̀ {\it sg.}, bɔ́má {\it pl.}   {\rm  ({\sc cl.}3)  `bad'}
  \ex\label{ex:GRM-qual-real}
dɪ́ɪ́ŋ {\it sg.}, dɪ́ɪ́má {\it pl.} {\rm  ({\sc cl.}3) `true, real'}

  
\z 
 \z


As shown in (\ref{ex:GRM-qual}), qualifiers agree in number with the head of the
noun phrase. 

\ea\label{ex:GRM-qual-agree}
 
  \ea\label{ex:GRM-qual-agree-sg}

\gll à  hã́-pɔ́lɪ̄ɪ̀\\
{\art} woman.{\sc cl.3.sg}-fat.{\sc cl.4.sg}\\
\glt `The fat woman' 


  \ex\label{ex:GRM-qual-agree-pl}

\gll à hã́-pɔ́lɪ̄ɛ̀.\\
{\art} woman.{\sc cl.3.pl}-fat.{\sc cl.4.pl}\\
\glt `The fat women'

  
\z 
 \z

The examples in (\ref{ex:GRM-qual-agree}) are complex stem nouns  of which the qualifier  `fat'  is 
a property of   the head `woman' (Section \ref{sec:GRM-comp-quality}). Many qualifiers are 
assigned to noun class 4, the reason being that qualifiers are often nominalized verbal stems 
(Section \ref{sec:GRM-der-agent}), e.g. {\sls pɔ́lɪ́ɪ́/pɔ́lɪ́ɛ́} ({\it qual}) 
`fat' $\leftarrow$ {\sls pɔ̀là}  ({\it v})  `fat (be)'.  Examples are provided in  
(\ref{ex:GRM-qual-cl4}).


\ea\label{ex:GRM-qual-cl4}
 
  \ea\label{ex:GRM-qual-cl4-call}
jɪ̀rà  {\rm `call'}  $>$ jɪ́rɪ́ɪ́ {\rm   `calling'}
\ex\label{ex:GRM-qual-cl4-give-birth}
lʊ̀là {\rm  `give birth'} $>$ lʊ́lɪ́ɪ́   {\rm `giving birth'}
\ex\label{ex:GRM-qual-cl4-die}
sʊ̀wà {\rm  `die'} $>$ sʊ́wɪ́ɪ́  {\rm  `corpse'}

  
\z 
 \z



Nonetheless, the two categories, noun, and qualifier, are differentiated by the
following characteristics: (i)  a qualifier must be semantically verbal (i.e. 
denoting a state or an event),
a noun must not necessarily be, and (ii) while a qualifier
modifies a noun,  a  noun functions as  the
nominal argument of the qualifier. The asymmetry is reflected in
(\ref{ex:GRM-qual-hot}).

\ea\label{ex:GRM-qual-hot}{\rm  /nʊm/ `hot'}
 
  \ea\label{ex:GRM-qual-hot-cmp-stem}
  \glll nɪ̀ɪ̀nʊ́ŋ ná.\\
 nɪɪ-nʊŋ na\\
  water-hot {\sc foc}\\
  \glt `It is HOT WATER.'

 \ex\label{ex:GRM-qual-hot-head}
  \glll  à nɪ́ɪ́ nʊ́mã́ʊ̃́.\\
 a  nɪɪ nʊma-ʊ\\
   {\sc art} water hot-{\sc pfv.foc}\\
  \glt `The water is HOT.'

 \ex\label{ex:GRM-qual-hot-qual}
  \glll  à nɪ́ɪ́ nʊ́mɪ́ɪ́ dʊ́á dé nɪ̄.\\
 [a nɪɪ nʊm-ɪ-ɪ]$_{NP}$ dʊa de nɪ \\
  {\sc art} water hot-{\nmlz}-{\sc cl.4} exist {\sc dem} {\sc postp}\\
  \glt `The hot water is there.'
  
\z 
 \z

In (\ref{ex:GRM-qual-hot-cmp-stem}) the stem {\sls nʊm} `hot' is part of the
complex stem noun {\sls nɪ̀ɪ̀nʊ́ŋ̀} `water-hot' (see Section
\ref{sec:GRM-com-stem-noun}).  In this morphological configuration, a
qualitative
modification is  established  between the stem {\sls nʊm} and the stem {\sls 
nɪɪ}.
In (\ref{ex:GRM-qual-hot-head}), {\sls nʊm}  functions as a verbal predicate in
the
intransitive clause, and the definite noun phrase {\sls a nɪɪ} `the water'
occupies
the argument position. In (\ref{ex:GRM-qual-hot-qual}) the stem {\sls nʊm} is
nominalized and the singular of  noun class 4 is suffixed. The word {\sls 
nʊ́mɪ́ɪ́} may be translated as  `the result of heat'. It is treated as a
qualifier since {\sls nɪɪ} `water' is  (the head of) the argument of the
predicate, and {\sls dʊa} is a predicate which needs   one core argument. Since 
{\sls nʊm}  can neither function as main predicate nor as head noun of the
argument phrase, and since {\sls nʊm}  is understood to be a property of the
entity
and not of the event, then {\sls nʊm} in (\ref{ex:GRM-qual-hot-qual}) is viewed 
as
a qualifier.


Given the arguments put forward, one could analyze the qualifiers as adjectives.
Both are  seen  categorically as nominals  and semantically as properties or
states.  However, there are no lexemes in Chakali  which can be assigned
the category adjective, that is, no lexeme which, in all  linguistic contexts,
can be identified as categorically distinct from nouns and verbs.  For instance,
the lexeme `intelligent' in English is an adjective in all linguistic contexts
and can `never' function as a noun or as  a verb.\footnote{Although
`intelligent' is a noun in a construction like `the
intelligent find cryptic crosswords challenging' (S. Foldvik, p. c.),  it would
not be
controversial to say that `intelligent'  undergoes a sort of  zero-derivation, 
i.e.
adjective $>$ noun.  As mentioned in footnote \ref{ft:GRM-fre-eng-deri},
distinctions between categories in English and French (and other Indo-European
languages) are often not formally signaled. }  There are no such lexemes in
Chakali. Qualifiers are
either derived
linguistic entities or idiomatic
expressions. More than one procedure is attested to construct qualifiers. In
(\ref{ex:GRM-qual-types}),   some types of qualifiers are provided.

\ea\label{ex:GRM-qual-types}
 
 \ea\label{ex:GRM-qual-t0} àbúmmò `black'  
  \ex\label{ex:GRM-qual-t1} àpʊ́lápʊ́lá `pointed, sharp'
  \ex\label{ex:GRM-qual-t2}  wɪ̀ɛ́zímíí  `wise' 


 
\z 
 \z


The expression {\sls bummo} in (\ref{ex:GRM-qual-t0}) is a nominal lexeme. When
it functions as a qualifier within a noun phrase,  the prefix vowel {\sls a-} is
suffixed to the nominal stem. Notice that this prefix vowel also occurs on
numerals (see Section \ref{sec:NUM}). The type of qualifier found in
(\ref{ex:GRM-qual-t1}) is often used to
describe perceived patterns, including color, texture, sound, manner of motion,
e.g. {\sls gã́ã́nɪ̀gã́ã́nɪ̀} `cloud state',  {\sls adʒìnèdʒìnè}
`yellowish-brown',  {\sls tùfútùfú} `smooth and soft'. Reduplication
characterises the form of this type of qualifiers. When a reduplicated qualifier
occurs in attributive function, i.e. following the head noun, it takes the
prefix {\sls a-} as well.\footnote{Although the prefix {\sls a-} on qualifier 
tends
to disappear in
normal speech. The prefix {\sls a-} is unacceptable in (\ref{ex:GRM-qual-t2}).}
The word in (\ref{ex:GRM-qual-t2}) is segmented as [[[{\sc
theme}-v]-{\sc nmlz}]-{\sc cl.4}]. The verbal stem {\sls zɪm} `know'   sees  its
theme argument incorporated, i.e.  {\sls wɪɛ-zɪm} `matters-know',  a structure
which is in turn nominalized by what I called in Section \ref{sec:GRM-der-agent}
event-nominalization.  The qualifiers in
(\ref{ex:GRM-qual-types}) are presented 
in corresponding syntactic positions in (\ref{ex:GRM-qual-types-bar}). 

%  The prefix {\sls a-} is also found with verbal state
% qualifiers.

[To do]
\ea\label{ex:GRM-qual-types-bar}
 

 \ea   {\sls  [X àbúmmò]$_{NP}$ dʊ̀à dé   tàɣàtà búmmò dʊ̄ā dē} {\rm `The black X  is
there.'}
  \ex   {\sls  [X àpʊ́lápʊ́lá]$_{NP}$ dʊ̀à dé  kísínʊ̃̀ʊ̃̀ pʊ́lápʊ́lá  dʊ̄ā dē}  {\rm 
`The
sharp  X is there'.}
\ex {\sls  [X wɪ̀ɛ́zímíí]$_{NP}$  dʊ̀à dé } {\rm  `The wise X is
there'.}
  
\z 
 \z
 
 tàɣàtà búmmò dʊ̄ā dē
 
 kísínʊ̃̀ʊ̃̀ pʊ́lápʊ́lá  dʊ̄ā dē



There are  limitations
on the number of qualifiers allowed within a noun phrase. Noun phrases with 
more than three qualifiers are often rejected by language consultants in
elicitation sessions.  The
language simply employs other strategies to stack properties. In fact noun
phrases with two qualifiers are rarely found in the texts
collected. 

The language has phrasal expressions which correspond to  monomorphemic
adjectives in some other languages. These expressions have the characteristic of
being metaphorical; their lexemic denotations may be seen as secondary, and
phrasal  denotations as non-compositional. For instance, a speaker must say 
{\sls ʊ̀ kpáɣá bàmbíí}, {\it lit.}`he has heart', if he/she wishes to express 
`he 
is
brave'. The word `brave' cannot be translated to {\sls bambii}, since its 
primary
meaning is `heart',  but to {\sls kpaɣa bambii}  `to be brave'. Another way of
expressing `brave'  is {\sls bàmbìì-tɪ́ɪ́ná}, {\it lit.} `owner of heart'. Other
examples  are {\sls síí-nʊ̀mà-tɪ́ɪ́nà}, {\it lit.} `eye-hot-owner', `wild,
violent  person'   and {\sls síí-tɪ́ɪ́ná}, {\it lit.} `eye-owner', `stingy,
greedy person'. These expressions are more frequently used as nouns in the
complement position of the identificational construction, such as in {\sls ʊ̀
jáá sísɪ́ámātɪ̄ɪ̄nā}, {\it lit.} she is eye-red-owner, `she is serious'.
As mentioned in Section \ref{sec:GRM-idiom},  it is often hard to establish
whether an expression is idiomatic when only one of the its components
is used in a non-literal sense.


%\begin{table}[h]
 %\begin{tabular}{lp{4cm}}
%Color & (see Section )\\
%Value & good, bad\\
 %Age & new, old, years old\\
 %Human propensity& mental state, physical state, behavior\\
 %Physical property& sense, consistency, texture, temperature, 
%edibility, sustantiality configuration\\
%Quantity&\\
 %\end{tabular}
%\caption{(frawley
%p.463) \label{tab:GRM-mod}}
%\end{table}

% Examples of 
% physical properties encoding
% 
% Sense
% Consistency
% Temperature
% Edibility
% Substantiality
% Configuration
% 




\subsection{Quantifiers}
\label{sec:GRM-quantifier}

%complex quantifier
%add nar as one and refer to section for other type ofwantifier sec:classifier

Quantifiers are expressions denoting quantities. They refer to the size of the
referent ensemble. The words {\sls muŋ} `all',   {\sls banɪɛ} `some' and {\sls 
tama}
`few, some' constitute the  monomorphemic quantifiers. The  former can be
expanded with a  nominal prefix. For instance, in {\sls ba-muŋ} `{\sc hum}-all'
and {\sls wɪ-muŋ} `{\sc abst}-all',  the prefixes identify the semantic class of
the entities which the expressions quantify (see Section \ref{sec:classifier}). 
The form of the quantifier {\sls banɪɛ} `some'  is  invariable: *{\sls anɪɛ}, 
*{\sls abanɪɛ} and *{\sls babanɪɛ} are unacceptable words.  The same can be 
said 
for the
word {\sls tama}
`few', which stays unchanged even  when it  modifies  nouns of different
semantic classes.  Another word treated as quantifier is {\sls máŋá} `only' as
in {\sls a nɪhããŋ maŋa kaa waa} `Only the woman is coming'.  



The 
expression {\sls kɪ̀ŋkáŋ̀} `a lot, many', which is made out of the  
classifier  {\sls kɪŋ-} and the quantitative verbal state lexeme {\sls kaŋ}  
`abundant'   (Sections  \ref{sec:classifier} and 
\ref{sec:GRM-verb-stative-active}, respectively) should not be confused with 
the 
intensifiers,  to which
we turn to examine the difference.

The expression {\sls kɪ̀ŋkáŋ̀} `a lot, many', which  is made out of the  
classifier  {\sls kɪŋ-} plus the quantitative verbal state lexeme {\sls kan}  
`abundant'   (Sections  \ref{sec:classifier} and 
\ref{sec:GRM-verb-stative-active}, respectively) should not be confused with 
the 
intensifiers  examined in Section \ref{sec:GRM-intensifier}.
The  lexeme {\sls kana} `abundant'  is semantically verbal but turns into a 
quantifier when {\sls kɪŋ-}  is prefixed to it.  Other evidence for its verbal 
status  is the utterance {\sls à kánã́ʊ̃́} `they are many' compared to {\sls 
à 
jáá tàmá} `they are few'.  In the former, {\sls kana} is the main verb of an 
intransitive perfective clause, while in the latter, {\sls tama} is the 
complement of the verb {\sls jaa} in an identificational construction  (Section 
\ref{sec:GRM-ident-cl}). Apart from {\sls  kɪŋkan}  `a lot, many',  other 
plurimorphemic (or complex) quantifiers are based on the suffixation the 
morpheme {\sls  -lɛɪ} `not'. The expression {\sls wɪ-muŋ-lɛɪ}, {\it lit.} {\sc 
abst}-all-not, as well as {\sls kɪŋ-muŋ-lɛɪ},  {\it lit.} {\sc conc}-all-not,  
both correspond to the English word `nothing' (Section \ref{sec:classifier} on 
negation).

%may be the word maŋa `only'?
%The word {\sls kɔta}`is a measure term from English quarter.

The meaning `a few' can be conveyed by  the word {\sls aŋmɛna} `how
much/many', which was introduced in Section \ref{sec:GRM-interg-pro} as an
interrogative
word. Example  
(\ref{ex:GRM-quant-int-only}) suggest that the word {\sls aŋmɛna} can also be
used in a non-interrogative way,  co-occurring here with {\sls maŋa} `only',  in
which case it is interpreted  as `amount' or `a certain number'.


\begin{exe}
 \ex\label{ex:GRM-quant-int-only}
\gll àŋmɛ̀nà máŋá tʃájɛ̄ɛ́.\\
   amount only remain.{\pfv}\\
\glt `Only a few are left.'
\z

Another way to express `(a) few'  is to duplicate the numeral {\sls dɪgɪɪ} 
`one',
e.g.
{\sls dɪgɪɪ-dɪgɪɪ ra} `There are just a few of them'.  The  examples in 
(\ref{ex:GRM-quant-mean}) show that the numeral {\sls dɪgɪɪ} `one' can
participate in the denotations of both total and partial quantities. 

\ea\label{ex:GRM-quant-mean}

 \ea {\sls mùŋ} {\rm `all' (total collective)}
 \ex  {\sls  dɪ́gɪ́ɪ́ mùŋ} {\rm `each' (total distributive)}
 \ex  {\sls  dɪ̄gɪ̄ɪ̄ dɪ́gɪ́ɪ́} {\rm `some, few' (partial distributive)}
 
\z 
 \z

The word {\sls galɪŋga} `waist' or `middle'  can also carry quantification. In
(\ref{ex:GRM-most}),  {\sls galɪŋga} is equivalent to {\sls bàkánà} (< {\sls 
bar-kaŋ},
{\it lit.} part-abound),  and means `most'.

\begin{exe}
 \ex\label{ex:GRM-most}
\gll   à kpã́ã́má  gálɪ̀ŋgà/bàkánà tʃájɛ̄ɛ́ à láʊ́ nɪ́.\\
{\art} yam.{\pl} most remain.{\pfv} {\art} farm.hut {\postp}\\
\glt  `Most of the yams remain/are left in the farm hut.'
\z

The word {\sls gba} `too' is treated as a quantifier and restricted to appear
after the subject, e.g. (\ref{ex:GRM-too-out-1})-(\ref{ex:GRM-too-out-4}). In
(\ref{ex:GRM-too-pos}), the speaker  considers himself/herself  as part of a
previously established set of individuals who beat their respective child. The
quantifier is additive such that  the denotation of the subject constituent is
added to this previously established set.  In (\ref{ex:GRM-too-neg}), it is
shown that negating the quantified expression results in an interpretation where
the speaker asserts that he/she is not a member of the set of individuals who
beat their child. Since generally there is only one `in focus' constituent in a
clause and that negation and focus cannot co-occur (see Sections
\ref{sec:GRM-foc-neg} and  \ref{sec:GRM-focus}), example (\ref{ex:GRM-too})
suggests that {\sls gba} is not a focus particle.


\ea\label{ex:GRM-too}

 \ea\label{ex:GRM-too-pos}
\gll ŋ̩̀ gbà máŋá m̩̀ bìè rē.\\
{\sc 1.sg} {\quant}.too  beat {\sc 1.sg.poss}  child {\foc}\\
\glt  `I beat my child too.' ({\it lit.} I too/as well/also beat my
child)

 \ex\label{ex:GRM-too-neg}
\gll ŋ̩̀ gbà lɛ̀ɪ́ máŋá  m̩̀ bìé.\\
{\sc 1.sg} {\quant}.too  {\neg} beat {\sc 1.sg.poss}  child\\
\glt  `I do not beat my child.' ({\it lit.} It is not me who  also
beat my child)


 \ex\label{ex:GRM-too-out-1}   \textasteriskcentered  gba m̩  maŋa a bie re
\ex \textasteriskcentered  m̩ maŋa gba a bie re
 \ex \textasteriskcentered  m̩ maŋa  a bie gba re
 \ex\label{ex:GRM-too-out-4} \textasteriskcentered  m̩ maŋa  a bie  re gba

\z 
 \z

\subsection{Intensifiers}
\label{sec:GRM-intensifier}

Although semantically it is a predicate modifier, marks a degree,  and make a 
statement stronger, I identify an intensifier on compatibility ground. Only 
color   and temperature terms have been found to be selected 
by intensifiers.


\ea\label{ex:intens-ideo} 

\ea  ásɪ̀àmā tʃʊ̃́ɪ̃́tʃʊ̃́ɪ̃́   {\rm `very/pure  red'}
\label{ex:BCTmod-prop-red}

\ex ábúmmò jírítí {\rm `very/pure black'}
\label{ex:BCTmod-prop-black} 

\ex  ápʊ̀mmá píópíó  {\rm `very/pure  white'}
\label{ex:BCTmod-prop-white}

\ex  sʊ́ɔ́nɪ̀ júlúllú  {\rm `very cold'}
\label{ex:BCTmod-prop-cold}

\ex   nʊ̀mà kpáŋkpáŋ  {\rm `very/pure hot'}
\label{ex:BCTmod-prop-hot}


\z
\z

The  intensifier ideophones  {\sls tʃʊ̃ɪ̃tʃʊ̃ɪ̃},  {\sls jiriti},  {\sls 
piopio},  
{\sls julullu},  and {\sls kpaŋkpaŋ} are translated into English as `very' 
(or 
 
`pure' in the case of color) in (\ref{ex:BCTmod-prop}). They are treated 
together as one kind of degree predicate modifier.  Note that no other 
properties have been found together with a (unique and) corresponding degree 
modifier. For instance, if one wishes to express `very X', where X refers to a 
color other than black, white, or red,   one has to employ the degree modifier 
{\sls pááá} following the term, which is a common expression in many 
Ghanaian 
languages. 


 \subsection{Numerals}
\label{sec:GRM-numeral}


\subsubsection{Atomic and Complex Numerals}
\label{sec:NUM-bas-comp}


Following \citet[263]{Gree78b}, I assume that  the simplest lexicalisation of a 
number is called a numeral
atom, whereas a complex numeral is an expression in which  one can infer at
least one arithmetical function.  A numeral atom can stand alone or can
be combined
with another numeral, either atomic or complex, to form a complex numeral. Atoms
are treated as  those forms which are not decomposable morpho-syntactically at a
synchronic level. Table \ref{tab:numeralatoms} displays the twelve
atoms of the numeral system.

 \begin{table}[!h]
  \caption{Atomic numerals from 1 to 8, 10, 20, 100, and 1000
\label{tab:numeralatoms}}
   \centering
  \begin{tabular}{llll}
\lsptoprule
Chakali &  English &  Chakali &  English\\ \midrule
 dɪ́ɡɪ́máŋá & one &   àlʊ̀pɛ̀   &seven\\    
álɪ́ɛ̀ &two   &   ŋmɛ́ŋtɛ́l &eight\\     
átòrò (HML) &three &   fí &ten\\   
ànáásɛ̀ &four & màtʃéó  &twenty\\     
 àɲɔ̃́ &five  &  kɔ̀wá (pl.  kɔ̀sá)   & hundred(s)\\  
  álòrò (HML)   &six &   tʊ́sʊ̀  (pl.  tʊ́sà) &thousand(s)\\    
 
\lspbottomrule
\end{tabular}
\end{table}

The term for `one'  is expressed  as  {\sls dɪ́gɪ́máŋá},  but  {\sls dɪ́gɪ́ɪ́}
alone  can also
be used. In general, the meaning associated with the morpheme {\sls máŋá} is
`only', e.g.  {\sls bahɪɛ̃ maŋa n̩ na} {\it old.man-only-I-saw} `I saw only an 
old
man'. 
 The number 8 is designated with  {\sls ŋmɛ́ŋtɛ́l}, an expression which is also
used to refer to the generic term for  `spider'.  Whether they are homonyms,
or whether
 their
meanings enter into a polysemous/heterosemous relationship is unclear. Another
characteristic is that the higher
numerals 100 and 1000  have their own plural form. To say a few words about 
some of the possible origins of higher numerals, the genesis of most of SWG 
higher numerals involves diffusion from non-Grusi sources, rather than from  
common SWG descents. I believe that higher numerals in the linguistic area 
where Chakali is spoken have two origins: one is Oti-Volta and the other is 
Gonja. The  forms for 100 and 1000  in Vagla and Dɛg  are similar
to Gonja's forms with the same
denotation, i.e. Gonja {\sls  kɪ̀làfá} `100' and  {\sls  kíɡ͡bɪ́ŋ} `1000'.  
Similar
form-denotation can be found in other North Guang languages (e.g.
Krache, Kplang, Nawuri, Dwang, and Chumburung) and {\sls lafa} is found in many
other Kwa languages, as well as  non-Kwa languages, e.g. Kabiye (Eastern
Grusi)  \citep{Chan09}. Borrowing is  supported by the claim that the Vaglas
and Dɛgas were where they are today before the arrival of the Gonjas
(\citealt[12-13]{Good54}; \citealt[516]{Ratt32a}), and the fact that they, but
mostly the Vaglas, are still in contact with the former conquerer, the Gonjas. 
Of all Western  Oti-Volta languages, the Tampulmas have had more contact with
Mampruli  than any other, whereas the Chakali and the Pasaale
with Waali, a language close to  Dagbani and Dagaare, all of them classified as
Western  Oti-Volta languages. Variations of Manessy's {\it oti-volta commun}
reconstructed forms {\sls *KO / *KOB}  `hundred'  and {\sls 
*TUS}  `thousand'  are found
distributed all over Northern
Ghana, cutting across genetic relationship.  It seems that the two high
numerals are areal features spread by Western  Oti-Volta languages,   and that
Chakali, Pasaale, and Tampulma speakers may  have borrowed them from languages 
with which they had the most contact, i.e.   Waali, Dagbani, Dagaare,  and 
Mampruli.

From the atoms,  the complex numerals are now examined. The arithmetical 
functions inferred are called operations. In Chakali three types of operation 
are found: addition, multiplication, and subtraction. An operation always has 
two arguments which are identified in \citet{Gree78b} as: 

\vspace{3ex}

\begin{tabular}{ll}
{ Augend:} & A value to which some other value is
added.\\
{ Addend:} & A value which is added to some other
value.\\
{ Multiplicand:} & A value to which some other
value multiplies.\\

{ Multiplier:} & A value which is multiplied to
some other value.\\

{ Subtrahend:}  & The number subtracted.\\
{ Minuend:}  & The number from which subtraction takes
place.\\
\end{tabular}
\vspace*{10pt}


The numeral {\sls   dɪ́ɡɪ́tūō} expresses the number 9. It is the only
expression associated with subtraction.  The subtrahend is the expression {\sls 
dɪɡɪɪ} `one'.   In {\sls   dɪ́ɡɪ́tūō},  the last syllable   is analyzed
as the
operation. It may originate from the state predicate  {\sls tùwó} which is
translated 
`not exist'  or `absent from' (Section \ref{sec:GRM-loc-cl}). Thus, assuming the
covert minuend 10, the numerali
expression receives the functional notaton [1 {\sc absent from} 10], or 10
minus 1.  The number 9 may also be expressed as {\sls sàndòsó}  (or
{\sls sandʊsə} in Tuosa and Katua). This expression is
used by some individuals in Ducie, Tuosa, and Katua, all of them from the most
senior generation.  One language consultant
associates  {\sls sàndòsó} with the language of women, but his claim is not
sustained by other language consultants. For the number 9, \citet[33]{Good54}
reports
{\sls saanese}
from the village Katua and  \citet[117]{Ratt32b} puts {\sls sandoso} as the form
 for 9 in Tampulma. 

 
A proper  treatment of  atomic versus  complex numerals   relies  on evidence as
to whether
a numeral is synchronically  decomposable. In  that spirit,  numbers from 
11 to 19 are expressed with  complex numerals:  one piece of evidence, which is
presented in Section \ref{sec:GRM-gender} and repeated in section
\ref{sec:NUM-npstruc}, comes from the gender agreement between the head of a
noun phrase and the
cardinal numeral functioning as modifier.  Table \ref{tab:numral11-19}
provides the  numerals from 11 to 19 with a common structure
[fi$_{10}$-d(ɪ)-X$_{1-9}$]. 




  \begin{table}[!h]
  \caption{Complex numerals from 11 to 19  \label{tab:numral11-19}}
   \centering

  \begin{tabular}{>{\slshape}ll}
\lsptoprule
 {\rm Chakali} & English\\ \midrule
 fídɪ̄dɪ́ɡɪ́ɪ́ & eleven\\
 fídáālìɛ̀ & twelve\\
  fídáātòrò &  thirteen\\
fídànáásɛ̀ &  fourteen\\
 fídàɲɔ̃́ &  fifteen\\
 fídáālòrò  & sixteen\\
fídālʊ̄pɛ̀ &  seventeen\\
  fídɪ̀ŋmɛ́ŋtɛ́l  &  eighteen\\
  fídɪ̀dɪ́ɡɪ́túò &  nineteen\\

\lspbottomrule
\end{tabular}
\end{table}

The criterion employed for  the distinction between augend and addend is that an
augend is serialized, that is, it is the expression which is constant in a
sub-progression. This expression is called the base. In the progression from 
eleven to nineteen shown  in  Table  \ref{tab:numral11-19},  the augend is 
{\sls 
fi} and the addends are the expressions for one to nine. Given the
above definition of a base,  the expression 
{\sls fi} is  the base in complex
numerals  from 11 to 19. The operator for
addition is {\sls dɪ} and its vowel surfaces only when the
 following word starts with a consonant (i.e. {\sls fídɪ̀ŋmɛ́ŋtɛ́l} `18', but
{\sls 
fídànáásɛ̀} `14').

Table \ref{tab:numeral21-99} provides the sequences of  numeral atoms forming the complex numerals 
referring to  numbers from 21 to 99. Some numeral  forms will come after an explanation of the 
table.



  \begin{table}[!h]
  \caption{Complex numerals from 21 to 99  \label{tab:numeral21-99}}
  \centering

  \begin{tabular}{lll}
\lsptoprule 
Number  & Numeral & Meaning\\ \midrule
 
21-29& atom {\sls anɪ} atom &  20  + (1 through 9)\\ 
30  &  atom  {\sls anɪ} atom  & 20  + 10\\   
31-39&  atom {\sls anɪ} complex  & 20  + (11 through 19)\\   
40 &  atom  atom & 20 $\times$ 2\\
41-49&   atom  atom  {\sls anɪ} atom &  20 $\times$  2  + (1 through 9)\\  
50 &  atom  atom  {\sls anɪ} atom & 20 $\times$ 2 + 10\\ 
51-59 & atom  atom  {\sls anɪ} complex &20 $\times$ 2  + (11 through 19)\\ 
60 & atom  atom & 20 $\times$ 3\\ 
61-69 & atom  atom {\sls anɪ} atom  &20 $\times$ 3 + (1 through 9)\\
70 &  atom  atom  {\sls anɪ} atom& 20 $\times$ 3 + 10\\ 
71-79 &atom  atom  {\sls anɪ} complex  &20 $\times$ 3   + (11 through 19)\\ 
80 & atom  atom  & 20 $\times$ 4\\ 
81-89 & atom  atom {\sls anɪ} atom&20 $\times$ 4 + (1 through 9)\\ 
90 &  atom  atom  {\sls anɪ} atom&20 $\times$ 4 + 10\\ 
91-99 & atom  atom  {\sls anɪ} complex& 20 $\times$ 4   + (11 through 19)\\  
 
\lspbottomrule
\end{tabular}
\end{table}


% Mathematics The number by which another number is multiplied. In 8 X 32, the
%multiplier is 8.

Table \ref{tab:numeral21-99} shows us that either (i) an atom can follow another
atom without any intervening particle  or (ii) the particle {\sls anɪ} can step 
in
between two atoms, or between one atom and one complex numeral. Case (i) is understood
as a phrase which multiplies the numerical values of  two atoms. For
instance, 
{\sls màtʃéó  ànáásɛ̀} [20 $times$
4] results in the product `eighty'.  All numeral phrases from 20 to 99 use {\sls 
matʃeo} `20'  in their formation. 
In case (ii),  the particle {\sls anɪ} is treated as an operator similar to the
semantics of  `and' in English numerals since it adds the value of each
argument, either atom or complex {\sls màtʃéó  ànáásɛ̀ ànɪ́  àlɪ̀ɛ̀} 
[20 $times$
4  $+$ 2] .  The same form is also found in noun phrases
expressing the union of two or more entities (see Section
\ref{sec:GRM-conjunc-nom}).
The vowels of {\sls anɪ} are reduced when preceded and followed by
vowels.

The same criterion applies for the distinction between multiplier and
multiplicand: the latter  is identified on the basis of what Greenberg calls
`serialization'. So, a  base may be   a serialized multiplicand as well since it
is the constant term in the complex expressions involved in a sub-progression.
The expression  {\sls matʃeo} `20' is therefore the base in complex numerals  
from 21
to 99. The composition of complex numerals is summarized in table
\ref{tab:threecompo}.


\begin{table}[h]
\caption{General structure of complex numerals  \label{tab:threecompo}}
  \centering

\begin{tabular}{lll}
\lsptoprule
  Argument & Meaning & Restriction\\
\midrule
 ($y$)   $x$   tuo  & subtraction  &$y={10}$\\
&& $x={1}$\\ 
   $x$ anɪ $y$ & addition  &$x>y$\\
$x$ dɪ $y$  & addition &$x={10}$\\
&& $y={1 \textrm{-}9}$\\ %left sister x is 10

$x y$ & multiplication &$x=20$\\
&& $y={2,3,4}$\\ %right sister y smaller than left sister x
$x y$  & multiplication  &$x={100}$\\
&& $y={2 \textrm{-}9}$\\ %left sister x is 10
$x y$  & multiplication  &$x={1000}$\\
&& $y={2 \textrm{-}999, 1000}$\\ %left sister x is 10
\lspbottomrule
\end{tabular}
\end{table}


As earlier mentioned, in subtraction  the minuend $y$ is covert. The only case
of subtraction 
is the numeral {\sls   dɪ́ɡɪ́tūō} `nine'.  Both
addition  and multiplication take two overt arguments  $x$ and
$y$. They
are presented in the first column  of Table \ref{tab:threecompo} with their
surface linear order. The operator for addition {\sls dɪ} is used only  for the
sum of 10 and numbers between 1 and 9. The form {\sls anɪ} is found in a variety
of structures, but it restricts the right sister $y$ to be lower than the left
sister $x$. In multiplication  the value of the argument $y$ 
depends on the
value of $x$. For the numerals designating  2000 and above, the argument $x$
must be
the atom {\sls tʊsʊ} `thousand' and $y$  any atom or complex numeral between 2 
and
999. There are no terms to express  `million' in Chakali. One can hear 
individuals at the
market  using the English word `million' when referring to  currency. According
to my consultants,  the expression {\sls  tʊsʊ tʊsʊ} [1000 $\cdot$ 1000]
`million' was
common, but became archaic even before the change of currency  in July 
2007. Examples of numerals are presented in (\ref{ex:diffstrings}).


   

%\begin{multicols}{2}


\ea\label{ex:diffstrings}
  \ea\label{ex:82}
\gll  màtʃéó  ànáásɛ̀ ànɪ́  àlɪ̀ɛ̀.\\
   {twenty} {four} {and} {two}\\
\glt `82'

\ex\label{ex:121}
\gll  kɔ̀wá  ànɪ́  màtʃéó  ànɪ́   dɪ́ɡɪ́máŋá.\\
    {hundred}  {and}  {twenty}  {and} {one}\\
\glt `121'

\ex\label{ex:232}
\gll  kɔ̀sá  àlɪ̀ɛ̀ ànɪ̄  màtʃéó  ànɪ́  fídáālìɛ̀.\\
    {hundred}  {two}  {and}   {twenty}  {and}  {twelve}\\
\glt `232'

\ex\label{ex:395}
\gll kɔ̀sá átòrò ànɪ́ màtʃéó ànáásɛ̀ ànɪ́  fídāāɲɔ̃̄.\\
 {hundred} {three}  {and} {twenty}  {four} {and}  {fifteen}\\
\glt `395'

\ex\label{ex:501}
\gll kɔ̀sá  áɲɔ̃̄ ànɪ́  dɪ́ɡɪ́máŋá.\\
    {hundred} {five}  {and}  {one}\\
\glt `501'

\ex\label{ex:1225}
\gll tʊ̀sʊ́  ànɪ́   kɔ̀sá  álɪ̀ɛ̀   ànɪ́  màtʃéó  ànɪ́  āɲɔ̃̀.\\
    {thousand}  {and}  {hundred}  {two}  {and}  {twenty}  {and} {five}\\
\glt `1225'

\ex\label{ex:21231}
\gll tʊ̀sʊ̀  màtʃéó   ànɪ́  dɪ́ɡɪ́máŋá ànɪ́  kɔ̀sá  ālɪ̀ɛ̀  ànɪ́ màtʃéó  ànɪ́     
fídɪ̄dɪ́ɡɪ́ɪ́ \\
    {thousand}   {twenty}   {and} {one}  {and}  {hundred}  {two}  {twenty} 
{and} {eleven}\\
\glt `21231'


\z
\z
%\end{multicols}

In summary,  the 
numeral system of Chakali is decimal (base-10) and vigesimal (base-20) and the 
base-20  operates throughout the formation of 20 to 99. In
\citet{Comr08}, numeral systems similar to the one described here are called
\textit{hybrid vigesimal-decimal}.


\subsubsection{Numerals as modifiers}
\label{sec:NUM-npstruc}

To a certain extent, Chakali offers a rigid word order within the noun phrase (Section 
\ref{sec:GRM-foc-neg}). The heading of (\ref{ex:all-num-NP}) offers an overview of  the linear 
order 
of elements in a noun phrase containing a numeral. Tha data shows that the numeral occurs following 
 
the head and the qualifier(s) and precedes the article, the demonstrative,  and the 
quantifier.\footnote{Note that the noun phrases in  (\ref{ex:all-num-NP}) and  
(\ref{ex:GRM-np-list}) were collected on an elicitation session. 
They were elicited in subject position of the sentence frame {\sls X 
wááwáʊ́/wááwá} `X has come'.}


\ea\label{ex:all-num-NP}{\rm  \textsc{art/poss}    \textsc{head}    \textsc{qual$_\textrm{1}$}    
\textsc{ qual$_\textrm{2}$}    \textsc {num}   \textsc{quant}      \textsc{dem}      
\textsc{foc/neg}\\}

  \ea\label{ex:all-w}
\gll à nɪ̀hã́ã́n-á mùŋ wááwáʊ́.\\
\textsc{art} {woman-\textsc{pl}} \textsc{quant}.all come.{\sc prf.foc}\\
\glt `All women came.'

 \ex\label{ex:all-ten-w}
\gll à nɪ̀hã́ã́n-á fí mùŋ wááwáʊ́.\\
\textsc{art} {woman-\textsc{pl}} \textsc{num} \textsc{quant}.all come.{\sc prf.foc}\\
\glt `All ten women came.'

\ex\label{ex:all-fat-ten-w}
\gll  à nɪ̀hã́pɔ́lɛ̄ɛ̄ fí mùŋ wááwáʊ́.\\
\textsc{art} {woman-\qual} \textsc{num} \textsc{quant}.all come.{\sc prf.foc}\\ 
\glt `All ten fat women came.'

\ex\label{ex:all-fat-blind-two-w}
\gll ʊ̀ nɪ̀hã́ɲʊ́lʊ́má pɔ̀lɛ̄ɛ̄  bàlɪ́ɛ́ mùŋ  wááwáʊ́.\\
\textsc{poss} {woman-\qual}  {\qual} \textsc{num} 
\textsc{quant}.all  come.{\sc prf.foc}\\ 
\glt `Both his two fat blind wives came.'

\ex\label{ex:all-fat-ten-w-those}
\gll  à nɪ̀hã́pɔ́lɛ̄ɛ̄ fí háŋ mùŋ wááwáʊ́.\\
\textsc{art} {woman-\qual} \textsc{num}  \textsc{dem} \textsc{quant}.all
  come.{\sc prf.foc}\\ 
\glt `Those all ten fat women came.'


\ex\label{ex:all-fat-ten-w-n}
\gll à nɪ̀hã́pɔ́lɛ̄ɛ̄ fí mùŋ lɛ̀ɪ̄ wááwá.\\
\textsc{art} {woman-\qual} \textsc{num} \textsc{quant}.all
\textsc{neg}  come.{\sc prf}  \\
\glt `Not all ten fat women came.'

 \ex\label{ex:full-temp}
\gll à nɪ̀hã́pɔ́lɛ̄ɛ̄ fí háŋ mùŋ  lɛ̀ɪ̄ wááwá. \\
\textsc{art} {woman-\qual}  \textsc{num}   \textsc{dem}  \textsc{quant}.all
\textsc{neg}   come.{\sc prf} \\
\glt `Not all those ten fat  women came.'

\z
\z

When they appear as noun modifiers  (or in predicative position),  a limited
number of numerals act as targets in gender agreement, i.e. only the forms 2-7. 
 This grammatical
phenomenon provides us with a  motivation to treat  the expressions for numbers
11-19 as complex numerals.
In Section \ref{sec:GRM-gender},  Chakali is analyzed as having two values for
the
feature gender (i.e. \textsc{g}{\sls a} or \textsc{g}{\sls b}, see also the personal pronouns in  
Section 
\ref{sec:GRM-personal-pronouns}). The assignment is based on the humanness 
property and
plurality of a referent.  Table \ref{tab:genders}(c) is repeated as Table 
\ref{tab:distagree} for convenience. 


\begin{table}[h]
\caption{Prefix forms on the numeral modifiers  from 2 
to 7\label{tab:distagree}}
\centering
 \begin{tabular}{lcc}
\lsptoprule
&\textsc{-hum}=\textsc{g}\textit{a}&\textsc{+hum}=\textsc{g}\textit{b}\\
\midrule
\textsc{sg}&a&a\\
\textsc{pl}&a&ba\\
\lspbottomrule
 \end{tabular} 


\end{table} 

The following examples display gender agreement between the numeral {\sls 
a-naasɛ}
`four' and the nouns {\sls bʊ̃́ʊ̃̀nà}  `goats' in (\ref{ex:NUM-domnumA.pl}), 
{\it
vííné} `cooking pots' in (\ref{ex:NUM-domnumH-.pl}), {\sls tàátá} 
`languages' in (\ref{ex:NUM-domabst.pl}) and {\sls bìsé} `children'  in
(\ref{ex:NUM-domnumH+.pl}). Again, the only numerals that agree in gender with
the noun they modify are {\sls álìɛ̀} `two', {\sls átòrò}  `three', {\it
ànáásɛ̀} `four', {\sls àɲɔ̃́} `five', {\sls álòrò}  `six',  and   {\sls 
àlʊ̀pɛ̀}
 `seven' (see examples \ref{ex:NUM-ungramhum-} and \ref{ex:NUM-ungramhum+}).
 The data in (\ref{ex:NUM-domnumA.pl})-(\ref{ex:NUM-domnumH+.pl}) tells us
that, when they function as controllers of agreement, nouns denoting non-human
animate, concrete inanimate and abstract entities  trigger the prefix form 
[{\sls a-}] on the modifying numeral whereas nouns denoting human entities 
trigger the
form [{\sls ba-}]. 


  \ea\label{ex:NUM-domnum}{\rm Agreement Domain: Numeral + Noun}
\ea\label{ex:NUM-domnumA.pl}
\gll   ŋ̩̀  kpágá   bʊ̃́ʊ̃́-ná {\bf à}-náásɛ̀ rā.\\
    \textsc{1.sg}  {have}  {goat(\textsc{g}\textit{a})-\textsc{pl}} 
{\textsc{3pl.g}\textit{a}-four}  \textsc{foc}  \\
\glt `I have four goats.'\\


\ex\label{ex:NUM-domnumH-.pl}
\gll     ŋ̩̀  kpágá víí-né {\bf à}-náásɛ̀ rā.\\
  \textsc{1.sg}  {have}  {pot(\textsc{g}\textit{a})-\textsc{pl}}  
{\textsc{3pl.g}\textit{a}-four}   \textsc{foc}  \\
\glt `I have four cooking pots.'\\


\ex\label{ex:NUM-domabst.pl}
\gll  ŋ̩̀ ŋmá  táá-tá {\bf à}-náásɛ̀ rā.\\
  \textsc{1.sg}  {speak}  {language(\textsc{g}\textit{a})-\textsc{pl}}  
{\textsc{3pl.g}\textit{a}-four}   \textsc{foc}  \\
\glt `I speak four languages.'\\


\ex\label{ex:NUM-domnumH+.pl}
\gll   ŋ̩̀  kpágá bì-sé  {\bf bà}-náásɛ̀ rā.\\
  \textsc{1.sg}  {have}  {child(\textsc{g}\textit{b})-\textsc{pl}}  
{\textsc{3pl.g}\textit{b}-four}  \textsc{foc}  \\
\glt `I have four children.'\\

\ex\label{ex:NUM-ungramhum-}
\gll   ŋ̩̀  kpágá víí-né   *aŋmɛŋtɛl/*adɪɡɪtʊʊ (ŋmɛ́ŋtɛ́l rā/dɪ́ɡɪ́tūō rō).\\
  \textsc{1.sg}  {have}  {pot(\textsc{g}\textit{a})-\textsc{pl}}  
{} {eight  \textsc{foc}/nine  \textsc{foc}}\\
\glt `I have eight/nine cooking pots.'\\

\ex\label{ex:NUM-ungramhum+}
\gll    ŋ̩̀  kpágá bì-sé   *baŋmɛŋtɛl/*badɪɡɪtʊʊ (ŋmɛ́ŋtɛ́l rā/dɪ́ɡɪ́tūō 
rō).\\
  \textsc{1.sg}  {have}  {child(\textsc{g}\textit{b})-\textsc{pl}} 
{} {eight  \textsc{foc}/nine  \textsc{foc}}\\
\glt `I have eight/nine children.'\\

\ex\label{ex:NUM-domnumH+.sg}
\gll  ŋ̩̀  kpágá víí-né   fídànáásɛ̀ rā.\\
  \textsc{1.sg}  {have}  {pot(\textsc{g}\textit{a})-\textsc{pl}}  
{fourteen}  \\
\glt `I have fourteen cooking pots.'\\


\ex\label{ex:NUM-domnumH+.sg.14}
\gll  ŋ̩̀  kpágá bì-sé   *fidanaasɛ/fídɪ́{\bf bà}náásɛ̀ rā.\\
    \textsc{1.sg}  {have}  {child(\textsc{g}\textit{b})-\textsc{pl}}  
{fourteen}  \textsc{foc}  \\
\glt `I have fourteen children.'\\
\z
\z

Recall that in Table \ref{tab:numral11-19} the numbers from 11 to 19 were all
presented with the form
{\sls  fid(ɪ)X} `Xteen'. Their treatment as complex numerals makes one 
crucial
prediction: since they   have a common structure
[fi$_{10}$-d(ɪ)-[X$_{1-9}$]$_{atom}$]$_{complex}$ and not [fid(ɪ)X]$_{atom}$,
 agreement   has
access to the atoms X$_{2-7}$ within {\sls fid(ɪ)X}. This is
illustrated with the examples (\ref{ex:NUM-domnumH+.sg}) and
(\ref{ex:NUM-domnumH+.sg.14}) using the word {\sls fidanaasɛ}
`fourteen'.
These two examples show that in cases where a controller is specified for
both \textsc{g}{\sls b} and \textsc{pl}, it must trigger the form
[ba-] on X$_{2-7}$   within the expressions referring to the numbers 12-17.



\subsubsection{Enumeration}
\label{sec:NUM-enum}

Chakali has enumerative forms. These  are numerals
with a purely sequential order characteristic and are used when one wishes
to count without 
any referential source or  to count off items one by one.

 \ea\label{ex:enum}
\gll dìèkèè, ɲɛ́wã́ã́, tòròò,  náásɛ̀,  ɲɔ̃́, lòrò, lʊ̀pɛ̀, 
ŋmɛ́ŋtɛ́l, dɪ́ɡɪ́tūo   (...)\\
one two three four five six seven eight nine {} \\
\glt `One,  two, three, four, five, six, seven,  eight, nine,  (...)' \\
\z

Basically, 
what
differentiate the forms in (\ref{ex:enum}) from  the forms in  Table \ref{tab:numeralatoms} are: 
(i) a specific enumerative use, (ii) the tendency to lengthen the last
vowel,\footnote{I also perceived lengthening  in Waali, Dɛg and
Vagla for the
corresponding enumerative
sequence.} (iii)  the numerals  expressing two, three, four, five,
six, and seven do not usually display the agreement prefix,  and
(iv) the forms for `one'
and `two' differ to a greater extent. The rest of the enumerative numerals, i.e. eight, nine, ten, 
etc., 
correspond entirely to those shown in table
\ref{tab:numeralatoms}.  In (\ref{ex:monkey}), an excerpt of a folk tale
displays the
enumerative use of numerals.

\begin{exe}
 \ex\label{ex:monkey}

\gll gbɪ̃̀ã́  píílí dìèkèè, ɲɛ́wã́ã́, tòròò,  náásɛ̀,  ɲɔ̃́, lòrò, lʊ̀pɛ̀, ànɪ́ 
háŋ̀ ŋ̩̀ kà sáŋɛ̄ɛ̄ nɪ́ŋ̀, dɪ́ɡɪ́tūo, fí\\
Monkey starts one two three four five six seven {\conn}  \textsc{dem}
\textsc{1.sg}
\textsc{egr} sit  \textsc{dxm} nine ten\\

\glt `The monkey started to count: one, two, three, four, five, six, seven, the
one I'm sitting on, nine, ten.' (CB 013)
\end{exe}



\subsubsection{Distribution}
\label{sec:NUM-distri}

Reduplication has several functions in Chakali and example 
(\ref{ex:NUM-distri1}) shows that the meaning of
distribution is expressed by the reduplication of a numeral.

\begin{exe}
\ex\label{ex:NUM-distri1}
 \gll  nɪ̀ɪ̀-tá álɪ̀ɛ̄-lɪ̀ɛ̄  ǹ̩  dɪ́ tɪ́ɛ́bá dɪgɪ-dɪgɪɪ.\\
  {water-\textsc{pl}} {two-two}   \textsc{1.sg}   \textsc{hest}   {give.\sc
3.pl} {one-one}\\
\glt  `Yesterday I gave two water bags to each individual.'\\
\end{exe}

%nɪ (human) is good instead of ba above

In (\ref{ex:NUM-distri1}) the phrase containing the thing distributed and
its quantity opens the utterance. The recipient of the giving event is covert
but is
understood here as being more than one individual. Both  forms express the
quantity of things distributed and the number of recipients per things
distributed. This is how the distributive reading is
encoded in the utterance. Compare (\ref{ex:NUM-distri2a-1010}) and
(\ref{ex:NUM-distri2b-1010}) with
(\ref{ex:NUM-distri2c-10}).

 
\begin{exe}
\ex\label{ex:NUM-distri2-10}
\begin{xlist}
 
\ex\label{ex:NUM-distri2a-1010}
\gll à kùórù  zʊ́ʊ́ zágá  múŋ  nò  à làà kpã́ã́má fí-fí.\\
  \textsc{art}  {chief}  {enter}  {compound.\textsc{sg}}   {all}  \textsc{foc} 
\textsc{conn}  {collect}  {yam.\textsc{pl}}  {ten-ten}\\
\glt  `From each house the chief takes 10 yams.'

\ex\label{ex:NUM-distri2b-1010}
\gll à  zágá  múŋ̀ tɪ́ɛ́  à  kùórù rō  kpã́ã́má fí-fí.\\
  \textsc{art} {compound} {all} {give}  \textsc{art}  {chief}  \textsc{foc}
yam.\textsc{pl}  {ten-ten}\\
\glt  `Each house gives 10 yams to the chief.'

\ex\label{ex:NUM-distri2c-10}
\gll à  zágá  múŋ̀ tɪ́ɛ́  à  kùórù rō kpã́ã́má fí.\\
  \textsc{art} {compound} {all} {give}  \textsc{art}  {chief}  \textsc{foc}
yam.\textsc{pl}  ten\\
\glt  `All the houses (the village) give 10 yams to the chief.'
\end{xlist}
\end{exe}


In (\ref{ex:NUM-distri2b-1010}) and (\ref{ex:NUM-distri2c-10}), the sources of
the giving event are kept constant. The reading in which
ten yams per house are being collected by the chief is accessible only
if the numeral {\sls fi}  `ten' is reduplicated (i.e.  {\sls fifi}).

\ea
\ea\label{ex:NUM-door-two-two}

 \gll  tɪ̀ɛ̀  à gár  nʊ̃́ã́ zènè  à nã́ɔ̃́ná  jáà  zʊ̄ʊ̄  álɪ̀ɛ̀-lɪ̀ɛ̀.\\
  {give}   \textsc{art}  {fence}  {mouth}   {big}  \textsc{art} 
{cow.\textsc{pl}}   {do} {enter} {two-two}\\
\glt  `Make the door large enough since the cows often enter two by two.'\\


\ex\label{ex:NUM-akee-apple-three-four}

 \gll  à tíí bánɪ̃́ɛ̃́ jāā  átò-tòrò  wō, à  bánɪ̃́ɛ̃́ jāā  àná-náásɛ̄.\\
 \textsc{art}  {akee.apple}  {some}   {three-three}  \textsc{foc}   \textsc{art}   {some}   
\textsc{ident} {four-four}\\
\glt  `Akee apples have sometimes  three seeds, sometimes four seeds.'\\

\z
\z


%Distribution requires two core arguments: a recepient of the distribution and a
%thing distributed. A distributive event  differs from a giving event by the
%inherent reitatative property of the event, that is distribution involves at
%least two succesive giving action involving either the same thing or the same
%recipent. 
 

%From the data presented in this section, it is hard to establish whether
%Chakali has partial or complete reduplication. 


The reduplication of the numeral {\sls álɪ̀ɛ̀} `two' in
(\ref{ex:NUM-door-two-two})
makes the
addressee understand that not only two cows might enter the cattle fence but a
possible sequence of  pairs. Similarly,   example 
(\ref{ex:NUM-akee-apple-three-four}) conveys a proposition which tells us that
the
fruit  {\sls tíì}  `Akee apple' (\textit{Blighia sapida}) can reveal sometimes
three
and sometimes
four seeds.


%(As the reference is to fruit tokens, the distributive reading is
%accessible in great part due to the precense of the quantifier banɪa 'some'.)
%\vspace{1.1cm}


\subsubsection{Frequency}
\label{sec:NUM-repet}

When the morpheme {\sls bɪ}  (Section
\ref{sec:GRM-preverb-iteration}) is prefixed to a cardinal numeral  stem, it
specifies the number of times an event happens. 

 
%(Mourelatos, 1981, p. 205)
%Mourelatos, Alexander (1981). "Events, processes, and states." In Syntax and
%Semantics: Tense and Aspect, edited by P. Tedeschi and A. Zaenen. New York:
%Academic Press.


\ea\label{ex:NUM-repet}{\rm A duty of the male's initiation for  funeral
caretaker}\\
 \gll jà wíré jà kɪ́ná rá àkà vàlà gó dùsèè múŋ nàvàl bɪ́-tòrò.\\
 \textsc{1.pl} undress  \textsc{1.pl.poss} thing \textsc{foc}  \textsc{conn} 
walk cross Ducie  \textsc{quant}.all  circuit \textsc{itr-num}\\
\glt  `We undress then walk around Ducie three times.'
\z


The meaning of {\sls bɪ}-{\sc num} corresponds to English `times' or  French 
`fois'.  Example (\ref{ex:NUM-repet}) illustrates a  case where  the morpheme 
{\sls bɪ} is prefixed to the numeral stem {\sls toro} `three' and translates 
into 
`three times'.



\subsubsection{Ordinals}
\label{sec:NUM-partitive}

Ordinal numerals are seen as those expressions conveying ranks or orders. The
investigation carried out  showed that the language
does not have a morphological marker or unique forms responsible for such a
phenomenon. Chakali expresses ranking and order by other
means.



\ea
\ea\label{ex:thirdmound}
\gll A: lìé ɪ̀ kà tá à pár?\\
 {} {where} \textsc{2.sg} \textsc{egr} {leave} \textsc{art} {hoe}\\
\glt   `Where did you leave the hoe?'

\gll B: ǹ̩ gɪ́lá à pár rá píé ātòrò tɪ̀n  gàntàl  nɪ̄.\\
 {}  \textsc{1.sg}  {leave} \textsc{art}   {hoe} \textsc{foc} 
\textsc{art} {yam.mound} {three} \textsc{art}  \textsc{reln}  \textsc{postp}\\
\glt   `I left the hoe behind the third yam mound.'

\z
\z


In example (\ref{ex:thirdmound}),  the expression {\sls a pie atoro  gantal nɪ} 
is
best translated as `behind the third yam mound' and not as `behind the three yam
mound'. In the context of B's utterance, there is no  salient set of three
mounds.  

The word  {\sls sɪnsagal} is frequently  used in combination with a numeral to
express a non-specific amount. For example  {\sls tʊ́sʊ̀ nɪ̄ sɪ́nsáɣál}  can 
be
translated into English as  `thousand and something'. In addition,  the word 
{\sls sɪnsagal} can be combined with a numeral to identify sibling ranks. In 
(\ref{ex:sibling})  {\sls sɪnsagal} is understood as `follower(s)'.  


\begin{exe}
\ex\label{ex:sibling}{\rm Sibling relationship}
\begin{xlist}

\ex\label{ex:sibling-a}
\gll ʊ̀ sɪ́nságál bátòrò jáá-ŋ̀.\\
   \textsc{3.sg.poss} {follower} {three} \textsc{ident-1.sg}\\
 \glt  `After him/her, I'm the third.'

\ex\label{ex:sibling-b}
\gll ǹ gàntàl tʊ́má jáá bàlɪ̀ɛ̀ wā.\\
 \textsc{1.sg.poss} {back} {owners} \textsc{ident}  {two}  \textsc{foc}\\
\glt   `I have two siblings younger than me.' 


\ex\label{ex:sibling-c}
\gll ǹ̩ sʊ́ʊ́ tʊ́má jáá bàlɪ̀ɛ̀ wā.\\
   \textsc{1.sg.poss} {front} {owners} \textsc{ident}  {two}  \textsc{foc}\\
 \glt  `I have two siblings older than me.'
\end{xlist}
\end{exe}



 Further, in a situation where a speaker wishes to express the fact that he/she 
won a race by getting to an a priori agreed goal, a natural way of expressing 
this would be  {\sls n̩  jaa dɪgɪmaŋa tɪɪna},  {\it lit:} I-is-1-owner,  `I am 
first'. The second and third (and so on) positions can also be expressed using 
the same construction (e.g. {\sls n̩  jaa anaasɛ tɪɪna}, {\it lit:} 
I-is-4-owner, 
 `I am fourth'). However,  there are other ways to express the same 
proposition: 
any of the expressions given in (\ref{ex:race}) is appropriate in this context.


\ea\label{ex:race}{\rm Position on a race}
\ea\label{ex:}
\gll à bàtʃʊ́álɪ́ɪ́ nɪ̄ ǹ̩ ná àlɪ̀ɛ̀  rā\\   
 \textsc{art}  {race} \textsc{postp}   \textsc{1.sg}  {see} {two}   \textsc{foc}
\\
 \glt  `At the race, I arrived second.'

\ex\label{ex:}
\gll mɪ́ŋ díjèē\\   
 \textsc{1.sg.st} {eat.\textsc{pfv}}\\
 \glt  `I arrived  first.' or `I won.'

\ex\label{ex:}
\gll mɪ́ŋ nɪ́ té sʊ̄ʊ̄, ɪ̀ sàɣà\\   
 \textsc{1.sg.st} {postp} {early} {front}  \textsc{2.sg} {be.on}\\
 \glt  `I arrived  first, you followed.'
\z
\z



Finally, the word {\sls búŋbúŋ} is translated into `first' and refers to a past 
state, its beginning or origin.

\ea\label{ex:seqevent}
 \gll   búŋbúŋ ní ǹ̩ fɪ́ wàà nʊ̃̄ã̄ sɪ̄ŋ̀.\\
first {\postp}  {\sc 1.sg} {\sc pst} {\sc neg} drink alcoholic.drink\\  
\glt  `At first, I was not drinking alcoholic beverage.'
\z


% \subsection{Number verbs (TO DO)}
% \label{sec:NUM-verb}
% 
% {\sls kpá pɛ̀} `to add'
% {\sls lɛ̀sɪ́tà} [{\sls lɛsta}] `subtract'
% {\sls bóntí}  `to divide'
% X
% {\sls ja } `equal'
% `to count'

 %A large quantity; a multitude,
%determine the number or amount of; count.
% total in number or amount; add up to.
%To constitute a group or number


\subsubsection{Miscellaneous usage of number concept} 
\label{sec:NUM-misc-usage}

In the performance of some rituals or customs, the number concepts 3
and 4 are associated with male and female respectively. Let us illustrate this
phenomenon 
with some examples. The Lobanɪɪ section of Ducie has a funeral song which is 
performed at
the death of a co-inhabitant. The song is repeated three times if the deceased
is a man and four in the case of a woman. When a person is initiated
to {\sls Sɪ́gmāá}, a male must drink the black medicine in three successive
occurrences and a female in four.  On the fifth day of the last funeral 
({\sls lúsɪ́nnà}), the children of the deceased are given food in a 
particular 
way
which involves offering the food and pulling  it back repeatedly: three times
for a male and four for a female. The same associations number-sex (i.e. {\it
three-male} and {\it
four-female})
are found in \citet[68-70]{Card27} where it is reported that, among the Kasena,
a woman must stay in her room three days after delivering a boy but four after
delivering a girl. Also,  the navel-string of a boy is twisted three times
 around her finger after being removed, but four times in the case of a
girl.

Two unusual phenomena involving numbers must be included. The first is
also found in neighboring languages (Dagaare, Waali, Buli, and probably 
others). 
The phrase {\sls tʃɔpɪsɪ alɪɛ} is used in greetings (Section \ref{}).   It 
literally means `two
days', yet it implies that the speaker has not met the addressee for a long
period  (i.e. days, weeks or years). In other languages, I have been informed 
that one
can say `two months' or `two years', but in Chakali, even if someone has not
seen another person for years it is appropriate to say  {\sls tʃɔpɪsɪ alɪɛ} `two
days'. The second concerns the reference to the number of puppies in a litter.
When a
speaker wishes to express the number of puppies a bitch has delivered, then
she/he
must add ten to the actual number. For example,  to express that a dog has given
birth to two puppies, one must say {\sls ʊ lʊla fidalɪɛ},  {\it lit.}  `She
give.birth twelve'. 




\subsubsection{Currency}
\label{sec:NUM-currency}

One peculiarity of Chakali appears when numerals are used in the domain of
currency. For example,  in (\ref{ex:70000}) the speaker needs to sell a
grasscutter (\textit{Thryonomys swinderianus}) for the price of seven Ghana
cedis.


\begin{exe}
\ex\label{ex:70000}
\gll kɔ̀sá átòrò ànɪ̀ mátʃéó àlɪ̀ɛ̀ ànɪ́ fī.\\
 hundred.\textsc{pl} three and twenty two and ten\\
\glt `Seven new Ghana Cedis, or seventy thousand old Ghana Cedis' ({\it lit:}
three
hundred and fifty)\\
\end{exe} 


Accounting for the reference to seven Ghana cedis with an expression literally
meaning three hundred and fifty (as was demonstrated in the previous
sections) is done in two steps.  First, Chakali speakers (still) refer
to the old Ghanaian currency (1967-2007), which after years of depreciation was
redenominated (July 2007). Today,  one new Ghana cedi ({\W ₵}) is worth 10,000
old Ghana cedis.\footnote{The term \textit{old} and \textit{new} were especially
used in the period of transition. The redenomination of July 2007 is the second
in the cedis history. The cedi was introduced by Kwame N'krumah in 1965,
replacing the British West African pound (2.4 cedis = 1 pound), but lasted only
two years. Thus,  the first redenomination actually occured in 1967.}  Secondly,
the Chakali word denoting `bag'  is  {\sls bʊ̀ɔ̀tɪ́à} 
(\textit{pl.} {\sls  bʊ̀ɔ̀tɪ̀sá},  \textit{etym.}  {\sls bʊɔ-tɪa} 
`hole-give').  
There is evidence
that the word has at least one additional sense in the language. In
(\ref{ex:price-market}) the prices of some items are presented.\footnote{The
prices are those recorded at the market in Ducie in
February 2008.}

%\begin{multicols}{2}

\ea\label{ex:price-market}

\ea\label{ex:yamtubers}
\gll bʊ̀ɔ̀tɪ̀à màtʃéó  átòrò ànɪ́ fí dɪ̀  àɲɔ̃́.\\
bag twenty three and ten and five\\
\glt `15,000' (for  three yam tubers)




\ex\label{ex:groundnutbag}
\gll bʊ̀ɔ̀tɪ̀à tʊ́sʊ̀.\\
bag thousand\\
\glt `200,000' (for a bag of groundnuts)


\ex\label{ex:driedcassava}
\gll bʊ̀ɔ̀tɪ̀à kɔ̀sá ālɪ̀ɛ̀.\\
bag hundred two\\
\glt `40,000' (for a basin of dried cassava)


\ex\label{ex:cassavabag}
\gll bʊ̀ɔ̀tɪ̀à kɔ̀sá ŋmɛ́ŋtɛ́l.\\
bag hundred eight\\
\glt `160,000' (for a bag of dried cassava) 


\ex\label{ex:ricebowl}
\gll bʊ̀ɔ̀tɪ̀à màtʃéó   ànáásɛ̀ ànɪ́ fī.\\
bag twenty four and ten\\
\glt `18,000' (for a bowl of rice) 

\ex\label{ex:Ricebag}
\gll bʊ̀ɔ̀tɪ̀à tʊ́sʊ̀ ànɪ́ kɔ̀sá  àɲɔ̃̄.\\
bag thousand and hundred five\\
\glt `300,000' (for a bag of rice) 


\z
\z
%\end{multicols}


In (\ref{ex:price-market}) the word {\sls bʊɔtɪa} initiates  each expression.
Since the
expressions refer solely to the amount of money, it is clear that the word {\it
bʊɔtɪa} does not have the  meaning `bag' but that  the
meaning of a numeral, i.e. 200 can be inferred. The distinction between {\it
bʊɔtɪa}$_{1}$ (=bag)
and {\sls bʊɔtɪa}$_{2}$  (=200) is supported by the following observations:  On 
some
occasions where  {\sls bʊɔtɪa} is used,  the word cannot refer to `bag' since
there are no potential referents available. In the position it occupies in
(\ref{ex:price-market}) {\sls bʊɔtɪa} is usually not pluralized, which is
obligatory for a modified noun. Further, the word {\sls kómbòrò} `half' can 
modify
{\sls 
bʊɔtɪa}$_{1}$  to mean `half a bag' (i.e. maize, groundnuts, etc), but  the
expression {\sls bʊ̀ɔ̀tɪ̀à kómbòrò} cannot mean `100 cedis' in the
language.\footnote{This claim was recently challenged by one of my consultants
who recalls his  mother using  {\sls bʊɔtɪa komboro} to mean `100 cedis'.  
Compare
this with English  where one can say \textit{half a grand} to mean 500
dollars. The reason why {\sls bʊɔtɪa komboro} was originally rejected was 
perhaps
that 100 old cedis was a very small sum  in 2008 and it was almost impossible
to hear the expression. In 2009,  another informant claimed never to have
heard such an expression to mean 100 old cedis.}  Going back to the form of the
expression given in
(\ref{ex:70000}),
it was also observed that in a conversation in which the reference to money is
understood, {\sls bʊɔtɪa}$_{2}$  is often not pronounced. One can use the 
utterance {\sls tʊ́sʊ̀}  `thousand' to refer to the price of a bag of 
groundnuts, that is an
amount of
two hundred thousand old cedis.\footnote{While a synchronic account of a sense
distinction for the form {\sls bʊɔtɪa} in Chakali is introduced, a diachronic 
one
is complicated by the reliability of oral sources and a lack of written records.
The origin of a sense distinction of the form {\sls bʊɔtɪa},  and its 
equivalent,   is found to be widespread in West Africa.  The lexical
item being discussed here is in Yoruba {\sls ʔàkpó}, Baatonum {\sls bʊɔrʊ}, 
Hausa
{\sls kàtàkù},  Dagbani {\sls kpaliŋa}, Dagaare {\sls bʊɔra}, Dagaare 
(Nandom 
dialect)
{\sls vʊɔra}, Sisaala {\sls bɔ̀tɔ́} and Waali {\sls bʊɔra}. Whether the word is
polysemous
in all these languages as it is in Chakali, I do not know. Akan and Gã had
something similar but seem to have lost the reference to currency: a study of
the words {\sls  bɔ̀tɔ́} and {\sls kotoko}/{\sls kɔtɔkɔ}  is needed.} 
Provisionally, I can say that the distinguishing characteristic of {\it
bʊɔtɪa}$_{1}$ is that it is a common noun and refers to `bag' and that {\it
bʊɔtɪa}$_{2}$ is an atomic (and a base) numeral. The latter is  a kind of hybrid
 numeral, a blend of a measure term and a numeral term,   which is only used in
the domain of currency.


 
\subsection{Demonstratives}
\label{sec:GRM-demons}


Unlike the pronominal demonstrative which acts as a noun phrase, 
a demonstrative
within the noun phrase modifies the head noun. The demonstratives in the noun
phrase are identical to the demonstrative pronouns introduced in Section
\ref{sec:GRM-demons-pro}, i.e.  {\sls haŋ}/{\sls hama}  ({\it sg.}/{\it pl.}).  



\begin{exe}
   \ex\label{ex:GRM-dem-sg}{\rm Priest talking to the shrine, holding a kola
nut above it}

\gll  má láá [kàpʊ́sɪ̀ɛ̀ háŋ̀]$_{NP}$ ká já mɔ̄sɛ̄ tɪ̀ɛ̀ wɪ́ɪ́ tɪ̀ŋ bà 
tàà búúrè\\
{\sc 2.pl} take kola.nut {\sc dem} {\sc conn} {\sc 1.pl} plead give matter {\sc
art} {\sc 3.pl.b} {\sc  egr} want\\
\glt   `Take this kola nut, we implore  you to give them what they desire.'

\z

Demonstrative  modifiers are mostly used in spatial deixis, but they do not
encode a proximal/distal distinction. Further, when a speaker uses {\sls haŋ}  
in
a non-spatial context, he/she tends to ignore the plural form (see example
(\ref{ex:GRM-dem-num}) below). In example (\ref{ex:GRM-dem-sg-non-spatial}), the
 demonstrative is placed before the quantifier,  which is not its canonical
position, as will be  shown in the summary examples in Section
\ref{sec:GRM-NP-sum}.\footnote{The plural form of {\sls tɔʊ} `village' in Katua 
is
{\sls tɔsɪ}. In the lect of Katua, the noun classes resemble the noun classes of
the Pasaale dialect, especially the lect of the villages  Kuluŋ and Yaala.} 



\begin{exe}
   \ex\label{ex:GRM-dem-sg-non-spatial}

\gll  dɪ́ ʊ̀ nʊ̃́ʊ̃́  dɪ́ [tʃàkàlɪ̀ tɔ́sá háŋ̀ mùŋ]$_{NP}$, dɪ́ bììsáà 
jáá nɪ́hɪ̃̀ɛ̃̂, bánɪ̃́ɛ̃́ 
ká bɪ̀ ŋmá dɪ́ sɔ̀ɣlá jáá nɪ́hɪ̃̀ɛ̃̂\\
{\sc comp} {\sc 3.sg}  hear {\sc comp} Chakali villages {\sc dem} {\sc quant}.all 
{\sc comp} Biisa {\sc ident} old some {\sc egr} {\sc itr}  say {\sc comp}
Sogla {\sc
ident} old\\
 \glt `He hears that of all  Chakali
settlements, some say that Biisa (Bisikan) is the oldest,  some
also say Sogola is the oldest.' ({\it Katua, 28/03/08, Jeo Jebuni})

\z


% %How does one makes the difference then? but   we notice that by adding the
% %article {\sls a} one can capture the meaning of the  proximal/distal
% % distinction. 

The examples in (\ref{ex:GRM-dem}) show that the typical position of  the
demonstrative is after the head noun and before the postposition, after the
numeral,  but before the article {\sls tɪŋ}. 


\ea\label{ex:GRM-dem} 
 
 
  \ea\label{ex:GRM-dem-n-postp} 
 \gll [tʃʊ̀ɔ̀sá háŋ̀]$_{NP}$ nɪ́ ǹ̩ǹ̩ dí kʊ̄ʊ̄ rā.\\
 morning {\sc dem} {\sc postp} {\sc 1.sg} eat t.z. {\sc foc}\\
 \glt  `This morning I ate T. Z.'

   \ex\label{ex:GRM-dem-num} 
 \gll [nárá bálɪ̀ɛ̀ háŋ̀]$_{NP}$ nā sɛ́wɪ́jɛ́ à mʊ́r.\\
person two {\sc dem}  {\sc foc} write {\sc art} story\\
\glt `THESE TWO MEN wrote the story.' 

   \ex\label{ex:GRM-dem-art}
 \gll làà [mʊ́sá záál háŋ̀ tɪ̀ŋ]$_{NP}$.\\
 collect Musa fowl  {\sc dem} {\sc art}\\
 \glt `Collect  Musah's  fowl'  


 
\z 
 \z


% \subsection{}
% \label{sec:GRM-}
%gba = too


\subsection{Focus and negation}
\label{sec:GRM-foc-neg}

When the focus is on a noun phrase, the free-standing particle {\sls ra} appears
to the right of the noun phrase (see Section \ref{sec:focus-forms} for the
various forms the focus particle can take). The   particle {\sls lɛɪ} `not' also  appears 
free-standing
to the right of the noun phrase, but it is part of the word in the case of a complex quantifier
(see Sections \ref{sec:GRM-quantifier}  and  \ref{sec:classifier}). Focus and
negation particles cannot co-occur together in a single noun phrase.  

\begin{exe}
 \ex\label{ex:GRM-foc-neg}{\rm Identification repair for sets of cats shown on an illutration}\\
 
 \gll [à dìèbísè hámà]$_{NP}$ lɛ̀ɪ́, [hámà]$_{NP}$ rā.\\
  {\sc art} cats {\sc dem.pl} {\sc neg}  {\sc dem.pl} {\sc foc}\\
 \glt   `Not these cats, THESE CATS.'
\z

In  (\ref{ex:GRM-foc-neg}), {\sls lɛɪ} `not' negates the noun phrase {\sls a 
diebise
hama} and {\sls ra} puts the focus on the demonstrative pronoun {\sls hama},
referring to a different set of cats.  Both focus and negation particles can be
thought as having scope over the noun phrases, functioning as discourse 
particles. 


\ea\label{ex:GRM-neg-cntrst}
  \ea\label{ex:GRM-neg-cntrst-nom}

 \gll  mòlìbíí lɛ́ɪ̀  kàà tɪ́ɛ̀ nárá tʊ̀ɔ̀rà\\
 money {\sc neg} {\sc ipfv} give people problem\\
 \glt  `It is not money that gives people problems.'

 
  \ex\label{ex:GRM-neg-cntrst-verb}
 \gll   mòlìbíí wàà tɪ́ɛ̀ nárá tʊ̀ɔ̀rà \\
 money {\sc neg.ipfv} give people problem\\
 \glt  `Money does not give people problems.'

\z
\z

Example ({\ref{ex:GRM-neg-cntrst}}) compares similar propositions involving negation. While 
({\ref{ex:GRM-neg-cntrst-nom}}) presupposes   it is something else than money  that gives  problems 
to people,  ({\ref{ex:GRM-neg-cntrst-verb}}) says that money does not give people problems.




\ea\label{ex:GRM-foc-lenght}
  \ea\label{ex:GRM-foc-w-lenght}
 \gll wáá/kàláá kpágá bʊ̀ɲɛ̃́.\\
 {\sc 3.sg.st}/K.{\foc}  have respect\\
 \glt  `HE/KALA has respect for others'

   \ex\label{ex:GRM-foc-n-lenght} 
 \gll ʊ̀/kàlá kpágá bʊ̀ɲɛ̃́ rá.\\
{\sc 3.sg}/K.  has respect {\foc}\\
\glt `He/Kala has RESPECT FOR OTHERS.' 

 \ex\label{ex:GRM-foc-w-lenght-2} 
\gll  wáá/bèléé kpágá záàl.\\
 {\sc 3.sg.st}/wild.dog{\foc}    catch fowl\\
 \glt   `IT/WILD DOG catches fowls.'

\ex\label{ex:GRM-foc-w-lenght-2} 
\gll  ʊ̀/bèlè kpágá záál là.\\
 {\sc 3.sg}/wild.dog catch fowl {\foc}\\
 \glt  `It/Wild dog catches FOWLS.'

\z
\z


The focus contrast offered in (\ref{ex:GRM-foc-lenght}) is still in need of validation:  one 
consultant insisted that if the focus particle does not appear after the object of {\sls kpaga} the 
subject --  in this case a pronoun or a proper noun -- needs to be lengthen and display high tone. 
This appears to co-relate to the distinction offered for personal pronoun in Section 
\ref{sec:GRM-personal-pronouns}.


\subsection{Coordination of nominals}
\label{sec:GRM-coord-nom}

\subsubsection{Conjunction of nominals}
\label{sec:GRM-conjunc-nom}



The coordination of nominals is accomplished by means of the conjunction
particle {\sls anɪ} ({\it gl.} {\sc conn}).  The vowels of the connective are 
heavily
centralized and the initial vowel is often dropped in fast speech.
The particle can be weakened to [nə], or simply [n̩], when the preceding and
following phonological material is vocalic.  A coordination of two indefinite
noun phrases is displayed in (\ref{ex:GRM-coor-ani}). 

\begin{exe}
 \ex\label{ex:GRM-coor-ani} 
 
 \gll váá ànɪ́ dìèbíè káá válà.\\
dog {\sc conn} cat {\sc  egr} walk\\
 \glt  `A dog and a cat are walking.'
\z

 The coordination of a sequence of
 more than two nouns is given in (\ref{ex:GRM-coor-sequen}). It is possible to
repeat the 
connective {\sls anɪ}, but a pause between the items in a
sequence is more
frequently found. 

\begin{exe}
 \ex\label{ex:GRM-coor-sequen} 
 
 \gll  bʊ̃́ʊ̃́ŋ, váà ànɪ́ dìèbíè káá válà.\\
 goat,  dog {\sc conn} cat {\sc  egr} walk\\
 \glt   `A goat, a dog and a cat are walking.'
\z

When a sequence of  two  modified nouns are conjoined, the head of the second
noun phrase may be omitted if it refers to the same kind of entity as
the first head noun. This is shown in (\ref{ex:GRM-coor-sequen-kind}).


\begin{exe}
 \ex\label{ex:GRM-coor-sequen-kind}
 
 \gll ǹ̩ kpáɣá tàɣtà zén nē ànɪ́ (tàɣtà) ábūmmò.\\
{\sc 1.sg} have shirt large  {\sc foc} {\sc conn} (shirt)  black\\
 \glt   `I got a large shirt and a black shirt.'
\z

If the conjoined noun phrase is definite, the article {\sls tɪŋ}
follows both conjuncts. This is shown in (\ref{ex:GRM-qual-conj}) where the
connective appears between two qualifiers.

\begin{exe}
 \ex\label{ex:GRM-qual-conj}
 
 \gll   à kór ábúmmò ànɪ́ ápʊ̀mmá tɪ̀ŋ.\\
{\sc art}  bench black {\sc conn} white {\sc art}\\
 \glt   `the black and white chair'
\z

When the weak personal pronouns are conjoined there are limitations on the order in which they can 
appear. The disallowed sequences seem to be caused by two constraints. First, consultants usually 
disapproved   the sequences where a singular pronoun is placed after a plural one. Examples are 
provided in (\ref{ex:GRM-conj-const-1}).



\ea\label{ex:GRM-conj-const-1}
\begin{multicols}{2}
\ea\label{ex:GRM-conj-const-1-g}{\rm Acceptable}\\
1.sg {\sc conn} 2.pl $>$ {\sls /n̩ anɪ ma/} [ǹnɪ́mā] \\
`I and you ({\it pl})'\\
1.sg {\sc conn} 3.pl.{\sc g}a  $>$   {\sls /n̩ anɪ a/} [ǹnánāā]\\
`I and they ({\it --hum})'\\
3.sg  {\sc conn} 2.pl $>$ {\sls /ʊ anɪ ma/} [ʊ̀nɪ́mā]\\
 `she and you ({\it pl})'\\
3.sg {\sc conn} 3.pl.{\sc g}b $>$   {\sls /ʊ anɪ ba/} [ʊ̀nɪ́bā]\\
`she and they ({\it +hum})'

\columnbreak 
\vfill

\ex\label{ex:GRM-conj-const-1-ng}{\rm Unacceptable}\\
2.pl {\sc conn} 1.sg  $>$ *{\sls /ma  anɪ n̩/}\\
3.pl.{\sc g}a   {\sc conn} 1.sg  $>$  *{\sls /a anɪ n̩/}\\
2.pl  {\sc conn} 3.sg $>$ */ma anɪ ʊ/\\
3.pl.{\sc g}b  {\sc conn} 3.sg $>$ *{\sls /ba anɪ ʊ/}\\

\z 
\end{multicols}
 \z

Secondly, the first person pronoun {\sls n̩} cannot be found after the 
conjunction, irrespective of the pronoun preceding it. The reason may be a 
constraint on the syllabification of two successive nasals.  In 
(\ref{ex:GRM-coor-sequen-nasal}), it is shown that the vowels of the 
conjunction 
{\sls anɪ} either  drop or assimilate the quality of the following vowel. In 
addition, a segment  {\sls n} is inserted between the conjunction and the 
following pronoun. 



\begin{exe}
 \ex\label{ex:GRM-coor-sequen-nasal}
/ʊ anɪ ʊ/   {\rm  3.sg.} {\sc conn}  {\rm   3.sg.} $>$ [ʊ̀nʊ́nʊ̀] {\rm  `she and she'}\\
/ʊ anɪ ɪ/   {\rm  3.sg.}  {\sc conn}  {\rm  2.sg.} $>$ [ʊ̀nɪ́nɪ̀]  {\rm   `she and you'}\\
/n̩ anɪ ʊ/   {\rm  1.sg.}  {\sc conn}  {\rm  3.sg.} $>$ [ǹ̩nʊ́nʊ̀] {\rm   `I and she'}\\
/n̩ anɪ ɪ/   {\rm  1.sg.}  {\sc conn}  {\rm  2.sg.} $>$ [ǹ̩nɪ́nɪ̀]  {\rm `I and you'}\\
/ɪ anɪ n̩/  {\rm   2.sg.}  {\sc conn}  {\rm  1.sg.} $>$ *[ɪn(V)nn̩]
\z



 If the first person pronoun {\sls n̩} were to follow the conjunction, there 
would 
be  (i) no vowel quality to assimilate, and (ii) three successive homorganic 
nasals, i.e. one from the conjunction, one inserted and one from the first 
person pronoun, which would give  rise to a sequence {\sls n(V)nn̩}. 

As shown in Table \ref{tab:GRM-conj-pron}, these problems do not arise when the 
strong pronouns ({\sc st}) are used. 



\begin{table}[htb!]
\small
 \caption[Conjunction of pronouns]{Conjunction of pronouns;  weak
pronoun ({\sc wk}) and   strong pronoun ({\sc st}) \label{tab:GRM-conj-pron}}

  \centering
  \begin{Itabular}{lllll}
\lsptoprule 
 & 3.sg. \& 3.sg. & 3.sg. \& 2.sg. & 3.sg. \& 1.sg. &
2.sg. \&1.sg.\\ \midrule

{\sc wk conn wk} &
ʊnʊnʊ & ʊnʊnɪ & \textasteriskcentered & \textasteriskcentered\\

{\sc wk conn wk} &
ʊnʊnʊ & ɪnʊnʊ &  n̩nʊnʊ &  n̩nɪnɪ\\

{\sc wk conn st} &
ʊnɪwa & ʊnɪhɪŋ &  {\sls ʊnɪmɪŋ} &  {\sls ɪnɪmɪŋ}\\

{\sc st conn wk} & 
wanʊnʊ & hɪnnʊnʊ & mɪnnʊnʊ & mɪnnɪnɪ\\

{\sc st conn st} &
wanɪwa & wanɪhɪŋ &  {\sls wanɪmɪŋ}  & mɪnnɪhɪŋ\\
\lspbottomrule
 
  \end{Itabular}
 
\end{table}


\paragraph{Apposition}
\label{sec:GRM-np-apposition}


\begin{exe}
 \ex\label{ex:GRM-coor-appo} 
 
 \gll kùórù bìnɪ̀hã́ã̀ŋ ŋmá tɪ̀ɛ̀ [ʊ̀ ɲɪ́ná kùórù]$_{NP}$ dɪ́ à
báàl párá   à kùó pétùù  (...)\\
chief young.girl say  give {\sc 3.sg.poss} father   chief that 
{\art} man farm {\art} farm   finish.{\foc}  {}\\

 \glt  `The chief's daughter told her father that the young
man had finished weeding the farm (...)' (CB 014)
\z


There is another conjunction-type of nominal coordination. The noun phrase 
{\sls 
ʊ ɲɪna kuoru} `her father chief'  in (\ref{ex:GRM-coor-appo}) is treated as two
noun phrases in apposition. In this case, apposition is represented as [[ʊ
ɲɪna]$_{NP}$ [kuoru]$_{NP}$]]$_{NP}$.
% in which  the definite noun phrase precedes the
%indefinite one.


\subsubsection{Disjunction of nomimals}
\label{sec:GRM-disjunct-nom}

In a disjunctive coordination, the language indicates a
contrast or a choice by means of a high tone and long {\sls káá}, 
equivalent
to 
English `or'. 
The connective {\sls káá}  is  placed between  two disjuncts. The alternative questions in 
(\ref{ex:GRM-disjct}) are given as examples.

\ea\label{ex:GRM-disjct}
\ea\label{ex:GRM-disjct-1}

\gll ɪ̀ búúrè tí rē káá kɔ́fɪ̀.\\
  {\sc 2.sg} want tea {\sc foc} {\sc conn} coffee\\
\glt  `Do you want tea or coffee?' 

\ex\label{ex:GRM-disjct-2}

\gll ɪ̀ búúrè tí rē káá kɔ́fɪ̀ rā ɪ̀ dɪ̀ búúrè.\\
  {\sc 2.sg} want tea {\sc foc} {\sc conn} coffee {\sc foc} {\sc 2.sg}
{\sc ipfv} want\\
\glt  `Do you want tea or do you want coffee?' 

\z 
 \z



This  connective
 should not be confused with the three conjunctions used to connect verb
phrases and clauses, i.e. {\sls aka}, {\sls ka} and {\sls a} (see Section
\ref{GRM-clause-coord}).   

%It can also be used at the end
%of a yes/no interrogative clause (see
%Section ).


\ea\label{ex:GRM-dij-vp4.5}

\gll ɪ̀ kàá tʊ̀mà tɪ̀ɛ̀ à kùórù ró zàáŋ káá tʃɪ́á.\\
  {\sc 2.sg} {\sc fut} work give {\sc art} chief  {\sc foc} today or
tomorrow\\
\glt  `Will you work for the chief today or tomorrow?' 
\z



Example (\ref{ex:GRM-dij-vp4.5}) shows that the same
particle may also occur between
 \is{temporal nominals}temporal nominals. 





\subsection{Two types of agreement}
\label{sec:GRM-agrrement}

Agreement is a phenomenon which operates
across word boundaries: it is a relation between a controller and a
target in a given syntactic domain. In \citet{Corb04, Corb06} 
  agreement is defined as follow: (i) the element which determines the
  agreement is the controller, (ii) the element whose form is determined by
  agreement is the target and (iii) the syntactic environment in which
  agreement occurs is the domain. Agreement features refer to the information
which is shared in an agreement domain. Finally there may be conditions on
  agreement, that  is, there is a particular type of agreement provided certain
  other conditions apply. Chakali has two types of agreement based on animacy.
They are presented in the two subsequent sections. 

\subsubsection{The gender system}
\label{sec:GRM-gender}

% The glosses for the gender values in the examples are indicated
% with {\sls {\sc g}a} for residuals and with {\sls {\sc g}b} for human. 

Gender is identified as the grammatical encoding of an agreement class.  
Chakali has four domains
in which agreement in gender can be observed; antecedent-anaphor,
possessive-noun, numeral-noun and  quantifier-noun.
The values shared reflect the humanness property of the referent,
dichotomizing the lexicon of nominals into a set of lexemes $a$ (i.e.
human-) and a set $b$ (i.e.  human+), thus {\sc gender} $a$ or
$b$   \citep{brin08c, brin08}.  The values for the feature {\sc gender} are presented in
Table \ref{tab:genders}. 


\begin{table}[htb!]
  \centering
  \caption{Gender in Chakali}
\label{tab:genders}


\subfloat[][Criteria for gender]{
 \begin{tabular}{llll}
\lsptoprule 
{\sc gender} & && Criteria\\ \midrule
\textit{a} &&& \textit{residuals}\\
\textit{b} &&& things that are categorized as human\\ 
\lspbottomrule  \end{tabular}
}

\subfloat[][Gender in weak  and strong third-person
pronouns]{
  \begin{tabular}{lll}
\lsptoprule
Pronoun & {\sc wk}  & {\sc st}\\
Grammatical function  & {\sc s|o}  &  {\sc s}\\ \midrule

{\it 3.sg.}  & {\sls ʊ} & {\sls waa}\\
{\it  3.pl.\textsc{g}a} & {\sls a} & {\sls awaa}\\
{\it  3.pl.\textsc{g}b} &  {\sls ba} &  {\sls bawaa}\\
 
\lspbottomrule
  \end{tabular}
}

\subfloat[][Agreement prefix forms]{
 \begin{tabular}{lcc}
\lsptoprule
&\textsc{-hum}=\textsc{g}\textit{a}&\textsc{+hum}=\textsc{g}\textit{b}\\
\midrule
{\it sg.}&{\sls a-} &{\sls a-}\\
{\it pl.}&{\sls a-} &{\sls ba-}\\
\lspbottomrule
 \end{tabular}
}

\end{table}



  In addition to the gender values proposed in Table \ref{tab:genders}(a),  a condition constrains 
the controller to be plural to observe the humanness distinction in agreement. As Table  
\ref{tab:genders}(b)  and   \ref{tab:genders}(c) show, the personal pronouns in the language do not 
distinguish humanness in the singular but only in the plural.
 
 The boundary separating human from non-human is subject to conceptual flexibility. In story 
telling 
non-human characters are ``humanized'', sometimes called personification, as (\ref{ex:antanaH+}) 
exemplifies: animals talk, are capable of thoughts and feelings, and can plan to go to funerals. If 
one compares the non-human referents in example (\ref{ex:antanaH+}) and (\ref{ex:domquantH-}), the 
former reflects personification, while the latter does not.

\begin{exe}
  \ex\label{ex:antanaH+}{\rm Domain: antecedent-anaphor}\\
\gll   váá  mã̀ã̀  sʊ́wá.   ʊ̀   ŋmá   dɪ́   ʊ̀  tʃɛ̀ná  
ŋmálɪ́ŋŋmɪ̃̀ɔ̃́ʊ̃̀   dɪ́   bá  káálɪ̀  ʊ̀ mã̀ã́  lúhò\\
    dog.{\sc sg} mother.{\sc sg} {die} {he} {said} {\sc comp} {his}
{friend} {bird's name} {\sc comp} {{\sc 3pl.g}{\it b}} {go} {his}
{mother} {funeral}\\
\glt `The Dog's mother died. Dog asks his friend Bird ({\it Ardea purpurea}) to 
accompany him to his
mother's funeral.'  ({\it lit}. that they should go to his mother's
funeral.) 
\z





In (\ref{ex:domquant}) the quantifier {\sls muŋ} `all'
agrees in gender with the nouns {\sls nɪbaala} `men' and
{\sls bɔlasa} `elephants'.  The form {\sls amuŋ} is
used with non-human, irrespective of the number value, and for human if
the referent is unique. The form  {\sls bamuŋ} can only  appear in such a phrase
if the referent is human and the number of the referent is greater than one. In
this example a contrast is being made between
human-reference and animal-reference to show that it is not animacy in general 
but
humanness
which presents an opposition in the language.


\ea\label{ex:domquant}{\rm Domain: Quantifier + Noun}\\


\ea\label{ex:domquantH+}

\gll   nɪ̀-báál-á  \textbf{bā}-mùŋ.\\
    {person({\sc g}{\it b})-male-{\sc pl}} {\sc g}{\it b}-{\sc all}\\
\glt `all men'\\

\ex\label{ex:domquantH-}

\gll   bɔ̀là-sá  \textbf{á}-mùŋ.\\
  {elephant({\sc g}{\it a})-{\sc pl}}  {\sc g}{\it a}-{\sc all}\\
\glt `all elephants'\\


\z 
 \z

In Section \ref{secːGRM-poss-pro}, it was shown   that the possessive pronouns
have the same forms as
the corresponding weak
pronouns.  In
(\ref{ex:domposs}),  the target pronouns agree with the covert
controller, which is the possessor of the possessive kinship relation.
The nouns referring to goat and human mothers, trigger
{\sc g(ender)}{\it a} and {\sc g(ender)}{\it b} respectively. In cases where 
the possessor is covert the proper
assignment of humanness is dependent on the humanness of the possessed
argument (i.e. `their child' is ambiguous in Chakali unless one can 
retrieve the relevant semantic  information of the possessed entity).

\ea\label{ex:domposs}{\rm Domain: Possessive (possessor) + Noun}\\


\ea\label{ex:dompossH+}

\gll  \textbf{bà}   bì-sé.\\
     {\sc poss.3pl.g}{\it b}
{child-{\sc pl}}\\
\glt `their children' (possessor = human mothers) 

\ex\label{ex:dompossH-}

\gll  \textbf{à}   bʊ̃́ʊ̃́n-á.\\
   {\sc poss.3pl.g}{\it a} {goat-{\sc
pl}}\\
\glt `their goats' (possessor = goat mothers) 





\z 
 \z
%\hfill{(Tampulma)}
 

%%%%%%%%%%HERE 05-10-09
Example (\ref{ex:domnum}) displays agreement between the numeral
{\sls à-náásɛ̀} `four' and the nouns  {\sls bʊ̃́ʊ̃̀nà}
({\sc cl.3}) `goats',  {\sls tàátá} ({\sc cl.7})
`languages',  {\sls vííné} ({\sc cl.5}) `cooking
pots' and  {\sls bìsé} ({\sc cl.1}) `children'. The  numerals that agree
in gender with the noun they
modify are  {\sls á-lɪ̀ɛ̀} `two',
 {\sls á-tòrò} `three',  {\sls à-náásɛ̀}
`four',  {\sls à-ɲɔ̃́} `five',
 {\sls à-lòrò} `six' and  {\sls á-lʊ̀pɛ̀}
`seven'. Here again, animate (other than human), concrete (inanimate) and
abstract entities on the one hand, and human on the other hand do not
trigger the same agreement pattern ({\sc anim} in (\ref{ex:domnumHA}),
 {\sc abst} in  (\ref{ex:domabst}),   {\sc conc} in (\ref{ex:domnumI})  vs. 
{\sc hum} in  (\ref{ex:domnumH+})). Clearly, as shown below, noun
class membership is not reflected in agreement ({\sls tàátá} ({\sc cl.7})
`languages' triggers {\sc g}a in (\ref{ex:domabst}) and 
 {\sls bìsé} ({\sc cl.1}) `children' triggers {\sc g}b in 
(\ref{ex:domnumH+})).


\ea\label{ex:domnum}{\rm Domain: Numeral + Noun}\\


\ea\label{ex:domnumHA}

\gll  ǹ̩ǹ̩  kpáɣá  bʊ̃́ʊ̃́n-á  \textbf{à}-náásɛ̀ rā.\\
    {\sc 1sg}  {have}  {goat({\sc g}{\it a})-{\sc pl}}  {{\sc 3pl.g}{\it
a}-four} {\sc foc}\\
\glt `I have four goats.'\\

\ex\label{ex:domabst}

\gll   ǹ̩ǹ̩  ŋmá  tàà-tá \textbf{à}-náásɛ̀ rā.\\
  {\sc 1sg}  {speak}  {language({\sc g}{\it a})-{\sc pl}}   {{\sc
3pl.g}{\it a}-four} {\sc foc}\\
\glt `I speak four languages.'\\



\ex\label{ex:domnumI}

\gll  ǹ̩ǹ̩  kpáɣá  víí-né   \textbf{à}-náásɛ̀ rā.\\
  {\sc 1sg}  {have}  {cooking.pot({\sc g}{\it a})-{\sc pl}}   {{\sc
3pl.g}{\it a}-four}  {\sc foc}\\
\glt `I have four cooking pots.'\\



\ex\label{ex:domnumH+}

\gll  ǹ̩ǹ̩   kpáɣá  bì-sé  \textbf{bà}-náásɛ̀  rā.\\
  {\sc 1sg}  {have}  {child({\sc g}{\it b})-{\sc pl}}   {{\sc 3pl.g}{\it
b}-four}  {\sc foc}\\
\glt `I have four children.'\\



\z 
 \z



Example (\ref{ex:all}) shows that in a coordination construction
involving the conjunction form {\sls (a)nɪ},  the targets display
consistently {\sc g}{\it b} when one of the conjuncts is
human-denoting.  In (\ref{ex:alla}) the noun
phrase {\sls a baal} `the man' and the noun phrase
 {\sls ʊ  kakumuso} `his donkeys' unite to form the noun
phrase acting as controller.  The noun phrase  {\sls a
 baal nɪ ʊ kakumuso} `the man and his
donkeys' triggers {\sc g}{\it b} on targets.  Consequently, the
form of the subject pronoun, the quantifier, the possessive pronoun
and the numeral must expose  {\sls ba} ({\sc 3.pl.}{\it b}).
The rule in (\ref{ex:rule}) constrains coordinate noun phrases to
trigger {\sc g}{\it b} if any of the conjuncts is specified as
{\sc g}{\it b}. No test has been applied to verify whether the
alignment of the conjunct noun phrases affects gender
resolution.






\ea\label{ex:all}{\rm Domain: Coordinate structure with {\it nɪ}}\\


\ea\label{ex:alla}

\gll  [à  báál   nɪ̀  ʊ̀ʊ̀  kààkúmò-sō]$_{NP}$  váláà  káálɪ̀  
tàmàlè
rā
\\
   {\sc art} {man} {\sc conn}  {\sc 3.sg.poss} {donkey-{\sc pl}} {walk}  {go}
{Tamale} {\foc}\\
\glt `The man and his donkeys walked to TAMALE'\\

\ex\label{ex:Tamanaphor}

\gll  \textbf{bà}  kʊ̃́ʊ̃́wã́ʊ̃́.\\
    {{\sc 3pl}.{\sc g}{\it b}} tire.{\pfv}.{\foc}\\
\glt `They are tired'\\

\ex\label{ex:Tamquant}

\gll   \textbf{bà}-mùŋ  nã̀ã̀sá tʃɔ́gáʊ́.\\
    {3.{\sc pl}.{\sc g}{\it b}-all} {feet.{\sc pl}} spoil.{\pfv}.{\foc}\\
\glt `All  feet were hurting'\\

\ex\label{ex:Tamposs}

\gll  \textbf{bà}  nã̀ã̀sá  tʃɔ́gáʊ́.\\
  {{\sc 3.pl.poss.g}{\it b}}  {feet.{\sc pl}}  spoil.{\pfv}.{\foc}\\
\glt `Their feet were hurting them'\\

\ex\label{ex:Tamnum}

\gll   \textbf{bà}  jáá \textbf{bà}-ɲɔ̃́  rā.\\
  {{\sc 3.pl.g}{\it b}} {\sc ident}
{3.{\sc pl}.{\sc g}{\it
b}-five} {\sc foc}\\
\glt `They were five altogether'\\


\ex\label{ex:rule}

{\sc resolution rule}: When unlike gender values are conjoined
(i.e. {\sc gender} {\it  a} and {\sc gender} {\it b}), the
coordinate noun phrase determines {\sc gender} {\it b} (i.e.
{\sc g}{\it a} + {\sc g}{\it a} = {\sc g}{\it a},
{\sc g}{\it a} + {\sc g}{\it b} = {\sc g}{\it b},
{\sc g}{\it b} + {\sc g}{\it a} = {\sc g}{\it b} and
{\sc g}{\it b} + {\sc g}{\it b} = {\sc g}{\it b}).




\z 
 \z

Examples (\ref{ex:antanaH+}) to (\ref{ex:all}) demonstrate how one can analyse
the humaness distinction as gender. The comparison between humans, animals,
concrete inanimate entities and abstract entities uncovers the sort of animacy
encoded in the language. Section \ref{sec:classifier}  presents a phenomenon 
which
shows
some similarity to gender agreement.



\subsubsection{The classifier system}
\label{sec:classifier}


 While there is abundant  literature describing Niger-Congo nominal
classifications and
agreement systems, the grammatical phenomenon  described in
this section  has not received much attention.  Consider the examples in
(\ref{ex:agr1}):   


\ea\label{ex:agr1}



\ea\label{ex:agrA}
\gll  dʒɛ̀tɪ̀ kɪ̀m-bɔ́n  ná.\\
  lion.{\sc sg}  {\sc anim}-dangerous.{\sc sg}  {\sc foc}\\
\glt  `A lion is DANGEROUS' (generic reading) 



\ex\label{ex:agrB}
\gll dʒɛ̀tɪ̀sá kɪ̀m-bɔ́má  rá.\\
  lion.{\sc pl}  {\sc  conc;anim}-dangerous.{\sc pl} {\sc foc}\\
\glt  `The lions are DANGEROUS' (individual reading) 



\ex\label{ex:agrD}
\gll   m̩̀ bɪ̀ɛ̀rəsá  nɪ̀-bɔ́má  rá.\\
{\sc poss.1.sg} {brother.{\sc pl}}  {\sc hum}-dangerous.{\sc pl} {\sc foc}\\
\glt  `My brothers are DANGEROUS'


\ex\label{ex:agrE}
\gll bà  jáá  nɪ̀-bɔ́má   rá.\\
{{\sc  3pl.g}{\it b}} {\sc ident} {\sc hum}-{dangerous.{\sc pl}} {\sc foc}\\
\glt  `They are DANGEROUS' (human participants) 



\ex\label{ex:agrF}
\gll   à   jáá   kɪ̀m-bɔ́má  rá.\\
{{\sc  3pl.g}{\it a}} {\sc ident}  {\sc  conc;anim}-{dangerous.{\sc pl}}
{\sc foc}\\
\glt  `They are DANGEROUS' (non-human, non-abstract participants) 



\ex\label{ex:agrG}
\gll záɪ́ɪ́   wɪ̀-bɔ́n ná.\\
 fly.{\sc nmlz} {\sc abst}-dangerous.{\sc sg}  {\sc foc}\\
\glt  `Flying is dangerous'\\




\ex\label{ex:agrH}
\gll à tʃígísíí wɪ̀-bɔ́má rá.\\
{\sc art}  turn.{\sc pv.nmlz}  {\sc abst}-dangerous.{\sc pl}  {\sc 
foc}
\\
\glt  `The turnings  are DANGEROUS' (repetitively turning clay bowls for
drying)\\


\z 
 \z

The sentences in (\ref{ex:agr1})  are made of two successive noun phrases. The referent of the 
first 
noun phrase is an entity or a process while the second noun phrase is semantically headed by a 
state 
predicate denoting a property.  Although speakers prefer the presence of   the  identificational 
verb {\sls jaa} `to be' between the two noun phrases, its  absence is acceptable and does not 
change 
the meaning of the sentence. In these  identificational constructions,  the comment identifies the 
topic as having a certain property, i.e. being bad, dangerous, or risky. The focus particle follows 
the second noun phrase, hence  $[$NP1 NP2 ra$]$  means `NP1 is NP2' in which salience or novelty of 
information comes from NP2. 


The form of  {\sls /bɔm/}   `bad' is determined by the number value of the 
first noun phrase. Irrespective of the animacy encoded in the referent, a  
singular noun phrase triggers the form {\sls [bɔŋ]} while a plural triggers 
{\sls [bɔma]} (i.e. {\sc cl.3}).  The number agreement is illustrated in 
(\ref{ex:agrA}) and (\ref{ex:agrB}).\footnote{Notice that the nominalized 
verbal lexemes in (\ref{ex:agrG}) and (\ref{ex:agrH}) each triggers a different 
form for {\sls /bɔm/}. The  form  {\sls tʃigisii}  `turning'  is analyzed as a 
nominalized pluractional verb (see Section \ref{sec:GRM-PluralVerb}).}  

Properties do not appear as  freestanding words in identificational 
constructions. To say `the lion is dangerous', the grammar has to combine the 
predicate with a nominal classifier that will license  a noun,  i.e. {\it lit.}  
`lion is {\it thing}-dangerous',  where {\it thing} stands for the slots where animacy is 
encoded. This is represented in (\ref{ex:frame}).  


\ea\label{ex:frame} [[ {\it thing}$_{animacy}$-property]$_{NP}$ {\sc foc}]$_{NP}$ \z 

There are three 
dummy substantives In  (\ref{ex:agr1}):   {\sls  nɪ-}, {\sls  wɪ-} and  {\sls  kɪn}-.  Each of them 
has a fully fledged noun counterpart; it can be pluralized, precede a demonstrative, etc. Those 
forms are {\sls kɪn}/{\sls  kɪna} ({\sc cl.3})  `thing',  {\sls  nar}/{\sls  nara} ({\sc cl.3}) 
`person' and {\sls wɪɪ}/{\sls  wɪɛ} ({\sc cl.4}) `matter, palaver, problem, etc.'.  Table 
\ref{tab:nounclassifier} provides the three possible
distinctions.  


\begin{table}[htb!]

  \caption{Classifiers and Nouns   \label{tab:nounclassifier}}
  \centering
  \begin{Itabular}[h]{lllll}
 \midrule 
 Classifier   & Animacy & Noun class  & Sing. & Plur.\\
\midrule  \midrule
   {nɪ-}/{na-} &  $[${\sc hum}$]$ & Class 3 & nár &  nárá\\ 
 {wɪ-} &  $[${\sc abst}$]$ & Class 4 & wɪ́ɪ́ &   wɪ́ɛ́\\
  {kɪn}-  &  $[${\sc conc;anim}$]$ & Class 3& kɪ̀n &   kɪ̀nà\\
 \midrule 
  \end{Itabular}
\end{table}

 %(see {\it forme radicale} in
%\citet[506]{Mane64})

Since there are form and sense compatibilities between the
inflecting noun pairs 
and the forms of the expressions preceding the qualitative predicate,  a common 
radical form for each is identified; {\sls kɪn}-
{\sc [conc;anim] } `concrete, non-human, non-abstract',  {\sls nɪ-} {\sc
 [hum] } `person, human being'  and  {\sls wɪ-} {\sc  [abst] } `non-concrete,
non-person' are the three classifiers. 
The  classifier and the semantic information encoded in the head of the
first noun phrase reflects one major analytical criterion for  
classifier systems \citep{Dixo86, Corb91, Grin00}.


All the sentences in (\ref{ex:agr1}) are ungrammatical without a classifier. The
three classifiers  combine with {\sls bɔŋ}/{\sls bɔma}  to  make  proper
constituents for an identificational construction. The complex unit made
out of  a property and a licensing classifier is schematically presented
in (\ref{ex:frame2}). 

\ea\label{ex:frame2}\textit{Identificational Construction with a classifier}\\

 [ [$^{topic}$ \ \ {\it head}$_{x}$  ]$_{NP}$ (jaa)[[$^{comment}${\sc
clf}$_{x}$-property]$_{NP}$ {\sc foc}]$_{NP}$]$_{S}$
\z

The structural setting  is the result of a combination of grammatical
constraints which specify that: (i) a property in predicative function cannot
stand on its own, (ii) in predicative function,  a property must be joined with
a classifier, (iii) the merging of the classifier and the property forms a
proper syntactic constituent for an identificational construction, and (iv) the
form of the classifier is dependent on the animacy encoded in the argument of a
qualitative predicate. 

Finally,  classifiers are also found in the formation of the words
meaning  `something' and `nothing'. Consider the examples in 
(\ref{ex:something}) and (\ref{ex:nothing}):


\begin{multicols}{2}
\ea


\ea\label{ex:somethingH}
\gll {ná-mùŋ-lɛ̀ɪ́}.\\
 {\sc hum}-all-not\\
\glt `no one'\\
\ex\label{ex:somethingC}
\gll  {wɪ́-mùŋ-lɛ̀ɪ́}.\\
 {\sc abst}-all-not\\
\glt `nothing'\\
\ex\label{ex:somethingA}
\gll  {kɪ́n-mùŋ-lɛ̀ɪ́}.\\
 {\sc  conc;anim}-all-not\\
\glt `nothing'\\

\z
\z

% 
\ea
 

\ea\label{ex:somethingH}
\gll {nɪ̀-dɪ́gɪ́ɪ́}.\\
 {\sc hum}-one\\
\glt `someone'\\
\ex\label{ex:somethingC}
\gll  {wɪ́-dɪ́gɪ́ɪ́}.\\
 {\sc abst}-one\\
\glt `something'\\
\ex\label{ex:somethingA}
\gll {kɪ̀n-dɪ́gɪ́ɪ́}.\\
 {\sc conc;anim}-one\\
\glt `something'\\

\z
\z

\end{multicols}

As with the role of classifiers in identificational constructions, here again
the classifiers narrows down the tracking of a  referent when one of those
quantifiers is used. The grammar of Chakali arranges the four animacies into
three categories, i.e.  {\sc abst}, {\sc conc;anim} and {\sc hum}.  A
distinction is also made in English between {\sc hum} (i.e. someone, no one) and
 {\sc anim;conc;abst} (i.e. something, nothing), however English does not have
a distinction which captures  specifically abstract entities.

\subsection{Summary}
\label{sec:GRM-NP-sum}

The term nominal in the present context was argued to represent two separate
notions. The first is  conceptual. Nominal stems denote classes of entities
whereas verbal stems denote events. The second notion is  formal. A nominal stem
was opposed to  a verbal stem in noun formation.  As a syntactic unit,  the
nominal  constitutes an obligatory support to the main predicate and was
presented above in  various forms:   as a pro-form, a single noun, or 
 noun phrases
consisting of a noun with a qualifier(s), an article(s), a demonstrative,  among
others.



\ea\label{ex:GRM-np-list} {\small 
\textsc{art/poss} \;  \textsc{head}\;   \textsc{qual}\;  \textsc{num}  
\;  
 \textsc{quant} \;    \textsc{dem} \;   \textsc{quant} \;   \textsc{art} \;  
\textsc{foc/neg} }\\


  \ea\label{ex:GRM-pro} 
ɪ̀   wááwáʊ́ {\rm `you came'}\\ {\sc head} 

  \ex\label{ex:GRM-h} 
hã́ã́ŋ wāāwāʊ̄ {\rm  `a woman came'}\\ 
{\sc head} 

  \ex\label{ex:GRM-ah} 
à hã́ã́ŋ  wāāwāʊ̄ {\rm  `the woman came'}\\  
{\sc art1} {\sc head} 

  \ex\label{ex:GRM-aha}
  à hã́ã́ŋ tɪ̀ŋ wāāwāʊ̄  {\rm  `the woman came'}\\  
{\sc art1} {\sc head} {\sc art2}

  \ex\label{ex:GRM-ph} 
ʊ̀ʊ̀ hã́ã́ŋ   wāāwāʊ̄  {\rm  `his woman came'}\\ 
 {\sc poss}  {\sc head}

  \ex\label{ex:GRM-pha} 
ʊ̀ʊ̀  hã́ã́ŋ tɪ̀ŋ   wāāwāʊ̄  {\rm  `his woman came'}\\
 {\sc poss} {\sc head} {\sc art2}  

 \ex\label{ex:GRM-dhq}
hámā mùŋ  wāāwāʊ̄ {\rm   `all these  came'}\\  
{\sc head} {\sc quant}{\rm .all}

 \ex\label{ex:GRM-hdq}
 nɪ̀hã́ã́ná hámā mùŋ  wāāwāʊ̄ {\rm  `all these women  came'}\\ 
  {\sc head} {\sc dem} {\sc quant}{\rm .all} 

  \ex\label{ex:GRM-hd} 
hã́ã́ŋ háŋ̀    wāāwāʊ̄ {\rm  `this woman  came'}\\ 
  {\sc head} {\sc dem}  

  \ex\label{ex:GRM-hq-all} 
 nɪ̀hã́ã́ná  mùŋ wááwáʊ́ {\rm  `all women  came'}\\  
{\sc head} {\sc quant}{\rm .all}

\ex\label{ex:GRM-hq-many} 
 nɪ̀hã́káná    wāāwāʊ̄  {\rm  `many women  came'}\\   
{\sc head-quant}  

  \ex\label{ex:GRM-hn} 
nárá bátòrò wāāwāʊ̄ {\rm  `three person  came'}\\  
  {\sc head} {\sc num}  

  \ex\label{ex:GRM-ahqln} 
à nɪ̀hã́ã́ná pɔ́lɛ̄ɛ̀ bálɪ̀ɛ̀   wāāwāʊ̄  {\rm  `the two fat women  came'}\\ 
 {\sc art1} {\sc head} {\sc qual} {\sc num}  

  \ex\label{ex:GRM-ahqlnd}
  à nɪ̀hã́ã́ná bálɪ̀ɛ̀ hámà wāāwāʊ̄    {\rm  `these two women  came'}\\ 
 {\sc art1} {\sc head} {\sc num} {\sc dem}  

  \ex\label{ex:GRM-ahqlqln}
à nɪ̀hã́ɲʊ́lʊ́má pɔ́lɛ̄ɛ̀ bálɪ̀ɛ̀ wāāwāʊ̄    {\rm  `the two fat blind women  came'} 
\\
  {\sc art1} {\sc head}  {\sc qual} {\sc qual}  {\sc num}  

\ex\label{ex:GRM-ahqlq} 
à nɪ̀hã́pɔ́lɛ̄ɛ̀    káná   wāāwāʊ̄  {\rm  `many fat women  came'}\\  
{\sc art1} {\sc head} {\sc qual} {\sc quant}{\rm .many}

 \ex\label{ex:GRM-ahqlqlq} 
à nɪ̀hã́pɔ́lɛ̄ɛ̀    ɲʊ́lʊ́ŋkáná   wāāwāʊ̄    {\rm  `many fat blind women  came'}   
\\ 
{\sc art1} {\sc head} {\sc qual} {\sc qual} {\sc quant}{\rm .many}  

  \ex\label{ex:GRM-ahqlqlqd} 
à nɪ̀hã́pɔ́lɛ̄ɛ̀   ɲʊ́lʊ́ŋkáná  hámà    wāāwāʊ̄ {\rm  `these many fat 
blind 
women  came'}\\ 
{\sc art1} {\sc head}-{\sc qual} {\sc qual}-{\sc quant}{\rm .many}  {\sc dem}


\ex\label{ex:GRM-phqlnq} 
à nɪ̀hã́pɔ́lɛ̄ɛ̀  ɲʊ́lʊ́má  fí bānɪ̃̄ɛ̃̄ wāāwāʊ̄    {\rm  `some of the ten fat 
blind women  came'}\\ 
{\sc art1} {\sc head}-{\sc qual} {\sc qual} {\sc num}  {\sc quant}{\rm .some} 


%  \ex\label{ex:GRM-phqlnq} 
% m̩̀m̩̀ párá áwíríjé átòrò bánɪ́ɛ́  {\rm  `some of my three good 
% hoes'}\\ 
% {\sc 1.sg.poss} {\sc head} {\sc qual} {\sc num} {\sc quant}.some

%\ex\label{ex:GRM-}
 
\z 
 \z
 
 
 To summarize, each noun phrase in (\ref{ex:GRM-np-list}) is a grammatically and semantically 
acceptable noun phrase and  respects the linear order offered on the first line. They were all 
elicited in a frame `X came'.  Certain orders are favored, but a strict linear order, especially 
among the qualifiers, needs further investigation.   Notice that each noun phrase in 
(\ref{ex:GRM-np-list}), except for the weak personal pronoun in (\ref{ex:GRM-pro}),  may or may not 
be in focus and may or may not be definite (i.e. accompanied by the article {\sls tɪŋ}). Also,  the 
slot {\sc head}  in (\ref{ex:GRM-np-list}) is not only represented in the examples by a noun or 
pronoun;  
example (\ref{ex:GRM-dhq}) is headed by a demonstrative pronoun. Needless to say, this list of 
possible distributions of nominal elements within the noun phrase is not exhaustive. Again, caution 
should be taken since the examples in (\ref{ex:GRM-np-list}), particularly those towards the end of 
the list, are the result of elicitation. Their order of appearance can only be  interpreted  as  an 
approximation of the noun phrase. 



\section{Verbals}
\label{sec:GRM-verbals}

% 
Any expression which can take
the place of  the predicate {\sc p} in  (\ref{ex:GRM-clause-frame-2}) is 
identified as \is{verbal} \textit{verbal}.


\begin{exe}
  \ex\label{ex:GRM-clause-frame-2}\label{ex:GRM-clause-frame-2}
 {\sc adj}  $\pm$ {\sc s|a}  $+$ {\sc p} $\pm$ {\sc o} $\pm$ {\sc adj} 
\end{exe}


The term  can also refer to a semantic notion at the lexeme 
level. The language is analyzed as exhibiting two types of verbal lexeme: the 
{\it stative} lexeme and the {\it active} lexeme were both shown in Section 
\ref{sec:GRM-der-agent} to take  part in 
nominalization processes. The verbal stem in (\ref{ex:verb-VP})  must be 
instantiated with a verbal lexeme. 

\ea\label{ex:verb-VP}
[[{\it preverb}]\textsubscript{EVG} [[{\it stem}]-[{\it 
suffix}]]\textsubscript{verb}]\textsubscript{VG}
\z


In addition, the term  can refer to the 
whole of the verbal constituent, including the verbal modifiers. The verbal 
group  \is{verbal group} (VG) illustrated in (\ref{ex:verb-VP})
consists of linguistic slots which encode   various aspects of an event  which
may be realized in an utterance. A free standing verb is the minimal requirement
to satisfy the role of a predicative expression. The verbal modifiers, which
are called preverbs (Section \ref{sec:GRM-precerv}),  are grammatical items 
which specify the event
according to various  semantic distinctions. They precede the  verb(s) and take
part in the expanded verbal group \is{expanded verbal group} (EVG). The 
expanded verbal group
identifies  a domain which excludes the main verb, so a  verbal group
without preverbs would  be equivalent to a verb or a series of verbs (see SVC in
Section \ref{sec:GRM-multi-verb-clause}).\footnote{The term and notion are 
inspired from
analyses of the verbal system of Gã \citep{Daku70}. A verbal group is unlike 
the
(traditional) verb phrase in that it does not include its internal argument,
i.e. direct object. I am aware of the obvious need to unify the descriptions of
the nominal constituent and the verbal constituent.} 


While a verbal stem provides the core meaning of the predication,  a suffix may 
supply information on  aspect, whether or not the verbal constituent is in 
focus 
and/or the index of participant(s) (i.e. {\sc o}-clitic, Section 
\ref{sec:GRM-morph-opro}).  Despite there being little focus on tone and 
intonation, attention on the tonal melody of the verbal constituent is 
necessary 
since this also affects the interpretation of the event. These characteristics 
are presented below in a brief overview of the verbal system. 

% 
% one melodies affecting not
% only elements of the verbal constituent but elements immediately preceding or
% following it, and (iii) affixes
%without its participant(s) and other peripheral expressions.



\subsection{Verbal lexeme}
\label{sec:GRM-verb-lexeme}


\subsubsection{Syllable structure and tonal melody}
\label{sec:GRM-verb-syll-und-tone}

There is a preponderance  of open syllables of type CV and CVV, and the  common 
syllable sequences found among the verbs are CV, CVV, CVCV, CVCCV, CVVCV, and 
CVCVCV.   In 
the dictionary,  monosyllabic verbs make up approximately 12\% of the verbs, 
bisyllabic 65\%,  and trisyllabic  22\%.  All segments 
are attested in onset position word initially, but only {\sls m, t, s, n, r, l, 
g, 
ŋ} and {\sls w} are found in onset position word-medially in bisyllabic verbs, 
and only {\sls  m, t, s, n,  l} and {\sls g} are found  in onset position 
word-medially in trisyllabic verbs.   All trisyllabic,  CVVCV,   
and CVCCV verbs 
have one of the front vowels (\{e, ɛ, i, ɪ\}) in the nucleus of their last 
syllable.  The data suggests that {\sc atr}-harmony is operative, but not   
{\sc ro}-harmony,  in these three environments, e.g. {\sls fùòlì} `whistle'. 
There is no restriction on vowel quality for the monosyllabic or bisyllabic 
verbs and both harmonies are operative.

 Table \ref{tab:GRM-verb-tone-melody} 
presents  verbs which are classified based on their syllable structures 
and  tonal melodies.  Despite the various attested melodies, instances of low 
tone CV verbs, CVV verbs other than low tone,  and rising or falling CVCV, 
CVCCV, and CVVCV verbs are marginal. 


%\clearpage
\begin{table}[htb]
\renewcommand{\arraystretch}{0.8}
\centering
\caption{Tonal melodies on verbs  \label{tab:GRM-verb-tone-melody}}
  
\begin{Itabular}{ll>{\slshape}ll}
\lsptoprule
Syllable type &  Tonal melody  & {\rm Form} & Gloss\\ [1ex] \midrule

CV  &  H & pɔ́ &plant\\
    &  L   &sɔ̀   &wash\\ [0.5ex]
\midrule

CVV   &  L & pàà   &   take\\
 & H &  kíí   &  forbid\\
 & LH &wòó &  vacant (be)\\
 & HL & gbáà &  herd\\
\midrule

CVCV    &  H  & kúló  &  tilt\\
&  L & bìlè  &  put\\
  &  HL & lúlò  &  leak\\
  &  HM & pílē &  cover\\[0.5ex]
  
  
\midrule

CVCCV   & H & bóntí  &  divide\\
 & H  & kámsɪ́&  blink\\
 &L &sùmmè  &  beg\\
 &L & zèŋsì  &  limp\\[0.5ex] 
  
\midrule

 CVVCV & H& píílí & start\\
 &H& tɪ́ásɪ́ & vomit\\
 &L &kààlɪ̀  & go\\
 &L &bùòlì  &  sing\\[0.5ex] 
 
\midrule

CVCVCV  & H  & zágálɪ́ &  shake\\ 
   & H  & vílímí  &  spin\\
    & L & hàrɪ̀gɪ̀&  try\\
   & L  &dʊ̀gʊ̀nɪ̀  &  chase\\ [0.5ex]

\lspbottomrule
\end{Itabular}   
\end{table} 
 
 [TO DO: CV are typically high while CVV are typically low, check others]
 The high (H) and low (L) register tones are assumed and two minimal pairs are 
identified:  one is {\sls télé} `lean on' and {\sls tèlè} `reach', but this 
may turn out to be a kind of  `aspectual switch' (from stative to process, or 
vice-versa), similar to the category switch discussed in Section 
\ref{sec:GRM-der-cat-switch}, and the other is {\sls pɔ̀} `protect' and  {\sls 
pɔ́} `plant'.  The mid tone (M)  is not contrastive and only describes a verb's 
intonation in the sentence frames compared.


 

\subsubsection{Verbal state and verbal process lexemes}
\label{sec:GRM-verb-stative-active}

A general distinction
between stative and non-stative events  is made: {\it verbal state} (stative
event) and {\it verbal process} (active event) 
lexemes are assumed. A verbal state lexeme can be identificational,
existential, possessive,  qualitative, quantitative, cognitive or  locative, and
refers more or less to a state or condition which is static, as opposed to
dynamic. The `copula' verbs {\sls jaa} and {\sls dʊa} (and its allolexe {\sls 
tuwo})
are treated as subtypes of verbal stative lexemes since they are the only verbal
lexemes which cannot function as a main verb in  a perfective intransitive
construction (see Section \ref{sec:GRM-verb-perf-intran}). Their meaning and
distribution was introduced in the sections concerned with the identificational
construction (Section \ref{sec:GRM-ident-cl}) and existential construction
(Section \ref{sec:GRM-loc-cl}).  The possessive verb
{\sls kpaga} `have'  is treated as  a verbal state lexeme as well (see 
possessive
clause in Section   \ref{sec:GRM-poss-cl}).  A qualitative verbal state lexeme
establishes a relation between an entity and a quality. Examples are given in
(\ref{ex:GRM-v-stat-qual}).


\ea\label{ex:GRM-v-stat-qual}{\rm Qualitative verbal state lexeme}\\

 {\sls bòró}  `short'  $>$ {\sls à dáá bóróó} {\rm  `The tree is short.'}\\
{\sls gòrò} `curved'  $>$ {\sls à dáá góróó} {\rm  `The wood is curved.'}\\
{\sls jɔ́gɔ́sɪ́} `soft'   $>$ {\sls   à bìé bàtɔ́ŋ jɔ́gɔ́sɪ̀jɔ̀ʊ̄} {\rm    `The 
baby's
skin is soft.'}
\z

Similarly, a quantitative verbal state lexeme  establishes a relation between an
entity and a quantity. Yet, in (\ref{ex:GRM-v-stat-quant}), the subject of   
{\sls maase} is the impersonal pronoun {\sls a} which refers to a situation and 
not an
individual. The verb {\sls hɪ̃ɛ̃}  `age' or `old'  is a quantitative verbal 
state
lexeme since it  measures  objective maturity between
two individuals, i.e. {\sls mɪŋ hɪ̃ɛ̃-ɪ}, {\it lit.} {\sc 1.sg.st} age-{\sc
2.sg.wk}, `I am older than you'. 


\ea\label{ex:GRM-v-stat-quant}{\rm Quantitative verbal state lexeme}\\

 {\sls kánà}  `abundant'  $>$ {\sls bà kánã́ʊ̃́} {\rm   `They are plenty 
(people).'} 
\\
{\sls mààsɪ̀} `enough'  $>$   {\sls à máásɪ́jʊ́} {\rm    `It is sufficient.'}\\
{\sls hɪ̃̀ɛ̃̀} `age' $>$ {\sls mɪ́ŋ hɪ̃́ɛ̃́ɪ̃̀} {\rm    `I am older than you'}
\z

Cognitive verbs such as {\sls liise} `think',  {\sls kʊ̃ʊ̃} `wonder, 
{\sls kisi} `wish', {\sls tʃii} `hate', etc.  are also treated as verbal 
state
lexemes. 

Verbal process lexemes denote non-stative events. They are often partitioned
along the
(lexical) aspectual distinctions of  \citet{Vend57}, i.e. activities, 
achievements, accomplishments. Such verbal categories did not formally emerge, 
so I am not in a position to categorize the verbal process lexemes at this 
point 
in the research (but see \citealt[51]{Bonv88} for a thorough description of a 
Grusi verbal system), although Section \ref{sec:GRM-verb-suffix} suggests that 
there is a system of verbal derivation  that  uses verbal process lexemes which 
needs to be uncovered.  Thus, verbs which express that the participant(s) is 
actively doing something, undergoes a process, performs an action, etc. all 
fall 
within the  set of verbal process lexemes. 



\subsubsection{Complex verb}
\label{sec:GRM-complex-verb}

A complex verb is  composed of more than one verbal lexeme. For
instance, when {\sls laa} `take' and {\sls di}
`eat' are brought together in a SVC (Section \ref{sec:GRM-multi-verb-clause}),
they denote separate taking and eating event. A complex verb denotes a single 
event.

\ea\label{ex:cpx-verb-laa-di}
\ea
 \gll ǹ̩ láá kúòsò díūū.\\
{\sc 1.sg} take G.  eat.{\foc}\\
\glt `I believe in God.'

\ex
 \gll  ǹ̩ láá bìé dʊ́ʊ̄.\\
{\sc 1.sg} take child put.{\foc}\\
\glt `I adopted a child.'
\z 
 \z
 
 The sequences  {\sls laa}+{\sls di} `believe'  and {\sls laa}+{\sls dʊ} 
`adopt'  
are  non-compositional, and less literal. Also, unlike complex stem nouns, but 
like SVCs, the elements which compose a complex verb must not necessarily be 
contiguous,  as  (\ref{ex:cpx-verb-laa-di}) shows. Other examples, among 
others, 
 are {\sls zɪ̀mà síí}, {\it lit.} know raise, `understand',  {\sls kpá 
tā}, 
{\it lit.}  take abandon, `drop' or `stop', and {\sls gɪ̀là zɪ̀mà}, {\it 
lit.} 
allow know, `prove'.



\subsubsection{Verb forms and aspectual distinction}
\label{sec:GRM-verb-word}

The inflectional system of Chakali verbs displays  few verb
forms and is closer  to neighbor Oti-Volta languages than, for instance,  a
`conservative' Grusi language like Kasem \citep[51]{Bonv88}.\footnote{Dagbani is
described as a language where the ``inflectional system  for verbs is relatively
poor''  \citep[96]{Olaw99}. It has an imperfective suffix {\sls -di}
\citep[97]{Olaw99} and  an imperative suffix {\sls -ma}/{\sls mi} 
\citep[101]{Olaw99}.
\citet[81]{Bodo97} writes that Dagaare has four verb forms: a dictionary
form, a perfective aspectual form, a perfective intransitive aspectual form and
an imperfective aspectual form. Also for Dagaare, \citet{Saan03}  talks about
four forms: perfective A and B, and Imperfective A  and B.}  Besides the
derivational suffixes (Section \ref{sec:GRM-deri-suff}), the verb in Chakali is
limited to two
inflectional suffixes and one assertive suffix:  (i) one signals negation in the
negative imperative clause (i.e.  {\sls  kpʊ́} `Kill',  {\sls tíí kpʊ̄ɪ̄} 
`Don't
kill'),  (ii) another attaches to some verb stems in the perfective intransitive
only, and (iii)  the other signals assertion and puts the verbal constituent in
focus. Since the negative imperative clause has already been presented in
Section
\ref{sec:GRM-imper-clause}, the perfective and imperfective intransitive
constructions are discussed next.  Both are recurrent clauses in data
elicitation. The former may contain both the perfective
suffix and the assertive suffix simultaneously, while the latter  displays the
 verb, with or without the assertive suffix.
 
 
 
\paragraph{Base form of a verb}
\label{sec:GRM-base-verb}

%what do you do with intransitive??

The form of the verb displayed in the dictionary is called the base form.  It  
is identified as the segmental sequence and melody which  would appear in a
positive imperfective transitive clause (Section \ref{sec:GRM-trans-intran}). 


\ea\label{ex:GRM-base-form}{\rm  Base form = Positive imperfective transitive}\\
\gll   bàà kʊ́ɔ́rɪ̀ sɪ̀ɪ̀máá rà.\\
3.{\sc pl}  make food {\sc foc}\\
\glt `They are making food.'\\\par
$\rightarrow$ {\rm kʊɔrɪ  (HL)}
 \z

 This sentence frame is one that does not affect the segmental sequence and 
melody the verb. The base form can also correspond  to a  verb elicited in 
isolation, although consultant are generally not at ease with verbs in 
isolation, unless they are framed in an utterance.

\paragraph{Perfective intransitive construction}
\label{sec:GRM-verb-perf-intran}

As its name suggests, a \is{perfective intransitive construction} perfective 
intransitive construction  lacks a 
grammatical
object and implies an event's completion or its 
reaching point.  In the case of \is{verbal state}verbal state,
the  \is{perfective}perfective  implies that the given state has been reached, 
or 
that the entity in subject position   satisfies the property encoded in
the \is{verbal state lexeme}verbal state lexeme. In 
(\ref{ex:GRM-intperfc-frame}),  two suffixes 
are
attached on  one verbal process stem and one  verbal state 
stem (see Section \ref{sec:nasalization-verb-suffix}
for the general phonotactics involved).\footnote{The presence of  a schwa
({\sls ə}) in a CVCəCV surface form, as in (\ref{ex:GRM-intperfc-frame-state}), 
is explained in Sections \ref{sec:epenthesis} and \ref{sec:PHO-weak-syll}.}


\ea\label{ex:GRM-intperfc-frame}{\rm Perfective intransitive construction}\\


\ea\label{ex:GRM-intperfc-frame-process}{{\it  Verbal process:} {\sc s}  $+$
{\sc p} }\\
\gll àfɪ̀á díōō.\\
A. {di-j[{\sc -lo, -hi, -ro}]-[{\sc +hi,+ro}]}\\

\glt `Afia ate.'

\ex  àfɪ̀á wá díjē   {\rm `Afia didn't eat.'}

\ex\label{ex:GRM-intperfc-frame-state}{{\it  Verbal state:} {\sc s}  $+$ {\sc p}
}\\
\gll à dáá télèjōó.\\
{\art} daa  {tele-j[{\sc -lo, -hi, -ro}]-[{\sc +hi,+ro}]}\\
\glt `The stick leans'

\ex à dáá wá tēlə̀jē  {\rm `The stick doesn't lean.'} %check

\z 
 \z

The first suffix to attach is the perfective suffix, i.e. -j[{\sc -lo, -hi, 
-ro}] or simply /jE/. Although it appears on every (positive and negative) stem 
in (\ref{ex:GRM-intperfc-frame}),  it does not surface on all verb stems. The 
information in Table \ref{tab:GRM-perf-suff} partly predicts whether or not a 
stem will surface with a suffix, and if it does, which form this suffix will 
have.


\begin{table}[htb]
 \centering
\caption{Perfective intransitive suffixes
\label{tab:GRM-perf-suff}}
\begin{Itabular}{p{2cm}p{2cm}p{2cm}}
\lsptoprule
Suffix /-jE/ & Suffix /-wA/ & No suffix\\[1ex]
\midrule

CV &  CVV & CVCV\textsuperscript{1}\\
 CVCV\textsuperscript{2} & &\\ 

 \lspbottomrule
\end{Itabular}
\end{table} 

Table \ref{tab:GRM-perf-suff} shows that, in a perfective intransitive
construction, a CV stem must
be suffixed with {\sls -jE} and  a CVV verb with {\sls -wA}. The examples in
(\ref{ex:GRM-jE-wA}) are negative in order to prevent the assertive
suffix from appearing (see Section \ref{sec:GRM-focus} on why negation and the
assertive suffix cannot co-occur).


\ea\label{ex:GRM-jE-wA}


\ea{\it CV}\\
po   $>$  àfíá wá pójē  {\rm `Afia didn't divide'}\\
pɔ  $>$ àfíá wá   pɔ́jɛ̄   {\rm `Afia didn't  plant'}\\
pu  $>$ àfíá wá  pújē   {\rm `Afia didn't  cover'}\\
pʊ  $>$ àfíá wá  pʊ́jɛ̄   {\rm `Afia didn't  spit'}\\
kpe  $>$ àfíá wá  kpéjē   {\rm `Afia didn't  crack and
remove'}\\
kpa  $>$ àfíá wá  kpájɛ̄   {\rm `Afia didn't  take'  }

\ex{\it CVV}\\
tuu $>$ àfíá wá  tūūwō   {\rm `Afia didn't  go down'}\\
tie $>$  àfíá wá   tīēwō {\rm `Afia didn't chew'}\\
sii  $>$  àfíá wá  sīīwō   {\rm  `Afia didn't  raise'}\\
jʊʊ   $>$  àfíá wá  jʊ̄ʊ̄wā  {\rm  `Afia didn't  marry'}\\
tɪɛ $>$  àfíá wá tɪ̄ɛ̄wā  {\rm  `Afia didn't  give'}\\
wɪɪ $>$  àfíá wá  wɪ̄ɪ̄wā  {\rm  `Afia is not  ill'}
 

\z 
 \z

The surface form of the perfective suffix which attaches to CV stems  is 
predicted by the {\sc atr}-harmony rule of Section
\ref{sec:vowel-harmony}. Notice that  {\sc ro}-harmony does not operate
in that domain. 

\begin{Rule}\label{PHO-rule-perf-wa}{Prediction  for perfective intransitive 
-/wA/ suffix}\\
If the vowel of a CVV stem is
{\sc +atr},
the vowel of the suffix is {\sc +ro}, and if the vowel of a CVV stem is {\sc
 --atr}, the vowel of the suffix is {\sc -ro}.\\
-/wA/ $>$  $\alpha${\sc ro}$_{suffix}$  /  $\alpha${\sc atr}$_{stem}$   
\end{Rule}

The CVV stems display  harmony between the stem
vowel(s) and the suffix vowel which is easily captured by a variable feature
alpha notation, as shown in Rule (\ref{PHO-rule-perf-wa}), which  assumes that 
the segment [{\sls o}] is the
[{\sc +ro, +atr}]-counterpart of [{\sls a}]. 




Predicting  which of 
set CVCV\textsuperscript{1} or set CVCV\textsuperscript{2} in Table 
\ref{tab:GRM-perf-suff}  a
stem falls  has proven unsuccessful. Provisionally,  I suggest that a CVCV
stem must be stored with such an information. One piece of evidence
supporting this claim comes from
the minimal pair {\sls tèlè} `reach' and  {\sls télé} `lean against':  the
former displays CVCV\textsuperscript{2} (i.e. tele-jE),  whereas the latter 
displays CVCV\textsuperscript{1}
(i.e. tele-\O).  The data shows that a  CVCV stem with round vowels is less 
likely to
behave like a CVCV\textsuperscript{1} stem, yet {\sls púmó} `hatch' is a 
counter-example, i.e.
{\sls a zal wa puməje} `the fowl didn't hatch'. The CVCCV, CVVCV, and CVCVCV 
stems
have  not been investigated, but {\sls kaalɪ} `go', a common  CVVCV verb, takes
the
/-jE/ suffix.  


\paragraph{Imperfective intransitive construction}
\label{sec:GRM-verb-perf-intran}

The imperfective  conveys the unfolding of an event, and it is often used to
describe an event taking place at the moment of speech. In addition, the
behavior of the egressive marker {\sls ka} (Section \ref{sec:GRM-EVC-egr-ingr})
suggest that the imperfective may be interpreted as a progressive event.  The 
imperfective is indicated by the base
form of a verb. As in the perfective intransitive, the assertive suffix may be
found attached to the verb stem. 


\ea\label{ex:GRM-assert-suff}
[[{\it verb stem}]-[{\sc +hi,+ro}]]$_\textrm{verb in focus}$
\z

Again, the constraints licensing the combination of the verb stem and the vowel
features  shown in (\ref{ex:GRM-assert-suff})   are (i) none of the other
constituents in the clause are in focus, (ii) the clause does not include
a negation element, and (iii) the clause is intransitive, that is, there is
no grammatical object. 

%[to do: note p. 199, + non-pronominal subject)

% , as opposed to an
% event perceived as bounded (i.e. perfective) or a hypothetical event (i.e.
% imperative)



\ea\label{ex:GRM-pos-neg-take}
\ea\label{ex:GRM-ipfv-out-pos}{\rm Positive}\\
 ʊ̀ kàá kpá  {\rm `She will take'}\\
   ʊ̀ʊ̀ kpáʊ̄ {\rm `She  is taking/takes'}

\ex\label{ex:GRM-ipfv-out-neg}{\rm Negative}\\
 ʊ̀ wàá kpā  {\rm  `She will not take'}\\
   ʊ̀ʊ̀   wàà   kpá {\rm `She  is not taking/does not take'}

   
   \ex\label{ex:GRM-ipfv-out-nfoc}
 \textasteriskcentered kalaa kpaʊ {\rm Kala is taking/takes'}\\
   \ex\label{ex:GRM-ipfv-out-stpro}
 \textasteriskcentered waa kpaʊ {\rm `SHE is taking/takes'}\\
    \ex\label{ex:GRM-ipfv-out-obj}
  \textasteriskcentered ʊ kpaʊ a bɪɪ  {\rm `She  is taking/takes the 
stone'}\\
 \ex\label{ex:GRM-ipfv-out-neg}
   \textasteriskcentered   ʊʊ   waa   kpaʊ {\rm `She  is not taking/does 
not take'}

\z 
 \z

In (\ref{ex:GRM-pos-neg-take}), the forms of the verb in the
intransitive imperfective take the assertive suffix to signal that the verbal
constituent is in focus, as opposed to the nominal argument. The 
assertive suffix cannot appear
when the subject is in focus (\ref{ex:GRM-ipfv-out-nfoc}) or when the strong
pronoun is used as subject (\ref{ex:GRM-ipfv-out-stpro}), when a grammatical
object follows the verb  (\ref{ex:GRM-ipfv-out-obj}), or when the negation
preverb {\sls waa} is present  (\ref{ex:GRM-ipfv-out-neg}).



\paragraph{Intransitive vs. transitive}
\label{sec:GRM-trans-intran}


Many verbs can occur in either  intransitive or transitive clauses. The subject 
of the intransitve  in (\ref{ex:GRM-clause-core-intrans}) and 
(\ref{ex:vp26.14.}) correspond to the subject of the transitive  in  
(\ref{ex:GRM-clause-core-trans}) and (\ref{ex:vp26.15.}), and the same verb is 
found with and without an object.


\begin{multicols}{2}
\ea\label{ex:GRM-clause-core}


 \ea\label{ex:GRM-clause-core-intrans}
\gll kàlá díjōō.\\
 Kala eat.{\sc pfv.foc}\\
\glt  `Kala ATE.' 
\ex\label{ex:GRM-clause-core-trans}
\gll kàlá dí sɪ̀ɪ̀máá rā.\\
Kala eat.{\sc pfv}  food {\sc foc}\\
\glt  `Kala ate FOOD' 



\ex\label{ex:vp26.14.}
\gll ʊ̀ʊ̀ búólùū.\\
    {\psg} sing.{\ipfv.\foc}\\
\glt  `He is SINGING.' 

\ex\label{ex:vp26.15.}
\gll  ʊ̀ʊ̀ búólù būōl lō.\\
    {\psg}  sing.{\ipfv} song {\foc}\\
\glt  `He is singing a SONG.' 

\z
 \z
 \end{multicols}


It is possible to promote a prototypical theme argument to the subject position.
However,  informants have difficulty with some nominals in the subject
position of
intransitive clauses.   The topic needs further investigation, although it is
certainly related to a semantic anomaly.  The data in
(\ref{ex:GRM-intran-theme-subj}), where the  prototypical {\sc o}(bject) is in
 {\sc a}-position, illustrates the problem. In order to concentrate on the 
 activities of  `goat
beating'- and `tree climbing'  and turn the two clauses
(\ref{ex:GRM-int-th-su-out-1}) and (\ref{ex:GRM-int-th-su-out-2}) into
acceptable utterances,  the optimal solution is to use the
impersonal pronoun {\sls ba} in subject position  (see impersonal
pronoun in
Section \ref{sec:GRM-impers-pro}).



\ea\label{ex:GRM-intran-theme-subj}

\ea
à bʊ̀ɔ̀ káá hírèū  {\rm `the hole is being dug'}
\ex\label{ex:GRM-int-th-su-out-1}
\textasteriskcentered a bʊ̃ʊ̃ŋ   kaa maŋãʊ̃  {\rm  `the goat is being beaten' }
$\rightarrow$ {\sls 
bàà máŋà à bʊ̃́ʊ̃́ŋ ná}
\ex\label{ex:GRM-int-th-su-out-2}
\textasteriskcentered a daa kaa zɪnãʊ̃  {\rm   `the tree is being climbed'}   
$\rightarrow$ {\sls 
bàà zɪ́ná à dáá rá}


\z 
 \z

%[to do: include image spectrogram]

Given that  the inflectional system of the verb is rather poor, and that the 
perfective
and assertive suffixes occur only in intransitive clauses,  how does one
encode a basic contrast like the one between a transitive perfective and
transitive imperfective? The paired examples in (\ref{ex:tra-pfv}) and
(\ref{ex:tra-ipfv})  illustrate 
 relevant contrasts.\nolinebreak 
 
  %[TO DO change tʃige sentence] 


\begin{multicols}{2}
\ea\label{ex:tra-pfv}{\rm Transitive perfective}\\

  \ea\label{ex:tra-pfv-eat}
ǹ̩ dí kʊ̄ʊ̄ rā\\
 `I ate T. Z..' 
 \ex\label{ex:tra-pfv-plant}
ǹ̩ pɔ́ dāā rā\\
`I planted a TREE.'
 \ex\label{ex:tra-pfv-cover}
ǹ̩ tʃígé vìì rē\\
`I turned a POT.' 
 \ex\label{ex:tra-pfv-tie}
ǹ̩ lómó bʊ̃́ʊ̃́ŋ ná\\
`I tied a GOAT.' 
 \ex\label{ex:tra-pfv-carry}
m̩̀ mɔ́ná díŋ né\\
`I carried  FIRE.' 

\z 
 \z

\ea\label{ex:tra-ipfv}{\rm Transitive imperfective}\\

 \ea\label{ex:tra-ipfv-eat}
ǹ̩ǹ̩ dí kʊ́ʊ́ rá\\
`I am eating T. Z..' 
 \ex\label{ex:tra-ipfv-plant}
m̩̀m̩̀ pɔ́ dáá rá\\
`I am planting a TREE.' 
 \ex\label{ex:tra-ipfv-cover}
ǹ̩ǹ̩ tʃígè vìì rē\\
`I am turning  a POT.' 

 \ex\label{ex:tra-ipfv-tie}
ǹ̩ǹ̩ lómò bʊ̃̄ʊ̃̄ŋ nā\\
`I am tying  a GOAT.' 
 \ex\label{ex:tra-ipfv-carry}
m̩̀m̩̀ mɔ́nà dīŋ nē\\
`I am carrying  FIRE.' 

\z 
 \z
 
\end{multicols}


Each pair in the verbal frames of  (\ref{ex:tra-pfv}) and (\ref{ex:tra-ipfv}) 
presents fairly regular patterns:  the high tone {\it versus} the falling tone 
on the CVCV verbs is one instance, the systematic change of the tonal melodies 
on the grammatical objects in the two CV-verb cases, and the length of the weak  
pronoun in the imperfective are identified. The data suggest that it is the 
tonal melody, and not exclusively the one associated with the verb, which 
supports aspectual function in this comparison. When the verb is followed by an 
argument, both perfective and the imperfective are expressed with the base form 
of the verb.  However,  the tonal melody alone  can determine whether a 
phonological string is to be understood as a bounded event which occurred in 
the 
past or an unbounded event unfolding at the moment of speech.



Tonal melody is crucial in the following examples as well. The examples in
(\ref{GRM-pfv-inter}) are three polar questions (see Section
\ref{sec:GRM-interr-polar}), one perfective  and two
imperfective. The two first have the
same segmental content, and the last contains the egressive preverb {\sls kaa}
with a rising tone indicating the future tense.  In order to signal a polar
question, each has  an extra-low tone and is slightly lengthened at the end of
the utterance. 

\ea\label{GRM-pfv-inter}


\ea\label{GRM-pfv-inter-pfv}
\glll {\T  } {\T    } {\T   } {\T    } {\T   }.\\
 ɪ   teŋesi  a  namɪ̃ã  raa\\
    {\sc 2.sg} {cut.{\sc pv}} {\sc art} {meat} {\sc foc}\\
\glt `Did you cut the meat (into pieces)?'\\



\ex\label{GRM-pfv-inter-impf}

\glll {\T   } {\T     } {\T } {\T    } {\T   }.\\
ɪ   teŋesi  a  namɪ̃ã  raa\\
    {\sc 2.sg} {cut.{\sc pv}} {\sc art} {meat} {\sc foc}\\
\glt `Are you cutting the meat (into pieces)?'\\


\ex\label{GRM-pfv-inter-impf-fut}

\glll {\T  } {\T    } {\T     } {\T } {\T    } {\T   }.\\
ɪ  kaa teŋesi  a  namɪ̃ã  raa\\
    {\sc 2.sg} {\sc ipfv.fut} {cut.{\sc pv}} {\sc art} {meat} {\sc foc}\\
\glt  `Will you (be) cut(ting) the meat (into pieces)?'\\
 
\z 
 \z

The only distinction perceived between (\ref{GRM-pfv-inter-pfv})  and
(\ref{GRM-pfv-inter-impf}) is a pitch difference on the third syllable of the
verb. The tonal melody associated with the verb in 
(\ref{GRM-pfv-inter-impf-fut}) is the
same as the one in (\ref{GRM-pfv-inter-impf}).



\paragraph{Ex-situ subject imperfective particle}
\label{sec:GRM-ipfv-part}

One topic-marking strategy is to prepose a non-subject constituent to the
beginning of the clause.  In  (\ref{ex:GRM-foc-top}),  the focus particle may or
may not
appear after the non-subjectival topic. Notice that one effect of this 
topic-marking strategy is that the particle {\sls dɪ} appears between the 
subject
and
the verb when the non-subject constituent is preposed and when the clause is
used to describe what is happening at the moment of speech. \nolinebreak

\begin{multicols}{2}
\ea\label{ex:GRM-foc-top}
 \ea\label{ex:GRM-foc-top-chew-prog-1}{\rm Imperfective}\\
\gll  sɪ́gá (rá)  ʊ̀ dɪ̀  tíē.\\
 bean  ({\foc}) {3.\sg} {\ipfv} chew\\
\glt `It is BEANS he is chewing'\\


 \ex\label{ex:GRM-foc-top-chew-perf-1}{\rm Perfective}\\
\gll  sɪ́gá (rá) ʊ̀   tìè.\\
 bean  ({\foc}) {3.\sg}  chew\\
\glt `It is BEANS he chewed'\\


\columnbreak


 \ex\label{ex:GRM-foc-top-go-prog-2}{\rm Imperfective}\\
\gll   wàà (rá) ʊ̀ dɪ̀  káálɪ̀.\\
Wa ({\foc}) {3.\sg} {\ipfv} go\\
\glt `It is to WA that he is going'\\


 \ex\label{ex:GRM-foc-top-go-perf-2}{\rm Perfective}\\
\gll   wàà (rá)  ʊ̀ kààlɪ̀.\\
Wa   ({\foc}) {3.\sg}  go\\
\glt `It is to WA that he went'\\
\z 
 \z
\end{multicols}

The position of {\sls dɪ} in  
(\ref{ex:GRM-foc-top-chew-prog-1}) and 
(\ref{ex:GRM-foc-top-go-prog-2}), that is between the subject and the verb, is 
generally occupied by linguistic items called  {\it preverbs},  to which the 
discussion turns in Section \ref{sec:GRM-precerv}.  Provisionally, the particle 
{\sls dɪ} may be treated as a preverb constrained to occur with  a preposed  
 non-subject constituent and an imperfective aspect.\footnote{I do not treat 
topicalization in this work, although the left-dislocation strategy in 
(\ref{ex:GRM-foc-top}) is the only one I know to exist.}


\paragraph{Subjunctive}
\label{sec:GRM-subjunctive}

In Section \ref{sec:GRM-desiderative mood} the preverbal {\sls ŋma}  is said to  convey a 
desiderative mood,  corresponding to the English modal expression `want to',  in a construction [NP 
{\sls ŋma} [NP VP]].  The embedded clause is said to be in the subjunctive mood, which is singled 
out by its high tone on the subject pronoun and  the non-actuality and potentiality of the event 
ihe clause.  In the examples (\ref{ex:GRM-subj-market}) and (\ref{ex:GRM-subj-mother}) a  
subjunctive is interpreted because it involves clauses expressing a future hypothetical time and 
realization. In all these cases, the clauses of which the high tone pronoun is subject seem to 
depend on and complement a more central event.

\ea
\ea\label{ex:GRM-subj-market}
\gll   ʊ̀ káálɪ̀ ʊ́ ká dí mɔ̀tìgú jáwà.\\
 {3.\sg} go  {3.\sg} {\sc ipfv} eat M. market \\
\glt `She is going to trade at the Motigu market.' ({\it lit.} eat-market,  `trade')\\

 
\ex\label{ex:GRM-subj-mother}
\gll zʊ̀ʊ̀ m̀m̀ mã̀ã̀ pé, ʊ́ kpá ǹǹ kʊ̀ʊ̀ fàlá tɪ́ɛ́ɪ́, ɪ́ kíínì.\\
 enter   {1.\sg} mother end  {3.\sg} take    {1.\sg\poss} t.z. bowl give.{\sc 2.sg} {\sc 
2.sg} clean.bowl\\
\glt `Go to my mother, she will give you my t.z. bowl so you can finish it.'\\

 \z
 \z
 
 In  (\ref{ex:GRM-subj-market}), according to the speaker, the trading activity is the intention of 
the woman and it will take place in all likelihood, and 
 in  (\ref{ex:GRM-subj-mother}),  the speaker tells about two situations that the 
addressee  will 
probably experience. 


% ŋ ŋma ŋ kaalɪ dusee tʃɪa
% ŋ ŋma ŋ tʃɪ  kaalɪ dusee
% ŋ tʃɪ kaalɪ dusee

\subsection{Preverb particles}
\label{sec:GRM-precerv}

Preverb particles  encode various event-related
meanings. They are part of the verbal domain  called the expanded verbal group
(EVG), discussed in (\ref{sec:GRM-verbals}) and schematized in  
(\ref{ex:verb-VP}). This domain  follows the 
subject and precedes the main verb(s) and is generally accessible  only to a 
limited set of linguistic items. These grammatical morphemes are
not verbs, in the sense that they do not contribute to SVCs as verbs do,  but as
`auxiliaries'. Still,  some of the preverbs may historically derive from verbs, 
and  some others may synchronically function as verbs.  Examples of the latter
are the egressive particle {\sls ka} and ingressive particle {\sls wa},  which 
are
discussed in Section \ref{sec:GRM-EVC-egr-ingr}. 

Nevertheless, given the data available,  it would not be incorrect to analyze 
some of the preverbs  as additional SVC verbs.  However, we will see  that a 
preverb differs from a verb in that it exposes functional categories,  cannot 
inflect for the perfective or assertive suffix,  and never takes  a complement, 
such as a grammatical object, or cannot be modified by  an adjunct. But again,  
a first verb in a SVC and a preverb are categories which can be hard to 
distinguish. Structurally and functionally, many of them may be analyzed as 
grammaticalized verbs in series. These characteristics are not special to 
Chakali; similar, but not identical, behavior are described for Gã and Gurene 
\citep{Daku07b, Daku08}.



\subsubsection{Egressive and ingressive particles}
\label{sec:GRM-EVC-egr-ingr}


The egressive particle {\sls ka(a)} ({\it gl.} {\sc egr})   `movement away from
the
deictic centre'  and   the ingressive
particle {\sls wa(a)} ({\it gl.} {\sc ingr})  `movement towards
the deictic centre' are  assumed to derive from the  verbs
{\sls kaalɪ} `go' and  {\sls waa} `come'.\footnote{A discussion on some aspects 
of
grammaticalization of  `come' and `go' can be read in  \citet{Bour92}. In the
literature, egressive  is also known as  {\it itive} (i.e. away from the
speakers,  `thither')  and  ingressive  is  known as {\it ventive} (i.e. towards
the speakers,   `hither'). }  Table
\ref{tab:deict-pre-verb} shows that  {\sls kaalɪ} `go' and {\sls waa} `come',  
like 
other verbs, change forms (and are acceptable) in these paradigms,  but {\sls 
ka(a)}  is  not.


\begin{table}[h]
\centering
\caption{Deictic verbs and preverbs \label{tab:deict-pre-verb}}

\begin{Itabular}{lllll}
\lsptoprule
Verb & $\sigma$  & Aspect & Positive & Negative\\[1ex] \midrule


{\sls waa} `come' & CVV   & {\sc pfv}   &  ʊ̀ wááwáʊ́   & ʊ̀ wà wááwá\\
    &&& `she came' & `she didn't come'\\

    &   & {\sc ipfv}   &  ʊ̀ʊ̀ wááʊ̄  & ʊ̀ wà wáá\\
    &&& `she is coming' & `she is not
coming'\\[1ex] \midrule




{\sls kaalɪ} `go' & CVVCV   & {\sc pfv}   &  ʊ̀ káálɪ́jʊ́   & ʊ̀ wà 
káálɪ́jɛ́\\
    &&& `she went' & `she didn't go'\\

    &   & {\sc ipfv}   &  ʊ̀ʊ̀ káálʊ̄ʊ̄  & ʊ̀ wà káálɪ́\\
    &&& `she is going' & `she is not
going'\\[1ex] \midrule


{\sls ka}  & CV   & {\sc pfv}   &  *ʊ kaʊ   & *ʊ wa kajɛ\\
  

    &   & {\sc ipfv}   &  *ʊ kaʊ  & *ʊ wa ka\\
  
\lspbottomrule


\end{Itabular}   
\end{table}


If the verbs {\sls kaalɪ} `go' and  {\sls waa} `come'
occur in a SVC,  they surface as {\sls ka} and {\sls wa} respectively. In
(\ref{GRM-prev}),  both {\sls ka} and {\sls wa} take part in  a two-verb SVC in which they are
first in the sequence.


\ea\label{GRM-prev}

\ea\label{GRM-prev-SVC-ka}
\glll gbɪ̃̀ã́     bààŋ   té   kà     sáŋá  
 à   
píé  {(...)}\\
monkey  quickly   early   go  sit {\art}
yam.mound.{\sc pl}   {(...)}\\
{} [[{\it pv} {\it pv}]$_{EVG}$  {\it v} {\it v}]$_{VP}$ {} {}
{(...)}\\
\glt `Monkey quickly went and sat on the (eighth) yam mounds (...)'  (LB 012)

\ex\label{GRM-prev-SVC-wa}
\glll ŋmɛ́ŋtɛ́l   làà nʊ̀ã̀  nɪ́ ká  ŋmá dɪ́ ʊ́ʊ́  wá  
ɲʊ̃̀ã̀ nɪ́ɪ́\\
spider collect mouth {\postp}  {\conn} say {\comp}  
{\sc 3.sg}  come   drink water\\
 {} {} {} {} {}  {} {} {} {\it v} {\it v} {}\\
\glt  `(Monkey went to spider's farm to greet him.)  Spider accepted
(the
greetings) and (Spider) asked him (Monkey) to come and drink water.'  (LB 011)

 
\z 
 \z


Because they derive from deictic verbs (historically or synchronically),  the
preverbs have the potential to indicate non-spatial  `event movement'  to or 
from a deictic centre.
This phenomenon is not uncommon cross-linguistically. \citet[62]{Nico07}
maintains  that when a movement verb becomes a tense marker, it may be reduced
to a verbal affix and its meaning can develop ``into meaning relating temporal
relations between events and reference times''. In Chakali, the  preverb {\sls 
kaa} contributes   temporal information to an expression. Consider in
(\ref{exe:GRM-crack-remove-attach}) the distribution and contribution of  {\sls 
kaa} to  the clauses headed by the verbs {\sls kpe} `crack a shell and remove a
seed from it' (henceforth `c\&r') and {\sls mara} `attach'.\footnote{In Gurene 
(Western Oti-Volta), it
is the ingressive particle which has a similar role. The ingressive  is 
commonly used before the verb, and can, among other things,  express future
tense \citep[see][59]{Daku07b}.}



\ea\label{exe:GRM-crack-remove-attach}
%\begin{multicols}{2}

\ea

 ʊ̀ kàá kpē  {\rm  `She will c\&r'}\\
   ʊ̀ʊ̀ kpéū  {\rm  `She  is c-\&r-ing/c-s\&r-s'}\\
   ʊ̀ kpéjòō   {\rm `She   c-\&r-ed'}\\
   kpé    {\rm  `C\&r!'}
\ex
 ʊ̀ kàá mārā   {\rm   `She will attach'}\\
   ʊ̀ʊ̀ máráʊ̄  {\rm  `She  is attaching/attaches'}\\
  ʊ̀ márɪ̀jʊ̄ {\rm `She   attached'}\\
   márá  {\rm  `Attach'}

%\end{multicols}
\z
\z

When the preverb particle {\sls kaa} is uttered with a rising pitch it situates 
the event in the future. The preverb particle {\sls kaa} can also be used to 
express that an event is
ongoing at the moment of speech, which I call the present 
progressive.   However,  when it is used to describe what is happening
now, {\sls kaa} can only appear when the subject is not a pronoun and its tone
melody differs from that of the future tense. These contrasts are given in
(\ref{exe:GRM-kaa-attach}).

\ea\label{exe:GRM-kaa-attach}
 ʊ̀ kàá mārā   {\rm `She will attach'}\\
   ʊ̀ʊ̀ máráʊ̄   {\rm  `She  is attaching'}\\
wʊ̀sá kàá mārā   {\rm  `Wusa will attach'}\\
wʊ̀sá káá   máráʊ̄  {\rm  `Wusa is attaching'}\\
\textasteriskcentered  wʊ̀sá   máráʊ̄   {\rm `Wusa is
attaching'}
\z

The paradigm in  (\ref{exe:GRM-kaa-attach}) shows that when the preverb 
particle {\sls kaa} appears with a rising tonal melody it  expresses the future 
tense, but  in
order to convey that a situation is ongoing at the time of speech (i.e. present
progressive), the preverb particle {\sls kaa} has a high tone. Thus, it is the
tonal melody on {\sls kaa} which distinguishes between the future and the 
present
progressive (both treated as imperfective),  plus the fact that pronouns cannot
co-occur with the preverb particle {\sls kaa} in the present progressive. 


In (\ref{x:GRM-tone-ipfv-L}) {\sls kaa}'s melody is shown to be affected by   
the 
pitch  of   the  preceding  noun {\sls bie} (LH) `child' and the demonstrative 
{\sls haŋ} (HL) `this'. 


\ea\label{x:GRM-tone-ipfv}
\ea\label{x:GRM-tone-ipfv-H}
\gll à bìé káá bīlīgī ʊ̀ʊ̀ nàál kɪ̀nkán nà.\\
{\art} child {\sc ipfv} touch {\sc poss.3.sg} grand.father {\sc quant} {\sc 
foc}\\
\glt `The child touches his grand-father.'

\ex\label{x:GRM-tone-ipfv-L}
\gll à bìè háŋ̀ kàà bīlīgī ʊ̀ʊ̀ nàál kɪ̀nkán nà.\\
{\art} child {\sc dem} {\sc ipfv} touch {\sc poss.3.sg} grand.father {\sc 
quant}.many  {\sc 
foc}\\
\glt `This child touches his grand-father.'

\z
\z



Although little evidence is available, the preverb {\sls wa} may also be used to
express a sort of hypothetical  mood.  In  (\ref{ex:GRM-prev-wa-hypo}), the
preverb {\sls wa} should be seen as contributing a supposition, or a 
hypothetical
circumstance where
someone would be found calling the number 8. 
%[to do: check both wa and waa]

\ea\label{ex:GRM-prev-wa-hypo}

\gll ŋmɛ́ŋtɛ́l   ŋmā dɪ̄, kɔ̀sánáɔ̃̀,   tɔ́ʊ́tɪ̄ɪ̄nā  ŋmá 
dɪ́, námùŋ   wá jɪ̀rà ŋmɛ́ŋtɛ́l sɔ́ŋ, bá  kpáɣʊ́ʊ̄ wàà 
bá kpʊ́\\
spider  say   {\comp}   buffalo land.owner say {\comp}  anyone   
{\ingr}   call eight  name {\sc 3.pl.hum+} catch.{\sc 3.sg} {\foc}
 {\sc 3.pl.hum+}  kill\\
\glt `Spider told Buffalo that landowner said anyone who calls the number 8
should be brought to him to be killed.' (LB 009)
\z



Finally, the example in (\ref{ex:GRM-verb-ta})  intends to show that some
elders of Ducie and
Gurumbele use  {\sls ta}  instead of {\sls ka(a)},  as a variant of the
preverb.\footnote{I gathered that  (i)   {\sls ta} is not a different
preverb (Gurene is said to have  a preverb {\sls ta}  signifying intentional
action  (M. E. K. Dakubu, p.\ c.),  and  (ii)   {\sls ta} can be heard in  
Ducie 
and
Gurumbele from people of the oldest generation, but somebody suggested to me
that {\sls ta} is the common form in Motigu (Mba Zien, p.\ c.).  The distinction
is  in need of further research. } 
 

\begin{exe}
   \ex\label{ex:GRM-verb-ta}{\rm Priest talking to the shrine, holding a kola
nut above it}

\gll  má láá kàpʊ́sɪ́ɛ̀ háŋ̀ ká jà mɔ́sɛ́ tɪ̀ɛ̀ wɪ́ɪ́ tɪ̀ŋ bà 
tà/kàà búúrè\\
{\sc 2.pl} take kola.nut {\sc dem} {\sc conn} {\sc 1.pl} plead give matter {\sc
art} {\sc 3.pl.}b {\sc  egr} want\\
\glt   `Take this kola nut, we implore  you to give them what they desire.'

\z


Unfortunately, since the relation between tense, aspect, and tonal melody is not
well-understood at this stage of research, the  egressive {\sls ka}   and the
ingressive  {\sls wa} are  broadly glossed as {\egr} and {\ingr} respectively, 
but
can also be associated with composite glosses such as {\ipfv .\fut} or  {\ipfv
.\pres}  in cases where a distinction is clear.




\subsubsection{Negation preverb}
\label{sec:GRM-verb-neg}

%check negative concord with nobodu, no one, all, nothing

%  Their
% lengths may vary depending on the speech rate, but  they are always long
% in 

There are three different particles of negation in the language:  the forms 
{\sls lɛɪ} and {\sls tɪ}   were discussed in Sections 
\ref{sec:GRM-imper-clause} 
and  \ref{sec:GRM-foc-neg}  respectively.  The negative preverb particle {\sls 
wa(a)} precedes the verb and is used in the verbal group (in non-imperative
mood). The same form is found in both  main and dependent clauses. 


\ea\label{ex:GRM-neg-pres-fut}

\ea
\gll ʊ̀  wàá pɛ̀.\\
   {\sc 3.sg}  {\neg} add\\
\glt  `She will not add.'

 \ex 
\gll  ʊ̀ʊ̀ wàà pɛ́.\\
  {\sc 3.sg} {\neg} add\\
\glt  `She is not adding.'


 \ex 
\gll  ʊ̀ wà pɛ́jɛ̄.\\
  {\sc 3.sg} {\neg} add\\
\glt  `She didn't  add.'

\z 
 \z

 The examples in (\ref{ex:GRM-neg-pres-fut}) show that a tonal quality on the 
negation particle and following verb  distinguishes between the present 
progressive and  the future,  as the preverb {\sls kaa} does (see example 
\ref{exe:GRM-kaa-attach}). The length of the negation particle can also 
function as a cue.

\ea\label{ex:neg-quant-any}
\ea\label{ex:neg-quant-any-1}
\gll námùŋ wà ná-ŋ̀.\\
 {\clf}.all {\neg} see-{1.\sg}\\
\glt  `Nobody saw me.' ({\it lit.} everyone not see me) 

\ex\label{ex:neg-quant-any-2}
\gll  ǹ wà ná námùŋ.\\
  {1.\sg}  {\neg}   see  {\clf}.all\\
\glt  `I did not see anyone.' ({\it lit.} I not see everyone) 

\z 
 \z

 Example (\ref{ex:neg-quant-any}) shows that when the negation particle {\sls 
wa(a)} and a quantifier appear in the same clause the quantifier is  in the 
positive. 




\ea\label{ex:GRM-neg-come}
 
  
\ea\label{}
\gll ʊ̀ wà wá dī.\\
{\sc 3.sg} {\neg} come eat\\
\glt `She did not come to eat.'
\ex\label{}
\gll ʊ̀ wàá wà dí.\\
{\sc 3.sg} {\neg} come eat\\
\glt `She will not come to eat.'

\z 
 \z
 
 The negative preverb  always precedes the verb {\sls waa} `come'. Although 
length (CV or CVV) is  hard to differentiate in natural speech, the examples in 
(\ref{ex:GRM-neg-come}) suggest that the tonal melody and length  establish 
meaning differences.

Assertion and negation seem to avoid one another and constrain the grammar  in
the following way:  {\it If a clause is negated,  none of its constituents can
be in focus.} In Section \ref{sec:GRM-personal-pronouns},  it was shown that (i)
negation cannot co-occur with the strong pronouns, and (ii) negation cannot
co-occur with an argument of the predicate in focus, i.e. with {\sls ra} or one 
of
its variants having scope over the noun phrase. The third non-occurrence of
negation concerns  the assertive form of the verb (Section
\ref{sec:GRM-focus}).  Consider the forms of the verb {\sls mara} `attach' in 
the
two paradigms in (\ref{ex:GRM-verb-neg-foc}).\footnote{The particle {\sls la} in
Dagaare has similar constraints. \citet[94]{Bodo97} calls it a {\it factitve}
particle.}





   \ea\label{ex:GRM-verb-neg-foc}

   \ea\label{ex:GRM-verb-neg-foc-pos}{\rm Positive}\\

 ʊ̀ kàá mārā   {\rm  `She will attach'}\\
   ʊ̀ʊ̀ máráʊ̄  {\rm  `She  is attaching/attaches'}\\
  ʊ̀ márɪ̀jʊ̄ {\rm `She   attached'}\\

  \ex\label{ex:GRM-verb-neg-foc-neg}{\rm Negative}\\

 ʊ̀ wàá mārā  {\rm `She will not attach'}\\
   ʊ̀ʊ̀ wàà márá {\rm  `She  is  not attaching/does not attach'}\\
  ʊ̀ wà márɪ̀jɛ̄   {\rm  `She   did not attach'}\\


\z 
 \z

The paradigms in (\ref{ex:GRM-verb-neg-foc})  suggest that the negation 
particle 
and the assertive suffix are in complementary distribution. 

\subsubsection{Tense, aspect, and mood preverbs}
\label{sec:GRM-verb-neg}


\paragraph{fɪ}

The preverb {\sls fɪ}   is identified with two different but interrelated 
meanings.  First, as (\ref{ex-preverb-fi-neut}) shows, the preverb {\sls fɪ}  
({\it gl.} {\sc pst}) is a neutral past tense particle (i.e.  as opposed to the 
specific  {\sls dɪ} of  Section \ref{sec:GRM-preverb-three-int-tense}), and the 
event referred to in the past can no longer be in effect in the present.


\ea\label{ex-preverb-fi-neut}

\ea
\gll ʊ̀ jáá  ǹ̩ǹ̩ tʃítʃà rā.\\
  {\sc 3.sg} {\ident} {\sc 3.sg.poss}  teacher {\foc}\\
\glt  `He is my TEACHER.' 

\ex
\gll   ʊ̀  fɪ̀ jáá  ǹ̩ǹ̩ tʃítʃà rà.\\
  {\sc 3.sg} {\pst} {\ident} {\sc 3.sg.poss}  teacher {\foc}\\  
\glt  `He was my TEACHER.' 

\z 
 \z 

 Secondly, the preverb {\sls fɪ}   ({\it gl.} {\sc mod}) can have  deontic
meaning.  


\ea\label{ex-preverb-fi-deonc}

\ea\label{ex-preverb-fi-deonc-pos}
\gll ʊ̀ fɪ̀ɪ́ jàà  ǹ̩ǹ̩ tʃítʃà rā.\\
  {\sc 3.sg}  {\mod}  {\ident} {\sc 3.sg.poss}  teacher {\foc}\\
\glt  `He should have been my TEACHER.' 

\ex
\gll ʊ̀ fɪ̀ wáá jàà  ǹ̩ǹ̩ tʃítʃà.\\
  {\sc 3.sg} {\mod} {\neg} {\ident} {\sc 3.sg.poss}  teacher\\  
\glt   `He should not have been my teacher.'  


\ex
 ʊ̀  fɪ̀ jáá  ǹ̩ǹ̩ tʃítʃà  rā  {\rm `He was my TEACHER.'}
\ex
 ʊ̀  fɪ̀ wà jáá  ǹ̩ǹ̩ tʃítʃà {\rm `He was not my teacher.'}

\z 
 \z 


In (\ref{ex-preverb-fi-deonc}),  its presence still conveys  past
tense, but in addition it expresses that the situation did not really occur, yet
it was objectively supposed to occur or subjectively expected to occur or
awaited. The lengthening of the preverb {\sls fɪ} in the positive  is not
accounted for, but I suspect it  signals the imperfective. 
Compare the first two
sentences in (\ref{ex-preverb-fi-deonc}) with the last two  which convey the
neutral past. 

\ea\label{ex-preverb-fi-pure-deonc}

\ea\label{ex:GRM-vp11.2}
\gll m̩̀m̩̀ mɪ̀bʊ̀à fɪ́  bɪ̀rgɪ̀.\\
 {\sc 1.sg.poss} life {\mod}  delay\\
\glt  `May I live long!' 

\ex\label{ex:GRM-vp11.3}
\gll tɪ̀ɛ̀ m̩̀m̩̀ mɪ̀bʊ̀à bɪ́rgɪ̀.\\
   give {\sc 1.sg.poss} life delay\\
\glt  `Let me live long!' 

\z 
 \z 



The positive sentence in (\ref{ex-preverb-fi-deonc-pos}) can receive  a
translation along these lines:  In a desirable possible world, he was my
teacher, but it is not what happened in
the real world. 


Finally, the preverb {\sls fɪ}  in (\ref{ex-preverb-fi-pure-deonc}) still 
conveys
 deontic modality, where the speaker prays or asks permission for a 
situation. Notice, however,  that it cannot refer to a past event. The two
sentences
in (\ref{ex-preverb-fi-pure-deonc}) have a corresponding meaning. Example
(\ref{ex:GRM-vp11.3}) is framed in an imperative clause (see \is{optative} {\it 
optative} in 
Section
\ref{sec:GRM-imper-clause}). 


\paragraph{Preverb three-interval tense}
\label{sec:GRM-preverb-three-int-tense}

Chakali encodes  in  preverbs  a type of time categorization  known as 
three-interval tense  \citep[366]{Fraw92}. It is possible to express that an 
event occurred specifically yesterday, as opposed to earlier today and the day 
before yesterday, i.e. {\it hesternal tense} ({\it gl.} {\sc hest}), or 
specifically tomorrow, as opposed to later today and the day after tomorrow, 
i.e. {\it crastinal tense}  ({\it gl.}  {\sc cras}). The hesternal tense 
particle {\sls dɪ}/{\sls de} ({\it gl.} {\sc hest})  refers to the day 
preceding 
the speech time.  It has the temporal nominal  counterpart  {\sls dɪare (tɪŋ)} 
`yesterday'.  



\ea\label{ex:vp2.11.a} 
\gll {(dɪ̀àrè tɪ̀n)} ʊ̀ nɪ́ ʊ̀ tʃɛ̀ná dɪ́ wāāwā  {(dɪ̀àrè tɪ̀n)}.\\
{(yesterday)} {\sc 3.sg} {\sc conn} {\sc 3.sg.poss} friend
{\sc hest}  come.{\sc pfv} {(yesterday)}\\ 
\glt  `He arrived with his friend yesterday.'
 \z



In (\ref{ex:vp2.11.a}),  the  phrase {\sls dɪare 
tɪn} `yesterday' is optional,  and  when it is used it must be expressed at the 
end or at the beginning of the clause.


\ea\label{ex:vp4.5} {\rm Will you work for the chief today or tomorrow?}\\
\gll  ǹ̩ tʃɪ́ kàá tʊ̀mà tɪ̄ɛ̄ʊ̄ rà, záàŋ,  ǹ̩ kàá hɪ̃̀ɛ̃̀sʊ̀ʊ̄.\\
 {\sc 1.sg} {\sc cras}  go  work give.{\sc 3.sg} {\sc foc},
today,   {\sc 1.sg}  {\sc egr} rest.{\sc foc}\\
\glt  `I shall work for
him tomorrow, today,  I shall rest.' 
 \z

The crastinal tense preverb {\sls tʃɪ} ({\it gl.} {\cras})  in  (\ref{ex:vp4.5})
functions as future particle,  but is limited to the day following the event
time.
In that sentence the event time referred to follows  the utterance
time by one day.  The temporal nominal counterpart  of {\sls tʃɪ} is {\sls  
tʃɪa} 
`tomorrow'. As
for the hesternal tense and the corresponding nominal,  the  nominal may or
may not co-occur with the crastinal tense particle. 




The hesternal tense particle {\sls dɪ} is homophonous with the ({\it ex-situ
subject}) imperfective particle  {\sls dɪ} discussed in Section
\ref{sec:GRM-ipfv-part}.  In addition, the question arises as to whether the
crastinal tense  is inherently future, and if so, whether or not it can
co-occur with the future-encoding egressive preverb discussed in Section
\ref{sec:GRM-EVC-egr-ingr}. Consider their distribution and meaning in the
examples given in (\ref{ex:GRM-prev-dist}).


\ea\label{ex:GRM-prev-dist}

\ea\label{ex:GRM-prev-dist-chew-presprog}{\rm Imperfective}\\
\gll  sɪ́gá (rá)  ʊ̀ dɪ̀  tíē.\\
 bean  ({\foc}) {3.\sg} {\ipfv} chew\\
\glt `It is BEANS he is chewing'

 \ex\label{ex:GRM-prev-dist-chew-past}{\rm Perfective/Past}\\
\gll  sɪ́gá (rá) ʊ̀   tìè.\\
 bean  ({\foc}) {3.\sg}  chew\\
\glt `It is BEANS he chewed'


 \ex\label{ex:GRM-prev-dist-chew-past}{\rm Hesternal past}\\
\gll  sɪ́gá (rá) ʊ̀ dɪ́ tìè.\\
 bean  ({\foc}) {3.\sg} {\hest}  chew\\
\glt `It is BEANS he chewed yesterday'


 \ex\label{ex:GRM-prev-dist-chew-past-pro}{\rm Hesternal past progressive}\\
\gll  sɪ́gá (ra) ʊ̀ dɪ́ɪ́ tīè.\\
 bean  ({\foc}) {3.\sg} {\hest}  chew\\
\glt `It is BEANS he was chewing yesterday'

 \ex\label{ex:GRM-prev-dist-chew-futprog}{\rm Future (progressive)}\\
\gll  sɪ́gá (rá) ʊ̀  kàá   tíē.\\
 bean  ({\foc}) {3.\sg} {\fut}  chew\\
\glt `It is BEANS he will be chewing / will chew'

 \ex\label{ex:GRM-foc-top-chew-crasfutprog}{\rm Crastinal future 
(progressive)}\\
\gll  sɪ́gá (rá) ʊ̀ tʃɪ́  kàá   tìè.\\
 bean  ({\foc}) {3.\sg} {\cras} {\fut}   chew\\
\glt `It is BEANS he will be chewing / will chew tomorrow '


\z 
 \z 
 
 
A specific tonal melody associated with  the sequence {\sls dɪ tie} can express 
either a present progressive, as in (\ref{ex:GRM-prev-dist-chew-presprog}),  or 
a hesternal past, as in  (\ref{ex:GRM-prev-dist-chew-past}). Lengthening the 
hesternal past particle allows one to express the tense associated with the 
particle, in addition to indicating  progressive 
(\ref{ex:GRM-prev-dist-chew-past-pro}). This strategy seems to correspond 
semantically  to the apparent syntactic anomaly *{\sls dɪ dɪ},  {\it lit.} {\sc 
 hest} {\sc ipfv}.  The example in (\ref{ex:GRM-foc-top-chew-crasfutprog}) 
shows that the crastinal tense particle and the egressive particle signaling  
future  can co-occur.  Inserting the imperfective particle {\sls  dɪ} between 
the egressive particle and the verb in  (\ref{ex:GRM-prev-dist-chew-futprog}) 
and (\ref{ex:GRM-foc-top-chew-crasfutprog}) is  unacceptable. It is unclear 
whether these two examples must be interpreted as progressive or not.  



\paragraph{te}
\label{sec:GRM-preverb-te}

Lacking a corresponding verb to capture its meaning, the verb {\sls te} is 
glossed with the English adverb `early'. Even though  it is attested as main 
verb,  {\sls te} can  function  as a preverb and it is indeed more common to 
find 
it in that function. 


 \ea\label{ex:GRM-prev-early}
 \ea \gll  ɪ̀ téjòō.\\
  {\sc 2.sg} early.{\foc}\\
\glt  `You are early.'


\ex\label{GRM-prev-SVC-ka-1}
\glll gbɪ̃̀ã́     bààŋ   té   kà     sáŋá   à   
píé  {(...)}\\
monkey quickly   early   go  sit {\art}
yam.mound.{\pl}   {(...)}\\
{} {\it pv} {\it pv}  {\it v} {\it v} {}  {} {}\\
\glt `Monkey quickly went and sat on the (eighth) yam mounds (...)'  (LB 012)
\z
\z

The main verb {\sls te} and the preverb {\sls te} are shown respectively in  
(\ref{ex:GRM-prev-early}).  It contributes a manner, one in which the event is 
carried out before the expected or usual time.  It cannot be used  in adjunct 
positions (Section \ref{sec:GRM-adjuncts}).  


\paragraph{zɪ}
\label{sec:GRM-preverb-after-then}

The preverb {\sls zɪ} is marginal in the corpus.\footnote{There is a formally 
similar particle, {\sls ze}  ({\it gl.} {\sc exp}),  which is still not
understood: (i) it occurs after the noun phrase, and  (ii) its meaning
corresponds to
 `expected (by both the speaker and the
addressee, or only by the speaker)'. It informs that the referent of
the noun phrase was anticipated before the utterance time (or relative time) by
the speaker and addressee (or only the speaker).  Consider  the following
example:

\ea\gll ba ze  waawaʊ.\\
{\sc 3.pl.b} {\sc  exp} come.{\sc pfv}\\
`They (the expected people) have come.'
\z
}  



\ea
\ea\label{ex:GRM-prev-zi-1}  {\rm A father is giving a sequence of tasks to
his son}

 \glll tʊ̀mà  à  zɪ̃́ɛ̃́  mʊ̃́ã̀  ká  kà  tʊ̀mà  kùó   àká   zɪ́ kà  
tʊ̀mà à  gár\\
  {work} \textsc{art} {wall} {before}  \textsc{conn} {\sc egr} {work}
{farm} 
\textsc{conn} {after}  {go} {work} \textsc{art} {cattle.fence}\\  
{} {} {} {} {}  {} {} {} {} {\it pv} {\it v} {\it v} {} {}\\
\glt  `First repair the wall, then go and farm, then repair the cattle fence.'


 \ex\label{ex:GRM-prev-zi-2}
  \glll kààlɪ̀ dɪ̀á ká zɪ́ kààlɪ̀ kùó.\\
go  house and then go farm\\
{} {} {}  {\it pv} {\it v} {}\\
 \glt `Go to the house and then go to the farm.'


\z 
 \z


There is no corresponding verb in the language.   It is used to express an 
order of events,  in such case words such as {\sls mʊ̃ã}  `before' and {\sls 
zɪ} `after' and the connective {\sls ka/aka}  `and/then' are used, as 
(\ref{ex:GRM-prev-zi-1}) shows. However,  as (\ref{ex:GRM-prev-zi-2}) 
illustrates,  the preceding event may be presupposed, so  it is not necessarily 
uttered.







\paragraph{baaŋ}
\label{sec:GRM-preverb-baang}

 The preverb  {\sls baaŋ}  ({\it gl.} {\sc mod})  is primarily modal and is  
usually translated into English `must', `immediately', `quickly'  or `just'. 


\ea\label{sec:GRM-prev-bg-must}
\ea\label{ex:GRM-7.17}
\gll  kùórù ŋmá dɪ́ ǹ̩ kàá bààŋ bɔ́ bʊ̃́ʊ̃́ná  fí rē.\\
 chief say {\comp} {\sc 1.sg} {\fut} {\mod}   pay  goat.{\pl} ten {\sc foc}\\
\glt  `The chief says that I must pay him ten goats.' 

\ex\label{ex:GRM-14.3}
\gll  ɪ̀ɪ̀ kàá bààŋ jáʊ́ rā.\\
{\sc 2.sg} {\fut} {\mod} do.{\sc 3.sg} {\foc}\\
\glt  `You must do it.'

 \z 
 \z
 
 First, the examples in (\ref{sec:GRM-prev-bg-must}) show that  the preverb  
{\sls baaŋ} conveys an obligation and the notion of temporality is secondary. 


\ea\label{sec:GRM-prev-bg-time}
\glll   {(...)} à kpá ʊ̀ʊ̀ néŋ à sàgà ʊ̀ʊ̀ nɪ̄ dɪ́ ʊ̀ bààŋ té 
bɛ̀rɛ̀gɪ̀ dʊ̃́ʊ̃̀\\
  {(...})  {\conn}  take {\sc 3.sg.poss} arm {\conn} {be.on} {\sc
3.sg}  {\postp} {\conn} {\sc 3.sg} {\mod} {early} turn.into python\\
{} {} {} {} {} {} {} {} {} {} {} {\it pv} {\it pv} {\it v} {}\\

\glt  `(...) then put his hand on her  and quickly turned into
a python.' (Python story 025)
\z
   
   Secondly, as illustrated in (\ref{sec:GRM-prev-bg-time}),  the preverb  
{\sls baaŋ} can express an  abrupt or
swift   manner. 




 \ea\label{ex:GRM-prev-bg-excerpt}
\ea\label{ex:FUS-mod}
\gll  ʊ̀ zɪ́má dɪ́ jà kàá ŋmá ʊ̀ʊ̀ wɪ́ɛ́ rá ʊ̀ʊ̀ bààŋ tʃùò dúò.\\
 {\sc 3.sg}  know {\sc comp}  {\sc 1.pl}  {\fut} talk   {\sc 3.sg.poss} matter 
{\foc} {\sc 3.sg} {\mod} lie sleep\\
\glt  `He knew that we would talk about him, so he quickly slept.'

\ex
\gll kàwàá bààŋ tàrɪ̀ kééééŋ.\\
pumpkin just creep {\advm}\\
\glt `A pumpkin just crept like that ...' 

\ex
\gll à kùò ní ʊ̀ bààŋ jírúú kéŋ néé à wà kʊ̀ʊ̀.\\
{\sc art} farm {\postp}  {\sc 3.sg} {\mod} call.{\ipfv} {\sc dxm} {\foc} {\sc 
conn} {\sc ingr} tire\\
\glt `At the farm he kept calling (for someone) but got tired (gave up).'


\ex
\gll díŋ bààŋ jàà dìŋtʊ́l̀.\\
fire  just {\ident} flame\\
\glt `The fire suddenly became flame.'

 \z 
 \z

 Finally, the preverb  {\sls baaŋ} may act as a discourse particle used mainly 
to 
emphasize or intensify the action carried out, reminiscent of  the use of 
`just' 
in some English registers.  It is often translated in text as `immediately', 
`suddenly', `then',  or simply `just'. Examples are given in 
(\ref{ex:GRM-prev-bg-excerpt}).


\paragraph{bɪ}
\label{sec:GRM-preverb-iteration}

The examples in (\ref{ex:GRM-prev-bi}) show that  the preverb particle {\sls bɪ}
expresses iteration, but also the single repetition of an event, and follows the
negation particle. 

% bɪ kuor ŋma
%  repeat
%  bɪ pɪlɪ
% start again
% start


\ea\label{ex:GRM-prev-bi}

\ea\label{ex:vp33.2.}
\gll ʊ̀ bɪ́ kʊ̀ɔ̀rɛ̀ sã̀ã̀ ʊ̀ʊ̀ dɪ̀à rá.\\
 {\sc 3.sg}  {\itr} make build {\sc 3.sg.poss} house {\foc}\\
\glt  `He rebuilt his hut.' 


\ex\label{ex:GRM-vp10.4}
\gll à bìtʃèlíí bɪ́ sīīú.\\
 {\art}  child.fall   {\itr} raise.{\foc}\\
\glt  `The fallen child gets up again.' 



\ex\label{ex:vp10.4.}
\gll ʊ̀ wà bɪ́ tùwō.\\
    {3.\sg} {\neg} {\itr} be.at\\
\glt  `She is not here again.' 

\z 
 \z 


Unlike other preverbs,  {\sls bɪ} may also appear within noun phrases to express
frequency time. This is shown in (\ref{ex:GRM-vp19.2.}) (see Section
\ref{sec:NUM-repet}).



\begin{exe} 
\ex\label{ex:GRM-vp19.2.}
\gll  ǹ̩ jáà  káálɪ̀ ùù pé rè tʃɔ̀pɪ̀sɪ̀ bíí mùŋ.\\
{\sc 1.sg} {\hab} go {\sc 3.sg.poss} end {\foc}  day.break {\itr} all\\
\glt  `I do visit him every day.' 

\z 



 \paragraph{bra}
\label{sec:GRM-preverb-return}

The preverb {\sls bra} has a corresponding verb with the same form. It is
primarily a motion verb which conveys a change of direction. 


\ea\label{ex:GRM-verb-bra}

\ea
\gll brà à káálɪ̀.\\
return {\conn} go\\
\glt `Go back.' (Hearer coming towards speaker, speaker ask addressee to turn 
and 
go back)

\ex
\gll brà àká tʃáʊ̀.\\
return {\conn} leave.{\sc 3.sg}\\
\glt `Return and leave him.' (Speaker ask addressee to turn and go away from 
the 
person the addressee is with)

\z 
 \z

The examples 
in (\ref{ex:GRM-verb-bra}) present the verb {\sls bra} in imperative clauses
separated by the connectives {\sls a} and {\sls aka}. 

\ea\label{ex:vp33.1.}
\gll ʊ̀ brá tʊ̀mà à tʊ́má tɪ́ŋ kà wà wíré kéŋ̀.\\
 {\sc 3.sg}  {again}  {work} {\art} {work}   {\art} {\egr} {\neg} well {\advm}\\
\glt  `He redid the work that was badly done.'
\z


When {\sls bra} functions as 
a preverb, as in (\ref{ex:vp33.1.}),  it loosely keeps its motion sense and
conveys in addition a sort of repetition. It differs from the morpheme {\sls bɪ}
introduced in
 Section \ref{sec:GRM-preverb-iteration} because it does not mean that an
action is
done
repeatedly.  Instead, the preverb {\sls bra} is associated with actions done 
`once
more', `over again',  or `anew'.

%regloss ka



\paragraph{ja}
\label{sec:GRM-preverb-hab}

The preverb {\sls ja(a)} ({\it gl.} {\sc hab})  indicates habitual aspect. It 
expresses that the subject's referent is accustomed to, familiar with, or 
routinely do the action described by the predicate.


\ea\label{ex:GRM-prev-hab-do}
\gll tʃɔ̀pɪ̀sɪ̀ bɪ́-múŋ̀ ʊ̀ʊ̀ jáà jááʊ̄.\\
 day.break {\itr}-all {\sc 3.sg} {\hab} do.{\sc 3.sg}\\
\glt `He does it every day.'
\z 

 A variation in  length and intonation suggest  an (im)perfective aspectual 
distinction. In   (\ref{ex:GRM-prev-hab-do})  there is a  vowel sequence {\sls 
aa} pronounced with a falling intonation. Compare this with the examples in 
(\ref{ex:GRM-prev-hab}). 



% % 
% % \ea\label{ex:habitual}
% % \ea\label{ex:}
% % \gll  ǹ̩  já kààlɪ̀ kùó.\\
% % {\sc 1.sg} {\sc hab} go farm\\
% % \glt `I do go to the farm.'
% % 
% % \ex\label{ex:}
% % \gll  ǹ̩  jáà káálɪ̀ kùó.\\
% % {\sc 1.sg} {\sc hab} go farm\\
% % \glt `I have been going to the farm.'
% % 
% % \z 
% %  \z
% %  
 

\ea\label{ex:GRM-prev-hab}
\ea\label{ex:GRM-prev-hab-do-pfv}
\gll  Kàlá já tùgòsì bísé ré.\\
 K.  {\hab}  beat.{\sc pl} child.{\sc pl} {\sc foc}\\
\glt `Kala beat children.' (He used to do it.)

\ex\label{ex:GRM-prev-hab-impv}
\gll  Kàlá jáà túgósì bísé ré.\\
K.  {\hab}  beat.{\sc pl} child.{\sc pl} {\sc foc}\\
\glt `Kala beat children.' (He regularly does it.)

\z 
 \z

The aspectual distinction in (\ref{ex:GRM-prev-hab}) is reflected by the 
preverb's vocalic length and intonation, but also on the following verb's 
intonation.




\paragraph{ha}
\label{sec:GRM-preverb-yet}

The morpheme {\sls ha} ({\it gl.} {\sc mod}) is similar in meaning to the 
English 
morpheme `yet'. 

% 
% % The morpheme  {\sls ha} is circumscribed to the expanded verbal group, 
% although I 
% % translate as `yet'  another expression in (\ref{ex:yet-conn}) which 
%functions 
% as 
% % connective. The expression {\sls haalɪ}, also found in many West African 
% % languages -- ultimately from Mande family --  is not frequent in the data 
% % available. 


\ea
\ea\label{ex:vp32.24}
\gll ʊ̀ʊ̀ háá díūū.\\
  {3.\sg}  {\mod} eat.\foc\\
\glt  `He is still eating.' 


\ex\label{ex:vp20.3.2.}
\gll ʊ̀ há wà díìjē.\\
 {3.\sg}  {\mod} {\neg} eat.{\pfv}\\
\glt  `He has not eaten yet.'


\ex\label{ex:vp21.2.1.}
\gll bà ɲíné ʊ̀ʊ̀ gɛ̀rɛ̀gá rá àká ʊ̀ʊ̀ háá wɪ̄ɪ̀.\\
 {\sc 3.pl.hum+} look {\sc 3.sg.poss} sickness {\foc} {\conn}  {\sc 3.sg}
{\mod} ill\\
\glt  `He has been cared for to no avail; he is still ill.' 


\ex\label{ex:vp20.1.1.}
\gll ʊ̀ há  wà wāā báàŋ múŋ̀.\\
    {3.\sg} {\mod}  {\neg} come {\dem} {\quant}.all\\
\glt  `He does not come here (ever).' 


\ex\label{ex:vp20.3.1.}
\gll ʊ̀ há wà wááwá.\\
    {3.\sg} {\mod}   {\neg} come.{\pfv}\\
\glt  `He has not come yet.' 

% % 
% % \ex\label{ex:yet-conn}
% % \gll ʊ̀ jɪ́rʊ́ʊ́ sāŋā mūŋ̀ ká ʊ̀ há wà wááwá.\\
% %  {3.\sg}  call.{\sc 3.sg} time all {\conn}   {\sc 3.sg} {\sc mod}  
% {\neg}
% % come.{\pfv}\\
% % \glt  `He called her long time ago, yet she has not
% % come.' 

\z 
 \z
 
 It is used when an event is or was anticipated and a speaker considers or considered probable the 
occurrence of the event. As  for the English `yet', it is frequently found in negative polarity. In 
such cases the morpheme {\sls ha} indicates that the event is expected to happen and the negative 
marker {\sls wa} indicates that the event has not unfolded or happened at the referred time. In the 
cases where {\sls ha} is found in a positive polarity,  it  conveys a continuative aspect, similar 
to English `still',  as in (\ref{ex:vp32.24}) and (\ref{ex:vp21.2.1.}). 



\paragraph{tu and zɪn}
\label{sec:GRM-preverb-up-down} 

The verbs {\sls tuu} and {\sls zɪna} are motion
expressions making reference to two opposite paths. 


\ea\label{ex:GRM-verb-up-down}
\ea
\gll ǹ̩ zɪ́nà sàl lá ḿ̩ páá tʃùònò.\\
{\sc 1.sg} go.up flat.roof {\foc} {\sc 1.sg} take.{\pv} shea.nut.seed.{\pl}\\
\glt  `I go up on the roof to collect my shea nuts.'

\ex
\gll ǹ túú dɪ̀à rá.\\
{\sc 1.sg} go.down house {\foc}\\
\glt I went down to the house.'
\z 
 \z

When they are used as main
predicate, as in example (\ref{ex:GRM-verb-up-down}),  they denote `go down' and
`go up' and  surface as {\sls tuu} and {\sls zɪna} respectively. 




\ea\label{ex:GRM-preverb-up-down}
\ea\label{ex:GRM-preverb-up}
\gll zɪ́ná tʃɔ́  à káálɪ̀.\\
   {go.up} run {\conn} go\\
\glt  `Go up,  run, and leave'  (*Run upwardly and go)

\ex\label{ex:GRM-preverb-down}
\gll tùù tʃɔ́  à káálɪ̀.\\
   {go.down} run {\conn} go\\
\glt  `Go down, run, and leave'  (*Run downwardly and go)

\z 
 \z
 
The verbal morphemes {\sls tuu} and {\sls zɪn} in 
(\ref{ex:GRM-preverb-up-down}) 
are not treated as preverbs, but first verbs in SVCs.  As explained at the 
beginning of  Section \ref{sec:GRM-precerv}, more criteria are required to be 
considered in order to categorize verbals  of that particular kind.


% % % % 6
% % % %  ́
% % % % The directional particles he (‘itive’, related to the homophonous verb 
% meaning
% % % % ‘go’ (departure from
% % % %  ́
% % % % deictic center or indexically determined location)) and va (‘ventive’, 
% related
% % % % to the homophonous verb meaning
% % % % ‘come’ (arrival at deictic center or indexically determined location)) 
% belong
% % % %to % the class of preverbs of Ewe.
% % % % These are forms that mark functional categories such as aspect, modality, 
% and
% % % % voice on verbs. Preverbs differ
% % % % from verbs in that they do not head VPs, do not inflect for habitual 
% aspect,
% % % %and % do not take NP or PP
% % % % complements (cf. Ameka 1991, 2005a,b, Ansre 1966).
% % % 
% % % 
% % % % The particle  {\sls ja} is
% % % % polyfunctional:  when it precedes a main verb it  means  either `do'   to
% % % % emphasize the event or conveys an habitual reading, or as, in the present
% % % %case,
% % % % it links two noun phrases. The latter case is glossed in example
% % % %(\ref{ex:agrE})
% % % % and (\ref{ex:agrF}) as {\sc ident}. 
% % % 
% % % 
% % % % --Dakubu
% % % % I wonder whether what you call IPFV is an egressive particle? such a 
% particle
% % % % derived from 'go' is quite common.  If it is incompletive / progressive 
% this
% % % % might have to do with the tone pattern



\subsection{Verbal suffixes}
\label{sec:GRM-verb-suffix}


In Section \ref{sec:GRM-verb-word}, two suffixes were introduced: the 
perfective 
intransitive suffix and the assertive suffix. It was shown that the perfective 
intransitive suffix surfaces either as {\sls -jE}, {\sls -wA} or {\sls  -\O} 
depending 
on  the verb stem.  The assertive suffix appears  in the imperfective and 
perfective  intransitive construction if  (i) none of the other constituents in 
the clause are in focus, (ii) the clause does not include negative polarity 
items, and (iii) the clause is intransitive, that is, there is no grammatical 
object. Also,  as mentioned in Section \ref{sec:GRM-imper-clause},  the suffix 
{\sls -ɪ}/{\sls -i} appears in the negative imperative. 

In this section,  the incorporated object index  ({\sc
o}-clitic), the pluractional  suffix, and  other derivative suffixes whose
functions are not fully understood are introduced.


\subsubsection{Incorporated object index}
\label{sec:GRM-morph-opro}


The object index  is represented as being incorporated into the verb,  and 
together they form a phonological word (e.g.  {\sls wʊ̀sá tɪ́ɛ́ń nā} < {\sls 
wʊ̀sá tɪɛ-n̩ na}  `Wusa gave-{\sc 1.sg} {\sc foc}').  For that reason I refer 
to this incorporated object index as the {\sc o}-clitic. Given the constraints 
governing the appearance of the perfective intransitive suffix and the 
assertive 
suffix, it is obvious that the {\sc o}-clitic cannot coexist with any of them. 
Recall that the  weak subject and object  pronouns are identical (see Section 
\ref{sec:GRM-personal-pronouns}).

\begin{table}[!htb]
\centering
\caption{Incorporated object index on  CV(V) stems\label{tab:object-clitic}}

\subfloat[tɪɛ `give']{
\begin{Itabular}{ll}
 wʊ̀sá tɪ́ɛ́-ń̩ nā & `Wusa gave ME'\\
wʊ̀sá tɪ́ɛ́-ɪ́ rā & `Wusa gave YOU'\\
 wʊ̀sá  tɪ́ɛ́-ʊ́ rā &  `Wusa gave HER'\\
 wʊ̀sá tɪ́ɛ́-já rā &  `Wusa gave US'\\
 wʊ̀sá tɪ́ɛ́-má rā & `Wusa gave YOU'\\
 wʊ̀sá tɪ́ɛ́-á rā & `Wusa gave THEM'\\
 wʊ̀sá tɪ́ɛ́-bá rā &  `Wusa gave THEM'\\
\end{Itabular} 
}
\quad
\subfloat[tie `cheat']{
\begin{Itabular}{ll}
 wʊ̀sá tíé-ń̩ nē & `Wusa cheated ME'\\
 wʊ̀sá tíé-í rē & `Wusa cheated YOU'\\
 wʊ̀sá tíé-ú rō &  `Wusa cheated HER'\\
 wʊ̀sá tíé-já rā &  `Wusa cheated US'\\
 wʊ̀sá tíé-má rā & `Wusa cheated YOU'\\
 wʊ̀sá tíé-á rā & `Wusa cheated THEM'\\
 wʊ̀sá tíé-bá rā &  `Wusa cheated THEM'\\
\end{Itabular} 
}
\quad
\subfloat[tie `cheat']{
\begin{Itabular}{ll}
 wʊ̀sá tíé-jé rē &  `Wusa cheated US'\\
 wʊ̀sá tíé-mé rē & `Wusa cheated YOU'\\
 wʊ̀sá tíé-é rē & `Wusa cheated THEM'\\
 wʊ̀sá tíé-bé rē &  `Wusa cheated THEM'\\
\end{Itabular} 
}
\quad
\subfloat[po `divide']{
\begin{Itabular}{ll}
 wʊ̀sá pó-jé rē &  `Wusa divided US'\\
 wʊ̀sá pó-mó rō & `Wusa divided YOU'\\ %pó-mó wō
 wʊ̀sá pó-á rā & `Wusa divided THEM'\\
 wʊ̀sá pó-bé rē &  `Wusa divided THEM'\\
\end{Itabular} 
}
\end{table}

Table \ref{tab:object-clitic} shows that the {\sc atr}-harmony 
operates in the domain produced by the {\sc
o}-clitic merging with a CV or CVV stem, but may or may not affect the
plural pronouns, as Tables \ref{tab:object-clitic}(b) and 
\ref{tab:object-clitic}(c) display. The form of the focus particle is determined
by
the preceding material (i.e. the phonological word  verb+{\sc
o}-clitic) and the harmony rules introduced in
Section
\ref{sec:focus-forms}.  The irregularities in Table \ref{tab:object-clitic}(d)
are not accounted for.  I did perceive rounding throughout in conversations
(i.e.  {\sls wʊ̀sá pómá rā} $>$ {\sls wʊ̀sá pómó wō} `Wusa divided you.{\sc 
pl}'), 
but
I was unable to get a consultant  produce it in an elicitation session. Table
\ref{tab:object-clitic}(d) should be seen as displaying various renditions,
i.e. with and without {\sc atr-}harmony or {\sc ro-}harmony.


A CVCV stem differs from a CV or CVV stem by exhibiting vowel apocope and/or 
vowel
coalescence.  Table \ref{tab:object-clitic-CVCV} provides paradigms for {\sls 
kpaga} `catch' and {\sls goro} `(go in) circle'. 



\begin{table}[!htb]
\centering
\caption{Incorporated object index on  CVCV stems
\label{tab:object-clitic-CVCV}}

\subfloat[kpaga `catch']{
\begin{Itabular}{ll}
 wʊ̀sá kpáɣń̩ nā & `Wusa caught ME'\\
 wʊ̀sá kpáɣɪ́ɪ́ rā & `Wusa caught YOU'\\
 wʊ̀sá kpáɣʊ́ʊ́ rā &  `Wusa caught HER'\\
 wʊ̀sá kpáɣə́já wā &  `Wusa caught US'\\
 wʊ̀sá kpáɣə́má wā & `Wusa caught YOU'\\
 wʊ̀sá kpáɣáá wā & `Wusa caught THEM'\\
 wʊ̀sá kpáɣə́bá wā &  `Wusa caught THEM'\\
\end{Itabular} 
}
\quad
\subfloat[goro `(go in) circle']{
\begin{Itabular}{ll}
wʊ̀sá górń̩ nō & `Wusa circled ME'\\
 wʊ̀sá góríí rē & `Wusa circled YOU'\\
 wʊ̀sá górúú rō &  `Wusa circled HER'\\
 wʊ̀sá górə́já wā/rā &  `Wusa circled US'\\
 wʊ̀sá górə́má wā/rā & `Wusa circled YOU'\\
 wʊ̀sá góráá wā/rā & `Wusa circled THEM'\\
 wʊ̀sá górə́bá wā/rā &  `Wusa circled THEM'\\
\end{Itabular} 
}
\end{table}

The schwas ({\sls ə}) in {\sls kpaɣəja} and  {\sls gorəja} are perceived as 
fronted,
and the ones in {\sls kpaɣəma} and {\sls gorəma}  as rounded. Although this is
certainly due to the following consonant, they are so weak that they can only be
heard when they are carefully pronounced (see Section \ref{sec:PHO-weak-syll}). 
The paradigm in Table  
\ref{tab:object-clitic-CVCV}(b) can also be uttered in the plural as 
{\sls górójé rē} ({\sc 1.pl}),  %
{\sls górémá rā} ({\sc 2.pl}), %
{\sls góráá rā} ({\sc 3.pl.-h}), and %
{\sls górébá rā} ({\sc 3.pl.+h}). 
 The focus particle {\sls wa} is a
variant of {\sls ra}. Some consultants  agree that these forms are in free 
variation,
yet the {\sls wa} form coexists only with  the plural in the paradigms elicited.
Nonetheless, such paradigm elicitations are particularly subject to
unnaturalness.\footnote{I personally believe that the alteration is
determined by some kind of sandhi, not number. As to why {\sls wa} appears only 
in
the plural, a scenario may be that (i) first, I install a routine by starting
with `ME' and ending with `THEM', (ii) in the process of eliciting, the passage
from third singular to first plural triggers  a different verb shape, i.e.
CVCVV/CVCN  to CVCVCV, and (iii)  although formally identical to the verb forms
of the singular, the reason why {\sls wa} follows the third plural non-human 
could
be explained by psychological habituation.}

\subsubsection{Pluractional suffixes}
\label{sec:GRM-PluralVerb}


A pluractional verb is defined as a verb which can (i) express the repetition of
an event,  (ii)   subcategorize for a plural object and/or  plural subject,
and/or  (iii)  be marked by the pluractional suffix {\sls -sI}, a derivative 
suffix whose  vowel quality is always high and
front
and  {\sc atr} value determined by the stem vowel(s).\footnote{An exposition of
the
`plural verbs' in Vagla can be found in \citet{Blen03}. \citet[viii]{daku07}
calls a similar morpheme `iterative' (i.e. Gurene {\sls -sɛ}).  Among the West
African
languages, it is the pluractional verbs in Hausa which have received most
attention \citep[see][]{Jose08}. [Storch forthcoming]}  According to (i) above, 
the iterativeness may
affect the interpretation of the number of participants of an event. Consider
the contrasts between the 
sentences in (\ref{ex:GRM-pv-cut}), where none of the arguments are in the
plural (i.e. contra (ii)).


\ea\label{ex:GRM-pv-cut}
  
 \ea\label{GRM-pv-cutsg}
\gll   ǹ̩  téŋé  à nàmɪ̃̀ã̀  rā.\\
    {\sc 1.sg} {cut}  {\sc art} {meat} {\sc foc}\\
\glt `I cut a piece of meat (i.e.  made a cut in the flesh or cut into two
pieces).'

\ex\label{GRM-pv-cutpl}
\gll ǹ téŋé-sí  à nàmɪ̃̀ã̀  rā.\\
    {\sc 1.sg} {cut-{\sc pv}} {\sc art} {meat} {\sc foc}\\
\glt `I cut the meat into pieces.'

 
\z 
 \z

In  (\ref{GRM-pv-cutpl}),  the formal distinction on the verb `cut',  compared
to (\ref{GRM-pv-cutsg}),  causes  the event to be interpreted as one which
involves the repetition of the `same'  sub-event.  The word {\sls namɪ̃ã} 
`meat'
is allowed in both the contexts of (\ref{GRM-pv-cutsg}) and
(\ref{GRM-pv-cutpl}), although one may argue that the word {\sls namɪ̃ã} is
inherently
plural but grammatically singular,  and that the word is appropriate in both
contexts. Despite the fact that  `meat' has indeed a plural form, i.e. {\sls 
nansa}, it is probably the mass term denotation of {\sls namɪ̃ã} which 
makes (\ref{GRM-pv-cutpl}) acceptable. 


\ea\label{GRM-pv-turn}
  
 \ea\label{GRM-pv-turnsg}
\gll   ǹ̩  tʃígé  à  hɛ̀ná  rá.\\
  {\sc 1.sg} {turn} {\sc art} {bowl.\sg} {\sc foc}\\
\glt `I turn (upside down) the bowl.'

 \ex\label{GRM-pv-turnpl1}
\gll   ǹ̩  tʃígé-sí  à  hɛ̀nsá  rá.\\
   {\sc 1.sg}   {turn-{\sc pv}} {\sc art} {bowl.\pl} {\sc foc}\\
\glt `I turn (upside down) the bowls (one after the other).'


 \ex\label{GRM-pv-turnpl2}
\gll {(?)}  n̩  tʃige-si   a  hɛna  ra.\\
    {}  {\sc 1.sg} {turn-{\sc pv}} {\sc art}  {bowl.\sg}  {\sc foc}\\
\glt `I turn (upside down in a repetitional fashion) the bowl.'

\z 
 \z

In (\ref{GRM-pv-turn}), however,  the 
grammatical object of a
pluractional verb {\sls tʃigesi} `turn iteratively' or `put on face
down iteratively'  must refer to individuated entities. Comparing  
(\ref{GRM-pv-turnsg}) and (\ref{GRM-pv-turnpl2}) with 
(\ref{GRM-pv-turnpl1}),   the pluractional verb cannot coexist with a singular
noun as grammatical object due to the fact that  some `turning' events cannot be
conceived as affecting the same object in a repetitive fashion. However, in
(\ref{GRM-pv-beat}) the `beating' can affect  one or several
individuals. 


\ea\label{GRM-pv-beat}
  
 \ea\label{GRM-pv-beat.sg}
\gll   ǹ̩   túgó  à bìè  rē.\\
   {\sc 1.sg}  {beat} {\sc art} {child.\sg} {\sc foc}\\
\glt `I beat the child.'

\ex\label{GRM-pv-beat.pl1}
\gll   ǹ̩ túgó-sí  à bìsé  ré.\\
   {\sc 1.sg}   {beat-{\sc pv}} {\sc art} {child.\pl} {\sc foc}\\
\glt ` I beat the children.'


\ex\label{GRM-pv-beat.pl2}
\gll ñ̩̀ túgó-sí  à  bìè  rē.\\
    ñ̩̀  {\sc 1.sg} {beat-{\sc pv}} {\sc art}  {child.\sg} {\sc foc} 
\\
\glt `I beat the child (more than once, over a short period of time).'


 
\z 
 \z

Whereas  (\ref{GRM-pv-beat.pl2})
has a possible interpretation, two language consultants
could not assign a meaning to (\ref{GRM-pv-catchout}) below. 




\ea\label{GRM-pv-catch}
  
 \ea\label{GRM-pv-catchsg}
\gll ŋ̩̀  kpágá  à  zál  là.\\
   {\sc 1.sg}   {caught} {\sc art} {chicken.\sg} {\sc foc}\\
\glt `I caught a chicken.'


 \ex\label{GRM-pv-catchpl1}
\gll ŋ̩̀ kpágá-sɪ́  à  zálɪ́ɛ́ rà.\\
    {\sc 1.sg} {caught-{\sc pv}} {\sc art} {chicken.\pl} {\sc foc}\\
\glt `I caught chickens (i.e. in repeated actions).'


 \ex\label{GRM-pv-catchpl2}
\gll   ŋ̩̀  kpágá  à  zálɪ́ɛ́ rà.\\
    {\sc 1.sg}  {caught} {\sc art} {chicken.\pl} {\sc foc}\\
\glt `I caught chickens (i.e. in one move).'


 \ex\label{GRM-pv-catchout}
\gll (?)  ŋ̩  kpaga-sɪ  a  zal  la.\\
  {}   {\sc 1.sg}  {caught-{\sc pv}} {\sc art} {chicken.\sg} {\sc foc}\\
\glt `I caught a chicken (i.e. after unsuccessful attempts until finally
succeeding with
one particular chicken).'

 
\z 
 \z


A pluractional verb usually denotes an action, but not a state. Therefore, in
(\ref{GRM-pv-catch}), the sense of {\sls kpaga}$_{1}$  is related to `catch', 
and 
not
to the  possessive sense of the verbal state lexeme   {\sls kpaga}$_{2}$
`have'.\footnote{Though I like to treat {\sls dʊasɪ} as a counterexample.  The
pluractional verb {\sls dʊasɪ} `be in a row'  may be  derived from the 
existential
predicate {\sls dʊa} `be on/at/in'.  For instance, the verbs {\sls tele} `lean' 
  
and {\sls telege} `lean' are determined 
by
the number value ({\it sg.}/{\it  pl.})  of the subject.  If more examples like
these  arise, {\it pluractional} would then loose its literal 
signification.} Beside {\sls /-sI/}, the suffix {\sls 
/-gE/} may also turn a verbal process lexeme into a pluractional verb, e.g.   
{\sls tɔtɪ} `pluck' $>$ {\sls  tɔrəgɛ} `pluck iteratively' and  {\sls keti} 
`break'  $>$
{\sls kerigi} `break iteratively'.

\ea\label{ex:GRM-kpa-paa}
  
 \ea\label{ex:GRM-kpa}
\gll kà kpá zál háŋ̀ tà.\\
go take.{\sc pl} fowl.{\sg} {\dem} let.free\\
\glt `Go and take this fowl away.'
   \ex\label{ex:GRM-paa}
\gll kà páá zálɪ́ɛ́ hámà tà.\\
go take.{\sc pl} fowl.{\pl} {\dem}.{\pl}  let.free\\
\glt `Go and take these fowls away.'
 
\z 
 \z


Finally, a pluractional verb must not necessarily display the
suffixation pattern
described above. This is confirmed by the pair  {\sls kpa}/{\sls 
paa} `take'  in
(\ref{ex:GRM-kpa-paa}).




\subsubsection{Possible derivational suffixes}
\label{sec:GRM-deri-suff}


\citet[37]{Daku09} and \citet[69]{Bonv88} identify some derivational suffixes 
in 
Gurene and Kasem respectively, but write that their signification is hard to 
establish.  However, their descriptions indicate that  derivational suffixes 
mainly encode aspectual distinctions.

As mentioned in Section \ref{sec:GRM-verb-syll-und-tone}, about 90\% of the
verbs are monosyllabic or bisyllabic, and  only the consonants {\sls m,
t, s, n,  l} and {\sls g} are found  in onset position word-medially in
trisyllabic verbs. This situation could suggest that 10\% of the verbs in the
current lexicon are the product of verbal derivation, and that the consonants
found  in onset position word-medially in trisyllabic verbs are part of
derivational suffixes. However, apart from the pluractional suffix discussed in
the previous section,  it is impossible at this stage of the research to
establish a systematic mapping between the third syllable of a trisyllabic verb
and a meaning.  


\ea
 \ea\label{ex:plur-ex}
\gll   ʊ̀ wʊ́rɪ́gɪ́ à hàɣlíbíé ré.\\
{\sc 3.sg} scatter {\sc art} block.{\sc pl} {\sc foc}\\ 
\glt `He scattered the mud blocks.' (they were piled and packed)

 \ex\label{ex:}
\gll  ʊ̀ wʊ́rá à hàɣlíbíí ré.\\
 {\sc 3.sg} move {\sc art} block  {\sc foc}\\
\glt `He moved a mud block.' (they are uneven, but still piled)
\z 
 \z
 
 
 
\begin{table}[!htb]
\small
\centering
\caption{Possible derivational suffixes [TO DO!]\label{tab:GRM-der-suff}}

\begin{tabular}{lllll}
\lsptoprule

 &&&{\sls -gV}&\\\midrule

wʊ̀rà {(v)}& `move, shift' & $>$ & wʊ̀rɪ̀gɪ̀ {(v)}& `scatter'\\
tàrà  {(v)}& `support' & $>$ &tàràgɛ̀ {(v)}& `pull'\\
%bɪla {(v)}& `turn repetitively' & $>$ & bɪlgɪ {(v)}& `clean'\\
brà {(v)}& `return' & $>$ & bɛ̀rɛ̀gɪ̀  {(v)}& `change direction'\\\midrule

&&&{\sls -mV} &\\\midrule

ɲàgà  {(v)} & `be sour' &$>$ & ɲàgàmɪ̀  {(v)}& `ferment'\\
víl {(n)} &`well' & $>$ &vílímí {(v)} & `whirl'\\
 mɪ̀là {(v)} & `turn round' & $>$ &mɪ̀lɪ̀mɪ̀ {(v)}& `turn'\\[1ex]\midrule

&&&{\sls -lV}&\\\midrule
 kàgà {(v)}& `choke'& $>$ & kàgàlɛ̀ {(v)} & `lie across'\\
 \lspbottomrule
\end{tabular}
\end{table}


The  example  provided in (\ref{ex:plur-ex}) and Table \ref{tab:GRM-der-suff} 
presents  some indications that {\sls m, l} and {\sls g}, i.e. CVCV{\sls  \{ m, l, g\} }V, 
are involved in some kinds of derivation, although the glosses (and part of 
speech categories) assigned to them clearly indicate that the next step would 
be 
to determine their exact meaning (and category).\footnote{The verb pair {\sls  go} 
`round'  and {\sls goro}  `(go in) circle'  is  manifestly a derivation as 
well, 
i.e. CV $>$ CV-rV.}


\section{Grammatical Pragmatic and Language Usage}
\label{sec:GRM-adjuncts}


\subsection{Manner deictics}
\label{sec:GRM-adv-pro}


Chakali has a two-term exophoric system of manner deixis \citep{koen12}. 
The  expressions  {\sls keŋ} and {\sls nɪŋ} are treated as  two manner deictics  
({\it gl.} {\sc dxm}).  Manner is a cover term since the content dimension 
appears to cover degree and  quality as well. Consider the 
examples in (\ref{ex:GRM-dxm-}).

\ea\label{ex:GRM-dxm-}
 \ea\label{ex:GRM-dxm-}
\gll kén nè bà já jāà\\
{\sc dxm} {\sc foc}  {\sc 3.pl.h+}  {\sc hab} do  \\
\glt `That's the way to do it (manner)'  

  \ex\label{ex:GRM-dxm-}
\gll  hàɣlɪ́kɪ́ŋ̀ zéné mààsɪ̀ nɪ́n nà\\
snake long equal {\sc dxm}  {\sc foc}\\
\glt `The snake was that/this big (degree)'  

  \ex\label{ex:GRM-dxm-}
  
  
\gll  kàlá máásɪ́ɪ́ nɪ́ŋ̀\\
K. equal.{\sc nmlz} {\sc dxm}\\
\glt `Kala is like that (quality) [of size]'  

  \ex\label{ex:GRM-dxm-}
  
  
\gll kàlá dʊ́nná kéŋ̀\\
K. type {\sc dxm}\\
\glt `Kala is like that (quality) [of nature]'  

\z 
 \z   

The expressions  {\sls keŋ} and {\sls nɪŋ} are very frequent  and bring
to mind the  English  `like this/that',  that is,  an expression which 
refers to something extralinguistic yet in  the context of the utterance. 
Example (\ref{ex:GRM-adv-pro-keng-ning}) illustrates this point.


\ea\label{ex:GRM-adv-pro-keng-ning}

 \ea\label{ex:GRM-adv-pro-ning}
\gll bàáŋ ɲʊ̃̀ã̀sá káá sìì báŋ̀ nɪ̄ nɪ̏ŋ.\\
{\q}  smoke  {\egr} rise {\dem} {\postp} {\advm}\\
\glt `What smoke is rising here like this?'  
%(Python story 059)
  \ex\label{ex:GRM-adv-pro-keng}
 \gll bàáŋ káá jāā kȅŋ?.\\
  {\sc q}.what {\sc egr} do  {\advm}\\ 
 \glt `What is doing like that?' (Reaction to a sound coming from inside a pot)

\z 
 \z   

The meaning difference between   {\sls nɪŋ}  and {\sls 
keŋ} seem to be 
motivated by the way they  encode a sort of psychological saliency on a
proximal/distal dimension. This distinction needs more evidence than the one I
provide,  but consider the conversation between A and B in
(\ref{ex:GRM-adv-pro-keng-AB}). 


\ea\label{ex:GRM-adv-pro-keng-AB}

 \ea\label{ex:GRM-adv-pro-A}
\gll A: nɪ́n nā bààbá ŋmȁ̠?\\
 {} {\advm} {\foc} B. say\\
\glt `Is this what Baaba said?'

  \ex\label{ex:GRM-adv-pro-B}
 \gll B: ɛ̃̀ɛ̃́ɛ̃̀ kén{\T ꜜ} né ʊ̀ ŋmá.\\
 {} yes {\advm} {\foc} {\sc 3.sg} say\\
\glt `Yes, that is what he said.'
 
 
\z 
 \z   

Similarly,  the (fictional) discourse excerpt in
(\ref{ex:GRM-adv-kapok}) concerns a father (A) addressing his son (B) on the
topic of  how to ignite kapok fiber. The sentence (\ref{ex:GRM-adv-kapok-A-2})
is accompanied with a demonstration on how to strike a cutlass on a stone.


\ea\label{ex:GRM-adv-kapok}
 
 
 \ea\label{ex:GRM-adv-kapok-A-1}
\gll A: kpá kóŋ à ŋmɛ̀nà díŋ.\\
{}  take kapok {\conn} ignite fire\\
\glt `Take some kapok and start a fire.'

 \ex\label{ex:GRM-adv-kapok-B}
\gll B:  ɲɪ̀nɪ̃̀ɛ̃́ bà já kà ŋmɛ̀nà.\\
{} {\q} {\sc 3.pl} do {\egr} ignite\\
\glt `How does one ignite.' 

 \ex\label{ex:GRM-adv-kapok-A-2}
\gll  A: ŋmɛ̀nà nɪ́ŋ̀.\\
{} ignite {\advm}\\
\glt `Ignite like this.'

 \ex\label{ex:GRM-adv-kapok-A-3}
 \gll  A: tʃɪ́á dɪ̀ tʃɪ́ wááwá ŋmɛ̀nà kéŋ̀.\\
{} tomorrow {\conn} {\cras} come.{\pfv} ignite {\advm}\\
\glt `Tomorrow when you come, ignite like that.'
 
\z 
 \z  
 
In the context of (\ref{ex:GRM-adv-kapok}), at the farm the next day, the boy
(B) would tell a colleague: {\sls ken ne ba ja ŋmɛna},  {\it lit.} like.that 
they
do ignite, `that is how one ignites'. 


\ea\label{ex:GRM-ning-prop-2}
 \gll nɪ́ŋ lɛ̀ɪ́ ʊ̀ʊ̀ dɪ̀à háŋ̀ já dʊ̀.\\
 {\advm} {\sc neg}  {\sc 3.sg.poss} house {\sc dem} {\sc hab} be\\
\glt  `This is not how his room used to be.'
%(Python story 078)

\z


In (\ref{ex:GRM-ning-prop-2}), {\sls nɪŋ} refers to the condition of  the room,
which is not a manner  but a property of the room. 
In addition, {\sls keŋ} and 
{\sls nɪŋ} can function as  discourse particles, whose
meanings resemble   English `like' in some registers \citep{Muff02}. In
(\ref{ex:GRM-keng-like}), {\sls keŋ} is considered superfluous since it does not
contribute to the manner of  motion or the state of the
participant.\footnote{Something identical to the translation of
(\ref{ex:GRM-keng-like}) may be heard in some varieties of  English spoken in 
Wa.
This suggest that Waali and/or Dagaare has at least  similar 
expression.} 
 
 \ea\label{ex:GRM-keng-like} 
 \gll ǹ̩ káálʊ̄ʊ̄ kéŋ̀.\\
 {\sc 1.sg} go.{\ipfv .\foc}  {\sc dxm}\\
 \glt `I am leaving like that'
\z

Also, depending on the intonation associated with it, and whether or not  the
focus
particle  is  present, {\sls keŋ} and {\sls nɪŋ} can function as
interjections used to convey comprehension or surprise. So a phrase like {\sls 
kén nȅȅ} could be roughly translated as `Is that so?', {\sls kén nè}   has 
a
similar function to the English  tag-question `Isn't it?', but {\sls kéēèŋ} 
or
{\sls kén né} could be translated as `yes, that is it'. 

Finally, \citet{Mcgi99} presents  {\sls nyɛ} and {\sls ɛɛ} (variant {\sls gɛɛ}) 
as
demonstrative pronouns in Pasaale, which can also modify an entire clause. The
former
corresponds to `this' and the latter to `that'. At this point, it is a matter of
comparing the two languages and the terminology employed.  Nonetheless, in the
majority of the examples provided by \citet{Mcgi99}, Chakali {\sls keŋ} and 
{\sls 
nɪŋ} seem to have the same function. 


\subsection{Spatial deictics {\sls bááŋ̀} and  {\sls dé}}
\label{sec:GRM-deic-adv}
 

A speaker-subjective,  two-way contrast  exists to locate entities in space. 
The spatial deixis demonstrative  {\sls bááŋ̀} designates the location of the 
speaker, while 
the spatial deixis demonstrative  {\sls dé} designates  where the
speaker is not located. They represent what is known as the `proximal' and
`distal' 
dimensions of  place deixis. In (\ref{ex:deic-adv-prox}) and
(\ref{ex:deic-adv-dist}),  they are translated as `here' and 'there'
respectively, and glossed {\sc dxl}, standing for `locative deixis'.  


\ea\label{ex:vp}


\ea\label{ex:deic-adv-prox}
\gll wàà  bááŋ̀.\\
  come {\dem} {\postp}\\
\glt  `Come here'

 \ex\label{ex:deic-adv-dist}
\gll ʊ̀ dʊ́á dé (nɪ̄).\\
    {\psg}  be.at  {\dem}  {\postp}\\
\glt  `He is there'


\z 
 \z

Notice that unlike the single demonstrative  modifier discussed in Section 
\ref{sec:GRM-demons},   {\sls bááŋ̀} and   {\sls dé}  encode a proximal/distal distinction.



\subsection{Focus}
\label{sec:GRM-focus}

Since the notion of focus has been discussed separately in connection with
nominals and verbals, this section offers a basic overview of what has been
stated.  \citet[326]{Dik97} writes that   ``the focal information in a
linguistic expression is
that
information which is relatively the most important or salient in the given
communicative setting''.  In Chakali, we saw  two ways in which a
speaker can integrate focal information, and both of them put `in focus' a
constituent.\footnote{The  terminology employed in the literature is probably
the result
of  complex and still obscure phenomena. For instance, for the
post-verbal particle {\sls lá} in Dagaare, \citet{Bodo97} uses the term
`factitive' and `affirmative' particle interchangeably, \citet{Daku05} uses
`(broad- and narrow-)  focus' and glosses it either as {\sc aff} or {\sc foc},
and
\citet{Saan03} uses post-verbal particle and glosses it as {\sc aff}. In-depth
accounts of focus in Grusi languages can only be found in \citet{blas90}, 
but see also \citet{Mcgi99}.
 Anne Schwarz has worked extensively on the topic in some Gur and Kwa
languages \citep{Schw10}.}   The first
encodes focal information in a particle which  always
 follows a nominal, i.e. {\sls ra} and variants. Its  phonological shape is
determined by the
preceding phonological material (see Sections \ref{sec:focus-forms} and 
\ref{sec:GRM-foc-neg}). The second, which was called the assertive suffix, takes
the form of vowel features which
are suffixed onto the verb  (see Sections \ref{sec:GRM-verb-perf-intran} and 
\ref{sec:GRM-verb-suffix}). It was claimed that  the assertive suffix surfaces
only if (i) none of the other constituents in the
clause are in focus, (ii) the clause does not include negative polarity items,
and (iii) the clause is intransitive.
The second criterion (ii) is applicable to the particle {\sls ra} as well: thus
focal
information can only exist in affirmative clauses, negation automatically
prevents information from being in focus.\footnote{\citet[94]{Bodo97} writes
(for
Dagaare) that
``[the factitive particle {\sls lá}] is in complementary distribution with the
negative polarity particles, as one would expect of an affirming particle".}  In
 (\ref{ex:GRM-focus}),  the
examples illustrate  how the  focal information is
encoded when the object (\ref{ex:GRM-focus-obj}), the subject
(\ref{ex:GRM-focus-subj}) and the predicate  (\ref{ex:GRM-focus-pred}) are
considered the most important piece of information. 


\ea\label{ex:GRM-focus}

 \ea\label{ex:GRM-focus-obj}{\rm Focus on object: What has the man chewed?}\\
\gll   à báál tíē sɪ́gá rá.\\
   {\art} man chew bean {\foc}\\
\glt `The man chewed BEANS'

\gll  kàlá tíē sɪ́gá rá.\\
   K.   chew bean {\foc}\\
\glt `Kala chewed BEANS'

\ex\label{ex:GRM-focus-subj}{\rm Focus on subject: Who has chewed the beans?}\\
\gll   à báál là  tíē sɪ́gá.\\
    {\art} man {\foc} chew bean\\
\glt `The MAN chewed beans'



\gll  kàláá tíē sɪ́gá.\\
   K.   chew bean\\
\glt `KALA chewed beans'

\ex\label{ex:GRM-focus-pred}{\rm Focus on predicate: What happened?}\\
\gll à báál tíéwóó.\\
   {\art} man chew.{\pfv .\foc}\\
\glt `The man CHEWED'


\z 
 \z

The focus particle does not differentiate between  grammatical functions and 
appears to be optional.  Also,  the assertive suffix is quite rare 
in narratives.  \citet[94]{blas90} is the only author to my knowledge 
who identifies the presence  of  evidentiality --  hearsay, more precisely -- 
in Gur languages. According to her the morpheme {\sls rɛ} in Sissala refers to 
reported or inferred information. This raise the question as to what extent the 
focus particle and the assertive suffix provide evidential 
information.\footnote{A promising avenue to follow in the study of focus 
would be the recent work of Anne Schwarz  who looks at 
the phenomenon from a perspective of encoding a thetic/
categorical distinction.}


\subsection{Linguistic taboos}
\label{sec:GRM-ling-taboo}

A linguistic \is{taboo} taboo is defined here as the avoidance of certain words 
on certain occasions due to  misfortune associated with those words. These 
circumstances depend on belief; they can be widespread or marginal. The 
avoidance of certain words may depend on the time of the day or action carried 
out. For instance, not only  is sweeping  not allowed when someone eats, but 
uttering the word {\sls tʃãã} `broom' is also forbidden. Also, mentioning 
certain animal names is excluded as they may either be tabooed by someone 
present, due to his/her animal totem and/or its meat is forbidden (Section 
Section \ref{sec:SOC-religion}),  or attract the animal's attention, i.e. the 
belief that the  animal may feel it is called out. The strategy is to 
substitute a word with another, often undertaking a metonymic strategy. 

\begin{table}[!htb]
\small
\centering
\caption{Taboo synonyms \label{ex:GRM-taboo-synonyms}}

\begin{tabular}{llll}
\lsptoprule
 Avoided word & Substitute word & Literal meaning & Gloss\\
\midrule

 {\sls bɔ̀là} & {\sls sé-zèŋ́}   & animal-big &   `elephant'\\

{\sls bɔ̀là} & {\sls néŋ-tɪ̄ɪ̄nā} &  arm|hand-owner  &  `elephant'\\

  {\sls dʒɛ̀tɪ̀} & {\sls ɲú-zéŋ-tɪ̄ɪ̄nā} &  head-big-owner &  `lion'\\
  
  {\sls bʊ́ɔ̀mánɪ́ɪ́} & {\sls ɲú-wíé-tɪ̄ɪ̄nā} &   head-small-owner &   `leopard'\\
  
  {\sls váà} & {\sls nʊ̃̀ã̀-tɪ́ɪ́ná}  & mouth-owner &   `dog'\\
  
  {\sls kɔ́ŋ} & {\sls nɪ́ɪ́-tɪ́ɪ́ná}   & water-owner &  `cobra'\\
  
  {\sls gbɪ̃̀ã́} & {\sls néŋ-gál-tɪ̄ɪ̄nā} &  arm|hand-left-owner &   `monkey'\\
  
  {\sls hèlé} & {\sls mùŋ-zɪ́ŋ-tɪ̄ɪ̄nā} & back-big-owner &  `type of squirrel'\\
  
  {\sls tébíŋ̀} & {\sls bárà-tʃɔ́gɔ́ʊ́}  & place-spoil.{\sc pfv.foc}&  `night'\\
  
  {\sls ɲʊ́lʊ́ŋ} & {\sls ɲú-bɪ́rɪ́ŋ-tɪ́ɪ́ná} &  head-full-owner &  `blind' \\
  
%   {\sls tʃã́ã́} & {\sls kɪ̀m-pɪ̀ɪ̀g-ɪ́ɪ̀} & thing-mark-{\sc nmlz}  &  `broom'\\
  
  {\sls búmmò} & {\sls dóŋ} &   dirt  &  `black'\\
  
%   {\sls dʊ̃̀ʊ̃̀wìé} & {\sls mábíé-wāá-tèlè-púsíŋ} &   sibling-will.not-reach-meet.me &   
%  `type of snake'\\

\lspbottomrule
 
\end{tabular}
\end{table}


The second column of  example (\ref{ex:GRM-taboo-synonyms}) contains expressions called \is{taboo 
synonyms} taboo \is{synonym} synonyms; they are substitutes to the words of the  the first column.  
The substitutes  are usually complex stem nouns with a transparent descriptive meaning. Most of 
them use the stem {\sls tɪɪna} `owner of', e.g. {\sls neŋ-tɪɪna}, {\it lit.} arm|hand-owner.of,  
`elephant',  the one with a big arm. 


\subsection{Ideophones\is{ideophone} and iconic strategies}
\label{sec:GRM-onoma}

 In Chakali, ideophones typically suggest the description of an abstract 
property or the manner in which an event unfolds.  The majority of ideophones 
functions like  qualifiers (Section \ref{sec:GRM-qualifier}) or adjunct 
adverbials  (Section \ref{sec:GRM-adjuncts}). Ideophones  tend to appear with a 
low tone. 
          

\ea\label{ex:GRM-ideo}

 \ea\label{ex:GRM-ideo-dxm}
\gll  à díŋ káá dīù gàlɪ̀gàlɪ̀gàlɪ̀/pèpèpè.\\
{\sc art} fire {\sc ipfv} eat  {\ideo}\\
  \glt `The fire is burning at an increasing rate.'


 \ex\label{ex:GRM-ideo-qual}
\gll à dʊ̃́ʊ̃́ síè jáá wə̀rwə̀rwə̀r.\\
{\sc art} python eye {\ident} {\ideo}\\
  \glt `The python's eyes are glittery.'

   \ex\label{ex:GRM-ideo-}
\gll à dáánɔ́ŋ márá bɪ̄jʊ̄ʊ́ lìgèlìgèlìgè.\\
{\sc art} tree.fruit well ripe.{\sc pfv}  {\ideo}\\
  \glt `The fruit is perfectly riped.'
  
  
  \ex\label{ex:GRM-ideo-}
\gll à sìbíé wàà márá bɪ̀ɪ̀ à dʊ́ nɪ̄ŋ wùròwùròwùrò.\\
{\sc art} beans  {\sc neg} well ripe {\sc conn} be   {\sc dxm} {\ideo}\\
  \glt `The bambara beans are not well cooked, they are still hard.'
  

\z 
 \z

The translations into English in (\ref{ex:GRM-ideo}) were not 
tested for consistency across many speakers.

An onomatopoeia is a type of \is{ideophone}ideophone which not only suggests 
the concept   it expresses with sound, but imitates  the actual sound of an 
entity or event.  Examples of onomatopoeia\is{onomatopoeia} are {\sls púpù} 
`motorbike', {\sls tʃétʃé} `bicycle', {\sls tʃɔ̀kɔ̃́ɪ̃́ tʃɔ̀kɔ̃́ɪ̃́} `sound 
of 
a 
guinea fowl',  {\sls kr̀r̀r̀r̀} `sound of running',  {\sls pã̀ã̀} `sound of an 
eruption caused by lighting a fire',  {\sls gbàgbá}  `duck',\footnote{The 
word for `duck' is probably borrowed from Waali. This bird was introduced 
recently and was hard to find in the villages visited.}   and {\sls 
kpókòkpókòkpókò} `sound of knocking on a clay pot'.


 Similarly, an iconic strategy to convey an amplified meaning or the idea of
continuity is to lengthen the sound of an existing word. 


 \ea\label{ex:GRM-lenght}
   \gll  kàwāá sìì tàrɪ̀ kéééééééŋ, àkà dʊ́á  bà dɪ̀ànʊ̃́ã́ nɪ̀.\\
pumpkin rise {creep} {\advm} {\conn} {be.at} {\sc 3.sg.poss} door {\postp}\\
\glt `The pumpkin crept, crept, crept, and crept up to their door mat.'
%\hfill{Python story (line 56)}
\z

In (\ref{ex:GRM-lenght}) the manner deictics {\sls keŋ} (Section
\ref{sec:GRM-adv-pro}) is stretched to simulate the extention in time of the
event, i.e. the pumpkin grew until it reached the door.\footnote{An equivalent
meaning may be expressed in some varieties of Gh. Eng.  with the 
adverbial
expression  {\sls ãããã}, as in `Today I worked  {\sls ãããã}, 
until night 
time.'}




%busabusa in Waali, from Akan ``unatteactive, dull, repulsive''
% tʃatʃara in Waali  ``a muddy area''
% apelepele in Waali ``clear''
% ganura ``paint/marks on sacks at market''



Reduplication\is{reduplication} of one or two syllables is the general 
structural shape of
\is{ideophone}Ideophones and onomatopoeias. A large set of visual 
perception expressions can be treated as
 ideophonic expressions (Section \ref{sec:GRM-qualifier}), all of which are 
reduplicated expression.  

\ea\label{ex:BCTreduplic}{\rm Visual perception expressions  and 
non-attested stems}\\

\ea {(kɪn|a)-hɔlahɔla}	  [áhɔ̀làhɔ̀là]   *hɔla   
\ex {(kɪn|a)-ahɔhɔla}		[áhɔ̀hɔ̀là] *hɔla   
\ex {(kɪn|a)-busabusa}	[ábùsàbùsà]	 *busa  
\ex {(kɪn|a)-adʒumodʒumo}	[ádʒùmòdʒùmò] *dʒumo  
\ex {(kɪn|a)-bʊɔbʊɔna}	[ábʊ̀ɔ̀nàbʊ̀ɔ̀nà]		 *bʊɔna  
\ex {(kɪn|a)-ʔileʔile}	[áʔìlèʔìlè]	 *ʔile 

\z
\z


If we accept that reduplication is a morphological process in which the root or 
stem is repeated (fully or partially), then it is questionable whether one can 
treat most of the naming data as reduplication. It is obvious from the examples in 
(\ref{ex:BCTreduplic}) that there is a `form-doubling' on the surface, yet such 
expressions  are not made out of attested stems (and they do not have loci in the chromatic space, 
\citealt[see][]{brin16}).  


\subsection{Interjection and formulaic language}
\label{sec:GRM-greet}


This section introduces some pieces of formulaic language, which is defined as
conventionalized and idiomatic words or phrases. It usually include greetings,
idioms, proverbs,  etc. \citep{Wray05}. First, common
interjections are introduced in Table \ref{tab:GRM-interj},\footnote{The 
etymology
of {\sls ʔàmé} has not been confirmed and {\sls gáfrà} is ultimately Hausa. 
The word
{\sls ʃɪ́ã̀ã̀} is equivalent to the
function  associated with the action of {\sls tʃuuse} in Chakali ({\sls tʃʊʊrɪ} 
in
Dagaare, {\sls tʃʊʊhɛ} in Waali, `puf' or `paf'  in Gh.  Eng.,   $<$
English `pout'), which
is a
fricative sound produced by a non-pulmonic, velarized ingressive airstream
mechanism, articulated with the lower lip and the upper front teeth while the
lips are protruded.} then some greetings and idioms are presented. 



\begin{table}[!htb]
\small
\centering
\caption{Selected interjections \label{tab:GRM-interj}}

\begin{tabular}{>{\slshape}lp{8cm}}
\lsptoprule
{\rm Interjection} & Gloss\\[1ex] \midrule
ʔàɪ́  &  express denial or refusal\\
ʔɛ̃ɛ̃ & express affirmation\\
gááfʊ̆̀rà    &  express excuse when interrupting or disturbing  ({\it from}  Hausa)\\
téò  &  express agreement or understanding  ({\it from}  Hausa)\\
ʔàmé   & so be it  ({\it etym.}  Amen?)\\
ʔóí  &  express surprise\\
fíó  &  express strong denial or refusal\\
ʔánsà  &  1) greet hospitably, welcome, 2) accept and thank ({\it from} Gonja)\\
ʔĩ́ĩ̀ĩ́  &  express disappreciation of an action
carried out by someone else\\
ʔàwó  &  reply to greetings, a sign of appraisal of the interlocutor's
concerns ({\it from} Gonja)\\
 ʔáábbà & express a  reaction to an unpalatable proposition, with disagreement and 
unexpectedness\\
ʃ̃ʊ̃́ɛ̃̀ɛ̃̀  &  express a disrespectful attitude towards what is being said and the 
one saying it \\
\lspbottomrule
\end{tabular}
\end{table}



Since they are conventionalized and idiomatic, the translation of  formulaic 
language in Table \ref{tab:GRM-interj}  is
 a rough  approximation. The dictionary offers various spelling since variations are 
regularly perceived.



\subsubsection{Greetings}
\label{sec:GRM-greet}

Compulsory prior to any communicative exchange, greetings trigger
both attention and respect. When meeting with elders, one should  squat or bend
forward hands-on-knees  while greeting. Clan names can be used in greetings,
e.g. {\sls ɪ́tʃà} `respect to you and to your clan'. In Table
\ref{tab:greetings},  typical greeting lines with some responses are provided.


\begin{table}[!htb]
\small
\centering
\caption{Greetings\label{tab:greetings}}

\begin{tabular}{l>{\slshape}lp{7cm}}
\lsptoprule
Time & {\rm Speaker A} & Followed by either speaker A or B\\ \midrule

Morning  & ánsùmōō  & {\sls ɪ̀ sìwȍȍ} `You stood?', {\sls ɪ̄ dɪ̀ 
tʃʊ́àwʊ̏ʊ̏}   `And your lying?', {\sls ɪ̀ bàtʃʊ̀àlɪ́ɪ̀ wīrȍȍ }  `You 
sleeping place was good?'\\[1ex]

Afternoon   & ántèrēē & {\sls ɪ́ wɪ́sɪ́ tèlȅȅ}   `Has the sun reached 
you?' {\sls  ɪ́ dɪ́á} `And your house?'  {\sls ɪ̄ bìsé mūŋ} `And all 
your children?'\\[1ex]
  

Evening & ɪ́ dʊ̀ànāā &  {\sls ɪ́ dʊ̄ɔ̄n tèlȅȅ}  `Your evening 
has reached', 
{\sls ɪ́ kúó} `And your farm?'\\
\lspbottomrule
\end{tabular} 
\end{table}


The  second singular plural 
{\sls ma} is added, i.e.  {\sls ánsùmōō} $\leftrightarrow$ {\sls māānsùmōō},
when there is more than one adressee or when there is  a single person but the
greetings
are intended to the entire house/family: thus  the number distinction {\sls ɪ}/{\sls 
ma}
does
not correspond to a politeness function, e.g.  French {\it tu}/{\it vous}. 
Chakali
 morning and afternoon greetings resemble those of Waali and other languages of
the area.
The response to various greetings such as {\sls ɪ́ dɪ́á} `and your house?',  
{\sls ʔánsà} `welcome, thanks' and many others is the multifunctional 
expression {\sls ʔàwó},  which is, among other things, a sign of appraisal of 
the
interlocutor's
concerns. The same expression is found in Gonja, but its functions may be different.
functions. I was told that the more extensive the greetings, the more
respect one shows the addressee.  For instance, the elders do not
appreciate the tendency of
the youths to morning-greet as {\sls ã̄sūmō}, but prefer something like 
{\sls 
áánsùùmōōō}. 

Other expressions often used are: {\sls tʃɔ̄pɪ̄sɪ́ ālɪ̀ɛ̀}  {\it lit.} morning two,  `long time no 
see';   {\sls bámùŋ kɔ́rɛ́ɪ́}  {\it lit.}  all.\textsc{+hum} extent? (Hausa), `how are all your 
people?', {\sls ànɪ́ mà wʊ̀zʊ́ʊ́rɪ́ tɪ̀ŋ} {\it lit.} and your day, used  after any bad event which 
happened to someone, e.g. referring to the funeral day, when the speaker has not seen the addressee 
since that day.


 

\subsubsection{Idioms}
\label{sec:GRM-idiom}

An idiom is a  composite expression which does not convey the literal  meaning 
of the composition  of its parts. Common among many African languages is a 
strategy by which  abstract nominals are expressed in idiomatic compounds. 
These compounds are made of stems whose meanings are disassociated from their 
ordinary usage.


Some examples have already been provided in section \ref{sec:GRM-qualifier}. In 
Chakali, words identifying mental states and habits/behaviors are often 
idiomatic, e.g. {\sls síínʊ̀màtɪ́ɪ́nà} ({síí-nʊ̀mà-tɪ́ɪ́nà}, {\it lit.} 
eye-hot-owner), `wild' or {\sls nʊ̀ã̀pʊ̀mmá} ({nʊã-pʊmma}, {\it lit.} 
mouth-white), `unreserved'. Even though the expression {\sls 
síínʊ̀màtɪ́ɪ́nà} 
is made out of three lexical roots, it is a `sealed' expression and is 
associated with the manner in which a person behaves, i.e. a wild person. The 
sequence {\sls jaa nʊ̃ã dɪgɪmaŋa} in (\ref{ex:GRM-idiom-mouth}), {\it lit.} 
do-mouth-one,  is also treated as an idiomatic expression.

\ea\label{ex:GRM-idiom-mouth}
   \gll   bà jáá nʊ̃̀ã̀ dɪ́gɪ́máŋá à sùmmè dɔ́ŋà.\\
{\sc 3.pl} do mouth one {\conn} help {\recp}\\
\glt `They should agree and help each other.'

\z

Needless to say, it is often difficult to  distinguish between an idiomatic
expression and  an expression in which only one of the  components is use in a
 non-literal sense. 
 
 %[to do {\sls di kaɲɪtɪ}]
 
 
% % % eat-patience
% % % However, it is often difficult to  distinguish between an idiomatic
% % % expression from an expression in which one of its component is use in a
% % % non-literal sense. For instance, one could treat {\sls di kaɲɪtɪ},  {\sls 
% lit.}
% % % eat-patience,  as an idiom, but the verb {\sls di} 
% % % Other idiomatic expressions are 
% % % {\S}, {\it lit.} , `'
% % % {\S}, {\it lit.} , `'
% % % {\S}, {\it lit.} , `'
% % % {\S}, {\it lit.} , `'






\subsection{Clicks}
\label{sec:GRM-greet}

\citet[151]{Nade89} writes that clicks\footnote{A click may be roughly defined 
as  the release of a pocket of air enclosed between two points of contact in the 
mouth. The air is rarefied by a sucking action of the tongue 
\citep[see][]{Lade93}.}  may be  heard in the Gur-speaking area to  mean an 
affirmative `yes', or `I'm listening'.  This also occurs in the villages where I 
stayed, but I noticed that one click usually means `yes', `I understand' or `I 
agree', whereas two clicks mean the opposite. The click is  palatal and produced 
with the lips closed.




